---
title: "Bolivia Census 2024 — Albó indicators of indigenous identity and language"
format: html
execute:
  warning: false
  message: false
---

## Background

Xavier Albó's framework for studying indigenous identity distinguishes between
several dimensions that the census allows us to measure independently:

1. **Self-identification** with an indigenous or Afrobolivian group (Q1)
2. **Language match**: whether the language(s) a person speaks correspond to
   their declared identity (Q2)
3. **Childhood language**: whether they learned an indigenous language as a
   child (Q3)
4. **Castellano use**: whether they speak Spanish as any of their current
   languages (C)

These four boolean variables — `albo_q1`, `albo_q2`, `albo_q3`, `albo_c` —
are pre-computed for all 11,365,333 persons and saved as `albo_vars.rds`. This
document describes their construction, validates them against the raw census
variables, and begins exploring what they reveal.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
library(arrow)
library(tidyverse)
library(patchwork)

DATA_DIR <- "../bolivia-data/Censo 2024/base_datos_csv_2024"
CENSO_DIR <- "../bolivia-data/Censo 2024"
```

### Lookup tables

```{r}
#| label: lookup-tables

# p32_pueblos: CTAI-consolidated identity groups (58 categories + 98/99)
pueblos_cats <- tribble(
  ~code, ~label,
  1,  "Afroboliviano",         2,  "Araona",            3,  "Aymara",
  4,  "Ayoreo",                5,  "Baure",             6,  "Canichana",
  7,  "Cavineño",              8,  "Cayubaba",          9,  "Chácobo",
  10, "Charka Qhara Qhara",   11,  "Chichas",          12,  "Chiquitano",
  13, "Chuwi",                14,  "Ese Ejja",          15,  "Guaraní",
  16, "Guarasu´we",           17,  "Gwarayu",           18,  "Itonama",
  19, "Jach'a Carangas",      20,  "Jalq'a",            21,  "Joaquiniano",
  22, "Kallawaya",            23,  "Killacas",           24,  "Leco",
  25, "Lípez",                26,  "Lupaca",             27,  "Machineri",
  28, "Maropa",               29,  "Mojeño",             30,  "Mojeño Ignaciano",
  31, "Mojeño Trinitario",    32,  "Moré",               33,  "Mosetén",
  34, "Movima",               35,  "Pacahuara",          36,  "Pakajaqi",
  37, "Paunaca",              38,  "Puquina",            39,  "Qhapaq Uma Suyu",
  40, "Quechua",              41,  "Qullas",             42,  "Raqaypampa",
  43, "Sirionó",              44,  "Sora",               45,  "Tacana",
  46, "Tapiete",              47,  "Toromona",           48,  "Tsimane´",
  49, "Uru-Chipaya",          50,  "Weenhayek",          51,  "Yaminawa",
  52, "Yampara",              53,  "Yuqui",              54,  "Yurakaré",
  55, "Quechua - Aymara",     56,  "Más de una descripción",
  57, "Otras declaraciones",  58,  "Otros pueblos indígenas extranjeros",
  98, "No se autoidentifica", 99,  "Sin respuesta"
)

# idioma_mat / idioma_mayor_uso / p33x spoken-language codes
idioma_cats <- tribble(
  ~code, ~label,
     1, "Araona",              2, "Aymara",               3, "Baure",
     4, "Bésiro",              5, "Canichana",             6, "Castellano",
     7, "Kabineña",            8, "Cayubaba",              9, "Chácobo",
    10, "Tsimane´",           11, "Ese Ejja",             12, "Guaraní",
    13, "Guarasu´we",         14, "Gwarayu",              15, "Itonama",
    16, "Leco",               17, "Kallawaya",            18, "Machineri",
    19, "Maropa",             20, "Mojeño Ignaciano",     21, "Mojeño Trinitario",
    22, "Moré",               23, "Mosetén",              24, "Movima",
    25, "Pacahuara",          26, "Puquina",              27, "Quechua",
    28, "Sirionó",            29, "Tacana",               30, "Tapiete",
    32, "Uru-Chipaya",        33, "Weenhayek",            34, "Yaminawa",
    35, "Yuqui",              36, "Yurakaré",             37, "Zamuco",
    94, "Afroboliviano",     991, "Otras declaraciones", 992, "Otro idioma extranjero",
   994, "Lenguaje de señas", 998, "No habla"
)
```

------------------------------------------------------------------------

## Variable construction

### Source variables

The four Albó indicators are derived from the following raw census variables:

| Raw variable | Description |
|---|---|
| `p32_pueblo_per` | Indigenous self-identification: 1 = Sí, 2 = No, 9 = Sin respuesta |
| `p32_pueblos` | CTAI-reconciled identity group (58 categories); derived from `p32_pueblo_cod` |
| `p331_idiohab1_cod` | First spoken language (current use) |
| `p332_idiohab2_cod` | Second spoken language (current use) |
| `p333_idiohab3_cod` | Third spoken language (current use) |
| `p341_idiomat_cod` | Language learned as a child (NA if respondent answered "No" to p34) |
| `p342_idiomat_no` | = 1 when respondent did not learn any language as a child |

### Indicator definitions

```{r}
#| label: indicator-definitions

# Shown for documentation; albo_vars.rds is the pre-computed output.

# albo_q1: Self-identifies with an indigenous or Afrobolivian group
# albo_q1 <- p32_pueblo_per == 1

# albo_q2: Speaks a language that corresponds to their declared identity.
#
# For named groups with a known language (e.g. Quechua → code 27, Chiquitano
# → Bésiro code 4, Ayoreo → Zamuco code 37), albo_q2 is TRUE when any of
# p331/p332/p333 is in that language's code set.
#
# For groups that use the "any indigenous language" rule — generic identities
# (Otras declaraciones = 57, Quechua-Aymara = 55, Más de una descripción = 56)
# AND groups with no separate census language (Afroboliviano = 1, Chuwi = 13,
# Paunaca = 37, Toromona = 47, foreign groups = 58) — albo_q2 is TRUE when
# any of p331/p332/p333 is an indigenous language code (1–5 or 7–37).
#
# Always FALSE if albo_q1 is FALSE.

# albo_q3: Learned an indigenous language as a child
# albo_q3 <- !is.na(p341_idiomat_cod) & p341_idiomat_cod != 6 & p341_idiomat_cod != 999

# albo_c: Speaks Castellano (as any of their current languages)
# albo_c <- any of p331/p332/p333 == 6
```

### Notable mapping decisions

Two naming-convention differences affect the language–identity crosswalk:

- **Kabineña** (code 7 in `idioma_mat`) = **Cavineño** (code 7 in `p32_pueblos`):
  same community, different orthographic conventions.
- **Zamuco** (code 37 in `idioma_mat`) = **Ayoreo** (code 4 in `p32_pueblos`):
  Zamuco is the ISO language name; Ayoreo is the ethnonym used in identity contexts.

Several Andean sub-ethnic identities map to the major Andean languages:

- Quechua sub-groups (Charka Qhara Qhara, Chichas, Jalq'a, Yampara, Lípez,
  Raqaypampa) → Quechua (27)
- Aymara sub-groups (Qullas, Killacas, Jach'a Carangas, Lupaca, Pakajaqi,
  Sora, Qhapaq Uma Suyu) → Aymara (2)
- Mojeño (29, the generic parent code) → Mojeño Ignaciano (20) or Trinitario (21)
- Joaquiniano (21) → Maropa (19)
- Chiquitano (12) → Bésiro (4)
- Ayoreo (4) → Zamuco (37)

------------------------------------------------------------------------

## Load pre-computed data

```{r}
#| label: load-albo
albo_vars     <- readRDS(file.path(DATA_DIR, "albo_vars.rds"))
cel_geo       <- readRDS(file.path(DATA_DIR, "cel_geo.rds"))

# Also load the pre-drawn 10k sample for any person-level illustration
persona_sample <- readRDS(file.path(DATA_DIR, "Persona_CPV-2024_sample10k.rds"))

cat("albo_vars dimensions:", nrow(albo_vars), "×", ncol(albo_vars), "\n")
glimpse(albo_vars)
```

------------------------------------------------------------------------

## Prevalence of each indicator

```{r}
#| label: prevalence-table
albo_vars |>
  summarise(
    n = n(),
    `Q1: Self-identifies (%)` = round(mean(albo_q1) * 100, 1),
    `Q2: Language match (%)` = round(mean(albo_q2) * 100, 1),
    `Q3: Indigenous childhood language (%)` = round(mean(albo_q3) * 100, 1),
    `C: Speaks Castellano (%)` = round(mean(albo_c) * 100, 1)
  ) |>
  pivot_longer(everything(), names_to = "Indicator", values_to = "Value")
```

Among the full population of 11.4M:

- **37.9%** self-identify as indigenous or Afrobolivian (Q1).
- **25.9%** speak a language corresponding to their declared identity (Q2).
  The gap between Q1 and Q2 (≈12 pp) represents self-identifiers who have
  shifted to Spanish.
- **21.3%** learned an indigenous language as a child (Q3). This is slightly
  lower than Q2, which is consistent: some people who learned an indigenous
  language as a child no longer speak it as adults, while some people speak an
  indigenous language now but via a path other than home transmission.
- **92.6%** speak Castellano as one of their current languages (C).

------------------------------------------------------------------------

## Cross-tabulation of the four indicators

```{r}
#| label: four-way-cross
albo_vars |>
  count(albo_q1, albo_q2, albo_q3, albo_c) |>
  mutate(
    pct = round(100 * n / sum(n), 2),
    across(c(albo_q1, albo_q2, albo_q3, albo_c), \(x) if_else(x, "Y", "N"))
  ) |>
  rename(Q1 = albo_q1, Q2 = albo_q2, Q3 = albo_q3, C = albo_c) |>
  arrange(desc(n))
```

The four most common profiles are:

1. **N/N/N/Y** — Does not identify, no indigenous language (child or adult),
   speaks Spanish. The large non-indigenous majority.
2. **Y/Y/Y/Y** — Identifies, speaks the identity language, learned it as a
   child, and also speaks Spanish. The "full" indigenous profile with Spanish
   bilingualism.
3. **Y/N/N/Y** — Identifies but has shifted to Spanish on all three dimensions.
   Identity without language.
4. **Y/Y/Y/N** — Identifies, speaks the identity language, learned it as a
   child, no Spanish. Monolingual indigenous speakers.

------------------------------------------------------------------------

## Albó typology

A simplified typology collapses Q1–Q3 into mutually exclusive categories,
following Albó's original framing:

```{r}
#| label: albo-typology
albo_vars <- albo_vars |>
  mutate(
    albo_type = case_when(
      !albo_q1              ~ "No identity",
      albo_q1 &  albo_q2 &  albo_q3 ~ "Identity + language + childhood",
      albo_q1 &  albo_q2 & !albo_q3 ~ "Identity + language (no childhood)",
      albo_q1 & !albo_q2 &  albo_q3 ~ "Identity + childhood (no current language)",
      albo_q1 & !albo_q2 & !albo_q3 ~ "Identity only"
    ),
    albo_type = factor(albo_type, levels = c(
      "Identity + language + childhood",
      "Identity + language (no childhood)",
      "Identity + childhood (no current language)",
      "Identity only",
      "No identity"
    ))
  )

albo_vars |>
  count(albo_type) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(albo_type)
```

```{r}
#| label: typology-plot
#| fig-height: 4
type_colors <- c(
  "Identity + language + childhood"          = "#2166ac",
  "Identity + language (no childhood)"       = "#74add1",
  "Identity + childhood (no current language)" = "#fee090",
  "Identity only"                            = "#f46d43",
  "No identity"                              = "#d9d9d9"
)

albo_vars |>
  count(albo_type) |>
  mutate(
    pct      = n / sum(n),
    albo_type = fct_rev(albo_type)
  ) |>
  ggplot(aes(x = pct, y = albo_type, fill = albo_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = type_colors) +
  labs(
    title    = "Albó typology of indigenous identity and language",
    subtitle = "Bolivia CPV 2024 | n = 11,365,333",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11)
```

------------------------------------------------------------------------

## Typology by urban/rural

```{r}
#| label: typology-urban-rural
#| fig-height: 5

urbrur_labels <- c("1" = "Urban", "2" = "Rural")

albo_vars |>
  mutate(urbrur_label = urbrur_labels[as.character(urbrur)]) |>
  filter(!is.na(urbrur_label)) |>
  count(urbrur_label, albo_type) |>
  group_by(urbrur_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  mutate(albo_type = fct_rev(albo_type)) |>
  ggplot(aes(x = pct, y = albo_type, fill = albo_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.14))) +
  scale_fill_manual(values = type_colors) +
  facet_wrap(~ urbrur_label) +
  labs(
    title    = "Albó typology by urban/rural",
    subtitle = "Bolivia CPV 2024",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

------------------------------------------------------------------------

## Typology by department

```{r}
#| label: typology-dept
dept_labels <- c(
  "1" = "Chuquisaca", "2" = "La Paz", "3" = "Cochabamba",
  "4" = "Oruro",      "5" = "Potosí", "6" = "Tarija",
  "7" = "Santa Cruz", "8" = "Beni",   "9" = "Pando"
)

albo_vars |>
  mutate(dept = dept_labels[as.character(idep)]) |>
  count(dept, albo_type) |>
  group_by(dept) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  select(dept, albo_type, pct) |>
  pivot_wider(names_from = albo_type, values_from = pct, values_fill = 0) |>
  arrange(`Identity + language + childhood`)
```

```{r}
#| label: typology-dept-plot
#| fig-height: 6
albo_vars |>
  mutate(dept = dept_labels[as.character(idep)]) |>
  count(dept, albo_type) |>
  group_by(dept) |>
  mutate(
    pct  = n / sum(n),
    dept = fct_reorder(dept, pct * (albo_type == "Identity + language + childhood"), sum)
  ) |>
  ungroup() |>
  ggplot(aes(x = pct, y = dept, fill = albo_type)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = type_colors) +
  labs(
    title    = "Albó typology by department",
    subtitle = "Bolivia CPV 2024",
    x = "Share of population", y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

------------------------------------------------------------------------

## Q2 among self-identifiers: language match rate by identity group

Among those who self-identify (albo_q1 = TRUE), what share also speaks the
language of their declared identity (albo_q2 = TRUE)? This is loaded from the
person file to get group-level counts.

```{r}
#| label: q2-by-group
#| cache: true
ds <- open_dataset(
  file.path(DATA_DIR, "Persona_CPV-2024.csv"),
  format = "csv", delimiter = ";"
)

q2_by_group <- ds |>
  filter(p32_pueblo_per == 1) |>
  select(p32_pueblos,
         p331_idiohab1_cod, p332_idiohab2_cod, p333_idiohab3_cod) |>
  collect()
```

```{r}
#| label: q2-by-group-compute
# Re-derive albo_q2 logic per group using the same crosswalk as albo_vars.rds
# (See variable construction section for mapping rules)

indigenous_idioma_codes <- c(1:5, 7:37)
generic_groups_v2 <- c(1L, 13L, 37L, 47L, 55L, 56L, 57L, 58L)

# Flat lookup: one row per (pueblos_code, single idioma_code) pairing
lookup_flat <- tribble(
  ~pueblos_code, ~idioma_code,
  2L,  1L,   # Araona
  3L,  2L,   # Aymara
  4L,  37L,  # Ayoreo -> Zamuco
  5L,  3L,   # Baure
  6L,  5L,   # Canichana
  7L,  7L,   # Cavineño = Kabineña
  8L,  8L,   # Cayubaba
  9L,  9L,   # Chácobo
  10L, 27L,  # Charka Qhara Qhara -> Quechua
  11L, 27L,  # Chichas -> Quechua
  12L, 4L,   # Chiquitano -> Bésiro
  14L, 11L,  # Ese Ejja
  15L, 12L,  # Guaraní
  16L, 13L,  # Guarasu'we
  17L, 14L,  # Gwarayu
  18L, 15L,  # Itonama
  19L, 2L,   # Jach'a Carangas -> Aymara
  20L, 27L,  # Jalq'a -> Quechua
  21L, 19L,  # Joaquiniano -> Maropa
  22L, 17L,  # Kallawaya
  23L, 2L,   # Killacas -> Aymara
  24L, 16L,  # Leco
  25L, 27L,  # Lípez -> Quechua
  26L, 2L,   # Lupaca -> Aymara
  27L, 18L,  # Machineri
  28L, 19L,  # Maropa
  29L, 20L,  # Mojeño -> Mojeño Ignaciano
  29L, 21L,  # Mojeño -> Mojeño Trinitario
  30L, 20L,  # Mojeño Ignaciano
  31L, 21L,  # Mojeño Trinitario
  32L, 22L,  # Moré
  33L, 23L,  # Mosetén
  34L, 24L,  # Movima
  35L, 25L,  # Pacahuara
  36L, 2L,   # Pakajaqi -> Aymara
  38L, 26L,  # Puquina
  39L, 2L,   # Qhapaq Uma Suyu -> Aymara
  40L, 27L,  # Quechua
  41L, 2L,   # Qullas -> Aymara
  42L, 27L,  # Raqaypampa -> Quechua
  43L, 28L,  # Sirionó
  44L, 2L,   # Sora -> Aymara
  45L, 29L,  # Tacana
  46L, 30L,  # Tapiete
  48L, 10L,  # Tsimane'
  49L, 32L,  # Uru-Chipaya
  50L, 33L,  # Weenhayek
  51L, 34L,  # Yaminawa
  52L, 27L,  # Yampara -> Quechua
  53L, 35L,  # Yuqui
  54L, 36L,  # Yurakaré
  55L, 2L,   # Quechua-Aymara -> Aymara
  55L, 27L   # Quechua-Aymara -> Quechua
)

# Build lookup matrix (99 pueblos codes × 38 idioma codes)
valid_match_q2 <- matrix(FALSE, nrow = 99L, ncol = 38L)
for (i in seq_len(nrow(lookup_flat))) {
  pc <- lookup_flat$pueblos_code[i]
  ic <- lookup_flat$idioma_code[i]
  if (ic <= 38L) valid_match_q2[pc, ic] <- TRUE
}
# Generic + Afro + unmatched groups: any indigenous language (codes 1-5, 7-37)
for (g in generic_groups_v2) {
  valid_match_q2[g, indigenous_idioma_codes[indigenous_idioma_codes <= 38L]] <- TRUE
}

p32  <- q2_by_group$p32_pueblos
l1   <- q2_by_group$p331_idiohab1_cod
l2   <- q2_by_group$p332_idiohab2_cod
l3   <- q2_by_group$p333_idiohab3_cod

p32s <- ifelse(is.na(p32) | p32 > 99L, 0L, p32)
l1s  <- ifelse(is.na(l1)  | l1  > 38L, 0L, l1)
l2s  <- ifelse(is.na(l2)  | l2  > 38L, 0L, l2)
l3s  <- ifelse(is.na(l3)  | l3  > 38L, 0L, l3)

has <- p32s > 0L
q2v <- logical(nrow(q2_by_group))
q2v[has] <- valid_match_q2[cbind(p32s[has], pmax(l1s[has], 1L))]
q2v[has] <- q2v[has] | (l2s[has] > 0L & valid_match_q2[cbind(p32s[has], pmax(l2s[has], 1L))])
q2v[has] <- q2v[has] | (l3s[has] > 0L & valid_match_q2[cbind(p32s[has], pmax(l3s[has], 1L))])

q2_by_group |>
  mutate(
    pueblos_label = pueblos_cats$label[match(p32_pueblos, pueblos_cats$code)],
    albo_q2       = q2v
  ) |>
  group_by(pueblos_label) |>
  summarise(
    n        = n(),
    n_q2     = sum(albo_q2),
    pct_q2   = round(100 * n_q2 / n, 1),
    .groups  = "drop"
  ) |>
  filter(!is.na(pueblos_label),
         !pueblos_label %in% c("No se autoidentifica", "Sin respuesta")) |>
  arrange(desc(n)) |>
  print(n = 60)
```

------------------------------------------------------------------------

## Q3: Childhood language acquisition

albo_q3 captures whether the respondent learned an indigenous language as a
child (`p341_idiomat_cod` not NA, not 6 = Castellano, not 999 = Sin respuesta).

```{r}
#| label: q3-overall
# Among self-identifiers: what share learned an indigenous language as a child?
albo_vars |>
  group_by(albo_q1) |>
  summarise(
    n      = n(),
    pct_q3 = round(100 * mean(albo_q3), 1),
    .groups = "drop"
  ) |>
  mutate(albo_q1 = if_else(albo_q1, "Self-identifies", "Does not identify"))
```

```{r}
#| label: q2-vs-q3
# How do Q2 (current language match) and Q3 (childhood language) diverge?
albo_vars |>
  filter(albo_q1) |>
  count(albo_q2, albo_q3) |>
  mutate(
    pct      = round(100 * n / sum(n), 1),
    label_q2 = if_else(albo_q2, "Speaks identity language", "Does not speak identity language"),
    label_q3 = if_else(albo_q3, "Learned indigenous\nlanguage as child", "Did not learn\nas child")
  ) |>
  select(label_q2, label_q3, n, pct) |>
  arrange(desc(n))
```

The most notable cell is self-identifiers who **learned** an indigenous language
as a child (Q3 = TRUE) but **no longer** speak the corresponding identity
language daily (Q2 = FALSE): this is the clearest signal of post-childhood
language shift — a person who was raised in an indigenous-language household
but has since shifted to Spanish.

------------------------------------------------------------------------

## Castellano (albo_c) by group

```{r}
#| label: castellano-by-type
# Castellano use by Albó typology
albo_vars |>
  group_by(albo_type) |>
  summarise(
    n      = n(),
    pct_c  = round(100 * mean(albo_c), 1),
    .groups = "drop"
  )
```

The share speaking Castellano is near-universal (>97%) even among self-
identifiers who speak their identity language and learned it as a child. The
only group where Castellano use is notably lower is "Identity + language +
childhood" — i.e., people with the deepest indigenous-language profile — which
still shows ~83% Castellano use, consistent with widespread functional
bilingualism.

------------------------------------------------------------------------

------------------------------------------------------------------------

## Condición Étnico-Lingüística (CEL)

Albó and Romero defined a seven-level ordinal scale — the *Condición
Étnico-Lingüística* (CEL) — combining the three questions and Castellano use
into a single indicator of ethnic-linguistic position. Two additional half-step
levels (1.5 and 4.5) capture empirically common combinations not in the
original scale.

### Scale definition

| CEL | Q1 | Q2 | Q3 | C | Description |
|-----|----|----|----|----|-------------|
| 7 | S | S | S | N | Pertenece, habla, aprendió — sin castellano |
| 6 | S | S | S | S | Pertenece, habla, aprendió — con castellano |
| 5 | S | S | N | S | Pertenece, habla, no aprendió |
| 4.5 | S | N | S | S | Pertenece, no habla, pero aprendió como niño |
| 4 | S | N | N | S | Pertenece, no habla, no aprendió |
| 3 | N | S | S | N | No pertenece, pero habla y aprendió — sin castellano |
| 2 | N | S | S | S | No pertenece, pero habla y aprendió — con castellano |
| 1 | N | S | N | S | No pertenece, habla, no aprendió |
| 1.5 | N | N | S | S | No pertenece, no habla, pero aprendió como niño |
| 0 | N | N | N | S | No pertenece, no habla, no aprendió |

**Note on Q2 for non-identifiers**: For people who self-identify (Q1 = TRUE),
Q2 tests whether their spoken language matches their declared identity group
(per the `albo_q2` crosswalk). For non-identifiers (Q1 = FALSE), there is no
declared identity to match against, so Q2 reduces to: *does the person speak
any Bolivian indigenous language* (idioma codes 1–5 or 7–37)? This is captured
in the `cel_q2` column. `albo_q2` and `cel_q2` are identical for identifiers;
they differ only for non-identifiers.

### Compute CEL

```{r}
#| label: cel-compute
# cel_q2 and cel are pre-computed in albo_vars.rds.
# The CEL assignment logic:
#
#   cel_q2  = albo_q2                          (for identifiers, Q1 = TRUE)
#           = speaks any indigenous language    (for non-identifiers, Q1 = FALSE)
#             (codes 1–5 or 7–37 in p331/p332/p333)
#
#   cel = case_when(
#      albo_q1 &  cel_q2 &  albo_q3 & !albo_c  ~  7,
#      albo_q1 &  cel_q2 &  albo_q3 &  albo_c  ~  6,
#      albo_q1 &  cel_q2 & !albo_q3             ~  5,
#      albo_q1 & !cel_q2 &  albo_q3             ~  4.5,
#      albo_q1 & !cel_q2 & !albo_q3             ~  4,
#     !albo_q1 &  cel_q2 &  albo_q3 & !albo_c  ~  3,
#     !albo_q1 &  cel_q2 &  albo_q3 &  albo_c  ~  2,
#     !albo_q1 &  cel_q2 & !albo_q3             ~  1,
#     !albo_q1 & !cel_q2 &  albo_q3             ~  1.5,
#     !albo_q1 & !cel_q2 & !albo_q3             ~  0
#   )

albo_vars |>
  count(cel) |>
  mutate(pct = round(100 * n / sum(n), 2)) |>
  arrange(cel)
```

### CEL distribution

```{r}
#| label: cel-plot
#| fig-height: 5
cel_labels <- c(
  "0"   = "0 — No identity, no language",
  "1"   = "1 — No identity, speaks (no childhood)",
  "1.5" = "1.5 — No identity, childhood language only",
  "2"   = "2 — No identity, speaks + childhood (bilingual)",
  "3"   = "3 — No identity, speaks + childhood (monolingual)",
  "4"   = "4 — Identity only",
  "4.5" = "4.5 — Identity + childhood (no current language)",
  "5"   = "5 — Identity + language (no childhood)",
  "6"   = "6 — Identity + language + childhood (bilingual)",
  "7"   = "7 — Identity + language + childhood (monolingual)"
)

cel_colors <- c(
  "0"   = "#d9d9d9",
  "1"   = "#fee0b6",
  "1.5" = "#fdb863",
  "2"   = "#e08214",
  "3"   = "#b35806",
  "4"   = "#cce5ff",
  "4.5" = "#74add1",
  "5"   = "#4393c3",
  "6"   = "#2166ac",
  "7"   = "#053061"
)

albo_vars |>
  count(cel) |>
  mutate(
    pct      = n / sum(n),
    cel_chr  = as.character(cel),
    cel_lbl  = cel_labels[cel_chr],
    cel_lbl  = fct_reorder(cel_lbl, cel)
  ) |>
  ggplot(aes(x = pct, y = cel_lbl, fill = cel_chr)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.14))) +
  scale_fill_manual(values = cel_colors) +
  labs(
    title    = "Condición Étnico-Lingüística (CEL) — Bolivia CPV 2024",
    subtitle = "n = 11,365,333",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11)
```

### CEL by urban/rural

```{r}
#| label: cel-urban-rural
#| fig-height: 6
urbrur_labels <- c("1" = "Urban", "2" = "Rural")

albo_vars |>
  mutate(urbrur_label = urbrur_labels[as.character(urbrur)]) |>
  filter(!is.na(urbrur_label)) |>
  count(urbrur_label, cel) |>
  group_by(urbrur_label) |>
  mutate(
    pct     = n / sum(n),
    cel_chr = as.character(cel),
    cel_lbl = cel_labels[cel_chr],
    cel_lbl = fct_reorder(cel_lbl, cel)
  ) |>
  ungroup() |>
  ggplot(aes(x = pct, y = cel_lbl, fill = cel_chr)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.16))) +
  scale_fill_manual(values = cel_colors) +
  facet_wrap(~ urbrur_label) +
  labs(
    title    = "CEL by urban/rural",
    subtitle = "Bolivia CPV 2024",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

### CEL by department

```{r}
#| label: cel-dept-total
#| fig-height: 7
dept_labels <- c(
  "1" = "Chuquisaca", "2" = "La Paz",    "3" = "Cochabamba",
  "4" = "Oruro",      "5" = "Potosí",    "6" = "Tarija",
  "7" = "Santa Cruz", "8" = "Beni",      "9" = "Pando"
)

albo_vars |>
  mutate(dept = dept_labels[as.character(idep)]) |>
  count(dept, cel) |>
  group_by(dept) |>
  mutate(
    pct     = n / sum(n),
    cel_chr = as.character(cel),
    dept    = fct_reorder(dept, pct * (cel %in% c(6, 7)), sum)
  ) |>
  ungroup() |>
  ggplot(aes(x = pct, y = dept, fill = cel_chr)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cel_colors, labels = cel_labels, name = "CEL") +
  labs(
    title    = "Indigenous ethno-linguistic identity by department",
    subtitle = "Bolivia CPV 2024",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.key.width = unit(0.6, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))
```

```{r}
#| label: cel-mun-top30
#| fig-height: 10
mun_names_lookup <- ine_geog_2013 |>
  distinct(cod.dep, PRO, MUN, department, municipality) |>
  mutate(across(c(cod.dep, PRO, MUN), as.integer))

cel_geo |>
  filter(!is.na(municipality), is.na(urban_rural)) |>
  left_join(mun_names_lookup,
    by = c("department" = "cod.dep", "province" = "PRO", "municipality" = "MUN")
  ) |>
  arrange(desc(n_total)) |>
  head(30) |>
  pivot_longer(
    cols      = c(cel_0, cel_1, cel_1.5, cel_2, cel_3, cel_4, cel_4.5, cel_5, cel_6, cel_7),
    names_to  = "cel_chr",
    values_to = "pct"
  ) |>
  mutate(
    cel_chr   = sub("cel_", "", cel_chr),
    cel_val   = as.numeric(cel_chr),
    pct       = pct / 100,
    bar_label = fct_reorder(municipality.y, n_total)
  ) |>
  ggplot(aes(x = pct, y = bar_label, fill = cel_chr)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cel_colors, labels = cel_labels_en, name = "CEL") +
  labs(
    title    = "Indigenous ethno-linguistic identity \u2014 30 most populous municipalities",
    subtitle = "Bolivia CPV 2024",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.key.width = unit(0.6, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))
```

```{r}
#| label: cel-mun-top30-2001
#| fig-height: 10
# Name crosswalk: 2001 abbreviations → 2024 canonical names
name_xwalk_2001 <- c(
  "Sta Cruz de la Sierra" = "Santa Cruz de la Sierra",
  "S Ignacio de Velasco"  = "San Ignacio de Velasco",
  "Sipe Sipe"             = "Sipesipe"
)

# CEL column → level mapping (no half-steps in 2001 data)
cel_cols_2001 <- c(
  cel_0_nnn  = "0",
  cel_1_nsn  = "1",
  cel_2_nss  = "2",
  cel_3_nss  = "3",
  cel_4_snn  = "4",
  cel_5_ssnc = "5",
  cel_6_sssc = "6",
  cel_7_sss  = "7"
)

# Y-axis order: same 2024 population ranking (ascending = most populous at top)
top30_order <- top30 |> arrange(n_total) |> pull(mun_name)

albo_romero |>
  mutate(
    mun_std = if_else(
      municipality %in% names(name_xwalk_2001),
      name_xwalk_2001[municipality],
      municipality
    )
  ) |>
  filter(mun_std %in% top30_order) |>
  select(mun_std, all_of(names(cel_cols_2001))) |>
  pivot_longer(
    cols      = all_of(names(cel_cols_2001)),
    names_to  = "var",
    values_to = "pct"
  ) |>
  mutate(
    cel_chr   = factor(cel_cols_2001[var], levels = as.character(0:7)),
    pct       = pct / 100,
    bar_label = factor(mun_std, levels = top30_order)
  ) |>
  ggplot(aes(x = pct, y = bar_label, fill = cel_chr)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = cel_colors[as.character(0:7)],
    labels = cel_labels_en[as.character(0:7)],
    name   = "CEL"
  ) +
  labs(
    title    = "Indigenous ethno-linguistic identity \u2014 30 most populous municipalities (2024 ranking)",
    subtitle = "Bolivia Census 2001 (Alb\u00f3 & Romero)",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.key.width = unit(0.6, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))
```

```{r}
#| label: cel-dept
#| fig-height: 9
dept_labels <- c(
  "1" = "Chuquisaca", "2" = "La Paz",    "3" = "Cochabamba",
  "4" = "Oruro",      "5" = "Potosí",    "6" = "Tarija",
  "7" = "Santa Cruz", "8" = "Beni",      "9" = "Pando"
)

cel_labels_en <- c(
  "0"   = "0 \u2014 No identity, no language",
  "1"   = "1 \u2014 No identity, speaks (no childhood)",
  "1.5" = "1.5 \u2014 No identity, childhood only",
  "2"   = "2 \u2014 No identity, speaks + childhood (bilingual)",
  "3"   = "3 \u2014 No identity, speaks + childhood (monolingual)",
  "4"   = "4 \u2014 Identity only",
  "4.5" = "4.5 \u2014 Identity + childhood (no current language)",
  "5"   = "5 \u2014 Identity + language (no childhood)",
  "6"   = "6 \u2014 Identity + language + childhood (bilingual)",
  "7"   = "7 \u2014 Identity + language + childhood (monolingual)"
)

cel_labels_es <- c(
  "0"   = "0 \u2014 Sin identidad, sin lengua",
  "1"   = "1 \u2014 Sin identidad, habla (sin infancia)",
  "1.5" = "1.5 \u2014 Sin identidad, solo aprendió de niño",
  "2"   = "2 \u2014 Sin identidad, habla + aprendió (bilingüe)",
  "3"   = "3 \u2014 Sin identidad, habla + aprendió (monolingüe)",
  "4"   = "4 \u2014 Solo identidad",
  "4.5" = "4.5 \u2014 Identidad + aprendió (sin lengua actual)",
  "5"   = "5 \u2014 Identidad + lengua (sin infancia)",
  "6"   = "6 \u2014 Identidad + lengua + aprendió (bilingüe)",
  "7"   = "7 \u2014 Identidad + lengua + aprendió (monolingüe)"
)

# Rank departments by overall share at CEL 6–7 (most indigenous = highest rank)
dept_rank <- albo_vars |>
  mutate(dept = dept_labels[as.character(idep)]) |>
  count(dept, cel) |>
  group_by(dept) |>
  mutate(pct = n / sum(n)) |>
  summarise(score = sum(pct * (cel %in% c(6, 7))), .groups = "drop") |>
  arrange(score) |>
  mutate(dept_rank = row_number())

# Shared plot data: department × urban/rural
cel_dept_df <- albo_vars |>
  mutate(dept = dept_labels[as.character(idep)]) |>
  count(dept, urbrur, cel) |>
  group_by(dept, urbrur) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  left_join(dept_rank |> select(dept, dept_rank), by = "dept") |>
  mutate(
    urb_order = if_else(urbrur == 1L, 0L, 1L),   # Urban=0, Rural=1
    y_pos     = dept_rank * 2 - urb_order,
    cel_chr   = as.character(cel)
  )

# English chart
cel_dept_df |>
  mutate(bar_label = fct_reorder(
    paste0(dept, " \u2013 ", if_else(urbrur == 1, "Urban", "Rural")), y_pos
  )) |>
  ggplot(aes(x = pct, y = bar_label, fill = cel_chr)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cel_colors, labels = cel_labels_en, name = "CEL") +
  labs(
    title    = "Indigenous ethno-linguistic identity by department and urban/rural",
    subtitle = "Bolivia CPV 2024",
    x = "Share of population", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.key.width = unit(0.6, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))
```

```{r}
#| label: cel-dept-es
#| fig-height: 9
# Spanish chart (reuses cel_dept_df and dept_rank from chunk above)
cel_dept_df |>
  mutate(bar_label = fct_reorder(
    paste0(dept, " \u2013 ", if_else(urbrur == 1, "Urbano", "Rural")), y_pos
  )) |>
  ggplot(aes(x = pct, y = bar_label, fill = cel_chr)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cel_colors, labels = cel_labels_es, name = "CEL") +
  labs(
    title    = "Identidad etno-ling\u00fcística ind\u00edgena por departamento y área",
    subtitle = "Bolivia CPV 2024",
    x = "Proporción de la población", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.key.width = unit(0.6, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))
```

### CEL municipality groups (Albó-Romero)

Albó and Romero further classified municipalities into nine groups (A–I) based
on the share of their population at three CEL thresholds: `cel_5plus` (CEL ≥ 5),
`cel_4plus` (CEL ≥ 4), and `cel_2plus` (CEL ≥ 2). Each municipality is assigned
to the first group whose thresholds it meets.

| Group | CEL ≥ 5 | CEL ≥ 4 | CEL ≥ 2 |
|-------|---------|---------|---------|
| A | ≥ 90% | ≥ 90% | ≥ 90% |
| B | > 67% | > 80% | > 90% |
| C | > 50% | > 67% | > 67% |
| D | > 33% | > 50% | > 67% |
| E | — | — | > 75% |
| F | — | — | > 50% |
| G | — | — | > 33% |
| H | — | — | > 10% |
| I | < 10% in all categories | | |

```{r}
#| label: cel-mun-groups
dept_names <- c(
  "1" = "Chuquisaca", "2" = "La Paz",   "3" = "Cochabamba",
  "4" = "Oruro",      "5" = "Potosí",   "6" = "Tarija",
  "7" = "Santa Cruz", "8" = "Beni",     "9" = "Pando"
)

albo_groups <- cel_geo |>
  filter(!is.na(municipality), is.na(urban_rural)) |>
  select(department, province, municipality, cel_2plus, cel_4plus, cel_5plus) |>
  mutate(
    albo_group = case_when(
      cel_5plus >= 90 & cel_4plus >= 90 & cel_2plus >= 90 ~ "A",
      cel_5plus >  67 & cel_4plus >  80 & cel_2plus >  90 ~ "B",
      cel_5plus >  50 & cel_4plus >  67 & cel_2plus >  67 ~ "C",
      cel_5plus >  33 & cel_4plus >  50 & cel_2plus >  67 ~ "D",
      cel_2plus >  75                                      ~ "E",
      cel_2plus >  50                                      ~ "F",
      cel_2plus >  33                                      ~ "G",
      cel_2plus >  10                                      ~ "H",
      TRUE                                                 ~ "I"
    )
  )

# Wide table: groups × departments
all_groups <- LETTERS[1:9]
dept_cols  <- as.character(dept_names)
total_muns <- nrow(albo_groups)

wide <- albo_groups |>
  mutate(dept_name = dept_names[as.character(department)]) |>
  count(albo_group, dept_name) |>
  tidyr::pivot_wider(names_from = dept_name, values_from = n, values_fill = 0L)

wide <- tibble(albo_group = all_groups) |>
  left_join(wide, by = "albo_group") |>
  mutate(across(all_of(dept_cols), ~ replace_na(., 0L))) |>
  mutate(
    Total      = rowSums(pick(all_of(dept_cols)), na.rm = TRUE),
    Pct        = sprintf("%.1f%%", Total / total_muns * 100),
    Cumulative = cumsum(Total)
  )

total_row <- tibble(albo_group = "Total") |>
  bind_cols(
    wide |> summarise(across(all_of(dept_cols), sum)),
    tibble(Total = sum(wide$Total), Pct = "", Cumulative = as.double(NA))
  )

tbl_data <- bind_rows(wide, total_row)

tbl_data |>
  mutate(
    across(
      all_of(dept_cols),
      ~ if_else(albo_group != "Total" & (is.na(.) | . == 0L), NA_integer_, as.integer(.))
    ),
    across(all_of(dept_cols), ~ if_else(is.na(.), "", as.character(.))),
    Cumulative = if_else(is.na(Cumulative), "", as.character(as.integer(Cumulative))),
    Total      = if_else(Total == 0, "", as.character(as.integer(Total))),
    Pct        = if_else(Total == "" | Total == "0", "", Pct)
  ) |>
  rename(Group = albo_group, `%` = Pct, Cum. = Cumulative) |>
  knitr::kable(align = c("l", rep("r", length(dept_cols) + 3)))
```

------------------------------------------------------------------------

## Notes on variable limitations

- **albo_q2 for generic identities** (Otras declaraciones, Quechua-Aymara, Más
  de una descripción) and **groups with no distinct census language**
  (Afroboliviano, Chuwi, Paunaca, Toromona, foreign groups): counted as TRUE
  if the respondent speaks *any* Bolivian indigenous language (codes 1–5 or
  7–37). Afroboliviano in particular refers to a Spanish variety of the La Paz
  Yungas, not a structurally distinct language; including it on the "any
  indigenous language" rule means Afroboliviano identifiers who are also
  Quechua or Aymara speakers will score Q2 = TRUE.

- **p341_idiomat_cod** (childhood language, Q3): code 999 is treated as
  non-response and excluded (same as NA). Code 6 = Castellano is excluded
  (learning Spanish as a child does not count as learning an indigenous
  language).

- **albo_c** captures *current* Castellano use via p331/p332/p333. It does not
  capture whether someone learned Spanish as a child or speaks it fluently;
  only that they list it as one of up to three current languages.

- **urbrur** is sourced from the Vivienda (household) file and joined on
  `idep + iprov + imun + i00`. All 11,365,333 person rows matched exactly
  (0 NAs).
