---
title: "Bolivia Municipalities — Flag & Coat of Arms Wikidata Review"
subtitle: "Record of P41 (flag image) and P94 (coat of arms image) additions to Wikidata — data frozen July 2026"
date: "2026-07-20"
format:
  html:
    embed-resources: false
    toc: true
    toc-depth: 2
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(purrr)
library(htmltools)

d <- readRDS(here::here("data", "wd_flag_coat_review.rds"))
list2env(d, envir = .GlobalEnv)
```

```{r amdecruz-lookup}
# Circular flag images scraped from https://amdecruz.com/asociados.php
# Only Santa Cruz department municipalities; sc4.svg (generic dept. logo) excluded.
amdecruz_lookup <- tibble::tribble(
  ~muni_name,                  ~amdecruz_url,
  "Buena Vista",               "https://amdecruz.com/images/municipios/Buenavista/buenavista4.svg",
  "Camiri",                    "https://amdecruz.com/images/municipios/Camiri/camiri4.svg",
  "Cotoca",                    "https://amdecruz.com/images/municipios/Cotoca/Cotoca4.svg",
  "El Carmen Rivero Tórrez",   "https://amdecruz.com/images/municipios/ElCarmenRiveroT/carmenrivero4.svg",
  "La Guardia",                "https://amdecruz.com/images/municipios/Laguardia/laguardia4.svg",
  "Mairana",                   "https://amdecruz.com/images/municipios/Mairana/mairana4.svg",
  "Mineros",                   "https://amdecruz.com/images/municipios/Minero/Minero4.svg",
  "Montero",                   "https://amdecruz.com/images/municipios/montero/monteroR.svg",
  "Pailón",                    "https://amdecruz.com/images/municipios/Pailon/Pailon4.svg",
  "Puerto Quijarro",           "https://amdecruz.com/images/municipios/PuertoQuijarro/puertoquijarro4.svg",
  "Puerto Suárez",             "https://amdecruz.com/images/municipios/Puertosuarez/PuertoSuarez4.svg",
  "Roboré",                    "https://amdecruz.com/images/municipios/Robore/robore4.svg",
  "San Carlos",                "https://amdecruz.com/images/municipios/SanCarlos/sancarlos4.svg",
  "San Juan de Yapacaní",      "https://amdecruz.com/images/municipios/SanJuan/sanjuan5.svg",
  "San Julián",                "https://amdecruz.com/images/municipios/SanJulian/sanjulian4.svg"
)
```

```{r helpers}
# ── URL builders ──────────────────────────────────────────────────────────────

# Thumbnail URL via Special:Redirect (avoids CORS issues with direct thumb URLs)
commons_thumb <- function(file, width = 120) {
  enc <- utils::URLencode(file, reserved = TRUE)
  paste0(
    "https://commons.wikimedia.org/w/index.php",
    "?title=Special:Redirect/file/", enc,
    "&width=", width
  )
}

# File description page on Commons
commons_link <- function(file) {
  enc <- utils::URLencode(file, reserved = TRUE)
  paste0("https://commons.wikimedia.org/wiki/File:", enc)
}

# Wikidata item page
wd_link <- function(qid) paste0("https://www.wikidata.org/wiki/", qid)

# ── Review table builder ──────────────────────────────────────────────────────
# Builds a full section: controls bar + scrollable table.
# section: short ID used in CSS class names and element IDs ("flag" or "coat")
# prop_id: Wikidata property being set ("P41" or "P94")

make_review_section <- function(data, section, prop_id, amdecruz_lkp = NULL) {
  n <- nrow(data)

  # Escape double-quotes inside HTML data attribute values
  attr_esc <- function(x) gsub('"', "&quot;", x, fixed = TRUE)

  # Pre-index amdecruz lookup for fast lookup by muni_name
  amdecruz_map <- if (!is.null(amdecruz_lkp)) {
    setNames(amdecruz_lkp$amdecruz_url, amdecruz_lkp$muni_name)
  } else NULL

  rows_html <- map_chr(seq_len(n), function(i) {
    row       <- data[i, ]
    qs_esc    <- attr_esc(row$quick_statement)
    thumb     <- commons_thumb(row$file)
    file_page <- commons_link(row$file)
    item_page <- wd_link(row$qid)
    fname_esc <- htmlEscape(row$file)

    # Build optional AMDECRUZ image cell
    amdecruz_cell <- if (!is.null(amdecruz_map)) {
      amd_url <- amdecruz_map[row$muni_name]
      if (!is.na(amd_url)) {
        paste0(
          '<td style="text-align:center;vertical-align:middle;padding:4px;width:90px">',
            '<a href="https://amdecruz.com/asociados.php" target="_blank" rel="noreferrer"',
               ' title="Source: amdecruz.com">',
              '<img src="', amd_url, '"',
              ' style="max-height:80px;max-width:80px;display:block;margin:auto"',
              ' loading="lazy"',
              ' onerror="this.replaceWith(document.createTextNode(\'—\'))">',
            '</a>',
          '</td>'
        )
      } else {
        '<td style="text-align:center;vertical-align:middle;padding:4px;width:90px;color:#aaa;font-size:11px">—</td>'
      }
    } else ""

    paste0(
      '<tr>',
        # Checkbox
        '<td style="text-align:center;vertical-align:middle;padding:6px 10px;width:36px">',
          '<input type="checkbox" class="', section, '-check"',
          ' data-qs="', qs_esc, '"',
          ' checked',
          ' onchange="updateCount(\'', section, '\')">',
        '</td>',
        # Image from Commons
        '<td style="text-align:center;vertical-align:middle;padding:4px;width:120px">',
          '<a href="', file_page, '" target="_blank" rel="noreferrer">',
            '<img src="', thumb, '"',
            ' style="max-height:90px;max-width:110px;display:block;margin:auto"',
            ' loading="lazy"',
            ' onerror="this.replaceWith(document.createTextNode(\'(no preview)\'))">',
          '</a>',
        '</td>',
        # AMDECRUZ circular image (only when lookup provided)
        amdecruz_cell,
        # Municipality → Wikidata link
        '<td style="vertical-align:middle;padding:6px 10px">',
          '<a href="', item_page, '" target="_blank" rel="noreferrer">',
            htmlEscape(row$muni_name),
          '</a>',
        '</td>',
        # Department
        '<td style="vertical-align:middle;padding:6px 10px;color:#555">',
          htmlEscape(row$department),
        '</td>',
        # Commons filename → file page link
        '<td style="vertical-align:middle;padding:6px 10px;font-size:11px;',
            'font-family:monospace;max-width:250px;word-break:break-all">',
          '<a href="', file_page, '" target="_blank" rel="noreferrer">',
            fname_esc,
          '</a>',
        '</td>',
      '</tr>'
    )
  })

  # Controls bar
  controls <- paste0(
    '<div style="display:flex;align-items:center;gap:10px;flex-wrap:wrap;',
         'margin-bottom:10px;padding:8px;background:#f6f8fa;border-radius:6px;',
         'border:1px solid #e1e4e8">',
      '<button class="btn btn-sm btn-outline-secondary"',
        ' onclick="selectSection(\'', section, '\',true)">Select all</button>',
      '<button class="btn btn-sm btn-outline-secondary"',
        ' onclick="selectSection(\'', section, '\',false)">Deselect all</button>',
      '<span id="', section, '-count" style="color:#586069;font-size:13px">',
        n, ' selected',
      '</span>',
      '<button id="', section, '-copy-btn"',
        ' class="btn btn-sm btn-primary"',
        ' style="margin-left:auto"',
        ' onclick="copySection(\'', section, '\')">',
        'Copy selected QuickStatements',
      '</button>',
    '</div>'
  )

  # Header: insert AMDECRUZ column only when lookup provided
  amdecruz_th <- if (!is.null(amdecruz_lkp)) {
    '<th>AMDECRUZ <span style="font-weight:normal;font-size:11px;color:#586069">(circular)</span></th>'
  } else ""

  table_html <- paste0(
    '<div style="overflow-x:auto">',
    '<table class="table table-sm table-hover" style="font-size:13px;min-width:700px">',
    '<thead class="table-light"><tr>',
      '<th></th>',
      '<th>Commons image</th>',
      amdecruz_th,
      '<th>Municipality</th>',
      '<th>Department</th>',
      '<th>', prop_id, ' value (Commons filename)</th>',
    '</tr></thead>',
    '<tbody>',
    paste(rows_html, collapse = ""),
    '</tbody></table></div>'
  )

  HTML(paste0(controls, table_html))
}

# ── Overlap / info table (no checkboxes) ────────────────────────────────────
make_info_table <- function(data, file_col) {
  if (nrow(data) == 0) return(HTML("<p><em>None.</em></p>"))

  rows_html <- map_chr(seq_len(nrow(data)), function(i) {
    row       <- data[i, ]
    file      <- row[[file_col]]
    thumb     <- commons_thumb(file, width = 80)
    file_page <- commons_link(file)
    item_page <- wd_link(row$qid)

    paste0(
      '<tr>',
        '<td style="text-align:center;padding:4px;width:100px">',
          '<a href="', file_page, '" target="_blank" rel="noreferrer">',
            '<img src="', thumb, '"',
            ' style="max-height:70px;max-width:90px;display:block;margin:auto"',
            ' loading="lazy"',
            ' onerror="this.replaceWith(document.createTextNode(\'(no preview)\'))">',
          '</a>',
        '</td>',
        '<td style="vertical-align:middle;padding:6px 10px">',
          '<a href="', item_page, '" target="_blank" rel="noreferrer">',
            htmlEscape(row$muni_name),
          '</a>',
        '</td>',
        '<td style="vertical-align:middle;padding:6px 10px;color:#555">',
          htmlEscape(row$department),
        '</td>',
        '<td style="vertical-align:middle;padding:6px 10px;font-size:11px;',
            'font-family:monospace;max-width:300px;word-break:break-all">',
          '<a href="', file_page, '" target="_blank" rel="noreferrer">',
            htmlEscape(file),
          '</a>',
        '</td>',
      '</tr>'
    )
  })

  HTML(paste0(
    '<div style="overflow-x:auto">',
    '<table class="table table-sm table-hover" style="font-size:13px">',
    '<thead class="table-light"><tr>',
      '<th>Image</th><th>Municipality</th><th>Department</th><th>Commons filename</th>',
    '</tr></thead>',
    '<tbody>', paste(rows_html, collapse = ""), '</tbody>',
    '</table></div>'
  ))
}
```

```{r javascript, results='asis'}
# Inline JavaScript for checkbox interactivity
cat('<script>
function updateCount(section) {
  var n = document.querySelectorAll("." + section + "-check:checked").length;
  var el = document.getElementById(section + "-count");
  if (el) el.textContent = n + " selected";
}

function selectSection(section, checked) {
  document.querySelectorAll("." + section + "-check").forEach(function(cb) {
    cb.checked = checked;
  });
  updateCount(section);
}

function copySection(section) {
  var checked = document.querySelectorAll("." + section + "-check:checked");
  if (checked.length === 0) { alert("Nothing selected."); return; }
  var lines = [];
  checked.forEach(function(cb) { lines.push(cb.dataset.qs); });
  var text = lines.join("\\n");
  var btn = document.getElementById(section + "-copy-btn");

  function flashSuccess() {
    btn.textContent = "\\u2713 Copied " + lines.length + " statement" +
      (lines.length === 1 ? "" : "s") + "!";
    btn.classList.replace("btn-primary", "btn-success");
    setTimeout(function() {
      btn.textContent = "Copy selected QuickStatements";
      btn.classList.replace("btn-success", "btn-primary");
    }, 3000);
  }

  if (navigator.clipboard && window.isSecureContext) {
    navigator.clipboard.writeText(text).then(flashSuccess).catch(function() {
      fallbackCopy(text, flashSuccess);
    });
  } else {
    fallbackCopy(text, flashSuccess);
  }
}

function fallbackCopy(text, callback) {
  var ta = document.createElement("textarea");
  ta.value = text;
  ta.style.cssText = "position:fixed;left:-9999px;top:-9999px";
  document.body.appendChild(ta);
  ta.focus(); ta.select();
  document.execCommand("copy");
  document.body.removeChild(ta);
  if (callback) callback();
}

document.addEventListener("DOMContentLoaded", function() {
  ["flag", "coat"].forEach(function(s) { updateCount(s); });
});
</script>')
```

## Status

The Wikidata statements shown in this document have been submitted. The review tables below are a permanent record of what was added and can be used to identify and correct any errors.

The statements were added in three QuickStatements batches on July 20, 2026:

- [Batch 42829](https://qs-dev.toolforge.org/batch/42829/)
- [Batch 42830](https://qs-dev.toolforge.org/batch/42830/)
- [Batch 42831](https://qs-dev.toolforge.org/batch/42831/)

All images load directly from Wikimedia Commons — an internet connection is required.

::: {.callout-note}
## Data sources

- **Additions** — cross-check of [Category:Flags of municipalities of Bolivia](https://commons.wikimedia.org/wiki/Category:Flags_of_municipalities_of_Bolivia) and [Category:Coats of arms of municipalities of Bolivia](https://commons.wikimedia.org/wiki/Category:Coats_of_arms_of_municipalities_of_Bolivia) against existing Wikidata P41/P94 claims for the `Q1062710` (Municipality of Bolivia) class.
- **Matching** — filenames decoded from Wikidata SPARQL; Commons filenames matched to municipalities by name extraction + fuzzy string match. A few ambiguous files were excluded (see [Unmatched files] below).
- **Data frozen** — July 20, 2026. The source file `data/wd_flag_coat_review.rds` reflects the state of Commons and Wikidata at that date and should not be refreshed, as the statements have already been submitted.
:::

## Impact

The goal of this review is to improve the coverage of municipal flags and coats of arms in Wikidata, which in turn will allow the generated infoboxes on English Wikipedia to include these images. If fully implemented (and you need to check for errors first):


Of the 195 Bolivian municipality articles currently on English Wikipedia, only 6 include a flag image and 1 includes a seal in the infobox. The generated infoboxes produced by bolivia-muni-infobox-en.qmd would cover all 340 municipalities, and by pulling images already linked in Wikidata, 107 of those would include a flag and 3 would include a coat of arms — a dramatic improvement in coverage, though 65 of those 107 flags correspond to municipalities that already have a Wikipedia article (the remaining 42 would be on new pages). If the new Commons-matched images are first added to Wikidata via the QuickStatements outputs, the generated infoboxes would expand further to 175 municipalities with flags and 41 with coats of arms — representing 22 of the 195 existing articles gaining a coat for the first time, and 96 total existing articles with a flag.


---

## Flags (P41) — `r nrow(review_flags)` additions {#sec-flags}

These `r nrow(review_flags)` municipalities had a flag image in the Commons category but no P41 claim in Wikidata. The QuickStatement added the Commons filename as the value of P41 (flag image).

::: {.callout-warning}
## Known errors in this batch

The following entries were submitted but are incorrect and should be fixed manually in Wikidata:

- **San Lorenzo** — wrong department (file belongs to a different San Lorenzo)
- **Apolo** — wrong image assigned
- **Camiri** — image is missing the sun symbol; a corrected file should be substituted
- **Tiquipaya** — flag is upside down; the correct orientation is at [File:Bandera-Tiquipaya.png](https://commons.wikimedia.org/wiki/File:Bandera-Tiquipaya.png)
:::

```{r flag-table, results='asis'}
make_review_section(review_flags, "flag", "P41", amdecruz_lkp = amdecruz_lookup)
```

---

## Coats of Arms (P94) — `r nrow(review_coats)` additions {#sec-coats}

These `r nrow(review_coats)` municipalities had a coat of arms / escudo in the Commons category but no P94 claim in Wikidata.

::: {.callout-warning}
## Known errors in this batch

- **San Ramón (Santa Cruz)** — [File:Escudo San Ramón, Beni.jpg](https://commons.wikimedia.org/wiki/File:Escudo_San_Ramón,_Beni.jpg) was mistakenly linked to the municipality of San Ramón in Santa Cruz department; it belongs to San Ramón in Beni. The P94 claim on the Santa Cruz item should be removed and the file correctly assigned to the Beni item.

Several other images in this batch have low-quality or unofficial-looking filenames (e.g. `ESCUDO OFICIAL-GOBIERNO AUTONOMO...`, `FB IMG escudo de colquechaca.jpg`) and may warrant further review.
:::

```{r coat-table, results='asis'}
make_review_section(review_coats, "coat", "P94")
```

---

## Already linked in Wikidata {#sec-overlap}

These files appear in the Commons category **and** are already referenced by a Wikidata P41/P94 claim. No action needed — shown here for confirmation.

### Flags (P41) — `r nrow(flag_overlap)` confirmed

```{r flag-overlap, results='asis'}
make_info_table(flag_overlap, "flag_file")
```

### Coats of Arms (P94) — `r nrow(coat_overlap)` confirmed

```{r coat-overlap, results='asis'}
make_info_table(coat_overlap, "coat_file")
```

---

## Wikidata claims outside this Commons category {#sec-wd-only}

These municipalities already have a P41 or P94 claim in Wikidata, but the linked file is **not** in the Commons category we checked. They may be in a subcategory, use a different filename, or be in a different category entirely. No action needed here.

- **`r nrow(wd_flags_not_in_commons)`** municipalities have P41 pointing to a flag not in this Commons category.
- **`r nrow(wd_coats_not_in_commons)`** municipalities have P94 pointing to a coat not in this Commons category.

```{r wd-flags-outside, results='asis'}
make_info_table(wd_flags_not_in_commons, "flag_file")
```

---

## Unmatched Commons files {#sec-unmatched}

These files in the Commons categories could not be reliably matched to a specific municipality and were excluded from the QuickStatements. Manual assignment may be possible for some.

### Flags

```{r unmatched-flags, results='asis'}
flags_unmatched |>
  mutate(
    Image = paste0(
      '<a href="', commons_link(file), '" target="_blank" rel="noreferrer">',
        '<img src="', commons_thumb(file, 80), '"',
        ' style="max-height:60px" loading="lazy"',
        ' onerror="this.replaceWith(document.createTextNode(\'—\'))">',
      '</a>'
    )
  ) |>
  select(Image, file, reason) |>
  knitr::kable(
    col.names = c("", "File", "Reason not assigned"),
    escape    = FALSE,
    format    = "html"
  ) |>
  kableExtra::kable_styling(font_size = 12, full_width = FALSE)
```

### Coats of Arms

```{r unmatched-coats, results='asis'}
coats_unmatched |>
  mutate(
    Image = paste0(
      '<a href="', commons_link(file), '" target="_blank" rel="noreferrer">',
        '<img src="', commons_thumb(file, 80), '"',
        ' style="max-height:60px" loading="lazy"',
        ' onerror="this.replaceWith(document.createTextNode(\'—\'))">',
      '</a>'
    )
  ) |>
  select(Image, file, reason) |>
  knitr::kable(
    col.names = c("", "File", "Reason not assigned"),
    escape    = FALSE,
    format    = "html"
  ) |>
  kableExtra::kable_styling(font_size = 12, full_width = FALSE)
```
