---
title: "Bolivia Municipalities on Wikipedia: Coverage by Language"
author: "Carwil Bjork-James"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
execute:
  warning: false
  message: false
---

## Overview

This document maps the completeness of Wikipedia coverage for Bolivia's 336
official municipalities across four Wikipedias: **English**, **Spanish**,
**Quechua**, and **Aymara**. Coverage is defined as having a Wikipedia article
linked from Wikidata via sitelinks.

The maps are designed to support Wikipedia editors who want to identify
municipalities lacking articles in one or more languages. They connect to the
deeper article-quality analysis in
[`bolivia-municipality-wikipedia.qmd`](bolivia-municipality-wikipedia.qmd),
which examines article content and structure on English Wikipedia.

---

## Setup

```{r setup}
library(tidyverse)
library(sf)
library(patchwork)
library(httr)
library(jsonlite)

source("wikidata_presence_matrix.R")
source("get_wikidata_instances.R")
```

---

## Fetch data from Wikidata

### Wikipedia presence matrix

We query Wikidata for all items that are instances of
[municipality of Bolivia (Q1062710)](https://www.wikidata.org/wiki/Q1062710)
and build a presence matrix indicating which Wikipedia language editions have an
article for each municipality.

Results are cached to `data/bo_municipality_wiki_presence.rds` so that
subsequent renders do not repeat the API calls.

```{r fetch-presence}
#| cache: false

presence_cache <- "data/bo_municipality_wiki_presence.rds"

if (file.exists(presence_cache)) {
  bo_municipality_wiki_presence <- readRDS(presence_cache)
} else {
  bo_municipality_wiki_presence <- wikidata_instance_wikipedia_presence(
    class_qid  = "Q1062710",   # municipality of Bolivia
    languages  = NULL,         # auto-discover all languages present
    limit      = 500,
    batch_size = 20
  )
  saveRDS(bo_municipality_wiki_presence, presence_cache)
}
```

### INE codes (P14142)

We retrieve the INE municipal code (property
[P14142](https://www.wikidata.org/wiki/Property:P14142)) for each municipality.
These 6-digit codes (e.g. `010101`) allow us to join Wikidata entities to the
GADM boundary file via a pre-built crosswalk.

```{r fetch-ine-codes}
#| cache: false

ine_cache <- "data/municipalities_wd_ine.rds"

if (file.exists(ine_cache)) {
  municipalities_wd_ine <- readRDS(ine_cache)
} else {
  municipalities_wd_ine <- get_wikidata_instances(
    class_qid      = "Q1062710",
    property       = "P14142",
    property_names = "ine_code",
    languages      = c("en", "es"),
    limit          = 500,
    batch_size     = 20
  )
  saveRDS(municipalities_wd_ine, ine_cache)
}
```

---

## Build map dataset

We join the presence matrix to INE codes, then match to GADM boundaries through
the pre-built lookup table.

```{r build-map-data}
gadm_sf     <- st_read("data/gadm41_BOL_3.gpkg", layer = "ADM_ADM_3",
                       quiet = TRUE) |>
  select(NAME_1, NAME_3, geom)
gadm_lookup <- readRDS("data/gadm_lookup.rds")

# Join presence matrix → INE codes → GADM crosswalk → spatial boundaries
map_data <- gadm_sf |>
  inner_join(
    gadm_lookup |> select(gadm_dep, gadm_mun, cod.mun),
    by = c("NAME_1" = "gadm_dep", "NAME_3" = "gadm_mun")
  ) |>
  left_join(
    bo_municipality_wiki_presence$data |>
      select(qid, en, es, qu, ay) |>
      left_join(
        municipalities_wd_ine |> select(qid, ine_code),
        by = "qid"
      ),
    by = c("cod.mun" = "ine_code")
  ) |>
  mutate(
    across(c(en, es, qu, ay),
           ~ factor(case_when(
             .x == 1L ~ "Has article",
             .x == 0L ~ "No article",
             TRUE      ~ "Not in Wikidata"
           ), levels = c("Has article", "No article", "Not in Wikidata")))
  )
```

---

## Coverage summary

```{r coverage-table}
# Count municipalities with / without an article per Wikipedia
coverage_summary <- map_data |>
  st_drop_geometry() |>
  summarise(across(c(en, es, qu, ay), list(
    has    = ~ sum(.x == "Has article",      na.rm = TRUE),
    no     = ~ sum(.x == "No article",       na.rm = TRUE),
    absent = ~ sum(.x == "Not in Wikidata" | is.na(.x), na.rm = TRUE)
  ))) |>
  pivot_longer(everything(),
               names_to  = c("language", ".value"),
               names_sep = "_") |>
  mutate(
    wikipedia = c(en = "English", es = "Spanish", qu = "Quechua", ay = "Aymara")[language],
    pct_has   = round(100 * has / (has + no + absent), 1)
  ) |>
  select(wikipedia, has, no, absent, pct_has)

coverage_summary |>
  rename(
    Wikipedia      = wikipedia,
    `Has article`  = has,
    `No article`   = no,
    `Not in WD`    = absent,
    `% covered`    = pct_has
  ) |>
  knitr::kable(caption = "Municipality article coverage by Wikipedia language edition")
```

---

## Maps

Colors indicate whether Wikidata links the municipality to a Wikipedia article
in each language. Grey polygons are municipalities present in GADM but not
matched through Wikidata.

```{r map-colors}
presence_colors <- c(
  "Has article"     = "#2166ac",
  "No article"      = "#f7f7f7",
  "Not in Wikidata" = "#cccccc"
)
```

```{r map-function}
#| include: false
make_presence_map <- function(data, col, title) {
  ggplot(data) +
    geom_sf(aes(fill = .data[[col]]), color = "white", linewidth = 0.12) +
    scale_fill_manual(
      values   = presence_colors,
      na.value = "#eeeeee",
      name     = NULL,
      drop     = FALSE
    ) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 11),
      legend.position = "bottom"
    )
}
```

```{r maps-individual}
#| fig-height: 22
#| fig-width: 6

p_en <- make_presence_map(map_data, "en", "English Wikipedia")
p_es <- make_presence_map(map_data, "es", "Spanish Wikipedia")
p_qu <- make_presence_map(map_data, "qu", "Quechua Wikipedia")
p_ay <- make_presence_map(map_data, "ay", "Aymara Wikipedia")

p_en / p_es / p_qu / p_ay +
  plot_annotation(
    title   = "Bolivia: Municipal Wikipedia Coverage by Language",
    caption = "Source: Wikidata (Q1062710 instances); boundaries: GADM 4.1",
    theme   = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )
```

---

## Municipalities missing across multiple Wikipedias

Municipalities with no article in any of the four languages are the highest-priority
targets for new articles. The table below lists them, ordered by INE code.

```{r missing-all-four}
map_data |>
  st_drop_geometry() |>
  filter(
    en == "No article",
    es == "No article",
    qu == "No article",
    ay == "No article"
  ) |>
  left_join(
    municipalities_wd_ine |> select(ine_code, label_en, label_es),
    by = c("cod.mun" = "ine_code")
  ) |>
  select(
    `INE code`  = cod.mun,
    `Name (EN)` = label_en,
    `Name (ES)` = label_es,
    English     = en,
    Spanish     = es,
    Quechua     = qu,
    Aymara      = ay
  ) |>
  arrange(`INE code`) |>
  knitr::kable(caption = "Municipalities with no article in English, Spanish, Quechua, or Aymara")
```

---

## Next steps

The deeper analysis of existing English Wikipedia articles — article length,
citations, infobox completeness, and census data quality — is in
[`bolivia-municipality-wikipedia.qmd`](bolivia-municipality-wikipedia.qmd),
with results saved to `data/bolivia_municipality_wikipedia.rds`.

Future work here will:

- Add a table of municipalities missing only in one language, to guide targeted
  translation campaigns
- Link article quality metrics (from `bolivia_municipality_wikipedia.rds`) to
  the coverage maps
- Track changes over time as new articles are created
