---
title: "Bolivia 2026 — Municipality Infoboxes"
subtitle: "{{Infobox settlement}} wikitext ready to copy to Wikipedia"
params:
  wiki_filter: "all"   # "existing", "new", or "all"
format:
  html:
    code-fold: true
    toc: false
    self-contained: true
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(purrr)
library(htmltools)

source("src/wikipedia-tools.R")
source("src/gh-copy-assets.R")

muni_lookup  <- readRDS(here::here("data", "muni_id_lookup_table.rds"))
prov_lookup  <- readRDS(here::here("data", "prov_id_lookup_table.rds"))
concejo_data  <- readRDS(here::here("data", "oep2026", "muni_concejo_comb.rds"))
alcaldes      <- readRDS(here::here("data", "oep2026", "alcaldes.rds"))
muni_roster   <- readRDS(here::here("data", "oep2026", "muni_roster.rds")) |>
  select(id_muni, page)

bo_municipality_wiki_presence <- readRDS(here::here("data", "bo_municipality_wiki_presence.rds"))
municipalities_wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds"))


# Population
pop_muni <- readRDS(here::here("data", "cpv2024", "population_municipality.rds")) |>
  select(id_muni = ine_code, municipality, department, province,
         pop_2024 = pop_2024_total, pop_2012 = pop_2012_total,
         pop_2001 = pop_2001_total)

pop_muni_ur <- readRDS(here::here("data", "cpv2024", "population_municipality_urb_rural.rds")) |>
  select(id_muni = ine_code, area, pop_2024 = pop_2024_total)

# Ethnic identification
auto_data <- readRDS(here::here("data", "cpv2024", "autoidentificacion_municipio.rds"))

# Municipality areas from admin boundaries
area_muni <- readxl::read_excel(here::here("maps", "bol_admin_boundaries.xlsx"),
                                sheet = "bol_admin3") |>
  mutate(id_muni = stringr::str_sub(adm3_pcode, 3, 8)) |>
  select(id_muni, area_sqkm)

# Municipality coordinate ranges from shapefile
admin3_shp <- sf::st_read(
  here::here("../ultimate-consequences/maps/humdata.org_cod-ab-bol/bol_admin_boundaries.shp/bol_admin3.shp"),
  quiet = TRUE
)
coord_ranges <- purrr::map_df(seq_len(nrow(admin3_shp)), function(i) {
  row <- admin3_shp[i, ]
  bbox <- sf::st_bbox(row$geometry)
  tibble::tibble(
    id_muni = substr(row$adm3_pcode, 3, 8),
    lat_min = bbox[["ymin"]],
    lat_max = bbox[["ymax"]],
    lon_min = bbox[["xmin"]],
    lon_max = bbox[["xmax"]]
  )
})

# Locator map filenames from Commons upload log
commons_log <- readr::read_csv(
  here::here("output", "locator_maps", "muni-maps-final", "commons_upload_log.csv"),
  show_col_types = FALSE
) |>
  select(id_muni, commons_filename) |>
  filter(!is.na(commons_filename))

# Wikidata QID ↔ INE crosswalk (reused for multiple lookups)
wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds")) |>
  select(qid, ine_code)

# Wikidata image properties: flag (P41), coat of arms (P94), seal (P158),
# inception (P571). Refreshed by running the block in bolivia-muni-infobox-en.qmd
# (see "Refresh Wikidata images" section) or manually via SPARQL.
# To refresh: run the SPARQL query at the top of the helpers chunk and saveRDS().
wd_images <- readRDS(here::here("data", "muni_wd_images.rds"))

# P402 (OSM relation ID) presence — from saved Wikidata snapshot
# To refresh: source("R/get_wikidata_instances.R"), then run:
#   get_wikidata_instances("Q1062710", property = "P402",
#     property_names = "osm_relation_id", country = "Q750") |>
#     saveRDS("data/municipalities_wd_p402.rds")
wd_p402 <- readRDS(here::here("data", "municipalities_wd_p402.rds"))
munis_with_p402 <- wd_p402 |>
  filter(!is.na(osm_relation_id)) |>
  left_join(wd_ine, by = "qid") |>
  filter(!is.na(ine_code)) |>
  pull(ine_code)

# Wikipedia presence filter
wiki_presence <- bo_municipality_wiki_presence$data |>
  left_join(wd_ine, by = "qid") |>
  filter(!is.na(ine_code))

munis_with_enwiki <- wiki_presence |> filter(en > 0) |> pull(ine_code)

# Add id_prov to muni_lookup
muni_lookup <- muni_lookup |>
  mutate(id_prov = stringr::str_sub(id_muni, 1, 4))

# Split into two subsets: municipalities without an existing EN article
# (rendered fresh) and those with one (rendered as a merge against the
# existing infobox). params$wiki_filter selects which subset(s) to show.
muni_lookup_new      <- muni_lookup |> filter(!id_muni %in% munis_with_enwiki)
muni_lookup_existing <- muni_lookup |> filter(id_muni %in% munis_with_enwiki)

# Raw wikitext infobox content for existing EN articles, keyed by id_muni.
# Uses parse_infobox_blocks() (verbatim field blocks, comments intact) rather
# than extract_infobox_as_list() or extract_infobox(), because those clean/
# flatten values into single-line strings — exactly the markup structure we
# need to preserve line-for-line when merging into the new infobox.
sa_article_quality <- readRDS(here::here("data", "cache", "sa_article_quality.rds"))

existing_wikitext <- muni_lookup_existing |>
  left_join(wd_ine, by = c("id_muni" = "ine_code")) |>
  select(id_muni, qid) |>
  left_join(
    sa_article_quality |> filter(lang == "en") |> select(qid, wikitext),
    by = "qid"
  ) |>
  filter(!is.na(wikitext))

# NULL / empty for municipalities whose existing EN article has no
# {{Infobox settlement}} at all (stub articles) — render_existing_blocks()
# routes those through build_infobox() (fresh scaffold) instead of a merge.
existing_infobox_blocks <- setNames(
  purrr::map(existing_wikitext$wikitext, function(wt) {
    parse_infobox_blocks(extract_infobox_wikitext(wt))
  }),
  existing_wikitext$id_muni
)
```

```{r}
#| label: helpers
#| include: false

# ── OEP 2026 source reference ────────────────────────────────────────────────
oep_source_ref <- paste0(
  "{{Cite report",
  "| publisher = Órgano Electoral Plurinacional",
  "| last1 = Órgano Electoral Plurinacional",
  "| last2 = Tribunal Supremo Electoral",
  "| title = Publicación de resultados Elecciones Subnacionales 2026: Elección de Autoridades Políticas Departamentales Regionales y Municipales 2026 (Primera y Segunda Vuelta Electoral)",
  "| series = Separata de información pública",
  "| date = 2026-04-23}}"
)

oep_ref_with_page <- function(page) {
  base <- sub("\\}\\}$", "", oep_source_ref)
  paste0("<ref>", base, "| page = ", page, "}}</ref>")
}

# ── INE 2024 census population reference ─────────────────────────────────────
ine_pop_ref       <- paste0(
  '<ref name="ine_pop">{{Cite',
  "| last = Instituto Nacional de Estadística",
  "| title = Población censada e indicadores demográficos: Tabulados por Municipio/TIOC",
  "| url = https://nube.ine.gob.bo/index.php/s/0OrfPjzMB16fATQ/download",
  '| date = 2025}}</ref>'
)
ine_pop_ref_reuse <- '<ref name="ine_pop" />'

# ── Canonical Infobox settlement field order ─────────────────────────────────
# Order follows https://en.wikipedia.org/wiki/Template:Infobox_settlement
# ("Complete empty syntax, with comments"), trimmed to fields we either
# generate ourselves or have observed in existing Bolivia EN municipality
# infoboxes. This lets us render both a fresh scaffold (new articles) and a
# merge with an existing infobox (existing articles) in one consistent order.
infobox_field_order <- c(
  "official_name", "other_name", "native_name", "native_name_lang", "name",
  "nickname", "motto", "settlement_type",
  "translit_lang1_info1",
  "image_skyline", "imagesize", "image_alt", "image_caption",
  "image_flag", "flag_size", "flag_alt",
  "image_seal", "seal_size",
  "image_shield", "shield_size", "shield_alt",
  "image_blank_emblem", "blank_emblem_type", "blank_emblem_size",
  "image_map", "mapsize", "map_caption",
  "image_map1", "mapsize1", "map_caption1",
  "image_dot_map",
  "pushpin_map", "pushpin_mapsize", "pushpin_label_position", "pushpin_map_caption",
  "coordinates", "coordinates_footnotes",
  "subdivision_type", "subdivision_name",
  "subdivision_type1", "subdivision_name1",
  "subdivision_type2", "subdivision_name2",
  "subdivision_type3", "subdivision_name3",
  "subdivision_type4", "subdivision_name4",
  "established_title", "established_date",
  "established_title1", "established_date1",
  "established_title2", "established_date2",
  "established_title3", "established_date3",
  "founder",
  "seat_type", "seat",
  "government_footnotes", "government_type",
  "leader_title", "leader_name",
  "leader_title1", "leader_name1",
  "leader_title2", "leader_name2",
  "leader_title3", "leader_name3",
  "unit_pref",
  "area_magnitude", "area_footnotes",
  "area_total_km2", "area_land_km2", "area_water_km2",
  "area_blank1_title", "area_blank1_km2",
  "area_note",
  "elevation_footnotes", "elevation_m",
  "population_footnotes", "population_as_of", "population_total",
  "population_density_km2",
  "population_urban_footnotes", "population_urban", "population_density_urban_km2",
  "population_rural_footnotes", "population_rural", "population_density_rural_km2",
  "population_blank1_title", "population_blank1", "population_density_blank1_km2",
  "population_blank2_title", "population_blank2",
  "population_demonym", "population_note",
  "demographics_type1", "demographics1_footnotes", "demographics1_title1", "demographics1_info1",
  "timezone", "utc_offset",
  "timezone1", "utc_offset1",
  "postal_code_type", "postal_code",
  "area_code_type", "area_code",
  "geocode",
  "blank_name", "blank_info",
  "website", "footnotes"
)

# Section-header comments, inserted before the first field of each section
# (only shown if that field is present in the rendered output).
infobox_section_headers <- c(
  official_name         = "<!--See the Table at Infobox Settlement for all fields and descriptions of usage-->\n<!-- Basic info  ---------------->",
  image_skyline         = "<!-- Images, maps and coordinates  ----------->",
  subdivision_type      = "<!-- Location ------------------>",
  established_title     = "<!-- Established -------------------->",
  seat_type             = "<!-- Seat of government / smaller parts -->",
  government_footnotes  = "<!-- Politics ----------------->",
  unit_pref             = "<!-- Area    --------------------->",
  elevation_footnotes   = "<!-- Elevation -------------------->",
  population_footnotes  = "<!-- Population   ----------------------->",
  demographics_type1    = "<!-- Demographics ----------------->",
  timezone              = "<!-- General information  --------------->",
  postal_code_type      = "<!-- Area/postal codes & others -------->",
  website               = "<!-- Website, footnotes ---------->"
)

# Render a named character vector of field values as {{Infobox settlement}}
# wikitext, in canonical order, with section-header comments inserted.
render_infobox_wikitext <- function(fields, order = infobox_field_order,
                                     section_headers = infobox_section_headers) {
  present <- order[order %in% names(fields)]
  # Any fields not in our canonical order (unexpected/unknown existing-article
  # fields) are appended at the end rather than silently dropped.
  extra   <- setdiff(names(fields), order)
  present <- c(present, extra)

  lines <- character(0)
  for (f in present) {
    if (f %in% names(section_headers)) lines <- c(lines, section_headers[[f]])
    lines <- c(lines, sprintf("|%-24s= %s", f, fields[[f]]))
  }
  paste0("{{Infobox settlement\n", paste(lines, collapse = "\n"), "\n}}")
}

# ── Coordinate range formatter ──────────────────────────────────────────────
# Converts decimal degree ranges to DMS format: "38°24′ N to 41°59′ N"
fmt_coord_range <- function(lat_min, lat_max, lon_min, lon_max) {
  dec_to_dms <- function(dec, pos_char, neg_char) {
    hemi <- if (dec >= 0) pos_char else neg_char
    dec  <- abs(dec)
    d    <- floor(dec)
    m    <- floor((dec - d) * 60)
    if (m == 60) { d <- d + 1; m <- 0 }
    sprintf("%d°%d′ %s", d, m, hemi)
  }
  lat_min_str <- dec_to_dms(lat_min, "N", "S")
  lat_max_str <- dec_to_dms(lat_max, "N", "S")
  lon_min_str <- dec_to_dms(lon_min, "E", "W")
  lon_max_str <- dec_to_dms(lon_max, "E", "W")
  
  list(
    Latitude = paste(lat_min_str, "to", lat_max_str),
    Longitude = paste(lon_min_str, "to", lon_max_str)
  )
}

# ── Top ethnicities ──────────────────────────────────────────────────────────
# Returns a wikitext string of ethnic groups declared by ≥ threshold of pop.
# Uses Wikipedia article names where a standard link exists.
ethnicity_wikilinks <- c(
  "Quechua"      = "[[Quechua people|Quechua]]",
  "Aymara"       = "[[Aymara people|Aymara]]",
  "Chiquitano"   = "[[Chiquitano people|Chiquitano]]",
  "Guaran\u00ed" = "[[Guaran\u00ed people|Guaran\u00ed]]",
  "Moje\u00f1o"  = "[[Moxo people|Moje\u00f1o]]",
  "Afroboliviano" = "[[Afro-Bolivians|Afroboliviano]]",
  "Uru"          = "[[Uru people|Uru]]",
  "Chipaya"      = "[[Chipaya people|Chipaya]]",
  "Yuracar\u00e9" = "[[Yuqui people|Yuracar\u00e9]]",
  "Movima"       = "[[Movima people|Movima]]",
  "Tacana"       = "[[Tacana people|Tacana]]",
  "Guaran\u00ed" = "[[Guaran\u00ed people|Guaran\u00ed]]",
  "Weenhayek"    = "[[Weenhayek people|Weenhayek]]",
  "Tsimane"      = "[[Tsimane|Tsimane]]",
  "Ayoreo"       = "[[Ayoreo people|Ayoreo]]",
  "Ese Ejja"     = "[[Ese Ejja people|Ese Ejja]]",
  "Leco"         = "[[Leco people|Leco]]",
  "Kallawaya"    = "[[Kallawaya people|Kallawaya]]",
  "Moseten"      = "[[Moset\u00e9n people|Moseten]]"
)

get_top_ethnicities <- function(id, auto_data, pop_total, threshold = 0.01) {
  row <- auto_data |> filter(ine_code == id) |> slice(1)
  if (nrow(row) == 0) return("")

  meta_cols <- c("ine_code", "municipio", "provincia", "departamento",
                 "area", "nivel", "total", "municipality", "province",
                 "department", "level", "row_num")
  grp_cols <- setdiff(names(row), meta_cols)

  groups <- tibble(
    group = grp_cols,
    n     = sapply(grp_cols, function(col) as.double(row[[col]]))
  )

  # "Campesino", "Originario", "Indígena", and "Sin especificar" are all
  # non-specific self-identification categories (rather than named ethnic
  # groups); combine them into a single "Unspecified Indigenous" total.
  is_unspecified <- groups$group %in% c("Campesino", "Originario", "Indígena") |
    grepl("Sin especificar", groups$group, fixed = TRUE)
  unspecified_n <- sum(groups$n[is_unspecified], na.rm = TRUE)

  groups <- groups |>
    filter(!is_unspecified)

  if (unspecified_n > 0) {
    groups <- groups |>
      add_row(group = "Unspecified Indigenous", n = unspecified_n)
  }

  groups <- groups |>
    filter(n / pop_total >= threshold) |>
    arrange(desc(n))

  if (nrow(groups) == 0) return("")

  # Translate and link group names, then format with count and percentage
  fmt_pct <- function(p) if (p >= 1) sprintf("(%d%%)", round(p)) else sprintf("(%.1f%%)", p)

  entries <- mapply(function(g, n) {
    label <- if (g == "Unspecified Indigenous") {
      g
    } else if (g %in% names(ethnicity_wikilinks)) {
      ethnicity_wikilinks[[g]]
    } else {
      trimws(sub("\\s*\\(.*", "", g))
    }
    pct <- n / pop_total * 100
    sprintf("%s: %s %s", label, format(round(n), big.mark = ",", scientific = FALSE), fmt_pct(pct))
  }, groups$group, groups$n, SIMPLIFY = TRUE, USE.NAMES = FALSE)

  ine_ref <- paste0(
    "<ref>{{Cite",
    "| last = Instituto Nacional de Estadística",
    "| title = Autoidentificación con alguna nación pueblo indígena originario campesino y afroboliviano: Tabulados por Municipio/TIOC",
    "| date = 2025",
    " |url = https://nube.ine.gob.bo/index.php/s/Mu3eStaUZfi97Vp/download",
    "}}</ref>"
  )

  paste0(paste(entries, collapse = "<br>"), ine_ref)
}

# ── Infobox field computation ─────────────────────────────────────────────────
# Returns a named character vector of field → value (including deliberately
# blank scaffold fields), or NULL if the municipality isn't found. This is the
# single source of truth for "what we know" about a municipality; both the
# fresh-scaffold render (new articles) and the merge-with-existing render
# (existing articles) build on top of it.
compute_infobox_fields <- function(id, muni_lookup, prov_lookup, pop_muni, pop_muni_ur,
                          area_muni, alcaldes, muni_roster, auto_data,
                          commons_log, munis_with_p402, coord_ranges,
                          wd_images = NULL, wd_ine = NULL) {
  ml <- muni_lookup |> filter(id_muni == id) |> slice(1)
  if (nrow(ml) == 0) return(NULL)

  muni_name  <- ml$muni_anexo
  department <- ml$department
  id_prov    <- ml$id_prov

  # Province name (use cpv2024 name if available, fall back to anexo)
  prov_row <- prov_lookup |> filter(id_prov == .env$id_prov) |> slice(1)
  prov_name <- if (nrow(prov_row) > 0) {
    coalesce(prov_row$province_cpv2024, prov_row$province_anexo)
  } else NA_character_

  # Population
  pop_row <- pop_muni |> filter(id_muni == id) |> slice(1)
  pop_total <- if (nrow(pop_row) > 0) pop_row$pop_2024 else NA_real_

  urban_pop <- pop_muni_ur |> filter(id_muni == id, area == "Urbana") |>
    pull(pop_2024) |> (\(x) if (length(x) == 0) NA_real_ else x)()
  rural_pop <- pop_muni_ur |> filter(id_muni == id, area == "Rural") |>
    pull(pop_2024) |> (\(x) if (length(x) == 0) NA_real_ else x)()

  # Area and coordinates
  geo <- area_muni |> filter(id_muni == id) |> slice(1)
  area_val   <- if (nrow(geo) > 0 && !is.na(geo$area_sqkm)) geo$area_sqkm else NA_real_
  area_km2   <- if (!is.na(area_val)) round(area_val) else ""
  
  # Coordinate ranges. Latitude/Longitude are not real Infobox settlement
  # parameters (the template uses a single |coordinates= field via {{Coord}}),
  # so lat_str/lon_str are computed here and reserved for future use (e.g. a
  # |coordinates= builder) but are NOT included in the returned field vector.
  lat_str <- ""
  lon_str <- ""
  coord_row <- coord_ranges |> filter(id_muni == id) |> slice(1)
  if (nrow(coord_row) > 0) {
    coord_fmt <- fmt_coord_range(coord_row$lat_min, coord_row$lat_max,
                                  coord_row$lon_min, coord_row$lon_max)
    lat_str <- coord_fmt$Latitude
    lon_str <- coord_fmt$Longitude
  }

  # Population density
  density_km2 <- if (!is.na(pop_total) && nrow(geo) > 0 && !is.na(geo$area_sqkm) && geo$area_sqkm > 0) {
    format(signif(pop_total / geo$area_sqkm, 2), scientific = FALSE, trim = TRUE)
  } else ""

  # Mayor (source data is ALL CAPS; convert to title case)
  alc <- alcaldes |> filter(id_muni == id) |> slice(1)
  mayor_str <- if (nrow(alc) > 0) {
    name_tc  <- stringr::str_to_title(alc$nombre)
    page_row <- muni_roster |> filter(id_muni == id) |> slice(1)
    ref_str  <- if (nrow(page_row) > 0) oep_ref_with_page(page_row$page) else ""
    paste0(name_tc, " (", alc$sigla, ", 2026)", ref_str)
  } else ""

  # Ethnicities
  eth_str <- if (!is.na(pop_total)) {
    get_top_ethnicities(id, auto_data, pop_total)
  } else ""

  # Build department and province wikilinks
  dept_link <- sprintf("[[%s Department|%s]]", department, department)
  prov_link <- if (!is.na(prov_name)) {
    sprintf("[[%s Province|%s]]", prov_name, prov_name)
  } else ""

  # Pushpin map caption
  pushpin_caption <- sprintf("Location of %s Municipality within Bolivia", muni_name)

  # Locator map from Commons upload log
  map_row <- commons_log |> filter(id_muni == id) |> slice(1)
  map_file    <- if (nrow(map_row) > 0) map_row$commons_filename else ""
  map_caption <- if (nzchar(map_file)) {
    sprintf("Location of the %s Municipality within Bolivia", muni_name)
  } else ""

  # QID for OSM maplink id= parameter
  qid_val <- if (!is.null(wd_ine)) {
    qr <- wd_ine |> filter(ine_code == id) |> slice(1)
    if (nrow(qr) > 0 && !is.na(qr$qid)) qr$qid else ""
  } else ""

  # Zoom level scaled by municipality area
  zoom_level <- if (!is.na(area_val)) {
    if      (area_val > 40000) 6L
    else if (area_val > 15000) 7L
    else if (area_val > 4000)  8L
    else if (area_val > 1000)  9L
    else                       10L
  } else 8L

  # OSM maplink: only include if Wikidata item has P402
  map1_str <- if (id %in% munis_with_p402) {
    id_part <- if (nzchar(qid_val)) paste0(" |id=", qid_val) else ""
    sprintf(
      "{{maplink |type=shape |stroke-width=3 |fill=#999999 |frame=yes |plain=yes |frame-align=center |zoom=%d%s}}",
      zoom_level, id_part
    )
  } else ""

  map1_caption <- if (nzchar(map1_str)) {
    sprintf("%s Municipality and its local surroundings", muni_name)
  } else ""

  # Format numbers with commas
  fn <- function(x) if (is.na(x) || x == "") "" else format(as.numeric(x), big.mark = ",", scientific = FALSE)

  # Wikidata image properties
  img_row      <- if (!is.null(wd_images)) wd_images |> filter(ine_code == id) |> slice(1) else tibble()
  flag_file    <- if (nrow(img_row) > 0 && !is.na(img_row$flag_file))    img_row$flag_file    else ""
  coat_file    <- if (nrow(img_row) > 0 && !is.na(img_row$coat_file))    img_row$coat_file    else ""
  seal_file    <- if (nrow(img_row) > 0 && !is.na(img_row$seal_file))    img_row$seal_file    else ""
  # inception_date (P571) excluded — not reliable enough to include
  # commons_cat is computed here and reserved for future use, but is no
  # longer included in the returned field vector — the |commons= infobox
  # field has been struck in favor of a {{Commons category}} template placed
  # directly in the article body (see bolivia-muni-generator-en.qmd).
  commons_cat  <- if (nrow(img_row) > 0 && !is.na(img_row$commons_cat))  img_row$commons_cat  else ""

  # INE code and OCHA P-code for |geocode =
  ocha_pcode   <- paste0("BO", id)
  geocode_str  <- paste0(
    id, " ([[National Institute of Statistics of Bolivia|INE code]])",
    "<br>",
    ocha_pcode, " ([[United Nations Office for the Coordination of Humanitarian Affairs|OCHA]] [[Place code|P-code]])"
  )

  c(
    official_name          = muni_name,
    other_name             = "",
    native_name            = "",
    nickname               = "",
    settlement_type        = "Municipality",
    motto                  = "",
    image_skyline          = "",
    image_flag             = flag_file,
    flag_size              = if (nzchar(flag_file)) "150px" else "",
    # "flag_alt" is the real Infobox settlement parameter for flag/shield
    # descriptive text (screen-reader alt text, not a visible caption —
    # this template has no visible-caption field for these images).
    flag_alt               = if (nzchar(flag_file)) paste0("Flag of ", muni_name, " Municipality") else "",
    image_seal             = seal_file,
    seal_size              = if (nzchar(seal_file)) "100px" else "",
    image_shield           = coat_file,
    shield_size            = if (nzchar(coat_file)) "100px" else "",
    shield_alt             = if (nzchar(coat_file)) paste0("Coat of arms of ", muni_name, " Municipality") else "",
    image_blank_emblem     = "",
    blank_emblem_type      = "",
    blank_emblem_size      = "",
    image_map              = map_file,
    mapsize                = "300px",
    map_caption            = map_caption,
    image_map1             = map1_str,
    mapsize1               = "",
    map_caption1           = map1_caption,
    image_dot_map          = "",
    pushpin_map            = "Bolivia",
    pushpin_label_position = "bottom",
    pushpin_map_caption    = pushpin_caption,
    subdivision_type       = "Country",
    subdivision_name       = "[[Image:Flag of Bolivia.svg|25px]] [[Bolivia]]",
    subdivision_type1      = "[[Departments of Bolivia|Department]]",
    subdivision_name1      = dept_link,
    subdivision_type2      = "[[Provinces of Bolivia|Province]]",
    subdivision_name2      = prov_link,
    subdivision_type4      = "[[Cantons of Bolivia|Cantons]]",
    subdivision_name4      = "",
    government_footnotes   = "",
    government_type        = "",
    leader_title           = "Mayor",
    leader_name            = mayor_str,
    area_magnitude         = "",
    unit_pref              = "Metric",
    area_footnotes         = "",
    area_total_km2         = fn(area_km2),
    area_land_km2          = "",
    area_water_km2         = "",
    population_as_of               = "2024",
    population_footnotes           = ine_pop_ref,
    population_note                = "[[2024 Bolivian census]]",
    population_total               = fn(pop_total),
    population_density_km2         = fn(density_km2),
    population_urban_footnotes     = ine_pop_ref_reuse,
    population_urban               = fn(urban_pop),
    population_rural_footnotes     = ine_pop_ref_reuse,
    population_rural               = fn(rural_pop),
    population_blank1_title        = "Indigenous ethnicities",
    population_blank1              = eth_str,
    timezone                = "[[Time in Bolivia|BOT]]",
    utc_offset              = "-4",
    postal_code_type        = "",
    postal_code             = "",
    area_code                = "",
    geocode                  = geocode_str,
    footnotes                = ""
  )
}

# Thin wrapper: fresh infobox scaffold for a municipality with no existing
# article. Identical to the old build_infobox() output.
build_infobox <- function(id, ...) {
  fields <- compute_infobox_fields(id, ...)
  if (is.null(fields)) return(NULL)
  render_infobox_wikitext(fields)
}

# ── Merge with an existing article's infobox ─────────────────────────────────
# Fields we deliberately force to blank regardless of what the existing
# article has, because municipalities are areas, not points.
infobox_force_blank_fields <- c("pushpin_map", "pushpin_label_position")

# Fields struck from the output entirely — omitted even if the existing
# article's infobox already contains them (unlike force_blank fields above,
# these are dropped, not rendered as an explicit blank line). Latitude/
# Longitude aren't real Infobox settlement parameters; commons has been
# replaced by a {{Commons category}} template in the article body instead
# (see bolivia-muni-generator-en.qmd).
infobox_excluded_fields <- c("Latitude", "Longitude", "commons")

# Combine our computed fields with an existing infobox's verbatim raw field
# blocks (from parse_infobox_blocks()), working line-by-line rather than on
# cleaned/flattened values. Rule ("only override if non-blank"): for any
# field, if we have a non-blank computed value, it wins (our data is
# fresher/sourced) and is rendered as a clean single line; otherwise, if the
# existing block has real content (ignoring any embedded HTML comments), its
# ORIGINAL raw text is carried over byte-for-byte — comments, spacing, and
# all, in whatever position they occupied in the source. Fields that are
# comment-only (no real value) or absent from both sides are rendered blank.
# The two pushpin fields are force-blanked regardless of existing content,
# because that's a deliberate content decision, not a data gap.
#
# One arithmetic correction: verbatim blocks already end in a "\n" (the line
# break that preceded the next "|" in the source); a single trailing "\n" is
# trimmed per block before joining with "\n" between lines/fields, otherwise
# every carried-over field would leave a spurious blank line behind it. This
# does not touch any interior newlines (e.g. a comment sitting on its own
# line mid-value is left exactly as written).
#
# Consecutive identical comment-only lines are also deduplicated — this
# happens when an existing article's own embedded section-header comment
# (carried over verbatim from a field) is textually identical to the
# canonical header we insert before the next field.
render_infobox_merge_lines <- function(new_fields, existing_blocks,
                                        order = infobox_field_order,
                                        section_headers = infobox_section_headers,
                                        force_blank = infobox_force_blank_fields,
                                        excluded = infobox_excluded_fields) {
  trim_one_trailing_newline <- function(x) sub("\n$", "", x)

  existing_by_field <- list()
  for (b in existing_blocks) existing_by_field[[b$field]] <- b$raw_value
  existing_by_field[excluded] <- NULL

  new_fields <- new_fields[setdiff(names(new_fields), excluded)]

  new_names      <- names(new_fields)
  existing_names <- names(existing_by_field)

  # Only emit a field if it has real content on at least one side (or is
  # explicitly force-blanked) -- don't pad the output with every canonical
  # field name, nor with placeholder fields that were already blank in the
  # source article.
  field_has_content <- function(f) {
    new_val <- if (f %in% new_names) new_fields[[f]] else NA
    has_new <- !is.na(new_val) && nzchar(trimws(new_val))
    if (has_new) return(TRUE)
    old_raw <- existing_by_field[[f]]
    !is.null(old_raw) && has_real_content(old_raw)
  }

  candidate_names <- union(new_names, existing_names)
  active_names    <- union(force_blank,
                            candidate_names[vapply(candidate_names, field_has_content, logical(1))])
  present_order   <- c(order[order %in% active_names], setdiff(active_names, order))

  lines_out <- character(0)
  origin    <- setNames(character(length(present_order)), present_order)

  # Append a line, skipping it if it's a comment-only line identical to the
  # immediately preceding line (collapses back-to-back repeated headers).
  append_line <- function(x) {
    is_comment_only <- grepl("^\\s*<!--.*-->\\s*$", x)
    if (is_comment_only && length(lines_out) > 0 &&
        trimws(lines_out[length(lines_out)]) == trimws(x)) {
      return(invisible(NULL))
    }
    lines_out <<- c(lines_out, x)
  }

  for (f in present_order) {
    header <- if (f %in% names(section_headers)) section_headers[[f]] else NULL

    if (f %in% force_blank) {
      if (!is.null(header)) append_line(header)
      append_line(sprintf("|%-24s= ", f))
      origin[[f]] <- "blanked"
      next
    }

    new_val <- if (f %in% new_names) new_fields[[f]] else NA
    has_new <- !is.na(new_val) && nzchar(trimws(new_val))

    if (has_new) {
      if (!is.null(header)) append_line(header)
      append_line(sprintf("|%-24s= %s", f, new_val))
      origin[[f]] <- "new"
      next
    }

    old_raw <- existing_by_field[[f]]
    has_old <- !is.null(old_raw) && has_real_content(old_raw)

    if (has_old) {
      if (!is.null(header)) append_line(header)
      field_line <- sprintf("|%-24s=%s", f, trim_one_trailing_newline(old_raw))
      sub_lines <- strsplit(field_line, "\n")[[1]]
      for (i in seq_along(sub_lines)) {
        if (i == 1) lines_out <- c(lines_out, sub_lines[i]) else append_line(sub_lines[i])
      }
      origin[[f]] <- "existing"
      next
    }

    if (!is.null(header)) append_line(header)
    append_line(sprintf("|%-24s= ", f))
    origin[[f]] <- "blank"
  }

  list(
    wikitext = paste0("{{Infobox settlement\n", paste(lines_out, collapse = "\n"), "\n}}"),
    origin   = origin
  )
}

# Prose (not list) summary of which fields were carried over unchanged from
# the existing article.
describe_preserved_fields <- function(origin) {
  preserved <- names(origin)[origin == "existing"]
  if (length(preserved) == 0) {
    return("No fields were preserved from the existing article — every overlapping field was refreshed with new sourced data.")
  }
  tagged <- sprintf("`%s`", preserved)
  field_txt <- if (length(tagged) == 1) {
    tagged
  } else if (length(tagged) == 2) {
    paste(tagged, collapse = " and ")
  } else {
    paste0(paste(tagged[-length(tagged)], collapse = ", "), ", and ", tagged[length(tagged)])
  }
  sprintf("Imported unchanged from the existing article: %s.", field_txt)
}

# ── Wikitext copy block ──────────────────────────────────────────────────────
wikitext_block <- function(id, label, content) {
  tagList(
    tags$button(
      class = "copy-btn",
      style = "margin-bottom: 6px;",
      onclick = sprintf("copyTableToClipboard('%s', this)", id),
      label
    ),
    tags$pre(
      tags$code(content),
      style = paste(
        "background-color: #f6f8fa; padding: 10px; border-radius: 6px;",
        "overflow-x: auto; font-size: 11px; white-space: pre; margin: 0;"
      )
    ),
    tags$textarea(
      id = id,
      style = "position: absolute; left: -9999px; height: 0; width: 0; overflow: hidden;",
      readonly = TRUE,
      content
    )
  )
}

# ── Municipality block ───────────────────────────────────────────────────────
muni_block <- function(id_muni, muni_name, department, infobox_wikitext, note = NULL) {
  # Read pre-generated full page wikitext if available
  page_wt_path <- here::here("output", "bolivia-municipality-wikitext",
                              paste0(id_muni, ".txt"))
  page_wt <- if (file.exists(page_wt_path)) {
    paste(readLines(page_wt_path, warn = FALSE), collapse = "\n")
  } else NULL

  tagList(
    tags$details(
      tags$summary(
        tags$strong(muni_name),
        tags$span(
          paste0(" \u2014 ", department),
          style = "color: #586069; font-weight: normal;"
        ),
        tags$code(
          paste0(" [", id_muni, "]"),
          style = "font-size: 11px; color: #999; margin-left: 6px;"
        )
      ),
      tags$div(
        style = "margin-top: 8px; font-size: 12px;",
        if (!is.null(note)) tags$p(
          note,
          style = "color: #586069; margin: 0 0 6px 0; font-style: italic;"
        ),
        wikitext_block(
          id      = paste0("infobox_", id_muni),
          label   = "Copy infobox",
          content = infobox_wikitext
        ),
        if (!is.null(page_wt)) tagList(
          tags$div(style = "margin-top: 8px;"),
          tags$button(
            class   = "copy-btn",
            onclick = sprintf("copyTableToClipboard('page_en_%s', this)", id_muni),
            "Copy page Wikitext"
          ),
          tags$textarea(
            id       = paste0("page_en_", id_muni),
            style    = "position: absolute; left: -9999px; height: 0; width: 0; overflow: hidden;",
            readonly = TRUE,
            page_wt
          )
        )
      )
    ),
    tags$hr(style = "border:none; border-top:1px solid #e1e4e8; margin:4px 0;")
  )
}

# ── Render blocks: municipalities without an existing article ────────────────
render_new_blocks <- function(lookup) {
  ids <- lookup |>
    arrange(id_muni) |>
    select(id_muni, muni_anexo, department)

  tagList(pmap(ids, function(id_muni, muni_anexo, department) {
    ib <- build_infobox(id_muni, muni_lookup, prov_lookup, pop_muni, pop_muni_ur,
                        area_muni, alcaldes, muni_roster, auto_data,
                        commons_log, munis_with_p402, coord_ranges,
                        wd_images = wd_images, wd_ine = wd_ine)
    if (is.null(ib)) return(NULL)
    muni_block(id_muni, muni_anexo, department, ib)
  }))
}

# ── Render blocks: municipalities with an existing article (merged) ─────────
render_existing_blocks <- function(lookup) {
  ids <- lookup |>
    arrange(id_muni) |>
    select(id_muni, muni_anexo, department)

  tagList(pmap(ids, function(id_muni, muni_anexo, department) {
    existing_blocks <- existing_infobox_blocks[[id_muni]]

    # No {{Infobox settlement}} found in this article at all (stub) — there
    # is nothing to merge against, so fall back to the fresh-scaffold path.
    if (is.null(existing_blocks) || length(existing_blocks) == 0) {
      ib <- build_infobox(id_muni, muni_lookup, prov_lookup, pop_muni, pop_muni_ur,
                          area_muni, alcaldes, muni_roster, auto_data,
                          commons_log, munis_with_p402, coord_ranges,
                          wd_images = wd_images, wd_ine = wd_ine)
      if (is.null(ib)) return(NULL)
      note <- "No existing infobox was found in this article \u2014 rendering a fresh scaffold instead of a merge."
      return(muni_block(id_muni, muni_anexo, department, ib, note = note))
    }

    new_fields <- compute_infobox_fields(id_muni, muni_lookup, prov_lookup, pop_muni, pop_muni_ur,
                        area_muni, alcaldes, muni_roster, auto_data,
                        commons_log, munis_with_p402, coord_ranges,
                        wd_images = wd_images, wd_ine = wd_ine)
    if (is.null(new_fields)) return(NULL)

    merged <- render_infobox_merge_lines(new_fields, existing_blocks)
    note   <- describe_preserved_fields(merged$origin)

    muni_block(id_muni, muni_anexo, department, merged$wikitext, note = note)
  }))
}
```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

::: {style="font-size: 11px; margin-bottom: 1em;"}

Each municipality has two copy buttons:

- **Copy infobox** — the `{{Infobox settlement}}` template with population, area, coordinates, mayor, and ethnic group data from the 2024 census. For municipalities with an existing English Wikipedia article, this is a *merge*: our computed fields take priority when non-blank, otherwise the existing article's field value is kept as-is (with original wikitext markup and citations intact). A short note above the copy button explains which fields were carried over unchanged.
- **Copy page Wikitext** *(when available)* — the full article text assembled in section order: lead sentence → `==Government==` (mayor paragraph, council size paragraph, council table) → `==Population==` (census population table, population pyramid) → `==Languages and Indigenous identification==` (ethnic identity table, languages table) → `==Communities==`. Generated by the [article content page](bolivia-muni-generator-en.html); render that page first to produce these files.

:::

```{r}
#| label: filter-summary
#| echo: false
#| results: asis
n_new      <- nrow(muni_lookup_new)
n_existing <- nrow(muni_lookup_existing)

if (params$wiki_filter == "existing") {
  cat(sprintf(
    "**Rendering merged infoboxes for %d municipalities with an existing English Wikipedia article.**\n\n",
    n_existing
  ))
} else if (params$wiki_filter == "new") {
  cat(sprintf(
    "**Rendering fresh infoboxes for %d municipalities without an English Wikipedia article.**\n\n",
    n_new
  ))
} else {
  cat(sprintf(
    "**Rendering fresh infoboxes for %d municipalities without an existing article, and merged infoboxes for %d municipalities with one.**\n\n",
    n_new, n_existing
  ))
}
```

```{r}
#| label: infobox-blocks
#| echo: false
#| results: asis
show_new      <- params$wiki_filter %in% c("new", "all")
show_existing <- params$wiki_filter %in% c("existing", "all")

if (show_new && show_existing) {
  cat("## Municipalities without an existing English Wikipedia article\n\n")
}
if (show_new) {
  render_new_blocks(muni_lookup_new)
}
if (show_new && show_existing) {
  cat("\n\n## Municipalities with an existing English Wikipedia article (merged infobox)\n\n")
}
if (show_existing) {
  render_existing_blocks(muni_lookup_existing)
}
```
