---
title: "Bolivia 2026 — Municipality Infoboxes"
subtitle: "{{Infobox settlement}} wikitext ready to copy to Wikipedia"
params:
  wiki_filter: "new"   # "existing", "new", or "all"
format:
  html:
    code-fold: true
    toc: false
    self-contained: true
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(purrr)
library(htmltools)

source("src/wikipedia-tools.R")
source("src/gh-copy-assets.R")

muni_lookup  <- readRDS(here::here("data", "muni_id_lookup_table.rds"))
prov_lookup  <- readRDS(here::here("data", "prov_id_lookup_table.rds"))
concejo_data <- readRDS(here::here("data", "oep2026", "muni_concejo_comb.rds"))
alcaldes     <- readRDS(here::here("data", "oep2026", "alcaldes.rds"))

bo_municipality_wiki_presence <- readRDS(here::here("data", "bo_municipality_wiki_presence.rds"))
municipalities_wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds"))


# Population
pop_muni <- readRDS(here::here("data", "cpv2024", "population_municipality.rds")) |>
  select(id_muni = ine_code, municipality, department, province,
         pop_2024 = pop_2024_total, pop_2012 = pop_2012_total,
         pop_2001 = pop_2001_total)

pop_muni_ur <- readRDS(here::here("data", "cpv2024", "population_municipality_urb_rural.rds")) |>
  select(id_muni = ine_code, area, pop_2024 = pop_2024_total)

# Ethnic identification
auto_data <- readRDS(here::here("data", "cpv2024", "autoidentificacion_municipio.rds"))

# Municipality areas from admin boundaries
area_muni <- readxl::read_excel(here::here("maps", "bol_admin_boundaries.xlsx"),
                                sheet = "bol_admin3") |>
  mutate(id_muni = stringr::str_sub(adm3_pcode, 3, 8)) |>
  select(id_muni, area_sqkm)

# Municipality coordinate ranges from shapefile
admin3_shp <- sf::st_read(
  here::here("../ultimate-consequences/maps/humdata.org_cod-ab-bol/bol_admin_boundaries.shp/bol_admin3.shp"),
  quiet = TRUE
)
coord_ranges <- purrr::map_df(seq_len(nrow(admin3_shp)), function(i) {
  row <- admin3_shp[i, ]
  bbox <- sf::st_bbox(row$geometry)
  tibble::tibble(
    id_muni = substr(row$adm3_pcode, 3, 8),
    lat_min = bbox[["ymin"]],
    lat_max = bbox[["ymax"]],
    lon_min = bbox[["xmin"]],
    lon_max = bbox[["xmax"]]
  )
})

# Locator map filenames from Commons upload log
commons_log <- readr::read_csv(
  here::here("output", "locator_maps", "muni-maps-final", "commons_upload_log.csv"),
  show_col_types = FALSE
) |>
  select(id_muni, commons_filename) |>
  filter(!is.na(commons_filename))

# Wikidata QID ↔ INE crosswalk (reused for multiple lookups)
wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds")) |>
  select(qid, ine_code)

# Wikidata image properties: flag (P41), coat of arms (P94), seal (P158),
# inception (P571). Refreshed by running the block in bolivia-muni-infobox-en.qmd
# (see "Refresh Wikidata images" section) or manually via SPARQL.
# To refresh: run the SPARQL query at the top of the helpers chunk and saveRDS().
wd_images <- readRDS(here::here("data", "muni_wd_images.rds"))

# P402 (OSM relation ID) presence — from saved Wikidata snapshot
# To refresh: source("R/get_wikidata_instances.R"), then run:
#   get_wikidata_instances("Q1062710", property = "P402",
#     property_names = "osm_relation_id", country = "Q750") |>
#     saveRDS("data/municipalities_wd_p402.rds")
wd_p402 <- readRDS(here::here("data", "municipalities_wd_p402.rds"))
munis_with_p402 <- wd_p402 |>
  filter(!is.na(osm_relation_id)) |>
  left_join(wd_ine, by = "qid") |>
  filter(!is.na(ine_code)) |>
  pull(ine_code)

# Wikipedia presence filter
wiki_presence <- bo_municipality_wiki_presence$data |>
  left_join(wd_ine, by = "qid") |>
  filter(!is.na(ine_code))

munis_with_enwiki <- wiki_presence |> filter(en > 0) |> pull(ine_code)

if (params$wiki_filter == "existing") {
  muni_lookup <- muni_lookup |> filter(id_muni %in% munis_with_enwiki)
} else if (params$wiki_filter == "new") {
  muni_lookup <- muni_lookup |> filter(!id_muni %in% munis_with_enwiki)
}

# Add id_prov to muni_lookup
muni_lookup <- muni_lookup |>
  mutate(id_prov = stringr::str_sub(id_muni, 1, 4))
```

```{r}
#| label: helpers
#| include: false

# ── Coordinate range formatter ──────────────────────────────────────────────
# Converts decimal degree ranges to DMS format: "38°24′ N to 41°59′ N"
fmt_coord_range <- function(lat_min, lat_max, lon_min, lon_max) {
  dec_to_dms <- function(dec, pos_char, neg_char) {
    hemi <- if (dec >= 0) pos_char else neg_char
    dec  <- abs(dec)
    d    <- floor(dec)
    m    <- floor((dec - d) * 60)
    if (m == 60) { d <- d + 1; m <- 0 }
    sprintf("%d°%d′ %s", d, m, hemi)
  }
  lat_min_str <- dec_to_dms(lat_min, "N", "S")
  lat_max_str <- dec_to_dms(lat_max, "N", "S")
  lon_min_str <- dec_to_dms(lon_min, "E", "W")
  lon_max_str <- dec_to_dms(lon_max, "E", "W")
  
  list(
    latitude = paste(lat_min_str, "to", lat_max_str),
    longitude = paste(lon_min_str, "to", lon_max_str)
  )
}

# ── Top ethnicities ──────────────────────────────────────────────────────────
# Returns a wikitext string of ethnic groups declared by ≥ threshold of pop.
# Uses Wikipedia article names where a standard link exists.
ethnicity_wikilinks <- c(
  "Quechua"      = "[[Quechua people|Quechua]]",
  "Aymara"       = "[[Aymara people|Aymara]]",
  "Chiquitano"   = "[[Chiquitano people|Chiquitano]]",
  "Guaran\u00ed" = "[[Guaran\u00ed people|Guaran\u00ed]]",
  "Moje\u00f1o"  = "[[Moxo people|Moje\u00f1o]]",
  "Afroboliviano" = "[[Afro-Bolivians|Afroboliviano]]",
  "Uru"          = "[[Uru people|Uru]]",
  "Chipaya"      = "[[Chipaya people|Chipaya]]",
  "Yuracar\u00e9" = "[[Yuqui people|Yuracar\u00e9]]",
  "Movima"       = "[[Movima people|Movima]]",
  "Tacana"       = "[[Tacana people|Tacana]]",
  "Guaran\u00ed" = "[[Guaran\u00ed people|Guaran\u00ed]]",
  "Weenhayek"    = "[[Weenhayek people|Weenhayek]]",
  "Tsimane"      = "[[Tsimane|Tsimane]]",
  "Ayoreo"       = "[[Ayoreo people|Ayoreo]]",
  "Ese Ejja"     = "[[Ese Ejja people|Ese Ejja]]",
  "Leco"         = "[[Leco people|Leco]]",
  "Kallawaya"    = "[[Kallawaya people|Kallawaya]]",
  "Moseten"      = "[[Moset\u00e9n people|Moseten]]"
)

get_top_ethnicities <- function(id, auto_data, pop_total, threshold = 0.01) {
  row <- auto_data |> filter(ine_code == id) |> slice(1)
  if (nrow(row) == 0) return("")

  meta_cols <- c("ine_code", "municipio", "provincia", "departamento",
                 "area", "nivel", "total", "municipality", "province",
                 "department", "level", "row_num")
  grp_cols <- setdiff(names(row), meta_cols)

  groups <- tibble(
    group = grp_cols,
    n     = sapply(grp_cols, function(col) as.double(row[[col]]))
  ) |>
    filter(n / pop_total >= threshold) |>
    arrange(desc(n))

  # Exclude non-specific / non-ethnic categories (substring match)
  groups <- groups |>
    filter(!grepl("Campesino|Originario", group))

  if (nrow(groups) == 0) return("")

  # Translate and link group names
  linked <- sapply(groups$group, function(g) {
    if (grepl("Sin especificar", g, fixed = TRUE)) {
      "Unspecified Indigenous"
    } else if (g %in% names(ethnicity_wikilinks)) {
      ethnicity_wikilinks[[g]]
    } else {
      # Strip any trailing parenthetical footnote marker
      trimws(sub("\\s*\\(.*", "", g))
    }
  }, USE.NAMES = FALSE)

  ine_ref <- paste0(
    "<ref>{{Cite",
    "| last = Instituto Nacional de Estadística",
    "| title = Autoidentificación con alguna nación pueblo indígena originario campesino y afroboliviano: Tabulados por Municipio/TIOC",
    "| date = 2025",
    " |url = https://nube.ine.gob.bo/index.php/s/Mu3eStaUZfi97Vp/download",
    "}}</ref>"
  )

  paste0(paste(linked, collapse = ", "), ine_ref)
}

# ── Infobox builder ──────────────────────────────────────────────────────────
build_infobox <- function(id, muni_lookup, prov_lookup, pop_muni, pop_muni_ur,
                          area_muni, alcaldes, auto_data,
                          commons_log, munis_with_p402, coord_ranges,
                          wd_images = NULL) {
  ml <- muni_lookup |> filter(id_muni == id) |> slice(1)
  if (nrow(ml) == 0) return(NULL)

  muni_name  <- ml$muni_anexo
  department <- ml$department
  id_prov    <- ml$id_prov

  # Province name (use cpv2024 name if available, fall back to anexo)
  prov_row <- prov_lookup |> filter(id_prov == .env$id_prov) |> slice(1)
  prov_name <- if (nrow(prov_row) > 0) {
    coalesce(prov_row$province_cpv2024, prov_row$province_anexo)
  } else NA_character_

  # Population
  pop_row <- pop_muni |> filter(id_muni == id) |> slice(1)
  pop_total <- if (nrow(pop_row) > 0) pop_row$pop_2024 else NA_real_

  urban_pop <- pop_muni_ur |> filter(id_muni == id, area == "Urbana") |>
    pull(pop_2024) |> (\(x) if (length(x) == 0) NA_real_ else x)()
  rural_pop <- pop_muni_ur |> filter(id_muni == id, area == "Rural") |>
    pull(pop_2024) |> (\(x) if (length(x) == 0) NA_real_ else x)()

  # Area and coordinates
  geo <- area_muni |> filter(id_muni == id) |> slice(1)
  area_km2   <- if (nrow(geo) > 0) round(geo$area_sqkm) else ""
  
  # Coordinate ranges (formatted as Latitude/Longitude fields)
  lat_str <- ""
  lon_str <- ""
  coord_row <- coord_ranges |> filter(id_muni == id) |> slice(1)
  if (nrow(coord_row) > 0) {
    coord_fmt <- fmt_coord_range(coord_row$lat_min, coord_row$lat_max,
                                  coord_row$lon_min, coord_row$lon_max)
    lat_str <- coord_fmt$latitude
    lon_str <- coord_fmt$longitude
  }

  # Population density
  density_km2 <- if (!is.na(pop_total) && nrow(geo) > 0 && !is.na(geo$area_sqkm) && geo$area_sqkm > 0) {
    round(pop_total / geo$area_sqkm)
  } else ""

  # Mayor (source data is ALL CAPS; convert to title case)
  alc <- alcaldes |> filter(id_muni == id) |> slice(1)
  mayor_str <- if (nrow(alc) > 0) {
    name_tc <- stringr::str_to_title(alc$nombre)
    paste0(name_tc, " (", alc$sigla, ", 2026)")
  } else ""

  # Ethnicities
  eth_str <- if (!is.na(pop_total)) {
    get_top_ethnicities(id, auto_data, pop_total)
  } else ""

  # Build department and province wikilinks
  dept_link <- sprintf("[[%s Department|%s]]", department, department)
  prov_link <- if (!is.na(prov_name)) {
    sprintf("[[%s Province|%s]]", prov_name, prov_name)
  } else ""

  # Pushpin map caption
  pushpin_caption <- sprintf("Location of %s Municipality within Bolivia", muni_name)

  # Locator map from Commons upload log
  map_row <- commons_log |> filter(id_muni == id) |> slice(1)
  map_file    <- if (nrow(map_row) > 0) map_row$commons_filename else ""
  map_caption <- if (nzchar(map_file)) {
    sprintf("Location of the %s Municipality within Bolivia", muni_name)
  } else ""

  # OSM maplink: only include if Wikidata item has P402
  map1_str <- if (id %in% munis_with_p402) {
    "{{maplink |type=shape |stroke-width=3 |fill=#999999 |frame=yes |plain=yes |frame-align=center |zoom=8}}"
  } else ""

  # Format numbers with commas
  fn <- function(x) if (is.na(x) || x == "") "" else format(as.numeric(x), big.mark = ",", scientific = FALSE)

  # Wikidata image properties
  img_row      <- if (!is.null(wd_images)) wd_images |> filter(ine_code == id) |> slice(1) else tibble()
  flag_file    <- if (nrow(img_row) > 0 && !is.na(img_row$flag_file))    img_row$flag_file    else ""
  coat_file    <- if (nrow(img_row) > 0 && !is.na(img_row$coat_file))    img_row$coat_file    else ""
  seal_file    <- if (nrow(img_row) > 0 && !is.na(img_row$seal_file))    img_row$seal_file    else ""
  inc_date     <- if (nrow(img_row) > 0 && !is.na(img_row$inception_date)) img_row$inception_date else ""
  commons_cat  <- if (nrow(img_row) > 0 && !is.na(img_row$commons_cat))  img_row$commons_cat  else ""

  paste0(
'{{Infobox settlement
<!--See the Table at Infobox Settlement for all fields and descriptions of usage-->
<!-- Basic info  ---------------->
|official_name          = ', muni_name, '
|other_name             =
|native_name            =
|nickname               =
|settlement_type        = Municipality
|motto                  =
<!-- images and maps  ----------->
|image_skyline          =
|imagesize              =
|image_caption          =
|image_flag             = ', flag_file, '
|flag_size              = ', if (nzchar(flag_file)) "150px" else "", '
|image_seal             = ', seal_file, '
|seal_size              = ', if (nzchar(seal_file)) "100px" else "", '
|image_shield           = ', coat_file, '
|shield_size            = ', if (nzchar(coat_file)) "100px" else "", '
|image_blank_emblem     =
|blank_emblem_type      =
|blank_emblem_size      =
|image_map              = ', map_file, '
|mapsize                = 300px
|map_caption            = ', map_caption, '
|image_map1             = ', map1_str, '
|mapsize1               =
|map_caption1           =
|image_dot_map          =
|pushpin_map            = Bolivia
|pushpin_label_position = bottom
|pushpin_map_caption    = ', pushpin_caption, '
<!-- Location ------------------>
|subdivision_type       = Country
|subdivision_name       = [[Image:Flag of Bolivia.svg|25px]] [[Bolivia]]
|subdivision_type1      = [[Departments of Bolivia|Department]]
|subdivision_name1      = ', dept_link, '
|subdivision_type2      = [[Provinces of Bolivia|Province]]
|subdivision_name2      = ', prov_link, '
|subdivision_type4      = [[Cantons of Bolivia|Cantons]]
|subdivision_name4      =
<!-- Smaller parts (e.g. boroughs of a city) and seat of government -->
|seat_type              = Seat
|seat                   =
<!-- Politics ----------------->
|government_footnotes   =
|government_type        =
|leader_title           = Mayor
|leader_name            = ', mayor_str, '
|leader_title1          =
|leader_name1           =
|established_title      = ', if (nzchar(inc_date)) "Established" else "", '
|established_date       = ', inc_date, '
<!-- Area    --------------------->
|area_magnitude         =
|unit_pref              = Metric
|area_footnotes         =
|area_total_km2         = ', fn(area_km2), '
|area_land_km2          =
|area_water_km2         =
<!-- Population   ----------------------->
|population_as_of               = 2024
|population_footnotes           =
|population_note                = [[2024 Bolivian census]]
|population_total               = ', fn(pop_total), '
|population_density_km2         = ', fn(density_km2), '
|population_urban               = ', fn(urban_pop), '
|population_blank1_title        = Indigenous ethnicities
|population_blank1              = ', eth_str, '
<!-- General information  --------------->
|timezone               = [[Time in Bolivia|BOT]]
|utc_offset             = -4
|latitude               = ', lat_str, '
|longitude              = ', lon_str, '
|elevation_footnotes    =
|elevation_m            =
<!-- Area/postal codes & others -------->
|postal_code_type       =
|postal_code            =
|area_code              =
|website                =
|commons                = ', commons_cat, '
|footnotes              =
}}')
}

# ── Wikitext copy block ──────────────────────────────────────────────────────
wikitext_block <- function(id, label, content) {
  tagList(
    tags$button(
      class = "copy-btn",
      style = "margin-bottom: 6px;",
      onclick = sprintf("copyTableToClipboard('%s', this)", id),
      label
    ),
    tags$pre(
      tags$code(content),
      style = paste(
        "background-color: #f6f8fa; padding: 10px; border-radius: 6px;",
        "overflow-x: auto; font-size: 11px; white-space: pre; margin: 0;"
      )
    ),
    tags$textarea(
      id = id,
      style = "position: absolute; left: -9999px; height: 0; width: 0; overflow: hidden;",
      readonly = TRUE,
      content
    )
  )
}

# ── Municipality block ───────────────────────────────────────────────────────
muni_block <- function(id_muni, muni_name, department, infobox_wikitext) {
  # Read pre-generated full page wikitext if available
  page_wt_path <- here::here("output", "bolivia-municipality-wikitext",
                              paste0(id_muni, ".txt"))
  page_wt <- if (file.exists(page_wt_path)) {
    paste(readLines(page_wt_path, warn = FALSE), collapse = "\n")
  } else NULL

  tagList(
    tags$details(
      tags$summary(
        tags$strong(muni_name),
        tags$span(
          paste0(" \u2014 ", department),
          style = "color: #586069; font-weight: normal;"
        ),
        tags$code(
          paste0(" [", id_muni, "]"),
          style = "font-size: 11px; color: #999; margin-left: 6px;"
        )
      ),
      tags$div(
        style = "margin-top: 8px; font-size: 12px;",
        wikitext_block(
          id      = paste0("infobox_", id_muni),
          label   = "Copy infobox",
          content = infobox_wikitext
        ),
        if (!is.null(page_wt)) tagList(
          tags$div(style = "margin-top: 8px;"),
          tags$button(
            class   = "copy-btn",
            onclick = sprintf("copyTableToClipboard('page_en_%s', this)", id_muni),
            "Copy page Wikitext"
          ),
          tags$textarea(
            id       = paste0("page_en_", id_muni),
            style    = "position: absolute; left: -9999px; height: 0; width: 0; overflow: hidden;",
            readonly = TRUE,
            page_wt
          )
        )
      )
    ),
    tags$hr(style = "border:none; border-top:1px solid #e1e4e8; margin:4px 0;")
  )
}

# ── Render all blocks ────────────────────────────────────────────────────────
render_muni_blocks <- function(lookup) {
  ids <- lookup |>
    arrange(id_muni) |>
    select(id_muni, muni_anexo, department)

  tagList(pmap(ids, function(id_muni, muni_anexo, department) {
    ib <- build_infobox(id_muni, muni_lookup, prov_lookup, pop_muni, pop_muni_ur,
                        area_muni, alcaldes, auto_data,
                        commons_log, munis_with_p402, coord_ranges,
                        wd_images = wd_images)
    if (is.null(ib)) return(NULL)
    muni_block(id_muni, muni_anexo, department, ib)
  }))
}
```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

::: {style="font-size: 11px; margin-bottom: 1em;"}

Each municipality has two copy buttons:

- **Copy infobox** — the `{{Infobox settlement}}` template with population, area, coordinates, mayor, and ethnic group data from the 2024 census.
- **Copy page Wikitext** *(when available)* — the full article text assembled in section order: lead sentence → `==Government==` (mayor paragraph, council size paragraph, council table) → `==Population==` (census population table, population pyramid) → `==Languages and Indigenous identification==` (ethnic identity table, languages table) → `==Communities==`. Generated by the [article content page](bolivia-muni-generator-en.html); render that page first to produce these files.

:::

```{r}
#| label: filter-summary
#| echo: false
#| results: asis
n_total <- nrow(muni_lookup)
filter_label <- switch(params$wiki_filter,
  existing = "with an existing English Wikipedia article",
  new      = "without an English Wikipedia article (new)",
  "all (no filter applied)"
)
cat(sprintf("**Rendering infoboxes for %d municipalities %s.**\n\n", n_total, filter_label))
```

```{r}
#| label: infobox-blocks
#| echo: false
#| results: asis
render_muni_blocks(muni_lookup)
```
