---
title: "Bolivia municipality text and figures for Wikipedia"
subtitle: "Wikitext ready to copy to Wikipedia — multiple cited sources"
params:
  wiki_filter: "all"   # "existing", "new", or "all"
format:
  html:
    code-fold: true
    toc: false
    self-contained: true
---

::: {style="font-size: 11px;"}

## Items produced per municipality

Each expanded municipality section contains the following copyable blocks.

---

**Copy lead Wikitext** — one to three sentences identifying the municipality's location, with optional sentences for population rank and province context:

> `'''[Municipality] Municipality'''` is the `[ordinal]` [[Municipalities of Bolivia\|municipal section]] of the `[[Province Province]]` in the `[[Department Department]]` in `[[Bolivia]]`.

*[If the municipality is the 1st or 2nd most populous in Bolivia:]*
> It is the `[most / second most]` populous municipality in `[[Bolivia]]`.

*[If not nationally top-2, but 1st or 2nd most populous in its department:]*
> It is the `[most / second most]` populous municipality in `[[Department Department]]`.

*[If not department top-2, and most populous in a province with 3 or more municipalities:]*
> It is the most populous municipality in `[[Province Province]]`.

*[If the only municipality in its province:]*
> `[Municipality]` Municipality is the only municipality in `[[Province Province]]` and its boundaries are identical.

*[If one of two municipalities in its province — four variants based on relative area and population:]*

| Larger area? | More populous? | Sentence |
|---|---|---|
| Yes | Yes | `[Municipality]` Municipality is the larger and more populous of the two municipalities in `[[Province Province]]`. |
| Yes | No | `[Municipality]` Municipality covers the larger area of the two municipalities in `[[Province Province]]`, but it is less populous. |
| No | Yes | `[Municipality]` Municipality is the more populous of the two municipalities in `[[Province Province]]`, but it covers a smaller area. |
| No | No | `[Municipality]` Municipality is the smaller and less populous of the two municipalities in `[[Province Province]]`. |

*[Omitted for provinces with 3 or more municipalities.]*

---

**Copy government Wikitext** — two paragraphs for a `==Government==` section:

> The Mayor of `[Municipality]` is `[Name]`, who was elected in the [March 2026 elections](https://en.wikipedia.org/wiki/2026_Bolivian_regional_elections).

> `[Municipality]` Municipality has `[5 / 7 / 9 / 11]` council seats, as do all `[municipalities with populations up to 15,000 / between 15,001 and 50,000 / between 50,001 and 75,000 / over 75,000 / department capitals]`. *[IOC municipalities only:]* One of those seats is an Indigenous seat assigned by the `[pueblo]` according to its own norms and procedures.

---

**Copy council table** — a wikitable of council members (one row per seat/type), for the `==Government==` section:

```
{| class="wikitable"
! Seat !! Type !! Name !! Party / Political Organization
|-
| rowspan="2" | [N] || Member || [Name] || rowspan="2" | [Party]
|-
| Alternate || ''[Alternate name]''
...
|}
```

---

**Copy population table** — a horizontal wikitable of census population for `==Population==`:

```
{| class="wikitable"
|+ Population in Recent Censuses
|-
!! [1992] !! 2001 !! 2012 !! 2024
|-
|| [pop_1992] || pop_2001 || pop_2012
|| pop_2024
|}
```

*[1992 column omitted for municipalities whose province was subdivided since 2002.]*

---

**Copy population pyramid** — a `{{Population pyramid 2}}` template for `==Population==`, with 5-year age bins from 0–4 to 85+ derived from the 2024 census:

```
{{Population pyramid 2 |year=2024 |m0=... |f0=... ... |mmax=... |fmax=... |source=2024 Bolivian Census}}
```

---

**Copy identity table** — a wikitable of indigenous self-identification from the 2024 census, for `==Languages and Indigenous identification==`.

**Copy language table** — a wikitable of languages spoken from the 2024 census, for `==Languages and Indigenous identification==`.

---

**Copy communities Wikitext** — a `==Communities==` section with all communities listed in three columns using `{{div col|colwidth=18em}}`. *[Omitted for municipalities with no communities in the 2013 INE geographic classification.]*

---

**Full page Wikitext** is saved to `output/bolivia-municipality-wikitext/en/[id].txt` and available as a single copy button on the [infobox page](bolivia-muni-infobox-en.html). It assembles all sections above in article order: lead → `==Government==` → `==Population==` → `==Languages and Indigenous identification==` → `==Communities==`.

---

:::


```{r}
#| label: setup
#| include: false

library(dplyr)
library(purrr)
library(htmltools)
library(english)
library(kableExtra)
library(glue)
library(wikitools)

# Language-aware article engine. This sources bolivia-builders.R (which sources
# the shared i18n layer, muni-i18n.R), plus muni-phrases.R and muni-references.R.
source("src/muni-article-functions.R")
source("src/gh-copy-assets.R")
source("src/generate-muni-poverty-tables.R")

# Language for this render.
lang <- "en"

muni_lookup       <- readRDS(here::here("data", "muni_id_lookup_table.rds"))
prov_lookup       <- readRDS(here::here("data", "prov_id_lookup_table.rds"))
prov_link_lookup  <- readRDS(here::here("data", "prov_link_lookup.rds"))
muni_roster       <- readRDS(here::here("data", "oep2026", "muni_roster.rds"))
concejo_data      <- readRDS(here::here("data", "oep2026", "muni_concejo_comb.rds"))
concejal_summary  <- readRDS(here::here("data", "oep2026", "concejal_summary.rds"))
muni_concejal_ioc <- readRDS(here::here("data", "oep2026", "muni_concejal_ioc.rds"))
alcaldes          <- readRDS(here::here("data", "oep2026", "alcaldes.rds"))

bo_municipality_wiki_presence <- readRDS(here::here("data", "bo_municipality_wiki_presence.rds"))
municipalities_wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds"))

# Commons category names (from Wikidata P373 via muni_wd_images.rds), used to
# populate the {{Commons category}} template placed at the end of the article
# body.
wd_images <- readRDS(here::here("data", "muni_wd_images.rds")) |>
  select(ine_code, commons_cat)

roster_pages <- muni_roster |> select(id_muni, page)

# Expected seat counts by census population — mirrors cleaning script section 3b
pop_muni <- readRDS(here::here("data", "cpv2024", "population_municipality.rds")) |>
  select(id_muni = ine_code, pop_2024 = pop_2024_total, pop_2012 = pop_2012_total)

census_pop   <- readRDS(here::here("data", "cpv2024", "census_pop_historical.rds"))
lang_data    <- readRDS(here::here("data", "cpv2024", "idiomas_hablados_muni.rds"))
auto_data    <- readRDS(here::here("data", "cpv2024", "autoidentificacion_municipio.rds"))
pop_age_muni <- readRDS(here::here("data", "cpv2024", "population_age_municipality.rds"))
dept_capital_ids <- paste0(sprintf("%02d", 1:9), "0101")

# Load poverty and basic services data for generate_poverty_service_tables()
poverty_nbi_muni <- readRDS(here::here("data", "cpv2024", "poverty_nbi_muni.rds"))
improved_water_muni <- readRDS(here::here("data", "cpv2024", "improved_water_muni.rds"))
improved_sanitation_muni <- readRDS(here::here("data", "cpv2024", "improved_sanitation_muni.rds"))

# Build departmental summary table (required by generate_poverty_service_tables)
final_table <- bind_rows(
  poverty_nbi_muni |>
    filter(is.na(area) | area == "") |>
    group_by(department) |>
    summarise(
      poor_2012 = sum(poor_total_2012, na.rm = TRUE),
      poor_2024 = sum(poor_total_2024, na.rm = TRUE),
      pop_2012 = sum(ref_pop_nbi_2012, na.rm = TRUE),
      pop_2024 = sum(ref_pop_nbi_2024, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      poverty_pct_2012 = (poor_2012 / pop_2012) * 100,
      poverty_pct_2024 = (poor_2024 / pop_2024) * 100,
      poverty_change = poverty_pct_2024 - poverty_pct_2012
    ) |>
    left_join(
      improved_water_muni |>
        filter(is.na(area) | area == "") |>
        group_by(department) |>
        summarise(
          water_lack_2001 = weighted.mean(pct_unimproved_water_2001, n_total_2001, na.rm = TRUE),
          water_lack_2024 = weighted.mean(pct_unimproved_water_2024, n_total_2024, na.rm = TRUE),
          .groups = "drop"
        ) |>
        mutate(water_change = water_lack_2024 - water_lack_2001),
      by = "department"
    ) |>
    left_join(
      improved_sanitation_muni |>
        filter(is.na(area) | area == "") |>
        group_by(department) |>
        summarise(
          sanitation_lack_2001 = weighted.mean(pct_unimproved_sanitation_2001, n_total_2001, na.rm = TRUE),
          sanitation_lack_2024 = weighted.mean(pct_unimproved_sanitation_2024, n_total_2024, na.rm = TRUE),
          .groups = "drop"
        ) |>
        mutate(sanitation_change = sanitation_lack_2024 - sanitation_lack_2001),
      by = "department"
    ) |>
    select(department, poverty_pct_2012, poverty_pct_2024, poverty_change,
           water_lack_2001, water_lack_2024, water_change,
           sanitation_lack_2001, sanitation_lack_2024, sanitation_change),
  # Add Bolivia national totals
  tibble(
    department = "Bolivia",
    poverty_pct_2012 = (sum(poverty_nbi_muni$poor_total_2012, na.rm = TRUE) /
                        sum(poverty_nbi_muni$ref_pop_nbi_2012, na.rm = TRUE)) * 100,
    poverty_pct_2024 = (sum(poverty_nbi_muni$poor_total_2024, na.rm = TRUE) /
                        sum(poverty_nbi_muni$ref_pop_nbi_2024, na.rm = TRUE)) * 100,
    water_lack_2001 = weighted.mean(improved_water_muni$pct_unimproved_water_2001,
                                    improved_water_muni$n_total_2001, na.rm = TRUE),
    water_lack_2024 = weighted.mean(improved_water_muni$pct_unimproved_water_2024,
                                    improved_water_muni$n_total_2024, na.rm = TRUE),
    sanitation_lack_2001 = weighted.mean(improved_sanitation_muni$pct_unimproved_sanitation_2001,
                                         improved_sanitation_muni$n_total_2001, na.rm = TRUE),
    sanitation_lack_2024 = weighted.mean(improved_sanitation_muni$pct_unimproved_sanitation_2024,
                                         improved_sanitation_muni$n_total_2024, na.rm = TRUE)
  ) |>
    mutate(
      poverty_change = poverty_pct_2024 - poverty_pct_2012,
      water_change = water_lack_2024 - water_lack_2001,
      sanitation_change = sanitation_lack_2024 - sanitation_lack_2001
    ) |>
    select(department, poverty_pct_2012, poverty_pct_2024, poverty_change,
           water_lack_2001, water_lack_2024, water_change,
           sanitation_lack_2001, sanitation_lack_2024, sanitation_change)
)

# Wikipedia presence: which municipalities have an English article?
wiki_presence <- bo_municipality_wiki_presence$data |>
  left_join(
    readRDS(here::here("data", "municipalities_wd_ine.rds")) |>
      select(qid, ine_code),
    by = "qid"
  ) |>
  filter(!is.na(ine_code))

munis_with_enwiki <- wiki_presence |> filter(en > 0) |> pull(ine_code)

# Filter muni_lookup based on params$wiki_filter
if (params$wiki_filter == "existing") {
  muni_lookup <- muni_lookup |> filter(id_muni %in% munis_with_enwiki)
} else if (params$wiki_filter == "new") {
  muni_lookup <- muni_lookup |> filter(!id_muni %in% munis_with_enwiki)
}

# Municipality areas from admin3 boundary file
area_muni <- readxl::read_excel(here::here("maps", "bol_admin_boundaries.xlsx"),
                                  sheet = "bol_admin3") |>
  mutate(id_muni = stringr::str_sub(adm3_pcode, 3, 8)) |>
  select(id_muni, area_sqkm)

# Add id_prov if not already present, then build province-context lookup
muni_lookup <- muni_lookup |>
  mutate(id_prov = stringr::str_sub(id_muni, 1, 4))

prov_counts <- muni_lookup |> dplyr::count(id_prov, name = "n_prov_munis")

muni_province_context <- muni_lookup |>
  select(id_muni, id_prov, muni_anexo) |>
  left_join(prov_counts, by = "id_prov") |>
  left_join(pop_muni,    by = "id_muni") |>
  left_join(area_muni,   by = "id_muni") |>
  left_join(prov_lookup |> select(id_prov, department), by = "id_prov") |>
  group_by(id_prov) |>
  mutate(
    area_rank = rank(-area_sqkm, ties.method = "first"),  # 1 = largest
    pop_rank  = rank(-pop_2024,  ties.method = "first")   # 1 = most populous
  ) |>
  ungroup() |>
  mutate(nat_pop_rank = rank(-pop_2024, ties.method = "first")) |>
  group_by(department) |>
  mutate(dept_pop_rank = rank(-pop_2024, ties.method = "first")) |>
  ungroup()

concejal_expected <- concejal_summary |>
  left_join(pop_muni, by = "id_muni") |>
  mutate(
    es_capital = id_muni %in% dept_capital_ids,
    sillas_esperadas = case_when(
      es_capital | pop_2024 > 75000 ~ 11L,
      pop_2024   > 50000            ~  9L,
      pop_2024   > 15000            ~  7L,
      !is.na(pop_2024)              ~  5L,
      TRUE                          ~ NA_integer_
    )
  )
```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

```{r}
#| label: filter-summary
#| echo: false
#| results: asis
n_total <- nrow(muni_lookup |> filter(id_muni %in% unique(concejo_data$id_muni)))
filter_label <- switch(params$wiki_filter,
  existing = "with an existing English Wikipedia article",
  new      = "without an English Wikipedia article (new)",
  "all (no filter applied)"
)
cat(sprintf("**Rendering %d municipalities %s.**\n\n", n_total, filter_label))
```

```{r}
#| label: council-tables
#| echo: false
#| results: asis
render_muni_blocks(muni_lookup, concejo_data, lang = lang)
```
