---
title: "Bolivia municipality text and figures for Wikipedia"
subtitle: "Wikitext ready to copy to Wikipedia — multiple cited sources"
params:
  wiki_filter: "all"   # "existing", "new", or "all"
format:
  html:
    code-fold: true
    toc: false
    self-contained: true
---

::: {style="font-size: 11px;"}

## Items produced per municipality

Each expanded municipality section contains the following copyable blocks.

---

**Copy lead Wikitext** — one to three sentences identifying the municipality's location, with optional sentences for population rank and province context:

> `'''[Municipality] Municipality'''` is the `[ordinal]` [[Municipalities of Bolivia\|municipal section]] of the `[[Province Province]]` in the `[[Department Department]]` in `[[Bolivia]]`.

*[If the municipality is the 1st or 2nd most populous in Bolivia:]*
> It is the `[most / second most]` populous municipality in `[[Bolivia]]`.

*[If not nationally top-2, but 1st or 2nd most populous in its department:]*
> It is the `[most / second most]` populous municipality in `[[Department Department]]`.

*[If not department top-2, and most populous in a province with 3 or more municipalities:]*
> It is the most populous municipality in `[[Province Province]]`.

*[If the only municipality in its province:]*
> `[Municipality]` Municipality is the only municipality in `[[Province Province]]` and its boundaries are identical.

*[If one of two municipalities in its province — four variants based on relative area and population:]*

| Larger area? | More populous? | Sentence |
|---|---|---|
| Yes | Yes | `[Municipality]` Municipality is the larger and more populous of the two municipalities in `[[Province Province]]`. |
| Yes | No | `[Municipality]` Municipality covers the larger area of the two municipalities in `[[Province Province]]`, but it is less populous. |
| No | Yes | `[Municipality]` Municipality is the more populous of the two municipalities in `[[Province Province]]`, but it covers a smaller area. |
| No | No | `[Municipality]` Municipality is the smaller and less populous of the two municipalities in `[[Province Province]]`. |

*[Omitted for provinces with 3 or more municipalities.]*

---

**Copy government Wikitext** — two paragraphs for a `==Government==` section:

> The Mayor of `[Municipality]` is `[Name]`, who was elected in the [March 2026 elections](https://en.wikipedia.org/wiki/2026_Bolivian_regional_elections).

> `[Municipality]` Municipality has `[5 / 7 / 9 / 11]` council seats, as do all `[municipalities with populations up to 15,000 / between 15,001 and 50,000 / between 50,001 and 75,000 / over 75,000 / department capitals]`. *[IOC municipalities only:]* One of those seats is an Indigenous seat assigned by the `[pueblo]` according to its own norms and procedures.

---

**Copy council table** — a wikitable of council members (one row per seat/type), for the `==Government==` section:

```
{| class="wikitable"
! Seat !! Type !! Name !! Party / Political Organization
|-
| rowspan="2" | [N] || Member || [Name] || rowspan="2" | [Party]
|-
| Alternate || ''[Alternate name]''
...
|}
```

---

**Copy population table** — a horizontal wikitable of census population for `==Population==`:

```
{| class="wikitable"
|+ Population in Recent Censuses
|-
!! [1992] !! 2001 !! 2012 !! 2024
|-
|| [pop_1992] || pop_2001 || pop_2012
|| pop_2024
|}
```

*[1992 column omitted for municipalities whose province was subdivided since 2002.]*

---

**Copy population pyramid** — a `{{Population pyramid 2}}` template for `==Population==`, with 5-year age bins from 0–4 to 85+ derived from the 2024 census:

```
{{Population pyramid 2 |year=2024 |m0=... |f0=... ... |mmax=... |fmax=... |source=2024 Bolivian Census}}
```

---

**Copy identity table** — a wikitable of indigenous self-identification from the 2024 census, for `==Languages and Indigenous identification==`.

**Copy language table** — a wikitable of languages spoken from the 2024 census, for `==Languages and Indigenous identification==`.

---

**Copy communities Wikitext** — a `==Communities==` section with all communities listed in three columns using `{{div col|colwidth=18em}}`. *[Omitted for municipalities with no communities in the 2013 INE geographic classification.]*

---

**Full page Wikitext** is saved to `output/bolivia-municipality-wikitext/[id].txt` and available as a single copy button on the [infobox page](bolivia-muni-infobox-en.html). It assembles all sections above in article order: lead → `==Government==` → `==Population==` → `==Languages and Indigenous identification==` → `==Communities==`.

---

:::


```{r}
#| label: setup
#| include: false

library(dplyr)
library(purrr)
library(htmltools)
library(english)
library(kableExtra)
library(glue)

source("src/wikipedia-tools.R")
source("src/gh-copy-assets.R")
source("src/generate-muni-poverty-tables.R")

dir.create(here::here("output", "bolivia-municipality-wikitext"), recursive = TRUE, showWarnings = FALSE)

# Title case wrapper that preserves lowercase "de" unless it's the first word
title_case_preserve_de <- function(x) {
  words <- strsplit(x, " ")[[1]]
  titled <- stringr::str_to_title(words)
  # If "De" appears after the first position, convert it back to lowercase
  for (i in seq_along(titled)) {
    if (i > 1 && tolower(titled[i]) == "de") {
      titled[i] <- "de"
    }
  }
  paste(titled, collapse = " ")
}

#' Convert Spanish Names and Places to Proper Title Case
#'
#' @param x A character vector of names or geographic entities
#' @return A character vector with proper Spanish title casing
#' @export
spanish_title_case <- function(x) {
  # Clean up leading/trailing whitespace to ensure regex boundaries work perfectly
  x <- trimws(x)
  
  # 1. Convert everything to standard title case
  titled <- stringr::str_to_title(x)
  
  # 2. Define the Spanish particles to be kept lowercase
  particles <- c("De", "Del", "La", "Las", "El", "Los", "Y")
  
  # 3. Build the regex pattern
  # (?<=\\s) : Lookbehind ensuring the particle is preceded by a space. 
  #            This prevents the very first word from being lowercased.
  # (...)    : Matches any of our specific capitalized particles.
  # \\b      : Word boundary ensuring we don't accidentally match prefixes of other words.
  pattern <- paste0("(?<=\\s)(", paste(particles, collapse = "|"), ")\\b")
  
  # 4. Replace matched particles with their lowercase equivalent
  stringr::str_replace_all(titled, pattern, tolower)
}

muni_lookup       <- readRDS(here::here("data", "muni_id_lookup_table.rds"))
prov_lookup       <- readRDS(here::here("data", "prov_id_lookup_table.rds"))
muni_roster       <- readRDS(here::here("data", "oep2026", "muni_roster.rds"))
concejo_data      <- readRDS(here::here("data", "oep2026", "muni_concejo_comb.rds"))
concejal_summary  <- readRDS(here::here("data", "oep2026", "concejal_summary.rds"))
muni_concejal_ioc <- readRDS(here::here("data", "oep2026", "muni_concejal_ioc.rds"))
alcaldes          <- readRDS(here::here("data", "oep2026", "alcaldes.rds"))

bo_municipality_wiki_presence <- readRDS(here::here("data", "bo_municipality_wiki_presence.rds"))
municipalities_wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds"))

# Commons category names (from Wikidata P373 via muni_wd_images.rds), used to
# populate the {{Commons category}} template placed at the end of the article
# body — this replaced the |commons= Infobox settlement field, which has been
# struck from bolivia-muni-infobox-en.qmd.
wd_images <- readRDS(here::here("data", "muni_wd_images.rds")) |>
  select(ine_code, commons_cat)

roster_pages <- muni_roster |> select(id_muni, page)

# Expected seat counts by census population — mirrors cleaning script section 3b
pop_muni         <- readRDS(here::here("data", "cpv2024", "population_municipality.rds")) |>
  select(id_muni = ine_code, pop_2024 = pop_2024_total, pop_2012 = pop_2012_total)

census_pop   <- readRDS(here::here("data", "cpv2024", "census_pop_historical.rds"))
lang_data    <- readRDS(here::here("data", "cpv2024", "idiomas_hablados_muni.rds"))
auto_data    <- readRDS(here::here("data", "cpv2024", "autoidentificacion_municipio.rds"))
pop_age_muni <- readRDS(here::here("data", "cpv2024", "population_age_municipality.rds"))
dept_capital_ids <- paste0(sprintf("%02d", 1:9), "0101")

# Load poverty and basic services data for generate_poverty_service_tables()
poverty_nbi_muni <- readRDS(here::here("data", "cpv2024", "poverty_nbi_muni.rds"))
improved_water_muni <- readRDS(here::here("data", "cpv2024", "improved_water_muni.rds"))
improved_sanitation_muni <- readRDS(here::here("data", "cpv2024", "improved_sanitation_muni.rds"))

# Build departmental summary table (required by generate_poverty_service_tables)
final_table <- bind_rows(
  poverty_nbi_muni |>
    filter(is.na(area) | area == "") |>
    group_by(department) |>
    summarise(
      poor_2012 = sum(poor_total_2012, na.rm = TRUE),
      poor_2024 = sum(poor_total_2024, na.rm = TRUE),
      pop_2012 = sum(ref_pop_nbi_2012, na.rm = TRUE),
      pop_2024 = sum(ref_pop_nbi_2024, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      poverty_pct_2012 = (poor_2012 / pop_2012) * 100,
      poverty_pct_2024 = (poor_2024 / pop_2024) * 100,
      poverty_change = poverty_pct_2024 - poverty_pct_2012
    ) |>
    left_join(
      improved_water_muni |>
        filter(is.na(area) | area == "") |>
        group_by(department) |>
        summarise(
          water_lack_2001 = weighted.mean(pct_unimproved_water_2001, n_total_2001, na.rm = TRUE),
          water_lack_2024 = weighted.mean(pct_unimproved_water_2024, n_total_2024, na.rm = TRUE),
          .groups = "drop"
        ) |>
        mutate(water_change = water_lack_2024 - water_lack_2001),
      by = "department"
    ) |>
    left_join(
      improved_sanitation_muni |>
        filter(is.na(area) | area == "") |>
        group_by(department) |>
        summarise(
          sanitation_lack_2001 = weighted.mean(pct_unimproved_sanitation_2001, n_total_2001, na.rm = TRUE),
          sanitation_lack_2024 = weighted.mean(pct_unimproved_sanitation_2024, n_total_2024, na.rm = TRUE),
          .groups = "drop"
        ) |>
        mutate(sanitation_change = sanitation_lack_2024 - sanitation_lack_2001),
      by = "department"
    ) |>
    select(department, poverty_pct_2012, poverty_pct_2024, poverty_change, 
           water_lack_2001, water_lack_2024, water_change, 
           sanitation_lack_2001, sanitation_lack_2024, sanitation_change),
  # Add Bolivia national totals
  tibble(
    department = "Bolivia",
    poverty_pct_2012 = (sum(poverty_nbi_muni$poor_total_2012, na.rm = TRUE) / 
                        sum(poverty_nbi_muni$ref_pop_nbi_2012, na.rm = TRUE)) * 100,
    poverty_pct_2024 = (sum(poverty_nbi_muni$poor_total_2024, na.rm = TRUE) / 
                        sum(poverty_nbi_muni$ref_pop_nbi_2024, na.rm = TRUE)) * 100,
    water_lack_2001 = weighted.mean(improved_water_muni$pct_unimproved_water_2001, 
                                    improved_water_muni$n_total_2001, na.rm = TRUE),
    water_lack_2024 = weighted.mean(improved_water_muni$pct_unimproved_water_2024, 
                                    improved_water_muni$n_total_2024, na.rm = TRUE),
    sanitation_lack_2001 = weighted.mean(improved_sanitation_muni$pct_unimproved_sanitation_2001, 
                                         improved_sanitation_muni$n_total_2001, na.rm = TRUE),
    sanitation_lack_2024 = weighted.mean(improved_sanitation_muni$pct_unimproved_sanitation_2024, 
                                         improved_sanitation_muni$n_total_2024, na.rm = TRUE)
  ) |>
    mutate(
      poverty_change = poverty_pct_2024 - poverty_pct_2012,
      water_change = water_lack_2024 - water_lack_2001,
      sanitation_change = sanitation_lack_2024 - sanitation_lack_2001
    ) |>
    select(department, poverty_pct_2012, poverty_pct_2024, poverty_change, 
           water_lack_2001, water_lack_2024, water_change, 
           sanitation_lack_2001, sanitation_lack_2024, sanitation_change)
)

# Wikipedia presence: which municipalities have an English article?
wiki_presence <- bo_municipality_wiki_presence$data |>
  left_join(
    readRDS(here::here("data", "municipalities_wd_ine.rds")) |>
      select(qid, ine_code),
    by = "qid"
  ) |>
  filter(!is.na(ine_code))

munis_with_enwiki <- wiki_presence |> filter(en > 0) |> pull(ine_code)

# Filter muni_lookup based on params$wiki_filter
if (params$wiki_filter == "existing") {
  muni_lookup <- muni_lookup |> filter(id_muni %in% munis_with_enwiki)
} else if (params$wiki_filter == "new") {
  muni_lookup <- muni_lookup |> filter(!id_muni %in% munis_with_enwiki)
}

# Municipality areas from admin3 boundary file
area_muni <- readxl::read_excel(here::here("maps", "bol_admin_boundaries.xlsx"),
                                  sheet = "bol_admin3") |>
  mutate(id_muni = stringr::str_sub(adm3_pcode, 3, 8)) |>
  select(id_muni, area_sqkm)

# Add id_prov if not already present, then build province-context lookup
muni_lookup <- muni_lookup |>
  mutate(id_prov = stringr::str_sub(id_muni, 1, 4))

prov_counts <- muni_lookup |> dplyr::count(id_prov, name = "n_prov_munis")

muni_province_context <- muni_lookup |>
  select(id_muni, id_prov, muni_anexo) |>
  left_join(prov_counts, by = "id_prov") |>
  left_join(pop_muni,    by = "id_muni") |>
  left_join(area_muni,   by = "id_muni") |>
  left_join(prov_lookup |> select(id_prov, department), by = "id_prov") |>
  group_by(id_prov) |>
  mutate(
    area_rank = rank(-area_sqkm, ties.method = "first"),  # 1 = largest
    pop_rank  = rank(-pop_2024,  ties.method = "first")   # 1 = most populous
  ) |>
  ungroup() |>
  mutate(nat_pop_rank = rank(-pop_2024, ties.method = "first")) |>
  group_by(department) |>
  mutate(dept_pop_rank = rank(-pop_2024, ties.method = "first")) |>
  ungroup()

concejal_expected <- concejal_summary |>
  left_join(pop_muni, by = "id_muni") |>
  mutate(
    es_capital = id_muni %in% dept_capital_ids,
    sillas_esperadas = case_when(
      es_capital | pop_2024 > 75000 ~ 11L,
      pop_2024   > 50000            ~  9L,
      pop_2024   > 15000            ~  7L,
      !is.na(pop_2024)              ~  5L,
      TRUE                          ~ NA_integer_
    )
  )
```


```{r}
#| label: helpers
#| include: false

# Process wikitext for in-page display:
#   - <ref>...</ref>  →  superscript [refN] (numbered from ref_offset + 1)
#   - [[Target|Disp]] →  HTML anchor
#   - everything else HTML-escaped so wikitext syntax shows literally
# Returns list(html = character, refs = character vector of ref contents).
process_for_display <- function(text, ref_offset = 0L) {
  refs <- character(0)

  # Step 1: swap <ref>...</ref> for safe placeholder tokens
  processed <- stringr::str_replace_all(text, "<ref>(.*?)</ref>", function(m) {
    vapply(m, function(single_m) {
      n    <- length(refs) + 1L
      refs <<- c(refs, stringr::str_match(single_m, "<ref>(.*?)</ref>")[1, 2])
      sprintf("REFSUP_%d_END", n + ref_offset)
    }, character(1), USE.NAMES = FALSE)
  })

  # Step 2: HTML-escape (turns < > & into entities, leaves placeholders intact)
  escaped <- htmltools::htmlEscape(processed)

  # Step 3: convert [[wikilinks]] to anchors
  escaped <- stringr::str_replace_all(
    escaped, "\\[\\[([^|\\]]+)\\|([^\\]]+)\\]\\]",
    function(m) {
      p <- stringr::str_match(m, "\\[\\[([^|\\]]+)\\|([^\\]]+)\\]\\]")
      paste0('<a href="', wikipedia_url_for(p[, 2]), '" target="_blank">', p[, 3], "</a>")
    }
  )
  escaped <- stringr::str_replace_all(
    escaped, "\\[\\[([^\\]]+)\\]\\]",
    function(m) {
      tgt <- stringr::str_match(m, "\\[\\[([^\\]]+)\\]\\]")[, 2]
      paste0('<a href="', wikipedia_url_for(tgt), '" target="_blank">', tgt, "</a>")
    }
  )

  # Step 4: restore placeholders as blue superscripts
  for (i in seq_along(refs)) {
    n_abs <- i + ref_offset
    escaped <- stringr::str_replace_all(
      escaped,
      sprintf("REFSUP_%d_END", n_abs),
      sprintf('<sup style="color:#06c;font-size:9px;">[ref%d]</sup>', n_abs)
    )
  }

  list(html = escaped, refs = refs)
}

# Render a footnotes section below a display block.
refs_footnotes <- function(refs, ref_offset = 0L) {
  if (length(refs) == 0) return(NULL)
  tags$div(
    style = "margin-top:5px; padding-top:4px; border-top:1px solid #e1e4e8;",
    lapply(seq_along(refs), function(i) {
      tags$p(
        tags$b(paste0("ref", i + ref_offset, ":")), " ", refs[[i]],
        style = "font-size:10px; font-family:monospace; margin:1px 0; color:#555; word-break:break-all;"
      )
    })
  )
}

# Wikitext copy block: button + display (with ref superscripts) + hidden textarea.
wikitext_block <- function(id, label, content) {
  disp <- process_for_display(content)
  tagList(
    tags$button(
      class = "copy-btn",
      style = "margin-bottom:6px;",
      onclick = sprintf("copyTableToClipboard('%s', this)", id),
      label
    ),
    tags$pre(
      HTML(disp$html),
      style = paste(
        "background-color:#f6f8fa; padding:10px; border-radius:6px;",
        "overflow-x:auto; font-size:11px; white-space:pre; margin:0;"
      )
    ),
    refs_footnotes(disp$refs),
    tags$textarea(
      id = id,
      style = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
      readonly = TRUE,
      content
    )
  )
}

# Citation for INE 2024 census population data, appended to all population sentences
.ine_pop_ref <- paste0(
  "{{Cite",
  "| last = Instituto Nacional de Estadística",
  "| title = Población censada e indicadores demográficos: Tabulados por Municipio/TIOC",
  "| url = https://nube.ine.gob.bo/index.php/s/0OrfPjzMB16fATQ/download",
  "| date = 2025}}"
)

.ine_lang_ref <- paste0(
  "{{Cite",
  "| last = Instituto Nacional de Estadística",
  "| title = Idioma en segundo orden de uso, idioma en tercer orden de uso e idiomas hablados (independientemente del orden de uso) de la población de 6 o más años de edad: Tabulados por Municipio/TIOC",
  "| url = https://nube.ine.gob.bo/index.php/s/BSHb8IVQGA997Pv/download",
  "| date = 2025}}"
)

.ine_identity_ref <- paste0(
  "{{Cite",
  "| last = Instituto Nacional de Estadística",
  "| title = Autoidentificación con alguna nación pueblo indígena originario campesino y afroboliviano: Tabulados por Municipio/TIOC",
  "| url = https://nube.ine.gob.bo/index.php/s/Mu3eStaUZfi97Vp/download",
  "| date = 2025}}"
)

# Compose a population-rank sentence for municipalities that are among the most
# populous nationally (top 2), in their department (top 2), or in their province
# (rank 1, only for provinces with 3+ municipalities). Returns NULL when none apply.
compose_pop_rank_sentence_en <- function(id_muni, muni_pc) {
  row <- muni_pc[muni_pc$id_muni == id_muni, ]

  # National top-2
  if (!is.na(row$nat_pop_rank) && row$nat_pop_rank <= 2) {
    ord <- if (row$nat_pop_rank == 1) "most" else "second most"
    return(paste0("It is the ", ord, " populous municipality in [[Bolivia]].",
                  "<ref>", .ine_pop_ref, "</ref>"))
  }

  # Department top-2
  if (!is.na(row$dept_pop_rank) && row$dept_pop_rank <= 2) {
    ord      <- if (row$dept_pop_rank == 1) "most" else "second most"
    dept_lnk <- paste0("[[", row$department, " Department]]")
    return(paste0("It is the ", ord, " populous municipality in ", dept_lnk, ".",
                  "<ref>", .ine_pop_ref, "</ref>"))
  }

  # Province most populous (only for provinces with 3+ municipalities)
  if (!is.na(row$pop_rank) && row$pop_rank == 1 && row$n_prov_munis >= 3) {
    id_prov   <- stringr::str_sub(id_muni, 1, 4)
    prov_name <- prov_lookup$province_anexo[prov_lookup$id_prov == id_prov]
    return(paste0("It is the most populous municipality in [[",
                  prov_name, " Province]].",
                  "<ref>", .ine_pop_ref, "</ref>"))
  }

  NULL
}

# Compose the province-context sentence appended to the lead, for municipalities
# in 1- or 2-municipality provinces. Returns NULL for 3+ municipality provinces.
compose_province_context_en <- function(id_muni, muni_pc, prov_lkp) {
  row       <- muni_pc[muni_pc$id_muni == id_muni, ]
  id_prov   <- stringr::str_sub(id_muni, 1, 4)
  prov_name <- prov_lkp$province_anexo[prov_lkp$id_prov == id_prov]
  muni_name <- row$muni_anexo
  prov_link <- paste0("[[", prov_name, " Province]]")
  n         <- row$n_prov_munis

  if (n == 1) {
    return(paste0(
      muni_name, " Municipality is the only municipality in ",
      prov_link, " and its boundaries are identical."
    ))
  }

  if (n == 2 && !is.na(row$area_rank) && !is.na(row$pop_rank)) {
    larger   <- row$area_rank == 1
    more_pop <- row$pop_rank  == 1
    return(dplyr::case_when(
       larger &  more_pop ~ paste0(muni_name, " Municipality is the larger and more populous of the two municipalities in ", prov_link, "."),
       larger & !more_pop ~ paste0(muni_name, " Municipality covers the larger area of the two municipalities in ", prov_link, ", but it is less populous."),
      !larger &  more_pop ~ paste0(muni_name, " Municipality is the more populous of the two municipalities in ", prov_link, ", but it covers a smaller area."),
      TRUE                ~ paste0(muni_name, " Municipality is the smaller and less populous of the two municipalities in ", prov_link, ".")
    ))
  }

  NULL
}

# Compose the Wikipedia lead sentence for a municipality (wikitext).
# Uses muni_anexo and province_anexo as display text; ordinal from last 2 digits
# of id_muni via english::ordinal().
compose_lead_sentence_en <- function(id_muni, muni_lkp, prov_lkp) {
  muni_row <- muni_lkp[muni_lkp$id_muni == id_muni, ]
  id_prov  <- substr(id_muni, 1, 4)
  prov_row <- prov_lkp[prov_lkp$id_prov == id_prov, ]
  muni_num <- as.integer(substr(id_muni, 5, 6))

  lead <- paste0(
    "'''", muni_row$muni_anexo, " Municipality''' is the ",
    english::ordinal(muni_num),
    " [[Municipalities of Bolivia|municipal section]] of the",
    " [[", prov_row$province_anexo, " Province]]",
    " in the [[", prov_row$department, " Department]]",
    " in [[Bolivia]]."
  )

  pop_rank <- compose_pop_rank_sentence_en(id_muni, muni_province_context)
  context  <- compose_province_context_en(id_muni, muni_province_context, prov_lkp)

  parts <- c(lead, pop_rank, context)
  paste(parts[!sapply(parts, is.null)], collapse = " ")
}

# Compose the mayor sentence for a municipality (wikitext).
compose_mayor_sentence_en <- function(id_muni, muni_name, alcaldes_data,
                                       page_number) {
  row    <- alcaldes_data[alcaldes_data$id_muni == id_muni, ]
  nombre <- spanish_title_case(row$nombre[1])
  ref    <- paste0("<ref>",
                   source_ref_with_page_number(page_number, source_ref_en),
                   "</ref>")
  paste0(
    "The Mayor of ", muni_name, " is ", nombre,
    ", who was elected in the ",
    "[[2026 Bolivian regional elections|March 2026 elections]].",
    ref
  )
}

# Compose the council intro sentence for a municipality.
# Returns one sentence for non-IOC municipalities, two for IOC ones.
compose_council_sentence_en <- function(id_muni, muni_name,
                                         concejal_exp, ioc_data) {
  row <- concejal_exp[concejal_exp$id_muni == id_muni, ]

  range_phrase <- dplyr::case_when(
    row$es_capital & row$pop_2024 > 75000 ~
      "department capitals and municipalities with populations over 75,000",
    row$es_capital ~
      "department capitals",
    row$pop_2024 > 75000 ~
      "municipalities with populations over 75,000",
    row$pop_2024 > 50000 ~
      "municipalities with populations between 50,001 and 75,000",
    row$pop_2024 > 15000 ~
      "municipalities with populations between 15,001 and 50,000",
    TRUE ~
      "municipalities with populations up to 15,000"
  )

  s1 <- paste0(
    muni_name, " has a municipal council with ", english::words(row$sillas_esperadas),
    " seats, as do all ", range_phrase, ".",
    "<ref>{{Cite news| title = A más población, los municipios tienen más concejales, según la ley| work = Los Tiempos| access-date = 2026-06-19| date = 2026-02-08| url = https://www.lostiempos.com/actualidad/pais/20260208/mas-poblacion-municipios-tienen-mas-concejales-ley}}</ref>"
  )

  ioc_rows <- ioc_data[ioc_data$id_muni == id_muni, ]
  if (nrow(ioc_rows) > 0) {
    pueblo <- spanish_title_case(ioc_rows$pueblo[1])
    s2 <- paste0(
      "One of those seats is an Indigenous seat assigned by the ",
      pueblo, " according to its own norms and procedures."
    )
    paste(s1, s2)
  } else {
    s1
  }
}

# Citations for census population table (2001/2012/2024 and 1992, distinct from .ine_pop_ref used in lead sentences)
.census_source_ref <- paste0(
  '{{Cite',
  '| last = Instituto Nacional de Estadística',
  '| title = Población censada e indicadores demográficos: Tabulados por Municipio/TIOC',
  '| date = 2025',
  '| url = https://nube.ine.gob.bo/index.php/s/0OrfPjzMB16fATQ/download',
  '}}'
)

.census_source_ref_1992 <- paste0(
  '{{Cite',
  '| last = Instituto Nacional de Estadística',
  '| title = CENSO NACIONAL DE POBLACIÓN Y VIVIENDA 1992: Ficha Municipal',
  '| access-date = 2026-06-22',
  '| date = 2018-12-17',
  '| url = https://anda.ine.gob.bo/index.php/catalog/47',
  '}}'
)

# Build a horizontal "Population in Recent Censuses" wikitable for one municipality.
# Includes 1992 only for municipalities whose province was not subdivided since 2002.
# References are passed separately: census_ref for 2001/2012/2024, census_ref_1992 for 1992.
build_census_pop_wikitable <- function(id, census_data, census_ref, census_ref_1992 = NULL) {
  row <- census_data[census_data$id_muni == id, ]
  if (nrow(row) == 0) return(NULL)
  row <- row[1, ]
  
  fmt <- function(x) formatC(x, format = "d", big.mark = ",")
  has_1992 <- !is.na(row$pop_1992)
  
  # Wrap bare citation templates in <ref> tags
  ref_main <- paste0("<ref>", census_ref, "</ref>")
  ref_1992 <- if (!is.null(census_ref_1992)) paste0("<ref>", census_ref_1992, "</ref>") else ""
  
  years <- if (has_1992) paste0("!! 1992", ref_1992, " !! 2001 !! 2012 !! 2024", ref_main) else paste0("!! 2001 !! 2012 !! 2024", ref_main)
  

  
  # Build the values row with references: 1992 gets its own ref if present
  if (has_1992) {
    vals <- paste0(
      "|| ", fmt(row$pop_1992),
      " || ", fmt(row$pop_2001),
      " || ", fmt(row$pop_2012), "\n|| ", fmt(row$pop_2024)
    )
  } else {
    vals <- paste0(
      "|| ", fmt(row$pop_2001), 
      " || ", fmt(row$pop_2012), "\n|| ", fmt(row$pop_2024)
    )
  }
  
  paste0(
    '{| class="wikitable"\n',
    '|+ Population in Recent Censuses\n',
    '|-\n',
    years, '\n',
    '|-\n',
    vals, '\n',
    '|}'
  )
}

# Build a {{Population pyramid 2}} wikitext template for one municipality.
build_pop_pyramid <- function(id, pop_data) {
  d <- pop_data |>
    filter(ine_code == id, area == "Total", age_group != "Total")
  if (nrow(d) == 0) return(NULL)

  total_pop <- pop_data |>
    filter(ine_code == id, area == "Total", age_group == "Total") |>
    pull(pop_2024_total)

  # Map age groups to template bin numbers; 85+ groups all map to bin 85
  bin_map <- c(
    "0 a 4" = 0, "5 a 9" = 5, "10 a 14" = 10, "15 a 19" = 15,
    "20 a 24" = 20, "25 a 29" = 25, "30 a 34" = 30, "35 a 39" = 35,
    "40 a 44" = 40, "45 a 49" = 45, "50 a 54" = 50, "55 a 59" = 55,
    "60 a 64" = 60, "65 a 69" = 65, "70 a 74" = 70, "75 a 79" = 75,
    "80 a 84" = 80, "85 a 89" = 85, "90 a 94" = 85, "95 o m\u00e1s" = 85
  )

  d <- d |>
    mutate(bin = bin_map[age_group]) |>
    summarise(
      male   = sum(pop_2024_male,   na.rm = TRUE),
      female = sum(pop_2024_female, na.rm = TRUE),
      .by = bin
    ) |>
    mutate(
      m_pct = round(male   / total_pop * 100, 1),
      f_pct = round(female / total_pop * 100, 1)
    ) |>
    arrange(bin)

  m_parts <- paste0("|m", d$bin, "=", d$m_pct, collapse = " ")
  f_parts <- paste0("|f", d$bin, "=", d$f_pct, collapse = " ")
  mmax <- ceiling(max(d$m_pct) * 2) / 2
  fmax <- ceiling(max(d$f_pct) * 2) / 2

  paste0(
    "{{Population pyramid\n",
    "|year=2024\n",
    m_parts, "\n",
    f_parts, "\n",
    "|mmax=", mmax, "\n",
    "|fmax=", fmax, "\n",
    "|source=2024 Bolivian Census\n",
    "|sourceref=",.census_source_ref,"\n",
    "}}"
  )
}


# I made the Qid for an online source that verifies the information:
#  Instituto Nacional de Estadística. “Clasificación de Ubicación
#  Geográfica a Nivel Comunidad.” Instituto Nacional de Estadística,
#  January 1, 2013. https://anda.ine.gob.bo/index.php/catalog/71/download/716.
#  reference_qid = "Q138354774"

# Build community list
ine_geog_2013 <- readxl::read_excel("data/CLASIF_UB_GEOG_COMUNIDAD.xlsx")
ine_geog_2013 <- ine_geog_2013 %>%
  rename(
    department = DEPARTAMENTO,
    province = PROVINCIA,
    municipality = MUNICIPIO,
    community = 8) %>%
  mutate(cod.prov = paste0(DEP, PRO),
         cod.mun = paste0(DEP, PRO, MUN),
         cod.com = Codigo) %>%
  rename(cod.dep = DEP)

ine_geog_2013_ref <- "{{Cite| last = Instituto Nacional de Estadística| title = Clasificación de Ubicación Geográfica a Nivel Comunidad| date = 2013-01-01 | url = https://anda.ine.gob.bo/index.php/catalog/71/download/716 }}"

n_muni_ine_geog_2013 <- ine_geog_2013 %>%
  distinct(department, province, municipality) %>%
  nrow()
  
community_list_df <- ine_geog_2013 %>%
  mutate(community = sapply(community, spanish_title_case)) %>%
  group_by(cod.mun) %>%
  summarise(n_communities = n_distinct(community),
             communities = paste0("* ", paste0(community, collapse = "\n* ")),
             communities_html = paste0("<ul style='column-count: 3; column-width: 18em; margin: 0; padding-left: 20px;'>", 
                                       paste0("<li>", community, "</li>", collapse = ""),
                                       "</ul>")) %>%
  ungroup()

# Note there is no check for applicability since there's only one new 
# municipality created since this dataset
compose_community_list_wikitext <- function(id_muni) {
  row <- community_list_df %>%
    filter(cod.mun == id_muni)
  
  if (nrow(row) == 0) {
    return(NULL)
  }
  
  if (id_muni %in% c("050405", "050401")) {
    row <- row %>%
      filter(cod.mun == "050401")
    
    return(list(
      wikitext = paste0(
        "== Communities ==\n",
        "Communities were divided between San Pedro de Macha and Colquechaca when ",
        "the municipalities were split in 2019. The following communities were in ",
        "the original municipality of Colquechaca:",
        "<ref>", ine_geog_2013_ref, "</ref>\n",
        "{{div col|colwidth=18em}}\n", row$communities,
        "\n{{div col end}}",
        "\n\n"
      ),
      html = paste0(
        "<h3>Communities</h3><p>Communities were divided between San Pedro de Macha and Colquechaca when the municipalities were split in 2019. The following communities were in the original municipality of Colquechaca:</p>",
        row$communities_html
      )
    ))
  }

    
  list(
    wikitext = paste0(
      "== Communities ==\n",
      "The following communities are located within the municipality:",
      "<ref>", ine_geog_2013_ref, "</ref>\n",
      "{{div col|colwidth=18em}}\n",
      row$communities,
      "\n{{div col end}}",
      "\n\n"
    ),
    html = paste0(
      "<h3>Communities</h3><p>The following communities are located within the municipality:</p>",
      row$communities_html
    )
  )
}


muni_block <- function(id_muni, muni_name, department, concejo) {
  page <- roster_pages$page[roster_pages$id_muni == id_muni]
  src  <- source_ref_with_page_number(page, source_ref_en)

  # Lead sentence
  lead_wt   <- compose_lead_sentence_en(id_muni, muni_lookup, prov_lookup)
  lead_disp <- process_for_display(lead_wt)

  # Government section
  mayor_wt   <- compose_mayor_sentence_en(id_muni, muni_name, alcaldes, page)
  council_wt <- compose_council_sentence_en(id_muni, muni_name,
                                             concejal_expected, muni_concejal_ioc)
  intro_wt   <- paste0(mayor_wt, "\n\n", council_wt)

  mayor_disp   <- process_for_display(mayor_wt, ref_offset = 0L)
  council_disp <- process_for_display(council_wt, ref_offset = length(mayor_disp$refs))
  all_intro_refs <- c(mayor_disp$refs, council_disp$refs)

  # Pre-compute all wikitext sections for individual buttons and full-page assembly
  council_wikitable <- build_council_wikitable_en(id_muni, data = concejo,
                                                   source = source_wikitext_en,
                                                   source_refs = src)
  autoident_wt  <- build_autoident_wikitable_en(id_muni, auto_data, pop_muni,
                                                 source = "2024 Bolivian Census",
                                                 source_ref = .ine_identity_ref)
  lang_wt       <- build_language_wikitable_en(id_muni, lang_data, pop_muni,
                                               source = "2024 Bolivian Census",
                                               source_ref = .ine_lang_ref)
  census_pop_wt <- build_census_pop_wikitable(id_muni, census_pop,
                                              .census_source_ref, .census_source_ref_1992)
  pyramid_wt    <- build_pop_pyramid(id_muni, pop_age_muni)
  comm          <- compose_community_list_wikitext(id_muni)
  reflist_wt <- "== References ==/n{{Reflist}}/n/n"

  # Commons category template — value from Wikidata (P373) when known,
  # otherwise left as a blank placeholder for manual completion.
  commons_cat_row <- wd_images |> filter(ine_code == id_muni) |> slice(1)
  commons_cat_val <- if (nrow(commons_cat_row) > 0 && !is.na(commons_cat_row$commons_cat)) {
    commons_cat_row$commons_cat
  } else " "
  commons_cat_wt <- sprintf("{{Commons category|%s}}", commons_cat_val)

  # Assemble full page wikitext in section order and save to file
  full_page_wt <- paste0(
    lead_wt, "\n\n",
    "==Government==\n",
    mayor_wt, "\n\n",
    council_wt, "\n\n",
    council_wikitable, "\n\n",
    "==Population==\n",
    if (!is.null(census_pop_wt)) paste0(census_pop_wt, "\n\n") else "",
    if (!is.null(pyramid_wt))    paste0(pyramid_wt,    "\n\n") else "",
    "==Languages and Indigenous identification==\n",
    autoident_wt, "\n\n",
    lang_wt,
    if (!is.null(comm)) paste0("\n\n", comm$wikitext) else "",
    reflist_wt, "\n",
    "==External links==\n",
    commons_cat_wt, "\n"
  )
  writeLines(full_page_wt,
             here::here("output", "bolivia-municipality-wikitext", paste0(id_muni, ".txt")))

  tagList(
    tags$details(
      tags$summary(
        tags$strong(muni_name),
        tags$span(
          paste0(" \u2014 ", department),
          style = "color: #586069; font-weight: normal;"
        ),
        tags$code(
          paste0(" [", id_muni, "]"),
          style = "font-size: 11px; color: #999; margin-left: 6px;"
        )
      ),
      # Lead sentence — separate copy block
      tags$div(
        style = "margin-top:8px; margin-bottom:10px;",
        tags$button(
          class = "copy-btn",
          style = "margin-bottom:6px;",
          onclick = sprintf("copyTableToClipboard('lead_en_%s', this)", id_muni),
          "Copy lead Wikitext"
        ),
        tags$p(HTML(lead_disp$html), style = "font-size:12px; margin:4px 0 0; color:#333;"),
        tags$textarea(
          id       = paste0("lead_en_", id_muni),
          style    = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
          readonly = TRUE,
          lead_wt
        )
      ),
      # Government section — mayor + council intro, one button copies both
      tags$div(
        style = "margin-top:8px; margin-bottom:10px;",
        tags$button(
          class = "copy-btn",
          style = "margin-bottom:6px;",
          onclick = sprintf("copyTableToClipboard('intro_en_%s', this)", id_muni),
          "Copy government Wikitext"
        ),
        tags$p(HTML(mayor_disp$html),   style = "font-size:12px; margin:4px 0 4px; color:#333;"),
        tags$p(HTML(council_disp$html),  style = "font-size:12px; margin:0 0 0;    color:#333;"),
        refs_footnotes(all_intro_refs),
        tags$textarea(
          id       = paste0("intro_en_", id_muni),
          style    = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
          readonly = TRUE,
          intro_wt
        )
      ),
      # Side-by-side: council kable | wikitable
      tags$div(
        style = "display:flex; gap:14px; align-items:flex-start; font-size:12px; margin-bottom:8px;",
        tags$div(
          style = "flex:1 1 0; min-width:0; overflow-x:auto;",
          build_council_kable_en(id_muni, data = concejo)
        ),
        tags$div(
          style = "flex:1 1 0; min-width:0;",
          wikitext_block(
            id      = paste0("concejo_en_", id_muni),
            label   = "Copy council table",
            content = council_wikitable
          )
        )
      ),
      # Communities section
      {
        if (!is.null(comm)) {
          tags$div(
            style = "margin-top:8px; margin-bottom:10px;",
            tags$button(
              class = "copy-btn",
              style = "margin-bottom:6px;",
              onclick = sprintf("copyTableToClipboard('communities_en_%s', this)", id_muni),
              "Copy communities Wikitext"
            ),
            HTML(comm$html),
            tags$textarea(
              id       = paste0("communities_en_", id_muni),
              style    = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
              readonly = TRUE,
              comm$wikitext
            )
          )
        }
      },
      # Side-by-side: ethnic identification kable | wikitable
      tags$div(
        style = "display:flex; gap:14px; align-items:flex-start; font-size:12px; margin-bottom:8px;",
        tags$div(
          style = "flex:1 1 0; min-width:0; overflow-x:auto;",
          build_autoident_kable_en(id_muni, auto_data, pop_muni)
        ),
        tags$div(
          style = "flex:1 1 0; min-width:0;",
          wikitext_block(
            id      = paste0("autoident_en_", id_muni),
            label   = "Copy identity table",
            content = autoident_wt
          )
        )
      ),
      # Side-by-side: language kable | wikitable
      tags$div(
        style = "display:flex; gap:14px; align-items:flex-start; font-size:12px; margin-bottom:8px;",
        tags$div(
          style = "flex:1 1 0; min-width:0; overflow-x:auto;",
          build_language_kable_en(id_muni, lang_data, pop_muni)
        ),
        tags$div(
          style = "flex:1 1 0; min-width:0;",
          wikitext_block(
            id      = paste0("lang_en_", id_muni),
            label   = "Copy language table",
            content = lang_wt
          )
        )
      ),
      # Census population table
      {
        if (!is.null(census_pop_wt)) {
          tags$div(
            style = "font-size:12px;",
            wikitext_block(
              id      = paste0("censuspop_en_", id_muni),
              label   = "Copy population table",
              content = census_pop_wt
            )
          )
        }
      },
      # Population pyramid wikitext
      {
        if (!is.null(pyramid_wt)) {
          tags$div(
            style = "font-size:12px;",
            wikitext_block(
              id      = paste0("pyramid_en_", id_muni),
              label   = "Copy population pyramid",
              content = pyramid_wt
            )
          )
        }
      },
      # Poverty and basic services section
      {
        poverty_tables <- generate_poverty_service_tables(
          ine_code = id_muni,
          muni_display_name = muni_name,
          dept_display_name = department
        )
        
        tags$div(
          style = "margin-top:16px; padding-top:12px; border-top:1px solid #e1e4e8;",
          tags$h4("Poverty and Basic Services", style = "margin:0 0 8px 0;"),
          tags$div(
            style = "display:flex; gap:14px; align-items:flex-start; font-size:12px;",
            tags$div(
              style = "flex:1 1 0; min-width:0; overflow-x:auto;",
              HTML(as.character(poverty_tables$kable))
            ),
            tags$div(
              style = "flex:1 1 0; min-width:0;",
              wikitext_block(
                id      = paste0("poverty_en_", id_muni),
                label   = "Copy poverty table",
                content = poverty_tables$wikitable
              )
            )
          )
        )
      }
    ),
    tags$hr(style = "border:none; border-top:1px solid #e1e4e8; margin:4px 0;")
  )
}

render_muni_blocks <- function(lookup, concejo) {
  ids <- lookup |>
    filter(id_muni %in% unique(concejo$id_muni)) |>
    arrange(id_muni) |>
    select(id_muni, muni_anexo, department)

  tagList(pmap(ids, function(id_muni, muni_anexo, department) {
    muni_block(id_muni, muni_anexo, department, concejo)
  }))
}
```

```{r source-data}
source_wikitext_en = "Supreme Electoral Tribunal."
source_wikitext_es = "Tribunal Supremo Electoral."

source_ref_en  <- "{{Cite report| publisher = Órgano Electoral Plurinacional| last1 = Órgano Electoral Plurinacional| last2 = Tribunal Supremo Electoral| title = Publicación de resultados Elecciones Subnacionales 2026: Elección de Autoridades Políticas Departamentales Regionales y Municipales 2026 (Primera y Segunda Vuelta Electoral)| series = Separata de información pública| date = 2026-04-23}}"
source_ref_es <- source_ref_en

source_ref_with_page_number <- function(page_number, source_ref) {
  source_ref_minus_end <- sub("\\}\\}$", "", source_ref)
  paste0(source_ref_minus_end, " | page = ", page_number, " }}")
}

```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

```{r}
#| label: filter-summary
#| echo: false
#| results: asis
n_total <- nrow(muni_lookup |> filter(id_muni %in% unique(concejo_data$id_muni)))
filter_label <- switch(params$wiki_filter,
  existing = "with an existing English Wikipedia article",
  new      = "without an English Wikipedia article (new)",
  "all (no filter applied)"
)
cat(sprintf("**Rendering %d municipalities %s.**\n\n", n_total, filter_label))
```

```{r}
#| label: council-tables
#| echo: false
#| results: asis
render_muni_blocks(muni_lookup, concejo_data)
```
