---
title: "Bolivia 2026 — Municipal Council Tables"
subtitle: "Wikitext ready to copy to Wikipedia"
format:
  html:
    code-fold: true
    toc: false
    self-contained: true
---

## Template

Each municipality section contains two copyable blocks:

**Copy intro** — two paragraphs:

> The Mayor of `[Municipality]` is `[Name]`, who was elected in the [March 2026 elections](https://en.wikipedia.org/wiki/2026_Bolivian_regional_elections).

> `[Municipality]` Municipality has `[5 / 7 / 9 / 11]` council seats, as do all `[municipalities with populations up to 15,000]` / `[between 15,001 and 50,000]` / `[between 50,001 and 75,000]` / `[over 75,000]` / `[department capitals]`. *(IOC municipalities only:)* One of those seats is an Indigenous seat assigned by the `[pueblo]` according to its own norms and procedures.

**Copy council table** — a wikitable of council members, one row per seat/type:

```
{| class="wikitable"
! Seat !! Type !! Name !! Party / Political Organization
|-
| rowspan="2" | [N] || Member || [Name] || rowspan="2" | [Party]
|-
| Alternate || ''[Alternate name]''
...
|}
```

---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(purrr)
library(htmltools)
library(english)

source("src/wikipedia-tools.R")
source("src/gh-copy-assets.R")

muni_lookup       <- readRDS(here::here("data", "muni_id_lookup_table.rds"))
prov_lookup       <- readRDS(here::here("data", "prov_id_lookup_table.rds"))
muni_roster       <- readRDS(here::here("data", "oep2026", "muni_roster.rds"))
concejo_data      <- readRDS(here::here("data", "oep2026", "muni_concejo_comb.rds"))
concejal_summary  <- readRDS(here::here("data", "oep2026", "concejal_summary.rds"))
muni_concejal_ioc <- readRDS(here::here("data", "oep2026", "muni_concejal_ioc.rds"))
alcaldes          <- readRDS(here::here("data", "oep2026", "alcaldes.rds"))

roster_pages <- muni_roster |> select(id_muni, page)

# Expected seat counts by census population — mirrors cleaning script section 3b
pop_muni         <- readRDS(here::here("data", "cpv2024", "population_municipality.rds")) |>
  select(id_muni = ine_code, pop_2024 = pop_2024_total)
dept_capital_ids <- paste0(sprintf("%02d", 1:9), "0101")

# Municipality areas from admin3 boundary file
area_muni <- readxl::read_excel(here::here("maps", "bol_admin_boundaries.xlsx"),
                                  sheet = "bol_admin3") |>
  mutate(id_muni = stringr::str_sub(adm3_pcode, 3, 8)) |>
  select(id_muni, area_sqkm)

# Add id_prov if not already present, then build province-context lookup
muni_lookup <- muni_lookup |>
  mutate(id_prov = stringr::str_sub(id_muni, 1, 4))

prov_counts <- muni_lookup |> dplyr::count(id_prov, name = "n_prov_munis")

muni_province_context <- muni_lookup |>
  select(id_muni, id_prov, muni_anexo) |>
  left_join(prov_counts, by = "id_prov") |>
  left_join(pop_muni,    by = "id_muni") |>
  left_join(area_muni,   by = "id_muni") |>
  left_join(prov_lookup |> select(id_prov, department), by = "id_prov") |>
  group_by(id_prov) |>
  mutate(
    area_rank = rank(-area_sqkm, ties.method = "first"),  # 1 = largest
    pop_rank  = rank(-pop_2024,  ties.method = "first")   # 1 = most populous
  ) |>
  ungroup() |>
  mutate(nat_pop_rank = rank(-pop_2024, ties.method = "first")) |>
  group_by(department) |>
  mutate(dept_pop_rank = rank(-pop_2024, ties.method = "first")) |>
  ungroup()

concejal_expected <- concejal_summary |>
  left_join(pop_muni, by = "id_muni") |>
  mutate(
    es_capital = id_muni %in% dept_capital_ids,
    sillas_esperadas = case_when(
      es_capital | pop_2024 > 75000 ~ 11L,
      pop_2024   > 50000            ~  9L,
      pop_2024   > 15000            ~  7L,
      !is.na(pop_2024)              ~  5L,
      TRUE                          ~ NA_integer_
    )
  )
```


```{r}
#| label: helpers
#| include: false

# Process wikitext for in-page display:
#   - <ref>...</ref>  →  superscript [refN] (numbered from ref_offset + 1)
#   - [[Target|Disp]] →  HTML anchor
#   - everything else HTML-escaped so wikitext syntax shows literally
# Returns list(html = character, refs = character vector of ref contents).
process_for_display <- function(text, ref_offset = 0L) {
  refs <- character(0)

  # Step 1: swap <ref>...</ref> for safe placeholder tokens
  processed <- stringr::str_replace_all(text, "<ref>(.*?)</ref>", function(m) {
    n    <- length(refs) + 1L
    refs <<- c(refs, stringr::str_match(m, "<ref>(.*?)</ref>")[1, 2])
    sprintf("REFSUP_%d_END", n + ref_offset)
  })

  # Step 2: HTML-escape (turns < > & into entities, leaves placeholders intact)
  escaped <- htmltools::htmlEscape(processed)

  # Step 3: convert [[wikilinks]] to anchors
  escaped <- stringr::str_replace_all(
    escaped, "\\[\\[([^|\\]]+)\\|([^\\]]+)\\]\\]",
    function(m) {
      p <- stringr::str_match(m, "\\[\\[([^|\\]]+)\\|([^\\]]+)\\]\\]")
      paste0('<a href="', wikipedia_url_for(p[, 2]), '" target="_blank">', p[, 3], "</a>")
    }
  )
  escaped <- stringr::str_replace_all(
    escaped, "\\[\\[([^\\]]+)\\]\\]",
    function(m) {
      tgt <- stringr::str_match(m, "\\[\\[([^\\]]+)\\]\\]")[, 2]
      paste0('<a href="', wikipedia_url_for(tgt), '" target="_blank">', tgt, "</a>")
    }
  )

  # Step 4: restore placeholders as blue superscripts
  for (i in seq_along(refs)) {
    n_abs <- i + ref_offset
    escaped <- stringr::str_replace_all(
      escaped,
      sprintf("REFSUP_%d_END", n_abs),
      sprintf('<sup style="color:#06c;font-size:9px;">[ref%d]</sup>', n_abs)
    )
  }

  list(html = escaped, refs = refs)
}

# Render a footnotes section below a display block.
refs_footnotes <- function(refs, ref_offset = 0L) {
  if (length(refs) == 0) return(NULL)
  tags$div(
    style = "margin-top:5px; padding-top:4px; border-top:1px solid #e1e4e8;",
    lapply(seq_along(refs), function(i) {
      tags$p(
        tags$b(paste0("ref", i + ref_offset, ":")), " ", refs[[i]],
        style = "font-size:10px; font-family:monospace; margin:1px 0; color:#555; word-break:break-all;"
      )
    })
  )
}

# Wikitext copy block: button + display (with ref superscripts) + hidden textarea.
wikitext_block <- function(id, label, content) {
  disp <- process_for_display(content)
  tagList(
    tags$button(
      class = "copy-btn",
      style = "margin-bottom:6px;",
      onclick = sprintf("copyTableToClipboard('%s', this)", id),
      label
    ),
    tags$pre(
      HTML(disp$html),
      style = paste(
        "background-color:#f6f8fa; padding:10px; border-radius:6px;",
        "overflow-x:auto; font-size:11px; white-space:pre; margin:0;"
      )
    ),
    refs_footnotes(disp$refs),
    tags$textarea(
      id = id,
      style = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
      readonly = TRUE,
      content
    )
  )
}

# Citation for INE 2024 census population data, appended to all population sentences
.ine_pop_ref <- paste0(
  "<ref>{{Cite",
  "| last = Instituto Nacional de Estadística",
  "| title = Población censada e indicadores demográficos: Tabulados por Municipio/TIOC",
  "| url = https://nube.ine.gob.bo/index.php/s/0OrfPjzMB16fATQ/download",
  "| date = 2025}}</ref>"
)

# Compose a population-rank sentence for municipalities that are among the most
# populous nationally (top 2), in their department (top 2), or in their province
# (rank 1, only for provinces with 3+ municipalities). Returns NULL when none apply.
compose_pop_rank_sentence_en <- function(id_muni, muni_pc) {
  row <- muni_pc[muni_pc$id_muni == id_muni, ]

  # National top-2
  if (!is.na(row$nat_pop_rank) && row$nat_pop_rank <= 2) {
    ord <- if (row$nat_pop_rank == 1) "most" else "second most"
    return(paste0("It is the ", ord, " populous municipality in [[Bolivia]].",
                  .ine_pop_ref))
  }

  # Department top-2
  if (!is.na(row$dept_pop_rank) && row$dept_pop_rank <= 2) {
    ord      <- if (row$dept_pop_rank == 1) "most" else "second most"
    dept_lnk <- paste0("[[", row$department, " Department]]")
    return(paste0("It is the ", ord, " populous municipality in ", dept_lnk, ".",
                  .ine_pop_ref))
  }

  # Province most populous (only for provinces with 3+ municipalities)
  if (!is.na(row$pop_rank) && row$pop_rank == 1 && row$n_prov_munis >= 3) {
    id_prov   <- stringr::str_sub(id_muni, 1, 4)
    prov_name <- prov_lookup$province_anexo[prov_lookup$id_prov == id_prov]
    return(paste0("It is the most populous municipality in [[",
                  prov_name, " Province]].", .ine_pop_ref))
  }

  NULL
}

# Compose the province-context sentence appended to the lead, for municipalities
# in 1- or 2-municipality provinces. Returns NULL for 3+ municipality provinces.
compose_province_context_en <- function(id_muni, muni_pc, prov_lkp) {
  row       <- muni_pc[muni_pc$id_muni == id_muni, ]
  id_prov   <- stringr::str_sub(id_muni, 1, 4)
  prov_name <- prov_lkp$province_anexo[prov_lkp$id_prov == id_prov]
  muni_name <- row$muni_anexo
  prov_link <- paste0("[[", prov_name, " Province]]")
  n         <- row$n_prov_munis

  if (n == 1) {
    return(paste0(
      muni_name, " Municipality is the only municipality in ",
      prov_link, " and its boundaries are identical."
    ))
  }

  if (n == 2 && !is.na(row$area_rank) && !is.na(row$pop_rank)) {
    larger   <- row$area_rank == 1
    more_pop <- row$pop_rank  == 1
    return(dplyr::case_when(
       larger &  more_pop ~ paste0(muni_name, " Municipality is the larger and more populous of the two municipalities in ", prov_link, "."),
       larger & !more_pop ~ paste0(muni_name, " Municipality covers the larger area of the two municipalities in ", prov_link, ", but it is less populous."),
      !larger &  more_pop ~ paste0(muni_name, " Municipality is the more populous of the two municipalities in ", prov_link, ", but it covers a smaller area."),
      TRUE                ~ paste0(muni_name, " Municipality is the smaller and less populous of the two municipalities in ", prov_link, ".")
    ))
  }

  NULL
}

# Compose the Wikipedia lead sentence for a municipality (wikitext).
# Uses muni_anexo and province_anexo as display text; ordinal from last 2 digits
# of id_muni via english::ordinal().
compose_lead_sentence_en <- function(id_muni, muni_lkp, prov_lkp) {
  muni_row <- muni_lkp[muni_lkp$id_muni == id_muni, ]
  id_prov  <- substr(id_muni, 1, 4)
  prov_row <- prov_lkp[prov_lkp$id_prov == id_prov, ]
  muni_num <- as.integer(substr(id_muni, 5, 6))

  lead <- paste0(
    "'''", muni_row$muni_anexo, " Municipality''' is the ",
    english::ordinal(muni_num),
    " [[Municipalities of Bolivia|municipal section]] of the",
    " [[", prov_row$province_anexo, " Province]]",
    " in the [[", prov_row$department, " Department]]",
    " in [[Bolivia]]."
  )

  pop_rank <- compose_pop_rank_sentence_en(id_muni, muni_province_context)
  context  <- compose_province_context_en(id_muni, muni_province_context, prov_lkp)

  parts <- c(lead, pop_rank, context)
  paste(parts[!sapply(parts, is.null)], collapse = " ")
}

# Compose the mayor sentence for a municipality (wikitext).
compose_mayor_sentence_en <- function(id_muni, muni_name, alcaldes_data,
                                       page_number) {
  row    <- alcaldes_data[alcaldes_data$id_muni == id_muni, ]
  nombre <- stringr::str_to_title(row$nombre[1])
  ref    <- paste0("<ref>",
                   source_ref_with_page_number(page_number, source_ref_en),
                   "</ref>")
  paste0(
    "The Mayor of ", muni_name, " is ", nombre,
    ", who was elected in the ",
    "[[2026 Bolivian regional elections|March 2026 elections]].",
    ref
  )
}

# Compose the council intro sentence for a municipality.
# Returns one sentence for non-IOC municipalities, two for IOC ones.
compose_council_sentence_en <- function(id_muni, muni_name,
                                         concejal_exp, ioc_data) {
  row <- concejal_exp[concejal_exp$id_muni == id_muni, ]

  range_phrase <- dplyr::case_when(
    row$es_capital & row$pop_2024 > 75000 ~
      "department capitals and municipalities with populations over 75,000",
    row$es_capital ~
      "department capitals",
    row$pop_2024 > 75000 ~
      "municipalities with populations over 75,000",
    row$pop_2024 > 50000 ~
      "municipalities with populations between 50,001 and 75,000",
    row$pop_2024 > 15000 ~
      "municipalities with populations between 15,001 and 50,000",
    TRUE ~
      "municipalities with populations up to 15,000"
  )

  s1 <- paste0(
    muni_name, " has a municipal council with ", row$sillas_esperadas,
    " seats, as do all ", range_phrase, ".",
    "<ref>{{Cite news| title = A más población, los municipios tienen más concejales, según la ley| work = Los Tiempos| access-date = 2026-06-19| date = 2026-02-08| url = https://www.lostiempos.com/actualidad/pais/20260208/mas-poblacion-municipios-tienen-mas-concejales-ley}}</ref>"
  )

  ioc_rows <- ioc_data[ioc_data$id_muni == id_muni, ]
  if (nrow(ioc_rows) > 0) {
    pueblo <- stringr::str_to_title(ioc_rows$pueblo[1])
    s2 <- paste0(
      "One of those seats is an Indigenous seat assigned by the ",
      pueblo, " according to its own norms and procedures."
    )
    paste(s1, s2)
  } else {
    s1
  }
}

muni_block <- function(id_muni, muni_name, department, concejo) {
  page <- roster_pages$page[roster_pages$id_muni == id_muni]
  src  <- source_ref_with_page_number(page, source_ref_en)

  # Lead sentence
  lead_wt   <- compose_lead_sentence_en(id_muni, muni_lookup, prov_lookup)
  lead_disp <- process_for_display(lead_wt)

  # Wikitext for both government paragraphs (combined into one copy target)
  mayor_wt   <- compose_mayor_sentence_en(id_muni, muni_name, alcaldes, page)
  council_wt <- compose_council_sentence_en(id_muni, muni_name,
                                             concejal_expected, muni_concejal_ioc)
  intro_wt   <- paste0(mayor_wt, "\n\n", council_wt)

  # Display: process each sentence separately so refs are numbered sequentially
  mayor_disp   <- process_for_display(mayor_wt, ref_offset = 0L)
  council_disp <- process_for_display(council_wt, ref_offset = length(mayor_disp$refs))
  all_intro_refs <- c(mayor_disp$refs, council_disp$refs)

  tagList(
    tags$details(
      tags$summary(
        tags$strong(muni_name),
        tags$span(
          paste0(" \u2014 ", department),
          style = "color: #586069; font-weight: normal;"
        ),
        tags$code(
          paste0(" [", id_muni, "]"),
          style = "font-size: 11px; color: #999; margin-left: 6px;"
        )
      ),
      # Lead sentence — separate copy block
      tags$div(
        style = "margin-top:8px; margin-bottom:10px;",
        tags$button(
          class = "copy-btn",
          style = "margin-bottom:6px;",
          onclick = sprintf("copyTableToClipboard('lead_en_%s', this)", id_muni),
          "Copy lead wikitext"
        ),
        tags$p(HTML(lead_disp$html), style = "font-size:12px; margin:4px 0 0; color:#333;"),
        tags$textarea(
          id       = paste0("lead_en_", id_muni),
          style    = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
          readonly = TRUE,
          lead_wt
        )
      ),
      # Government section — mayor + council intro, one button copies both
      tags$div(
        style = "margin-top:8px; margin-bottom:10px;",
        tags$button(
          class = "copy-btn",
          style = "margin-bottom:6px;",
          onclick = sprintf("copyTableToClipboard('intro_en_%s', this)", id_muni),
          "Copy government wikitext"
        ),
        tags$p(HTML(mayor_disp$html),   style = "font-size:12px; margin:4px 0 4px; color:#333;"),
        tags$p(HTML(council_disp$html),  style = "font-size:12px; margin:0 0 0;    color:#333;"),
        refs_footnotes(all_intro_refs),
        tags$textarea(
          id       = paste0("intro_en_", id_muni),
          style    = "position:absolute; left:-9999px; height:0; width:0; overflow:hidden;",
          readonly = TRUE,
          intro_wt
        )
      ),
      # Side-by-side kable + wikitable
      tags$div(
        style = "display:flex; gap:14px; align-items:flex-start; font-size:12px;",
        tags$div(
          style = "flex:1 1 0; min-width:0; overflow-x:auto;",
          build_council_kable_en(id_muni, data = concejo)
        ),
        tags$div(
          style = "flex:1 1 0; min-width:0;",
          wikitext_block(
            id      = paste0("concejo_en_", id_muni),
            label   = "Copy council table",
            content = build_council_wikitable_en(id_muni, data = concejo,
                                                  source = source_wikitext_en,
                                                  source_refs = src)
          )
        )
      )
    ),
    tags$hr(style = "border:none; border-top:1px solid #e1e4e8; margin:4px 0;")
  )
}

render_muni_blocks <- function(lookup, concejo) {
  ids <- lookup |>
    filter(id_muni %in% unique(concejo$id_muni)) |>
    arrange(id_muni) |>
    select(id_muni, muni_anexo, department)

  tagList(pmap(ids, function(id_muni, muni_anexo, department) {
    muni_block(id_muni, muni_anexo, department, concejo)
  }))
}
```

```{r source-data}
source_wikitext_en = "Supreme Electoral Tribunal."
source_wikitext_es = "Tribunal Supremo Electoral."

source_ref_en  <- "{{Cite report| publisher = Órgano Electoral Plurinacional| last1 = Órgano Electoral Plurinacional| last2 = Tribunal Supremo Electoral| title = Publicación de resultados Elecciones Subnacionales 2026: Elección de Autoridades Políticas Departamentales Regionales y Municipales 2026 (Primera y Segunda Vuelta Electoral)| series = Separata de información pública| date = 2026-04-23}}"
source_ref_es <- source_ref_en

source_ref_with_page_number <- function(page_number, source_ref) {
  source_ref_minus_end <- sub("\\}\\}$", "", source_ref)
  paste0(source_ref_minus_end, " | page = ", page_number, " }}")
}

```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

```{r}
#| label: council-tables
#| echo: false
#| results: asis
render_muni_blocks(muni_lookup, concejo_data)
```
