---
title: "Bolivia 2026 — Municipality Infoboxes"
subtitle: "{{Infobox settlement}} wikitext ready to copy to Wikipedia"
params:
  wiki_filter: "all"   # "existing", "new", or "all"
format:
  html:
    code-fold: true
    toc: false
    self-contained: true
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(purrr)
library(htmltools)
library(wikitools)

# Language-aware infobox engine (sources muni-article-functions.R ->
# bolivia-builders.R / muni-i18n.R / muni-phrases.R / muni-references.R).
source("src/muni-infobox-functions.R")
source("src/gh-copy-assets.R")

lang <- "en"

muni_lookup  <- readRDS(here::here("data", "muni_id_lookup_table.rds"))
prov_lookup  <- readRDS(here::here("data", "prov_id_lookup_table.rds"))
alcaldes     <- readRDS(here::here("data", "oep2026", "alcaldes.rds"))
muni_roster  <- readRDS(here::here("data", "oep2026", "muni_roster.rds")) |>
  select(id_muni, page)

bo_municipality_wiki_presence <- readRDS(here::here("data", "bo_municipality_wiki_presence.rds"))

# Population (total and urban/rural split)
pop_muni <- readRDS(here::here("data", "cpv2024", "population_municipality.rds")) |>
  select(id_muni = ine_code, municipality, department, province,
         pop_2024 = pop_2024_total, pop_2012 = pop_2012_total,
         pop_2001 = pop_2001_total)

pop_muni_ur <- readRDS(here::here("data", "cpv2024", "population_municipality_urb_rural.rds")) |>
  select(id_muni = ine_code, area, pop_2024 = pop_2024_total)

# Ethnic identification
auto_data <- readRDS(here::here("data", "cpv2024", "autoidentificacion_municipio.rds"))

# Municipality areas from admin boundaries
area_muni <- readxl::read_excel(here::here("maps", "bol_admin_boundaries.xlsx"),
                                sheet = "bol_admin3") |>
  mutate(id_muni = stringr::str_sub(adm3_pcode, 3, 8)) |>
  select(id_muni, area_sqkm)

# Locator map filenames from Commons upload log
commons_log <- readr::read_csv(
  here::here("output", "locator_maps", "muni-maps-final", "commons_upload_log.csv"),
  show_col_types = FALSE
) |>
  select(id_muni, commons_filename) |>
  filter(!is.na(commons_filename))

# Wikidata QID <-> INE crosswalk (reused for multiple lookups)
wd_ine <- readRDS(here::here("data", "municipalities_wd_ine.rds")) |>
  select(qid, ine_code)

# Wikidata image properties: flag (P41), coat of arms (P94), seal (P158).
wd_images <- readRDS(here::here("data", "muni_wd_images.rds"))

# P402 (OSM relation ID) presence — from saved Wikidata snapshot
wd_p402 <- readRDS(here::here("data", "municipalities_wd_p402.rds"))
munis_with_p402 <- wd_p402 |>
  filter(!is.na(osm_relation_id)) |>
  left_join(wd_ine, by = "qid") |>
  filter(!is.na(ine_code)) |>
  pull(ine_code)

# Wikipedia presence filter
wiki_presence <- bo_municipality_wiki_presence$data |>
  left_join(wd_ine, by = "qid") |>
  filter(!is.na(ine_code))

munis_with_enwiki <- wiki_presence |> filter(en > 0) |> pull(ine_code)

# Add id_prov to muni_lookup (used by the engine)
muni_lookup <- muni_lookup |>
  mutate(id_prov = stringr::str_sub(id_muni, 1, 4))

# Split into two subsets: municipalities without an existing EN article
# (rendered fresh) and those with one (rendered as a merge).
muni_lookup_new      <- muni_lookup |> filter(!id_muni %in% munis_with_enwiki)
muni_lookup_existing <- muni_lookup |> filter(id_muni %in% munis_with_enwiki)

# Raw wikitext infobox blocks for existing EN articles, keyed by id_muni.
# parse_infobox_blocks() preserves verbatim field markup (comments intact) —
# exactly what the merge needs. NULL/empty for stub articles with no infobox,
# which render_existing_blocks() routes through the fresh-scaffold path instead.
sa_article_quality <- readRDS(here::here("data", "cache", "sa_article_quality.rds"))

existing_wikitext <- muni_lookup_existing |>
  left_join(wd_ine, by = c("id_muni" = "ine_code")) |>
  select(id_muni, qid) |>
  left_join(
    sa_article_quality |> filter(lang == "en") |> select(qid, wikitext),
    by = "qid"
  ) |>
  filter(!is.na(wikitext))

existing_infobox_blocks <- setNames(
  purrr::map(existing_wikitext$wikitext, function(wt) {
    parse_infobox_blocks(extract_infobox_wikitext(wt))
  }),
  existing_wikitext$id_muni
)
```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

::: {style="font-size: 11px; margin-bottom: 1em;"}

Each municipality has two copy buttons:

- **Copy infobox** — the `{{Infobox settlement}}` template with population, area, coordinates, mayor, and ethnic group data from the 2024 census. For municipalities with an existing English Wikipedia article, this is a *merge*: our computed fields take priority when non-blank, otherwise the existing article's field value is kept as-is (with original wikitext markup and citations intact). A short note above the copy button explains which fields were carried over unchanged.
- **Copy page Wikitext** *(when available)* — the full article text assembled in section order: lead sentence → `==Government==` (mayor paragraph, council size paragraph, council table) → `==Population==` (census population table, population pyramid) → `==Languages and Indigenous identification==` (ethnic identity table, languages table) → `==Communities==`. Generated by the [article content page](bolivia-muni-generator-en.html); render that page first to produce these files.

:::

```{r}
#| label: filter-summary
#| echo: false
#| results: asis
n_new      <- nrow(muni_lookup_new)
n_existing <- nrow(muni_lookup_existing)

if (params$wiki_filter == "existing") {
  cat(sprintf(
    "**Rendering merged infoboxes for %d municipalities with an existing English Wikipedia article.**\n\n",
    n_existing
  ))
} else if (params$wiki_filter == "new") {
  cat(sprintf(
    "**Rendering fresh infoboxes for %d municipalities without an English Wikipedia article.**\n\n",
    n_new
  ))
} else {
  cat(sprintf(
    "**Rendering fresh infoboxes for %d municipalities without an existing article, and merged infoboxes for %d municipalities with one.**\n\n",
    n_new, n_existing
  ))
}
```

```{r}
#| label: infobox-blocks
#| echo: false
#| results: asis
show_new      <- params$wiki_filter %in% c("new", "all")
show_existing <- params$wiki_filter %in% c("existing", "all")

no_infobox_note <- "No existing infobox was found in this article \u2014 rendering a fresh scaffold instead of a merge."

if (show_new && show_existing) {
  cat("## Municipalities without an existing English Wikipedia article\n\n")
}
if (show_new) {
  render_new_blocks(muni_lookup_new, lang)
}
if (show_new && show_existing) {
  cat("\n\n## Municipalities with an existing English Wikipedia article (merged infobox)\n\n")
}
if (show_existing) {
  render_existing_blocks(muni_lookup_existing, lang, no_infobox_note)
}
```
