---
title: "Bolivia: Poverty and Basic Services Trends (2001–2024)"
format: 
  html:
    page-layout: full
    code-tools: true
---

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
library(kableExtra)
library(htmltools)
source("src/gh-copy-assets.R")
```

```{r}
#| label: assets
#| echo: false
#| results: asis
include_table_assets()
```

This document presents a comparative overview of poverty and basic service access across Bolivia's departments. Data can be exported in HTML or Wikitable (MediaWiki) format.

**Note on temporal coverage**: Water and sanitation data are available for 2001 and 2024. Poverty (NBI) data is available for 2012 and 2024.

```{r load-data}
# Load the English datasets generated from censo-2024-poverty.qmd
poverty_nbi_dept <- readRDS("data/cpv2024/poverty_nbi_dept.rds")
improved_water_dept <- readRDS("data/cpv2024/improved_water_dept.rds")
improved_sanitation_dept <- readRDS("data/cpv2024/improved_sanitation_dept.rds")
```

```{r process-data}
# 1. Extract Poverty (NBI) percentages (2012 and 2024)
poverty_tab <- poverty_nbi_dept |>
  select(
    department,
    poor_2012 = poor_total_2012,
    poor_2024 = poor_total_2024,
    pop_2012 = ref_pop_nbi_2012,
    pop_2024 = ref_pop_nbi_2024
  ) |>
  mutate(
    poverty_pct_2012 = (poor_2012 / pop_2012) * 100,
    poverty_pct_2024 = (poor_2024 / pop_2024) * 100,
    poverty_change = poverty_pct_2024 - poverty_pct_2012
  ) |>
  select(department, poverty_pct_2012, poverty_pct_2024, poverty_change)



# 3. Extract Water Lack (Unimproved) percentages (2001 and 2024)
water_tab <- improved_water_dept |>
  select(
    department,
    water_lack_2001 = pct_unimproved_water_2001,
    water_lack_2024 = pct_unimproved_water_2024
  ) |>
  mutate(
    water_change = water_lack_2024 - water_lack_2001
  ) |>
  select(department, water_lack_2001, water_lack_2024, water_change)

# 4. Extract Sanitation Lack (Unimproved) percentages (2001 and 2024)
sanitation_tab <- improved_sanitation_dept |>
  select(
    department,
    sanitation_lack_2001 = pct_unimproved_sanitation_2001,
    sanitation_lack_2024 = pct_unimproved_sanitation_2024
  ) |>
  mutate(
    sanitation_change = sanitation_lack_2024 - sanitation_lack_2001
  ) |>
  select(department, sanitation_lack_2001, sanitation_lack_2024, sanitation_change)

# 5. Extract national totals from raw poverty data (aggregate all departments)
country_poverty <- poverty_nbi_dept |>
  summarise(
    poor_2012 = sum(poor_total_2012),
    pop_2012 = sum(ref_pop_nbi_2012),
    poor_2024 = sum(poor_total_2024),
    pop_2024 = sum(ref_pop_nbi_2024)
  ) |>
  mutate(
    poverty_pct_2012 = (poor_2012 / pop_2012) * 100,
    poverty_pct_2024 = (poor_2024 / pop_2024) * 100,
    poverty_change = poverty_pct_2024 - poverty_pct_2012,
    department = "Bolivia"
  ) |>
  select(department, poverty_pct_2012, poverty_pct_2024, poverty_change)

country_water <- improved_water_dept |>
  summarise(
    n_unimproved_2001 = sum(n_unimproved_water_2001),
    n_total_2001 = sum(n_total_2001),
    n_unimproved_2024 = sum(n_unimproved_water_2024),
    n_total_2024 = sum(n_total_2024)
  ) |>
  mutate(
    water_lack_2001 = (n_unimproved_2001 / n_total_2001) * 100,
    water_lack_2024 = (n_unimproved_2024 / n_total_2024) * 100,
    water_change = water_lack_2024 - water_lack_2001,
    department = "Bolivia"
  ) |>
  select(department, water_lack_2001, water_lack_2024, water_change)

country_sanitation <- improved_sanitation_dept |>
  summarise(
    n_unimproved_2001 = sum(n_unimproved_sanitation_2001),
    n_total_2001 = sum(n_total_2001),
    n_unimproved_2024 = sum(n_unimproved_sanitation_2024),
    n_total_2024 = sum(n_total_2024)
  ) |>
  mutate(
    sanitation_lack_2001 = (n_unimproved_2001 / n_total_2001) * 100,
    sanitation_lack_2024 = (n_unimproved_2024 / n_total_2024) * 100,
    sanitation_change = sanitation_lack_2024 - sanitation_lack_2001,
    department = "Bolivia"
  ) |>
  select(department, sanitation_lack_2001, sanitation_lack_2024, sanitation_change)

country_row <- country_poverty |>
  left_join(country_water, by = "department") |>
  left_join(country_sanitation, by = "department")

# 6. Combine all (excluding marginality) and add country row at bottom
final_table <- poverty_tab |>
  left_join(water_tab, by = "department") |>
  left_join(sanitation_tab, by = "department") |>
  arrange(department) |>
  bind_rows(country_row) |>
  # Keep departments sorted, Bolivia always at bottom
  mutate(sort_key = if_else(department == "Bolivia", 999, 0)) |>
  arrange(sort_key, department) |>
  select(-sort_key)

# Helper for consistent formatting
format_pct <- function(x) {
  ifelse(is.na(x), "-", sprintf("%.0f%%", x))
}

format_chg <- function(x) {
  if(is.na(x)) return("-")
  prefix <- if(x > 0) "+" else ""
  paste0(prefix, sprintf("%.0f%%", x))
}

# Function to create Wikitext shade syntax
# color: "red" (poverty), "blue" (water), "green" (sanitation)
wikitext_shade <- function(value_pct, color) {
  if(is.na(value_pct)) return("-")
  # Extract numeric value and format
  numeric_val <- floor(as.numeric(gsub("%", "", value_pct)))
  sprintf("{{shade|color=%s|%s}}%%", color, numeric_val)
}
```

```{r render-table}
#| echo: false

# Prepare formatted columns for HTML display
table_display <- final_table |>
  mutate(
    p_12 = map_chr(poverty_pct_2012, format_pct),
    p_24 = map_chr(poverty_pct_2024, format_pct),
    p_ch = map_chr(poverty_change, format_chg),
    w_01 = map_chr(water_lack_2001, format_pct),
    w_24 = map_chr(water_lack_2024, format_pct),
    w_ch = map_chr(water_change, format_chg),
    s_01 = map_chr(sanitation_lack_2001, format_pct),
    s_24 = map_chr(sanitation_lack_2024, format_pct),
    s_ch = map_chr(sanitation_change, format_chg)
  ) |>
  mutate(
    # Bold Bolivia row using HTML (since kable doesn't process markdown)
    department = if_else(department == "Bolivia", paste0("<b>", department, "</b>"), department)
  ) |>
  select(
    department,
    p_12, p_24, p_ch,
    w_01, w_24, w_ch,
    s_01, s_24, s_ch
  ) |>
  rename(`Department` = department)

# Render HTML table with nested headers and vertical dividers
table_display |>
  kbl(
    caption = "Comparative Trends in Poverty and Basic Service Access by Department",
    col.names = c(
      "Department", 
      "2012", "2024", "Change",
      "2001", "2024", "Change",
      "2001", "2024", "Change"
    ),
    align = "lcccccccccc",
    escape = FALSE
  ) |>
  kable_classic(full_width = FALSE) |>
  add_header_above(c(
    " " = 1, 
    "Poverty (NBI)" = 3, 
    "Lack of Water Access" = 3, 
    "Lack of Sanitation Access" = 3
  )) |>
  column_spec(1, border_right = TRUE) |>
  column_spec(4, border_right = TRUE) |>
  column_spec(7, border_right = TRUE)

```

## Municipality Examples

Below are poverty and basic services statistics for six representative municipalities, shown alongside their department and national context.

```{r load-muni-data}
# Load municipality-level data (all three metrics)
poverty_nbi_muni <- readRDS("data/cpv2024/poverty_nbi_muni.rds")
improved_water_muni <- readRDS("data/cpv2024/improved_water_muni.rds")
improved_sanitation_muni <- readRDS("data/cpv2024/improved_sanitation_muni.rds")

# Extract data for 6 representative municipalities
muni_list <- list(
  list(muni = "Sucre",       dept = "Chuquisaca",  ine_code = "010101"),
  list(muni = "La Paz",      dept = "La Paz",      ine_code = "020101"),
  list(muni = "Cochabamba",  dept = "Cochabamba",  ine_code = "030101"),
  list(muni = "Oruro",       dept = "Oruro",       ine_code = "040101"),
  list(muni = "Santa Cruz",  dept = "Santa Cruz",  ine_code = "070101"),
  list(muni = "Trinidad",    dept = "Beni",        ine_code = "080101")
)

# Build municipality data tables with all metrics (2012, 2024, change)
muni_poverty_data <- poverty_nbi_muni |>
  filter(ine_code %in% map_chr(muni_list, ~.$ine_code)) |>
  select(
    ine_code,
    municipality,
    department,
    poor_2012 = poor_total_2012,
    poor_2024 = poor_total_2024,
    pop_2012 = ref_pop_nbi_2012,
    pop_2024 = ref_pop_nbi_2024
  ) |>
  mutate(
    poverty_pct_2012 = (poor_2012 / pop_2012) * 100,
    poverty_pct_2024 = (poor_2024 / pop_2024) * 100,
    poverty_change = poverty_pct_2024 - poverty_pct_2012
  ) |>
  select(ine_code, municipality, department, poverty_pct_2012, poverty_pct_2024, poverty_change)

muni_water_data <- improved_water_muni |>
  filter(ine_code %in% map_chr(muni_list, ~.$ine_code)) |>
  select(
    ine_code,
    municipality,
    water_lack_2001 = pct_unimproved_water_2001,
    water_lack_2024 = pct_unimproved_water_2024
  ) |>
  mutate(
    water_change = water_lack_2024 - water_lack_2001
  ) |>
  select(ine_code, municipality, water_lack_2001, water_lack_2024, water_change)

muni_sanitation_data <- improved_sanitation_muni |>
  filter(ine_code %in% map_chr(muni_list, ~.$ine_code)) |>
  select(
    ine_code,
    municipality,
    sanitation_lack_2001 = pct_unimproved_sanitation_2001,
    sanitation_lack_2024 = pct_unimproved_sanitation_2024
  ) |>
  mutate(
    sanitation_change = sanitation_lack_2024 - sanitation_lack_2001
  ) |>
  select(ine_code, municipality, sanitation_lack_2001, sanitation_lack_2024, sanitation_change)
```

```{r render-muni-examples}
#| echo: false
#| results: asis

# Function to create context table for a municipality (with municipality, department, and Bolivia)
create_muni_context_table <- function(muni_name, dept_name, ine_code) {
  # Get municipality data by INE code to avoid name-matching failures
  muni_pov <- muni_poverty_data |> filter(ine_code == !!ine_code) |> slice(1)
  muni_wat <- muni_water_data |> filter(ine_code == !!ine_code) |> slice(1)
  muni_san <- muni_sanitation_data |> filter(ine_code == !!ine_code) |> slice(1)
  
  # Get department data
  dept_pov <- final_table |> filter(department == dept_name) |> slice(1)
  
  # Get Bolivia data
  bolivia_pov <- final_table |> filter(department == "Bolivia")
  
  # Combine into context table: municipality | department | Bolivia
  context_table <- tibble(
    location = c(muni_name, dept_name, "Bolivia"),
    poverty_pct_2012 = c(muni_pov$poverty_pct_2012[1], dept_pov$poverty_pct_2012[1], bolivia_pov$poverty_pct_2012[1]),
    poverty_pct_2024 = c(muni_pov$poverty_pct_2024[1], dept_pov$poverty_pct_2024[1], bolivia_pov$poverty_pct_2024[1]),
    poverty_change = c(muni_pov$poverty_change[1], dept_pov$poverty_change[1], bolivia_pov$poverty_change[1]),
    water_lack_2001 = c(muni_wat$water_lack_2001[1], dept_pov$water_lack_2001[1], bolivia_pov$water_lack_2001[1]),
    water_lack_2024 = c(muni_wat$water_lack_2024[1], dept_pov$water_lack_2024[1], bolivia_pov$water_lack_2024[1]),
    water_change = c(muni_wat$water_change[1], dept_pov$water_change[1], bolivia_pov$water_change[1]),
    sanitation_lack_2001 = c(muni_san$sanitation_lack_2001[1], dept_pov$sanitation_lack_2001[1], bolivia_pov$sanitation_lack_2001[1]),
    sanitation_lack_2024 = c(muni_san$sanitation_lack_2024[1], dept_pov$sanitation_lack_2024[1], bolivia_pov$sanitation_lack_2024[1]),
    sanitation_change = c(muni_san$sanitation_change[1], dept_pov$sanitation_change[1], bolivia_pov$sanitation_change[1])
  )
  
  return(context_table)
}

# Display each municipality context table
for (i in seq_along(muni_list)) {
  muni_ctx <- create_muni_context_table(muni_list[[i]]$muni, muni_list[[i]]$dept, muni_list[[i]]$ine_code)
  
  # Format for display (same as departmental table)
  display_table <- muni_ctx |>
    mutate(
      p_12 = sprintf("%.0f%%", poverty_pct_2012),
      p_24 = sprintf("%.0f%%", poverty_pct_2024),
      p_ch = if_else(poverty_change > 0, sprintf("+%.0f%%", poverty_change), sprintf("%.0f%%", poverty_change)),
      w_01 = sprintf("%.0f%%", water_lack_2001),
      w_24 = sprintf("%.0f%%", water_lack_2024),
      w_ch = if_else(water_change > 0, sprintf("+%.0f%%", water_change), sprintf("%.0f%%", water_change)),
      s_01 = sprintf("%.0f%%", sanitation_lack_2001),
      s_24 = sprintf("%.0f%%", sanitation_lack_2024),
      s_ch = if_else(sanitation_change > 0, sprintf("+%.0f%%", sanitation_change), sprintf("%.0f%%", sanitation_change)),
      # Add " Department" if department name matches municipality name
      location = case_when(
        location == muni_list[[i]]$dept & location == muni_list[[i]]$muni ~ paste0(location, " Department"),
        TRUE ~ location
      )
    ) |>
    select(location, p_12, p_24, p_ch, w_01, w_24, w_ch, s_01, s_24, s_ch) |>
    rename(`Location` = location)
  
  cat("\n### ", muni_list[[i]]$muni, " (", muni_list[[i]]$dept, " Department)\n\n")
  
  print(
    display_table |>
      kbl(
        caption = paste("Poverty and Basic Services:", muni_list[[i]]$muni, "Municipality"),
        col.names = c(
          "Location",
          "2012", "2024", "Change",
          "2001", "2024", "Change",
          "2001", "2024", "Change"
        ),
        align = "lcccccccccc",
        escape = FALSE
      ) |>
      kable_classic(full_width = FALSE) |>
      add_header_above(c(
        " " = 1,
        "Poverty (NBI)" = 3,
        "Lack of Water Access" = 3,
        "Lack of Sanitation Access" = 3
      )) |>
      column_spec(1, border_right = TRUE) |>
      column_spec(4, border_right = TRUE) |>
      column_spec(7, border_right = TRUE)
  )
  cat("\n")
}

```

## Municipality Wikitable Exports

Six representative municipalities with their departmental and national context.

```{r define-wt-helpers}
#| echo: false

# Helper functions for Wikitable formatting (defined here for use in municipality tables)
format_change_wt <- function(x) {
  if(is.na(x)) return("-")
  if(x > 0) {
    sprintf("{{Increase}} +%.0f%%", x)
  } else if(x < 0) {
    sprintf("{{Decrease}} %.0f%%", x)
  } else {
    "0.0%"
  }
}

wikitext_shade <- function(value_pct, color) {
  if(is.na(value_pct)) return("-")
  sprintf("{{shade|color=%s|%s}}", color, value_pct)
}
```

```{r render-muni-wikitables}
#| echo: false
#| results: asis

# For each municipality, create a Wikitable export
for (i in seq_along(muni_list)) {
  muni_ctx <- create_muni_context_table(muni_list[[i]]$muni, muni_list[[i]]$dept, muni_list[[i]]$ine_code)
  
  # Build Wikitable rows with same formatting as departmental table
  wikitable_rows_muni <- muni_ctx |>
    mutate(
      p_12_wt = map_chr(poverty_pct_2012, ~wikitext_shade(sprintf("%.0f%%", .), "red")),
      p_24_wt = map_chr(poverty_pct_2024, ~wikitext_shade(sprintf("%.0f%%", .), "red")),
      p_ch_wt = map_chr(poverty_change, format_change_wt),
      w_01_wt = map_chr(water_lack_2001, ~wikitext_shade(sprintf("%.0f%%", .), "blue")),
      w_24_wt = map_chr(water_lack_2024, ~wikitext_shade(sprintf("%.0f%%", .), "blue")),
      w_ch_wt = map_chr(water_change, format_change_wt),
      s_01_wt = map_chr(sanitation_lack_2001, ~wikitext_shade(sprintf("%.0f%%", .), "green")),
      s_24_wt = map_chr(sanitation_lack_2024, ~wikitext_shade(sprintf("%.0f%%", .), "green")),
      s_ch_wt = map_chr(sanitation_change, format_change_wt)
    ) |>
    select(
      location,
      p_12_wt, p_24_wt, p_ch_wt,
      w_01_wt, w_24_wt, w_ch_wt,
      s_01_wt, s_24_wt, s_ch_wt
    )
  
  # Build Wikitable header
  muni_wt_header <- paste0(
    "{| class=\"wikitable\"\n",
    "|+ ", muni_list[[i]]$muni, " Municipality: Poverty and Basic Service Access\n",
    "|-\n",
    "! align=left | Location !! colspan=3 | Poverty (NBI) !! colspan=3 | Lack of Water Access !! colspan=3 | Lack of Sanitation Access\n",
    "|-\n",
    "! align=left | !! 2012 !! 2024 !! Change !! 2001 !! 2024 !! Change !! 2001 !! 2024 !! Change\n"
  )
  
  # Build rows
  muni_wt_rows <- wikitable_rows_muni |>
    mutate(
      row_num = row_number(),
      # Bold municipality name; add " Department" to department row if names match
      dept_display = case_when(
        row_num == 1 ~ paste0("'''", location, "'''"),  # First row: municipality in bold
        row_num == 2 & location == muni_list[[i]]$muni ~ paste0(location, " Department"),  # Second row: dept with " Department" if it matches muni name
        TRUE ~ location
      )
    ) |>
    glue::glue_data("|-\n| align=left | {dept_display} || {p_12_wt} || {p_24_wt} || {p_ch_wt} || {w_01_wt} || {w_24_wt} || {w_ch_wt} || {s_01_wt} || {s_24_wt} || {s_ch_wt}")
  
  # Source row
  source_text <- "NBI indicates a Unsatisfied Basic Needs definition for poverty. '''Source:''' [[National Institute of Statistics of Bolivia|INE Bolivia]], 2024 Bolivian Census"
  source_refs <- '{{Cite| last = Instituto Nacional de Estadística| title = Pobreza: Tabulados por Municipio/TIOC| access-date = 2026-07-22| date = 2025| url = https://nube.ine.gob.bo/index.php/s/KQSoTrfdRQ5jldM/download}}<br/>{{Cite| last = Instituto Nacional de Estadística| title = Servicios Básicos: Tabulados por Municipio/TIOC| access-date = 2026-07-22| date = 2025| url = https://nube.ine.gob.bo/index.php/s/dEJeTU4j0czYIdG/download}}'
  source_line <- paste0('|-\n|colspan="10"|', source_text, '<ref>', source_refs, '</ref>')
  
  muni_wt_footer <- paste0("\n", source_line, "\n|}")
  
  wikitable_muni_text <- paste0(muni_wt_header, paste(muni_wt_rows, collapse = "\n"), muni_wt_footer)
  
  # Display with copy button
  cat("\n### ", muni_list[[i]]$muni, "\n\n")
  
  print(
    tagList(
      copy_to_clipboard_button_wt(wikitable_muni_text, "Copy Wikitable to Clipboard"),
      tags$pre(
        style = "background-color: #f6f8fa; padding: 12px; border-radius: 6px; overflow-x: auto; max-height: 300px;",
        wikitable_muni_text
      )
    )
  )
  cat("\n")
}

```

## Wikitable Export

Export this table to Wikipedia in Wikitext format with color-coded percentages.

```{r render-wikitable}
#| echo: false

# Build Wikitable rows with shade-colored percentages and change templates
wikitable_rows <- final_table |>
  mutate(
    p_12_wt = map_chr(poverty_pct_2012, ~wikitext_shade(format_pct(.), "red")),
    p_24_wt = map_chr(poverty_pct_2024, ~wikitext_shade(format_pct(.), "red")),
    p_ch_wt = map_chr(poverty_change, format_change_wt),
    w_01_wt = map_chr(water_lack_2001, ~wikitext_shade(format_pct(.), "blue")),
    w_24_wt = map_chr(water_lack_2024, ~wikitext_shade(format_pct(.), "blue")),
    w_ch_wt = map_chr(water_change, format_change_wt),
    s_01_wt = map_chr(sanitation_lack_2001, ~wikitext_shade(format_pct(.), "green")),
    s_24_wt = map_chr(sanitation_lack_2024, ~wikitext_shade(format_pct(.), "green")),
    s_ch_wt = map_chr(sanitation_change, format_change_wt)
  ) |>
  select(
    department,
    p_12_wt, p_24_wt, p_ch_wt,
    w_01_wt, w_24_wt, w_ch_wt,
    s_01_wt, s_24_wt, s_ch_wt
  ) 

# Build the Wikitable string with align=right
wt_header <- "{| class=\"wikitable\" align=right\n|+ Comparative Trends in Poverty and Basic Service Access by Department\n|-\n! align=left | Department !! colspan=3 | Poverty (NBI) !! colspan=3 | Lack of Water Access !! colspan=3 | Lack of Sanitation Access\n|-\n! align=left | !! 2012 !! 2024 !! Change !! 2001 !! 2024 !! Change !! 2001 !! 2024 !! Change\n"

wt_rows <- wikitable_rows |>
  mutate(
    dept_display = if_else(department == "Bolivia", paste0("'''", department, "'''"), department)
  ) |>
  glue::glue_data("|-\n| align=left | {dept_display} || {p_12_wt} || {p_24_wt} || {p_ch_wt} || {w_01_wt} || {w_24_wt} || {w_ch_wt} || {s_01_wt} || {s_24_wt} || {s_ch_wt}")

# Source row with citations
source_text <- "NBI indicates a Unsatisfied Basic Needs definition for poverty.  '''Source:''' [[National Institute of Statistics of Bolivia|INE Bolivia]], 2024 Bolivian Census"
source_refs <- '{{Cite| last = Instituto Nacional de Estadística| title = Pobreza: Tabulados por Municipio/TIOC| access-date = 2026-07-22| date = 2025| url = https://nube.ine.gob.bo/index.php/s/KQSoTrfdRQ5jldM/download}}<br/>{{Cite| last = Instituto Nacional de Estadística| title = Servicios Básicos: Tabulados por Municipio/TIOC| access-date = 2026-07-22| date = 2025| url = https://nube.ine.gob.bo/index.php/s/dEJeTU4j0czYIdG/download}}'
source_line <- paste0('|-\n|colspan="10"|', source_text, '<ref>', source_refs, '</ref>')

wt_footer <- paste0("\n", source_line, "\n|}")

wikitable_text <- paste0(wt_header, paste(wt_rows, collapse = "\n"), wt_footer)

# Display with copy button
tagList(
  h3("Copy WikiText Below"),
  copy_to_clipboard_button_wt(wikitable_text, "Copy Wikitable to Clipboard"),
  tags$pre(
    style = "background-color: #f6f8fa; padding: 12px; border-radius: 6px; overflow-x: auto; max-height: 400px;",
    wikitable_text
  )
)

```
