---
title: "Bolivia Municipalities on Wikipedia: Coverage by Language"
author: "Carwil Bjork-James"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
execute:
  warning: false
  message: false
---

## Overview

This document maps the completeness of Wikipedia coverage for Bolivia's 336
official municipalities across six Wikipedias: **English**, **Spanish**,
**Quechua**, **Aymara**, **German**, and **Portuguese**. 
Coverage is defined as having a Wikipedia article
linked from Wikidata via sitelinks.

The maps are designed to support Wikipedia editors who want to identify
municipalities lacking articles in one or more languages. They connect to the
deeper article-quality analysis in
[`bolivia-municipality-wikipedia.qmd`](bolivia-municipality-wikipedia.qmd),
which examines article content and structure on English Wikipedia.

---

## Setup

```{r setup}
library(tidyverse)
library(sf)
library(patchwork)
library(httr)
library(jsonlite)
library(rvest)
library(wikitools)

source("wikidata_presence_matrix.R")
source("R/get_wikidata_instances.R")
```

---

## Fetch data from Wikidata

### Wikipedia presence matrix

We query Wikidata for all items that are instances of
[municipality of Bolivia (Q1062710)](https://www.wikidata.org/wiki/Q1062710)
and build a presence matrix indicating which Wikipedia language editions have an
article for each municipality.

Results are cached to `data/bo_municipality_wiki_presence.rds` so that
subsequent renders do not repeat the API calls.

```{r fetch-presence}
#| cache: false

presence_cache <- "data/bo_municipality_wiki_presence.rds"

if (file.exists(presence_cache)) {
  bo_municipality_wiki_presence <- readRDS(presence_cache)
} else {
  bo_municipality_wiki_presence <- wikidata_instance_wikipedia_presence(
    class_qid  = "Q1062710",   # municipality of Bolivia
    languages  = NULL,         # auto-discover all languages present
    limit      = 500,
    batch_size = 20
  )
  saveRDS(bo_municipality_wiki_presence, presence_cache)
}
```

### INE codes (P14142)

We retrieve the INE municipal code (property
[P14142](https://www.wikidata.org/wiki/Property:P14142)) for each municipality.
These 6-digit codes (e.g. `010101`) allow us to join Wikidata entities to the
GADM boundary file via a pre-built crosswalk.

```{r fetch-ine-codes}
#| cache: false

ine_cache <- "data/municipalities_wd_ine.rds"

if (file.exists(ine_cache)) {
  municipalities_wd_ine <- readRDS(ine_cache)
} else {
  municipalities_wd_ine <- get_wikidata_instances(
    class_qid      = "Q1062710",
    property       = "P14142",
    property_names = "ine_code",
    languages      = c("en", "es"),
    limit          = 500,
    batch_size     = 20
  )
  saveRDS(municipalities_wd_ine, ine_cache)
}
```

### Category membership

As a second strategy we pull pages directly from Wikipedia categories. This
catches articles that exist and are correctly categorised but whose Wikidata
item lacks a sitelink — a common gap on the Spanish and Quechua editions.

The category structures differ by language edition:

- **English** (`en`): flat — all municipalities in one category.
- **Quechua** (`qu`): mixed — some pages are directly in *Munisipyu (Buliwya)*
  and others in its subcategories; we fetch both.
- **Portuguese** (`pt`): flat — one category, relatively sparse.

Spanish uses a different approach (see below): the curated table on
*Anexo:Municipios de Bolivia* lists every article with its INE code,
letting us join directly without going through Wikidata QIDs.

`muni_cat_specs` is the canonical registry; add entries for other editions as
needed. For each article found, the Wikidata QID is resolved via the MediaWiki
API and merged into the presence matrix.

```{r fetch-category-presence}
#| cache: false

# Helper: collect all article titles from a category, optionally descending
# one level into its subcategories.
# - include_self    : include pages directly in `category`
# - include_subcats : also collect pages from every subcategory
# Subcategory titles returned by the API carry a namespace prefix
# (e.g. "Categoría:", "Katiguriya:"); strip it before the recursive call
# because get_wp_category_members() adds the correct prefix internally.
.get_muni_cat_pages <- function(category, lang,
                                include_self    = TRUE,
                                include_subcats = FALSE) {
  empty <- tibble(pageid = integer(0), ns = integer(0), title = character(0))

  pages <- if (include_self) {
    get_wp_category_members(category, type = "page", lang = lang)
  } else {
    empty
  }

  if (include_subcats) {
    subcats <- get_wp_category_members(category, type = "subcat", lang = lang)
    if (nrow(subcats) > 0) {
      sub_pages <- map_dfr(subcats$title, function(full_title) {
        # Strip local namespace prefix ("Categoría:", "Katiguriya:", etc.)
        cat_base <- sub("^[^:]+:", "", full_title)
        get_wp_category_members(cat_base, type = "page", lang = lang)
      })
      pages <- bind_rows(pages, sub_pages)
    }
  }

  distinct(pages, title, .keep_all = TRUE)
}

# Per-language fetch specs:
#   category       : top-level category name (no namespace prefix needed)
#   include_self   : collect pages directly in that category
#   include_subcats: also collect pages from each subcategory
muni_cat_specs <- list(
  en = list(category = "Municipalities of Bolivia",
            include_self = TRUE,  include_subcats = FALSE),
  qu = list(category = "Munisipyu_(Buliwya)",
            include_self = TRUE,  include_subcats = TRUE),
  pt = list(category = "Municípios da Bolívia",
            include_self = TRUE,  include_subcats = FALSE)
)

cat_cache <- "data/bo_municipality_cat_presence.rds"

if (file.exists(cat_cache)) {
  cat_presence <- readRDS(cat_cache)
} else {
  cat_rows <- imap(muni_cat_specs, function(spec, lang) {
    message("Fetching ", lang, ".wikipedia: ", spec$category, " ...")
    members <- .get_muni_cat_pages(
      spec$category, lang,
      include_self    = spec$include_self,
      include_subcats = spec$include_subcats
    )
    if (nrow(members) == 0) {
      message("  (empty — skipping)")
      return(tibble(qid = character(0)))
    }
    message("  ", nrow(members), " pages; resolving Wikidata QIDs ...")
    info <- get_page_info_batch(members$title, lang = lang)
    tibble(qid = na.omit(info$wikidata_qid)) |>
      mutate(!!rlang::sym(lang) := 1L)
  })

  # Full-join per-language tibbles; NA → 0 where a QID is absent for a language
  non_empty <- keep(cat_rows, ~ nrow(.x) > 0)
  cat_presence <- if (length(non_empty) == 0) {
    tibble(qid = character(0))
  } else if (length(non_empty) == 1) {
    non_empty[[1]]
  } else {
    reduce(non_empty, full_join, by = "qid") |>
      mutate(across(-qid, ~ replace_na(.x, 0L)))
  }

  saveRDS(cat_presence, cat_cache)
}
```

### Spanish Wikipedia

The Spanish Wikipedia has a curated page —
[Anexo:Municipios de Bolivia](https://es.wikipedia.org/wiki/Anexo:Municipios_de_Bolivia)
— that lists every municipality with its *Código INE* and a direct link to
the article. Because the INE code is already present, we can join directly to
the GADM boundaries without resolving Wikidata QIDs. Every row in the table
carries an article link, so presence equals membership in the table.

```{r fetch-es-presence}
#| cache: false

es_cache <- "data/bo_es_ine_presence.rds"

if (file.exists(es_cache)) {
  es_ine_presence <- readRDS(es_cache)
} else {
  page  <- read_html(
    "https://es.wikipedia.org/w/index.php?title=Anexo:Municipios_de_Bolivia"
  )
  rows  <- page |> html_node("table.wikitable") |> html_nodes("tr")

  # Skip header row; extract INE code (col 1) and Nombre href (col 2)
  es_ine_presence <- map_dfr(rows[-1], function(row) {
    cells <- row |> html_nodes("td")
    if (length(cells) < 2) return(NULL)
    ine  <- cells[[1]] |> html_text(trim = TRUE)
    href <- cells[[2]] |> html_nodes("a") |> html_attr("href")
    if (length(href) == 0 || !grepl("^\\d{6}$", ine)) return(NULL)
    tibble(ine_code = ine, es_article = href[[1]])
  }) |>
    mutate(es = 1L)

  message(nrow(es_ine_presence), " municipalities found in Anexo:Municipios_de_Bolivia")
  saveRDS(es_ine_presence, es_cache)
}
```

### Combine sources

Take the union of the Wikidata sitelink and category-membership strategies for
English, Quechua, and Portuguese. Spanish is handled separately via the Anexo
table (INE-code join). Aymara uses Wikidata only.

```{r combine-presence}
cat_langs <- setdiff(names(cat_presence), "qid")

# Start from Wikidata data (excluding es, which comes from the Anexo table).
# Full-join category presence so QIDs found only in categories are also retained.
wd_data <- bo_municipality_wiki_presence$data |>
  select(qid, any_of(c("en", "qu", "ay", "de", "pt")))

presence_combined <- wd_data |>
  full_join(
    cat_presence |> rename_with(~ paste0("cat_", .), .cols = -qid),
    by = "qid"
  )

# Union: for each language with a category counterpart, prefer 1 over 0/NA
for (lang in cat_langs) {
  cat_col  <- paste0("cat_", lang)
  lang_sym <- rlang::sym(lang)
  if (lang %in% names(presence_combined)) {
    presence_combined <- presence_combined |>
      mutate(!!lang_sym := pmax(.data[[lang]], .data[[cat_col]], na.rm = TRUE))
  } else {
    # Language present only in category data — add the column
    presence_combined <- presence_combined |>
      mutate(!!lang_sym := replace_na(.data[[cat_col]], 0L))
  }
}
presence_combined <- presence_combined |> select(-starts_with("cat_"))

# Diagnostic: articles added by the category strategy
lang_labels <- c(en = "English", es = "Spanish", qu = "Quechua",
                 ay = "Aymara", de = "German", pt = "Portuguese")

gains <- map_dfr(intersect(cat_langs, setdiff(names(wd_data), "qid")), function(lang) {
  tibble(
    language      = lang,
    Wikipedia     = lang_labels[lang],
    wikidata_only = sum(wd_data[[lang]] == 1L, na.rm = TRUE),
    combined      = sum(presence_combined[[lang]] == 1L, na.rm = TRUE),
    gained        = sum(
      presence_combined[[lang]] == 1L &
        (is.na(wd_data[[lang]]) | wd_data[[lang]] == 0L),
      na.rm = TRUE
    )
  )
})

gains |>
  select(Wikipedia, `Via Wikidata` = wikidata_only,
         `Combined` = combined, `Added by category` = gained) |>
  knitr::kable(caption = "Articles added by the category-membership strategy")
```

---

## Build map dataset

We join the presence matrix to INE codes, then match to GADM boundaries through
the pre-built lookup table.

```{r build-map-data}
gadm_sf     <- st_read("data/gadm41_BOL_3.gpkg", layer = "ADM_ADM_3",
                       quiet = TRUE) |>
  select(NAME_1, NAME_3, geom)
gadm_lookup <- readRDS("data/gadm_lookup.rds")

# Join presence matrix → INE codes → GADM crosswalk → spatial boundaries.
# en/qu/ay/pt come from presence_combined (Wikidata + category union), joined
# via Wikidata QID → INE code. Spanish comes from es_ine_presence joined
# directly on INE code, bypassing QID resolution.
map_data <- gadm_sf |>
  inner_join(
    gadm_lookup |> select(gadm_dep, gadm_mun, cod.mun),
    by = c("NAME_1" = "gadm_dep", "NAME_3" = "gadm_mun")
  ) |>
  left_join(
    presence_combined |>
      select(qid, any_of(c("en", "qu", "ay", "de", "pt"))) |>
      left_join(municipalities_wd_ine |> select(qid, ine_code), by = "qid"),
    by = c("cod.mun" = "ine_code")
  ) |>
  left_join(
    es_ine_presence |> select(ine_code, es),
    by = c("cod.mun" = "ine_code")
  ) |>
  # es = 0 (not NA) for municipalities absent from the Anexo table
  mutate(es = replace_na(es, 0L)) |>
  mutate(
    across(c(en, es, qu, ay, de, pt),
           ~ factor(case_when(
             .x == 1L ~ "Has article",
             .x == 0L ~ "No article",
             TRUE      ~ "Not in Wikidata"
           ), levels = c("Has article", "No article", "Not in Wikidata")))
  )
```

---

## Coverage summary

```{r coverage-table}
# Count municipalities with / without an article per Wikipedia
coverage_summary <- map_data |>
  st_drop_geometry() |>
  summarise(across(c(en, es, qu, ay, de, pt), list(
    has    = ~ sum(.x == "Has article",      na.rm = TRUE),
    no     = ~ sum(.x == "No article",       na.rm = TRUE),
    absent = ~ sum(.x == "Not in Wikidata" | is.na(.x), na.rm = TRUE)
  ))) |>
  pivot_longer(everything(),
               names_to  = c("language", ".value"),
               names_sep = "_") |>
  mutate(
    wikipedia = c(en = "English", es = "Spanish", qu = "Quechua", ay = "Aymara",
                  de = "German", pt = "Portuguese")[language],
    pct_has   = round(100 * has / (has + no + absent), 1)
  ) |>
  select(wikipedia, has, no, absent, pct_has)

coverage_summary |>
  rename(
    Wikipedia      = wikipedia,
    `Has article`  = has,
    `No article`   = no,
    `Not in WD`    = absent,
    `% covered`    = pct_has
  ) |>
  knitr::kable(caption = "Municipality article coverage by Wikipedia language edition")
```

---

## Maps

Colors indicate whether Wikidata links the municipality to a Wikipedia article
in each language. Grey polygons are municipalities present in GADM but not
matched through Wikidata.

```{r map-colors}
presence_colors <- c(
  "Has article"     = "#2166ac",
  "No article"      = "#f7f7f7",
  "Not in Wikidata" = "#aaaaaa"
)
```

```{r map-helpers}
#| include: false

make_presence_map <- function(data, col, title) {
  bb <- sf::st_bbox(data)
  ggplot(data) +
    geom_sf(aes(fill = .data[[col]]), color = "white", linewidth = 0.12) +
    scale_fill_manual(
      values   = presence_colors,
      na.value = "#aaaaaa",
      name     = NULL,
      drop     = FALSE
    ) +
    coord_sf(xlim = bb[c("xmin", "xmax")], ylim = bb[c("ymin", "ymax")],
             expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 22),
      legend.position = "none"
    )
}

make_legend_panel <- function(data) {
  legend_plot <- ggplot(data) +
    geom_sf(aes(fill = en), color = "white", linewidth = 0.12) +
    scale_fill_manual(
      values   = presence_colors,
      na.value = "#aaaaaa",
      name     = NULL,
      drop     = FALSE
    ) +
    theme_void() +
    theme(
      legend.position  = "inside",
      legend.position.inside = c(0.5, 0.5),
      legend.text      = element_text(size = 16),
      legend.key.size  = unit(1.8, "cm"),
      legend.key.spacing = unit(0.5, "cm")
    )
  wrap_elements(full = cowplot::get_legend(legend_plot))
}
```

:::{.column-page-inset-right}
```{r map-panel}
#| fig-height: 18
#| fig-width: 24

p_en <- make_presence_map(map_data, "en", "English Wikipedia")
p_es <- make_presence_map(map_data, "es", "Spanish Wikipedia")
p_qu <- make_presence_map(map_data, "qu", "Quechua Wikipedia")
p_ay <- make_presence_map(map_data, "ay", "Aymara Wikipedia")
p_de <- make_presence_map(map_data, "de", "German Wikipedia")
p_pt <- make_presence_map(map_data, "pt", "Portuguese Wikipedia")
p_lg <- make_legend_panel(map_data)

# Flat layout: legend in row 1 col 4, spacer in row 2 col 4
design <- "
ABCD
####
EFG#
"

p_en + p_es + p_qu + p_lg +
  p_ay + p_de + p_pt +
  plot_layout(design = design, heights = c(1, 0.08, 1)) +
  plot_annotation(
    title   = "Bolivia: Municipal Wikipedia Coverage by Language",
    caption = "Sources: Wikidata sitelinks (Q1062710 instances) + Wikipedia category membership; boundaries: GADM 4.1",
    theme   = theme(
      plot.title = element_text(face = "bold", size = 28, hjust = 0.5)
    )
  )

```
:::
---

## Municipalities missing across multiple Wikipedias

Municipalities with no article in any of the four languages are the highest-priority
targets for new articles. The table below lists them, ordered by INE code.

```{r missing-all-four}
map_data |>
  st_drop_geometry() |>
  filter(
    en == "No article",
    es == "No article",
    qu == "No article",
    ay == "No article"
#    de == "No article",
#    pt == "No article"
  ) |>
  left_join(
    municipalities_wd_ine |> select(ine_code, label_en, label_es),
    by = c("cod.mun" = "ine_code")
  ) |>
  select(
    `INE code`  = cod.mun,
    `Name (EN)` = label_en,
    `Name (ES)` = label_es,
    English     = en,
    Spanish     = es,
    Quechua     = qu,
    Aymara      = ay,
    German      = de,
    Portuguese  = pt
  ) |>
  arrange(`INE code`) |>
  knitr::kable(caption = "Municipalities with no article in English, Spanish, Quechua, or Aymara")
```

---

## Spanish Wikipedia: settlement vs. municipality articles

The coverage count for Spanish Wikipedia understates a structural pattern on
es.wikipedia. Of the 340 articles listed in
[Anexo:Municipios de Bolivia](https://es.wikipedia.org/wiki/Anexo:Municipios_de_Bolivia),
only a minority link to Wikidata items classified as *municipio de Bolivia*
(Q1062710). The majority are linked instead to the Wikidata item for the
**municipal capital** — a city, town, or human settlement — which serves double
duty as the de facto article about both the urban centre and the surrounding
municipality.

This is not a coverage gap but a structural divergence: es.wikipedia treats the
settlement and the municipality as a single article subject, while Wikidata
models them as distinct entities. The Wikidata-only strategy therefore
undercounts Spanish coverage severely, which is why the Anexo-table approach
(joined by INE code) is used instead.

```{r es-p31-summary}
p31_per_article <- readRDS("data/bo_es_anexo_p31.rds")

# Summary by instance type
p31_per_article |>
  tidyr::separate_rows(instance_of, sep = "; ") |>
  count(instance_of, name = "n_articles") |>
  arrange(desc(n_articles)) |>
  knitr::kable(
    col.names = c("Instance of (Wikidata type)", "Articles"),
    caption   = paste0(
      "Wikidata instance types for the ", nrow(p31_per_article),
      " Anexo articles whose items are not classified as Q1062710 ",
      "(municipality of Bolivia)"
    )
  )
```

Out of the 340 entries in the Anexo:

- **`r nrow(p31_per_article)`** resolve to Wikidata items classified as
  settlements rather than municipalities — overwhelmingly *human settlement*
  and *city*, with towns, big cities, and border cities also represented.
- **`r sum(is.na(page_info$wikidata_qid))`** have no Wikidata item linked to
  their es.wikipedia article at all.
- **`r 340L - nrow(p31_per_article) - sum(is.na(page_info$wikidata_qid))`**
  link directly to a Q1062710 municipality item.

The per-article breakdown (article title, QID, and instance types) is saved to
`data/bo_es_anexo_p31.rds` for use in downstream analyses.

```{r es-p31-per-article}
p31_per_article |>
  arrange(title) |>
  knitr::kable(
    col.names = c("es.wikipedia article", "Wikidata QID", "Instance of"),
    caption   = "es.wikipedia Anexo articles linked to non-municipality Wikidata items"
  )
```

---

## Next steps

The deeper analysis of existing English Wikipedia articles — article length,
citations, infobox completeness, and census data quality — is in
[`bolivia-municipality-wikipedia.qmd`](bolivia-municipality-wikipedia.qmd),
with results saved to `data/bolivia_municipality_wikipedia.rds`.

Future work here will:

- Add a table of municipalities missing only in one language, to guide targeted
  translation campaigns
- Link article quality metrics (from `bolivia_municipality_wikipedia.rds`) to
  the coverage maps
- Track changes over time as new articles are created
