---
title: "Bolivia Municipality Wikipedia Pages"
author: "Carwil Bjork-James"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
editor: visual
---

# Overview

This document scrapes English Wikipedia for all pages in the [Municipalities of Bolivia](https://en.wikipedia.org/wiki/Category:Municipalities_of_Bolivia) category tree, extracts structured information from each page, and compares the coverage with Wikidata's inventory of Bolivian municipalities.

```{r setup}
#| message: false
library(tidyverse)
library(httr)
library(jsonlite)
library(WikipediR)

source("src/wikipedia-tools.R")
source("R/get_wikidata_instances.R")
```

# 1. Discover municipality pages from Wikipedia category

We start by listing the subcategories of `Category:Municipalities of Bolivia`, keeping only those whose name begins with "Municipalities" (i.e. excluding "Categories by city in Bolivia" and other non-municipal subcategories).

```{r get-subcategories}
top_cat <- "Category:Municipalities_of_Bolivia"
subcats <- get_subcategories(top_cat)
subcats

# Keep only subcategories beginning with "Municipalities"
muni_subcats <- subcats |>
  filter(str_detect(title, "^Category:Municipalities"))

muni_subcats
```

Now collect all article pages from each subcategory, plus any pages listed directly under the top-level category.

```{r get-all-pages}
# Pages directly in the top-level category
top_pages <- get_category_pages(top_cat)

# Pages in each "Municipalities of ..." subcategory
subcat_pages <- map_dfr(muni_subcats$title, function(cat) {
  Sys.sleep(0.3)
  get_category_pages(cat) |>
    mutate(source_category = cat)
})

# Combine and deduplicate
all_muni_pages <- bind_rows(
  top_pages |> mutate(source_category = top_cat),
  subcat_pages
) |>
  distinct(title, .keep_all = TRUE) |>
  # Drop any remaining category or template pages
  filter(ns == 0)

message(nrow(all_muni_pages), " unique municipality article pages found")
```

# 2. Fetch page metadata (page length + Wikidata QID)

```{r page-info}
page_info <- get_page_info_batch(all_muni_pages$title)

muni_df <- all_muni_pages |>
  select(-pageid) |>
  left_join(page_info, by = "title")
```

# 3. Cache and parse wikitext

We cache each page's raw wikitext to `wikipedia-pages/bolivia/` so that subsequent runs don't re-download.

```{r cache-wikitext}
cache_dir <- "wikipedia-pages/bolivia"
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

# Fetch / load cached wikitext for every page
muni_df <- muni_df |>
  mutate(wikitext = map_chr(title, function(t) {
    txt <- cache_wikitext(t, cache_dir)
    if (is.null(txt)) NA_character_ else txt
  }, .progress = "Fetching wikitext"))

muni_df <- muni_df |>
  mutate(wikipedia_url = paste0("https://en.wikipedia.org/wiki/", gsub(" ", "_", title)))

```

# 4. Extract structured fields from each page

## 4a. Character count, citations, and census years

```{r extract-counts}
muni_df <- muni_df |>
  mutate(
    n_chars = nchar(wikitext),
    n_citations = map_int(wikitext, count_citations),
    n_refs = map_int(wikitext, count_refs),
    census_years = map(wikitext, extract_census_years),
    census_years_str = map_chr(census_years, ~ paste(.x, collapse = ", "))
  )
```

## 4b. Infobox fields

Parse the main infobox from each article and extract every key–value pair. We first collect all infobox data as a list column, then pivot the most common fields into their own columns.

```{r extract-infoboxes}
muni_df <- muni_df |>
  mutate(
    infobox_wikitext = map_chr(wikitext, function(w) {
      txt <- extract_infobox_wikitext(w)
      if (is.null(txt)) NA_character_ else txt
    }),
    infobox_raw = map(wikitext, extract_infobox)
  )

is_blank <- \(v) v == "" | trimws(v) == "" | startsWith(trimws(v), "<!--")

muni_df <- muni_df |>
  mutate(infobox_fields = map(infobox_raw, \(x) names(x)[!is_blank(x)])) |>
  mutate(n_fields = lengths(infobox_fields)) |>
  mutate(infobox_values = map(infobox_raw, \(x) unname(x[!is_blank(x)])))

# Identify which infobox keys appear most often
all_keys <- muni_df |>
  filter(!map_lgl(infobox_raw, is.null)) |>
  pull(infobox_raw) |>
  map(names) |>
  unlist() |>
  # Normalise key names to lowercase with underscores
  str_to_lower() |>
  str_replace_all("\\s+", "_")

filled_keys <- muni_df |> 
  pull(infobox_fields) |> 
  unlist() 

key_freq <- tibble(key = filled_keys) |>
  count(key, sort = TRUE)
```

## 4c. Infobox structure and field coverage

### Template used

Nearly all Bolivia municipality articles use the general-purpose `{{Infobox settlement}}` template rather than a Bolivia-specific template. This template defines a large number of fields — most of which are left blank in any given article. The raw wikitext for **Vacas Municipality** (Cochabamba) illustrates the structure:

```{r vacas-example}
#| echo: false
muni_df |>
  filter(str_detect(title, "Vacas")) |>
  pull(infobox_wikitext) |>
  cat()
```

The template includes many placeholder fields with no value — lines like `|nickname =` or `|image_flag =`. These are excluded from `infobox_fields` and `infobox_values` by the `is_blank()` filter (which also drops values that are only whitespace or HTML comments). In the Vacas example, the substantively filled fields include `official_name`, `native_name`, `settlement_type`, the subdivision hierarchy, `seat`, `leader_title`/`leader_name`, `established_date`, `area_total_km2`, `population_total`, `population_as_of`, `coordinates`, `elevation_m`, and `area_code`.

### Field frequency across all articles

The table and chart below show how many of the `r nrow(muni_df)` municipality articles contain a non-blank value for each field.

```{r key-freq-table}
n_articles <- sum(!map_lgl(muni_df$infobox_raw, is.null))

key_freq |>
  mutate(pct = round(100 * n / n_articles, 1)) |>
  rename(field = key, articles = n, `% of articles` = pct) |>
  print(n = 40)
```

```{r key-freq-plot}
key_freq |>
  filter(n >= 5) |>
  mutate(key = fct_reorder(key, n)) |>
  ggplot(aes(x = n, y = key)) +
  geom_col() +
  labs(
    x = "Number of articles with a non-blank value",
    y = NULL,
    title = "Infobox field coverage across Bolivia municipality articles"
  )
```

A few observations stand out:

- **Universal fields** (≥185 of 187 articles): `settlement_type`, `subdivision_type`/`subdivision_name` (country, department, province), and `official_name`. These form the reliable core of the infobox data.
- **Near-universal but slightly less complete**: `unit_pref`, `utc_offset`, `timezone`, `pushpin_map_caption`, and `seat_type` — present in \~178 articles, suggesting a handful of older or stub articles lack them.
- **Geographic/demographic fields with partial coverage**: `coordinates` (157), `population_total` (153), `leader_name` (92), `elevation_m` (79), `image_skyline` (78), `area_total_km2` (73).
- **Sparse fields**: `established_date` (10), `population_density_km2` (16), and `native_name` (33) are present in only a minority of articles. Population density in particular appears to be rarely maintained.

```{r pivot-infobox}
# For each page, create a one-row tibble of cleaned infobox values
infobox_tbl <- muni_df |>
  transmute(
    title,
    infobox_row = map(infobox_raw, function(ib) {
      if (is.null(ib)) return(tibble())
      # Normalise key names
      names(ib) <- names(ib) |>
        str_to_lower() |>
        str_replace_all("\\s+", "_")
      # Clean values
      ib <- map(ib, clean_infobox_value)
      as_tibble(ib)
    })
  ) |>
  unnest(infobox_row, keep_empty = TRUE)

# Join back to main table (keeping the infobox columns separate for clarity)
muni_result <- muni_df |>
  select(-infobox_raw, -wikitext) |>
  left_join(infobox_tbl, by = "title")
```

# 5. Compare with Wikidata

Query Wikidata for all items that are instances of **municipality of Bolivia** ([Q1062710](https://www.wikidata.org/wiki/Q1062710)), then compare with the Wikipedia category listing.

```{r wikidata-query}
# Q1062710 = "municipality of Bolivia"
wd_munis <- get_wikidata_instances(
  "Q1062710",
  property = c("P131"),
  property_names = c("located_in"),
  country = "Q750",
  languages = c("en", "es"),
  limit = 500
)

# Extract English Wikipedia article title from the sitelinks
wd_munis <- wd_munis |>
  mutate(
    en_wiki_title = map_chr(wikipedia_articles, function(arts) {
      if (length(arts) == 0) return(NA_character_)
      en_art <- arts[str_detect(arts, "^en:")]
      if (length(en_art) == 0) return(NA_character_)
      str_replace(en_art[1], "^en:\\s*", "")
    })
  )
```

```{r compare-lists}
# Pages found in Wikipedia category but not in Wikidata
wp_only <- muni_result |>
  filter(!wikidata_qid %in% wd_munis$qid | is.na(wikidata_qid))

# Wikidata items with no Wikipedia category page
wd_only <- wd_munis |>
  filter(!qid %in% muni_result$wikidata_qid)

# Both
in_both <- muni_result |>
  filter(wikidata_qid %in% wd_munis$qid)

cat(glue::glue(
  "Wikipedia category pages: {nrow(muni_result)}\n",
  "Wikidata 'municipality of Bolivia' items: {nrow(wd_munis)}\n",
  "In both: {nrow(in_both)}\n",
  "Wikipedia only: {nrow(wp_only)}\n",
  "Wikidata only (no category page): {nrow(wd_only)}\n"
))
```

### Pages in Wikipedia category but not Wikidata

```{r wp-only-table}
wp_only |>
  select(title, wikidata_qid, n_chars, n_citations) |>
  arrange(title)
```

### Wikidata items with no Wikipedia category page

```{r wd-only-table}
wd_only |>
  select(qid, label_en, label_es, en_wiki_title) |>
  arrange(label_en)
```

# 6. Infobox completeness maps

These choropleths use the GADM boundary file and the same join infrastructure as [`bolivia-wikipedia-completeness.qmd`](bolivia-wikipedia-completeness.qmd), linking `muni_result` to spatial boundaries via Wikidata QIDs and INE codes.

White = no English Wikipedia article; grey = has article but field absent or not parseable; colors = field value as described per map.

```{r map-setup}
library(sf)

gadm_sf      <- st_read("data/gadm41_BOL_3.gpkg", layer = "ADM_ADM_3",
                         quiet = TRUE) |>
  select(NAME_1, NAME_3, geom)
gadm_lookup  <- readRDS("data/gadm_lookup.rds")
municipalities_wd_ine <- readRDS("data/municipalities_wd_ine.rds")

# Join muni_result fields to INE codes
map_info <- municipalities_wd_ine |>
  select(qid, ine_code) |>
  left_join(
    muni_result |>
      select(wikidata_qid, n_fields, population_as_of, leader_name) |>
      mutate(
        has_article = TRUE,
        pop_year = case_when(
          str_detect(population_as_of, "2024") ~ "2024",
          TRUE ~ population_as_of
        ) |> factor(levels = c("No article", "2001", "2010", "2011",
                               "2012", "2020", "2024")),
        leader_year = str_extract(leader_name, "\\b(20\\d{2})\\b") |>
          factor(levels = c("No article", "2007", "2008", "2010",
                            "2021", "2022", "2026"))
      ),
    by = c("qid" = "wikidata_qid")
  ) |>
  replace_na(list(has_article = FALSE))

# Build spatial dataset
infobox_map_data <- gadm_sf |>
  inner_join(
    gadm_lookup |> select(gadm_dep, gadm_mun, cod.mun),
    by = c("NAME_1" = "gadm_dep", "NAME_3" = "gadm_mun")
  ) |>
  left_join(map_info, by = c("cod.mun" = "ine_code")) |>
  mutate(
    # Recode discrete variables so "No article" renders as white
    # rather than collapsing with grey (field absent)
    pop_year    = if_else(!has_article, factor("No article", levels(pop_year)),
                          pop_year),
    leader_year = if_else(!has_article, factor("No article", levels(leader_year)),
                          leader_year)
  )
```

```{r map-theme}
#| include: false
theme_bolivia_map <- function() {
  theme_void(base_size = 11) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 12),
      plot.caption    = element_text(hjust = 0.5, size = 8, color = "grey50"),
      legend.position = "bottom"
    )
}
```

## 6a. Number of infobox fields filled

White = no Wikipedia article; grey = article exists but no infobox found. Color runs from pale (few fields filled) to dark blue (many fields).

```{r map-n-fields}
#| fig-height: 6
#| fig-width: 7

ggplot() +
  # White base for all polygons (= "no article" default)
  geom_sf(data = infobox_map_data, fill = "white", color = "grey80",
          linewidth = 0.12) +
  # Colored layer only for municipalities with an article
  geom_sf(data = infobox_map_data |> filter(has_article),
          aes(fill = n_fields), color = "white", linewidth = 0.12) +
  scale_fill_distiller(
    palette   = "Blues",
    direction = 1,
    na.value  = "#cccccc",
    name      = "Fields filled",
    limits    = c(0, NA)
  ) +
  labs(
    title   = "Bolivia: Infobox fields filled per municipality article",
    caption = "Source: English Wikipedia; boundaries: GADM 4.1.\nWhite = no article; grey = article with no infobox."
  ) +
  theme_bolivia_map()
```

## 6b. Census year used for population figures

Most articles cite the **2001** census. A small number have been updated to 2012 or 2024.

```{r map-pop-year}
#| fig-height: 6
#| fig-width: 7

pop_year_colors <- c(
  "No article" = "white",
  "2001"       = "#d1e5f0",
  "2010"       = "#92c5de",
  "2011"       = "#4393c3",
  "2012"       = "#2166ac",
  "2020"       = "#fddbc7",
  "2024"       = "#d6604d"
)

ggplot(infobox_map_data) +
  geom_sf(aes(fill = pop_year), color = "grey80", linewidth = 0.12) +
  scale_fill_manual(
    values   = pop_year_colors,
    na.value = "#cccccc",
    name     = "Census year",
    drop     = FALSE
  ) +
  labs(
    title   = "Bolivia: Census year cited in Wikipedia population figures",
    caption = "Source: English Wikipedia (population_as_of infobox field); boundaries: GADM 4.1.\nWhite = no article; grey = article with field absent."
  ) +
  theme_bolivia_map()
```

## 6c. Mayor election year

The year in parentheses after the mayor's name in `leader_name` indicates when that person's term began. Most articles reflect the **2007** Bolivian municipal elections and have not been updated since. Bolivia's most recent municipal elections were in **2021**; national elections in **2026**.

```{r map-leader-year}
#| fig-height: 6
#| fig-width: 7

leader_year_colors <- c(
  "No article" = "white",
  "2007"       = "#4575b4",
  "2008"       = "#74add1",
  "2010"       = "#abd9e9",
  "2021"       = "#fdae61",
  "2022"       = "#f46d43",
  "2026"       = "#d73027"
)

# Labels: append * to years present in scale but absent from data
leader_year_labels <- c(
  "No article" = "No article",
  "2007"       = "2007",
  "2008"       = "2008",
  "2010"       = "2010",
  "2021"       = "2021*",
  "2022"       = "2022",
  "2026"       = "2026*"
)

ggplot(infobox_map_data) +
  geom_sf(aes(fill = leader_year), color = "grey80", linewidth = 0.12) +
  scale_fill_manual(
    values   = leader_year_colors,
    labels   = leader_year_labels,
    na.value = "#cccccc",
    name     = "Election year",
    drop     = FALSE
  ) +
  # override.aes forces correct colors for factor levels with no mapped data;
  # append the na.value color so the count matches (7 levels + NA = 8 entries)
  guides(fill = guide_legend(
    override.aes = list(fill = c(unname(leader_year_colors), "#cccccc"))
  )) +
  labs(
    title   = "Bolivia: Election year of mayor named in Wikipedia infobox",
    caption = "Source: English Wikipedia (leader_name infobox field); boundaries: GADM 4.1.\nWhite = no article; grey = article with no dated leader. * = no municipalities currently listed."
  ) +
  theme_bolivia_map()
```

## 6d. Infobox field count summary

```{r n-fields-table}
muni_result |>
  filter(title != "Municipalities of Bolivia") |>
  select(title, wikidata_qid, n_fields) |>
  arrange(desc(n_fields)) |>
  mutate(rank = row_number()) |>
  filter(rank <= 10 | rank >= (n() - 9)) |>
  mutate(group = if_else(rank <= 10, "Most complete", "Least complete")) |>
  select(group, title, n_fields) |>
  knitr::kable(
    col.names = c("", "Municipality", "Fields filled"),
    caption   = "Bolivia municipality Wikipedia articles ranked by infobox fields filled (top 10 and bottom 10)"
  )
```

# 7. Summary table

```{r save-results}
# Save the main result
saveRDS(muni_result, "data/bolivia_municipality_wikipedia.rds")
write_csv(muni_result |> select(-census_years),
          "data/bolivia_municipality_wikipedia.csv")

muni_result |>
  select(title, wikidata_qid, page_length, n_chars, n_citations, n_refs,
         census_years_str) |>
  arrange(title) |>
  head(20)
```
