---
title: "Bolivia 2024 Census: Population and Ethnic Identification by Geography"
format: html
---

```{r setup}
library(readxl)
library(tidyverse)
```

## Part 1: Population by Geography (`POBLACION.xlsx`)

Source: sheet "1" of `POBLACION.xlsx`. The header spans Excel rows 5–7; data begins at row 9. Excel column A is blank and is dropped automatically by `read_excel`, so 24 columns remain (B–Y).

- **Cols 1–6** (B–G): row counter, geographic level, departamento, provincia, municipio, area
- **Cols 7–15** (H–P): population counts for 2001 / 2012 / 2024 × (total, hombre, mujer)
- **Cols 16–24** (Q–Y): derived percentages, same structure

Within each geographic level, rows come in triples: the entity total (area = entity name), then "Urbana", then "Rural".

```{r load-population}
raw <- read_excel(
  "../bolivia-data/Censo 2024/POBLACION.xlsx",
  sheet     = "1",
  col_names = FALSE,
  skip      = 8   # rows 1–8: title, three header rows, blank separator
)

col_names <- c(
  "row_num",
  "nivel",
  "departamento",
  "provincia",
  "municipio",
  "area",
  "pop_2001_total",  "pop_2001_hombre",  "pop_2001_mujer",
  "pop_2012_total",  "pop_2012_hombre",  "pop_2012_mujer",
  "pop_2024_total",  "pop_2024_hombre",  "pop_2024_mujer",
  "pct_2001_total",  "pct_2001_hombre",  "pct_2001_mujer",
  "pct_2012_total",  "pct_2012_hombre",  "pct_2012_mujer",
  "pct_2024_total",  "pct_2024_hombre",  "pct_2024_mujer"
)

names(raw) <- col_names

pop_raw <- raw |>
  mutate(
    row_num = as.integer(row_num),
    across(starts_with("pop_") | starts_with("pct_"), as.numeric)
  ) |>
  filter(!is.na(row_num))
```

### Disaggregate by geographic level

Each level has "total" rows (area = entity name) and subdivision rows (area = "Urbana" or "Rural").

```{r disaggregate-population}
subdivisions <- c("Urbana", "Rural")

pop_country        <- pop_raw |> filter(nivel == "País")

pop_dept           <- pop_raw |>
  filter(nivel == "Departamento", !area %in% subdivisions)

pop_dept_urb_rural <- pop_raw |>
  filter(nivel == "Departamento",  area %in% subdivisions)

pop_prov           <- pop_raw |>
  filter(nivel == "Provincia",     !area %in% subdivisions)

pop_prov_urb_rural <- pop_raw |>
  filter(nivel == "Provincia",      area %in% subdivisions)

pop_muni           <- pop_raw |>
  filter(nivel == "Municipio/TIOC", !area %in% subdivisions)

pop_muni_urb_rural <- pop_raw |>
  filter(nivel == "Municipio/TIOC",  area %in% subdivisions)
```

### Quick checks

```{r checks-population}
cat("Country rows:           ", nrow(pop_country),       "\n")
cat("Dept totals:            ", nrow(pop_dept),           "\n")
cat("Dept urban+rural rows:  ", nrow(pop_dept_urb_rural), "\n")
cat("Province totals:        ", nrow(pop_prov),           "\n")
cat("Province urban+rural:   ", nrow(pop_prov_urb_rural), "\n")
cat("Municipality totals:    ", nrow(pop_muni),           "\n")
cat("Municipality urb+rural: ", nrow(pop_muni_urb_rural), "\n")
```

```{r preview-dept}
pop_dept |>
  select(departamento, pop_2024_total, pop_2024_hombre, pop_2024_mujer)
```

```{r preview-muni}
pop_muni |>
  select(departamento, provincia, municipio, pop_2024_total, pop_2024_hombre, pop_2024_mujer) |>
  head(12)
```

### Save population outputs

```{r save-population}
#| eval: false
#| echo: true
# This code will not be run, since the data is already saved.
saveRDS(pop_raw,           "data/poblacion_raw.rds")
saveRDS(pop_country,       "data/poblacion_pais.rds")
saveRDS(pop_dept,          "data/poblacion_departamento.rds")
saveRDS(pop_dept_urb_rural,"data/poblacion_departamento_urb_rural.rds")
saveRDS(pop_prov,          "data/poblacion_provincia.rds")
saveRDS(pop_prov_urb_rural,"data/poblacion_provincia_urb_rural.rds")
saveRDS(pop_muni,          "data/poblacion_municipio.rds")
saveRDS(pop_muni_urb_rural,"data/poblacion_municipio_urb_rural.rds")
```

---

## Part 2: Ethnic Self-Identification (`AUTOIDENTIFICACION.xlsx`, sheet 3)

Source: sheet "3" of `AUTOIDENTIFICACION.xlsx`. This table reports, for each geographic
unit and urban/rural subdivision, the number of people who declared belonging to each of
133 indigenous nations, pueblos, and ethnic categories in the 2024 census. The `total`
column is the count of all declarations (a person who identified with multiple groups is
counted once per declaration).

The header spans two rows (Excel rows 5–6):

- **Row 5** geographic keys + `TOTAL` + merged header "DECLARACIONES DE AUTOIDENTIFICACIÓN
  CENSO 2024" spanning all ethnic-group columns.
- **Row 6** individual ethnic-group names (133 names, cols 7–139).

Data begins at row 7. Footnote rows at the bottom of the sheet (with long `nivel` strings)
are removed during cleaning.

```{r load-autoidentificacion}
auto_path <- "../bolivia-data/Censo 2024/AUTOIDENTIFICACION.xlsx"

# Row 6 of the sheet holds the 133 ethnic-group names; those become the trailing
# column names. Read it separately before reading the data.
ethnic_group_names <- read_excel(
  auto_path,
  sheet     = "3",
  col_names = FALSE,
  skip      = 5,   # skip rows 1–5; row 6 is the one we want
  n_max     = 1
) |>
  unlist(use.names = FALSE)

# Assemble the full column-name vector: 6 geographic keys + 133 ethnic groups.
auto_col_names <- c(
  "nivel", "departamento", "provincia", "municipio", "area", "total",
  ethnic_group_names
)

# Read data rows (starting at Excel row 7).
auto_raw <- read_excel(
  auto_path,
  sheet     = "3",
  col_names = FALSE,
  skip      = 6
)
names(auto_raw) <- auto_col_names

auto_raw <- auto_raw |>
  mutate(across(c(total, all_of(ethnic_group_names)), as.numeric)) |>
  # Drop footnote rows: they have very long nivel strings, not short level names
  filter(!is.na(nivel), nchar(nivel) <= 20)
```

### Disaggregate by geographic level

The row structure mirrors Part 1: each geographic entity appears three times (total,
Urbana, Rural). The `nivel` labels used here are identical to those in `POBLACION.xlsx`.

```{r disaggregate-autoidentificacion}
auto_country        <- auto_raw |> filter(nivel == "País",            !area %in% subdivisions)

auto_dept           <- auto_raw |>
  filter(nivel == "Departamento", !area %in% subdivisions)

auto_dept_urb_rural <- auto_raw |>
  filter(nivel == "Departamento",  area %in% subdivisions)

auto_prov           <- auto_raw |>
  filter(nivel == "Provincia",     !area %in% subdivisions)

auto_prov_urb_rural <- auto_raw |>
  filter(nivel == "Provincia",      area %in% subdivisions)

auto_muni           <- auto_raw |>
  filter(nivel == "Municipio/TIOC", !area %in% subdivisions)

auto_muni_urb_rural <- auto_raw |>
  filter(nivel == "Municipio/TIOC",  area %in% subdivisions)
```

### Quick checks

```{r checks-autoidentificacion}
cat("Country rows:           ", nrow(auto_country),       "\n")
cat("Dept totals:            ", nrow(auto_dept),           "\n")
cat("Dept urban+rural rows:  ", nrow(auto_dept_urb_rural), "\n")
cat("Province totals:        ", nrow(auto_prov),           "\n")
cat("Province urban+rural:   ", nrow(auto_prov_urb_rural), "\n")
cat("Municipality totals:    ", nrow(auto_muni),           "\n")
cat("Municipality urb+rural: ", nrow(auto_muni_urb_rural), "\n")
```

```{r preview-auto-dept}
# Total declarations by department (top 5 groups)
top5 <- auto_muni |>
  select(all_of(ethnic_group_names)) |>
  summarise(across(everything(), sum, na.rm = TRUE)) |>
  pivot_longer(everything(), names_to = "grupo", values_to = "n") |>
  slice_max(n, n = 5)

auto_dept |>
  select(departamento, total, Quechua, Aymara, Chiquitano, Guaraní, Mojeño) |>
  arrange(desc(total))
```

```{r preview-auto-muni}
auto_muni |>
  select(departamento, provincia, municipio, total, Quechua, Aymara) |>
  head(12)
```

### Save ethnic identification outputs

```{r save-autoidentificacion}
saveRDS(auto_raw,           "data/autoidentificacion_raw.rds")
saveRDS(auto_country,       "data/autoidentificacion_pais.rds")
saveRDS(auto_dept,          "data/autoidentificacion_departamento.rds")
saveRDS(auto_dept_urb_rural,"data/autoidentificacion_departamento_urb_rural.rds")
saveRDS(auto_prov,          "data/autoidentificacion_provincia.rds")
saveRDS(auto_prov_urb_rural,"data/autoidentificacion_provincia_urb_rural.rds")
saveRDS(auto_muni,          "data/autoidentificacion_municipio.rds")
saveRDS(auto_muni_urb_rural,"data/autoidentificacion_municipio_urb_rural.rds")
```
