---
title: "Bolivia 2024 Census: Population by Geography"
format: html
---

```{r setup}
library(readxl)
library(tidyverse)
```

## Load and clean

Source: sheet "1" of `POBLACION.xlsx`. The header spans Excel rows 5–7; data begins at row 9. Excel column A is blank and is dropped automatically by `read_excel`, so 24 columns remain (B–Y).

- **Cols 1–6** (B–G): row counter, geographic level, departamento, provincia, municipio, area
- **Cols 7–15** (H–P): population counts for 2001 / 2012 / 2024 × (total, hombre, mujer)
- **Cols 16–24** (Q–Y): derived percentages, same structure

Within each geographic level, rows come in triples: the entity total (area = entity name), then "Urbana", then "Rural".

```{r load}
raw <- read_excel(
  "../bolivia-data/Censo 2024/POBLACION.xlsx",
  sheet     = "1",
  col_names = FALSE,
  skip      = 8   # rows 1–8: title, three header rows, blank separator
)

col_names <- c(
  "row_num",
  "nivel",
  "departamento",
  "provincia",
  "municipio",
  "area",
  "pop_2001_total",  "pop_2001_hombre",  "pop_2001_mujer",
  "pop_2012_total",  "pop_2012_hombre",  "pop_2012_mujer",
  "pop_2024_total",  "pop_2024_hombre",  "pop_2024_mujer",
  "pct_2001_total",  "pct_2001_hombre",  "pct_2001_mujer",
  "pct_2012_total",  "pct_2012_hombre",  "pct_2012_mujer",
  "pct_2024_total",  "pct_2024_hombre",  "pct_2024_mujer"
)

names(raw) <- col_names

pop_raw <- raw |>
  mutate(
    row_num = as.integer(row_num),
    across(starts_with("pop_") | starts_with("pct_"), as.numeric)
  ) |>
  # Drop footnote/blank rows at the bottom (no row number)
  filter(!is.na(row_num))
```

## Disaggregate by geographic level

Each level has "total" rows (area = entity name) and subdivision rows (area = "Urbana" or "Rural").

```{r disaggregate}
subdivisions <- c("Urbana", "Rural")

pop_country        <- pop_raw |> filter(nivel == "País")

pop_dept           <- pop_raw |>
  filter(nivel == "Departamento", !area %in% subdivisions)

pop_dept_urb_rural <- pop_raw |>
  filter(nivel == "Departamento",  area %in% subdivisions)

pop_prov           <- pop_raw |>
  filter(nivel == "Provincia",     !area %in% subdivisions)

pop_prov_urb_rural <- pop_raw |>
  filter(nivel == "Provincia",      area %in% subdivisions)

pop_muni           <- pop_raw |>
  filter(nivel == "Municipio/TIOC", !area %in% subdivisions)

pop_muni_urb_rural <- pop_raw |>
  filter(nivel == "Municipio/TIOC",  area %in% subdivisions)
```

## Quick checks

```{r checks}
cat("Country rows:           ", nrow(pop_country),       "\n")
cat("Dept totals:            ", nrow(pop_dept),           "\n")
cat("Dept urban+rural rows:  ", nrow(pop_dept_urb_rural), "\n")
cat("Province totals:        ", nrow(pop_prov),           "\n")
cat("Province urban+rural:   ", nrow(pop_prov_urb_rural), "\n")
cat("Municipality totals:    ", nrow(pop_muni),           "\n")
cat("Municipality urb+rural: ", nrow(pop_muni_urb_rural), "\n")
```

```{r preview-dept}
pop_dept |>
  select(departamento, pop_2024_total, pop_2024_hombre, pop_2024_mujer)
```

```{r preview-muni}
pop_muni |>
  select(departamento, provincia, municipio, pop_2024_total, pop_2024_hombre, pop_2024_mujer) |>
  head(12)
```

## Save outputs

```{r save}
saveRDS(pop_raw,           "data/poblacion_raw.rds")
saveRDS(pop_country,       "data/poblacion_pais.rds")
saveRDS(pop_dept,          "data/poblacion_departamento.rds")
saveRDS(pop_dept_urb_rural,"data/poblacion_departamento_urb_rural.rds")
saveRDS(pop_prov,          "data/poblacion_provincia.rds")
saveRDS(pop_prov_urb_rural,"data/poblacion_provincia_urb_rural.rds")
saveRDS(pop_muni,          "data/poblacion_municipio.rds")
saveRDS(pop_muni_urb_rural,"data/poblacion_municipio_urb_rural.rds")
```
