---
title: "Bolivia 2024 Census: Population and Ethnic Identification by Geography"
format: html
---

```{r setup}
library(readxl)
library(tidyverse)
```

## Part 1: Population by Geography (`POBLACION.xlsx`)

Source: sheet "1" of `POBLACION.xlsx`. The header spans Excel rows 5–7; data begins at row 9. Excel column A is blank and is dropped automatically by `read_excel`, so 24 columns remain (B–Y).

- **Cols 1–6** (B–G): row counter, geographic level, departamento, provincia, municipio, area
- **Cols 7–15** (H–P): population counts for 2001 / 2012 / 2024 × (total, hombre, mujer)
- **Cols 16–24** (Q–Y): derived percentages, same structure

Within each geographic level, rows come in triples: the entity total (area = entity name), then "Urbana", then "Rural".

```{r load-population}
raw <- read_excel(
  "../bolivia-data/censo_2024/POBLACION.xlsx",
  sheet     = "1",
  col_names = FALSE,
  skip      = 8   # rows 1–8: title, three header rows, blank separator
)

col_names <- c(
  "row_num",
  "nivel",
  "departamento",
  "provincia",
  "municipio",
  "area",
  "pop_2001_total",  "pop_2001_hombre",  "pop_2001_mujer",
  "pop_2012_total",  "pop_2012_hombre",  "pop_2012_mujer",
  "pop_2024_total",  "pop_2024_hombre",  "pop_2024_mujer",
  "pct_2001_total",  "pct_2001_hombre",  "pct_2001_mujer",
  "pct_2012_total",  "pct_2012_hombre",  "pct_2012_mujer",
  "pct_2024_total",  "pct_2024_hombre",  "pct_2024_mujer"
)

names(raw) <- col_names

pop_raw <- raw |>
  mutate(
    row_num = as.integer(row_num),
    across(starts_with("pop_") | starts_with("pct_"), as.numeric)
  ) |>
  filter(!is.na(row_num))
```

### Disaggregate by geographic level

Each level has "total" rows (area = entity name) and subdivision rows (area = "Urbana" or "Rural").

```{r disaggregate-population}
subdivisions <- c("Urbana", "Rural")

pop_country        <- pop_raw |> filter(nivel == "País")

pop_dept           <- pop_raw |>
  filter(nivel == "Departamento", !area %in% subdivisions)

pop_dept_urb_rural <- pop_raw |>
  filter(nivel == "Departamento",  area %in% subdivisions)

pop_prov           <- pop_raw |>
  filter(nivel == "Provincia",     !area %in% subdivisions) |> 
  filter(provincia != "Territorio Indígena Multiétnico")

pop_prov_urb_rural <- pop_raw |>
  filter(nivel == "Provincia",      area %in% subdivisions) |> 
  filter(provincia != "Territorio Indígena Multiétnico")

pop_muni           <- pop_raw |>
  filter(nivel == "Municipio/TIOC", !area %in% subdivisions)

pop_muni_urb_rural <- pop_raw |>
  filter(nivel == "Municipio/TIOC",  area %in% subdivisions)
```

```{r expand-prov-lookup-table}
#| eval: false
prov_id_lookup_table <- readRDS("data/prov_id_lookup_table.rds")
prov_id_lookup_long <- prov_id_lookup_table |> 
  pivot_longer(cols = starts_with("province"), names_to = "prov_list", values_to = "province") 
prov_id_lookup <- prov_id_lookup_long |>
  distinct(id_prov, department, province)

pop_prov_with_id <- pop_prov |> left_join(prov_id_lookup, by = c("provincia" = "province", 
                                             "departamento" = "department")) 
pop_prov_with_id |> filter(is.na(id_prov)) |> select(departamento, provincia, id_prov) ->
  pop_prov_missing_id

prov_id_no_match <- prov_id_lookup |> 
  anti_join(pop_prov, by = c("province" = "provincia", 
                             "department" = "departamento"))

# Manual matches: pop_prov_missing_id → id_prov (same dept, similar name)
# Beni / Territorio Indígena Multiétnico is a TIOC with no standard province code
manual_prov_match <- tribble(
  ~provincia,                    ~departamento,  ~id_prov,
  "Sur Yungas",                  "La Paz",       "0211",  # Sud Yungas
  "Sur Lípez",                   "Potosí",       "0510",  # Sur Lipez / Sud Lípez
  "General Bernardino Bilbao",   "Potosí",       "0513",  # General Bilbao
  "Avilez",                      "Tarija",       "0604",  # Avilés
  "Valle Grande",                "Santa Cruz",   "0708",  # Vallegrande
  "Obispo Santisteban",          "Santa Cruz",   "0710",  # Obispo Santistéban
  "Ñuflo de Chávez",             "Santa Cruz",   "0711",  # Ñuflo De Chávez
  "Ángel Sandoval",              "Santa Cruz",   "0712"   # Angel Sandoval
)

# Build cpv2024 name → id_prov lookup from auto-matched + manual rows
cpv_prov_name_lookup <- pop_prov_with_id |>
  filter(!is.na(id_prov)) |>
  select(id_prov, prov_cpv2024 = provincia) |>
  bind_rows(manual_prov_match |> select(id_prov, prov_cpv2024 = provincia)) |>
  distinct(id_prov, .keep_all = TRUE)

# Add prov_cpv2024 column to prov_id_lookup_table
prov_id_lookup_table <- readRDS("data/prov_id_lookup_table.rds") # reload for clean merge
prov_id_lookup_table <- prov_id_lookup_table |>
  left_join(cpv_prov_name_lookup, by = "id_prov") |> 
  relocate(prov_cpv2024, .before=department) |> 
  rename(province_cpv2024 = prov_cpv2024) |>
  # recount distinct province names per department to identify duplicates
  rowwise() |> mutate(
    n_unique = n_distinct(c(province_gb2014, province_anexo, province_census, 
                            province_gadm, province_cpv2024), na.rm = TRUE)
  ) |> 
  ungroup()

# saveRDS(prov_id_lookup_table, "data/prov_id_lookup_table.rds")
# 
# source("src/update-versioned-archive.R")
# update_versioned_archive(prov_id_lookup_table, "../ultimate-consequences/data/prov_id_lookup_table.rds")

```

```{r add-ine-code}
muni_id_lookup_table <- readRDS("data/muni_id_lookup_table.rds")
crosswalk_muni <- muni_id_lookup_table |>
  select(muni_cpv2024, department, id_muni) |>
  rename(ine_code = id_muni)

prov_id_lookup_table <- readRDS("data/prov_id_lookup_table.rds")
crosswalk_prov <- prov_id_lookup_table |>
  select(province_cpv2024, department, id_prov) |>
  rename(ine_code = id_prov)

dep_crosswalk <- crosswalk_prov |>
  mutate(id_dep = str_extract(ine_code, "^\\d{2}")) |>
  select(-province_cpv2024, -ine_code) |> 
  distinct()

add_ine_code_muni <- function(df) {
  df |>
    left_join(crosswalk_muni, by = c("municipio" = "muni_cpv2024", "departamento" = "department")) |>
    relocate(ine_code, municipio, provincia, departamento, area, nivel)
}

add_ine_code_prov <- function(df) {
  df |>
    left_join(crosswalk_prov, by = c("provincia" = "province_cpv2024", "departamento" = "department")) |>
    relocate(ine_code, provincia, departamento, area, nivel) |>
    select(-municipio)
}

add_ine_code_dept <- function(df) {
  df |>
    left_join(dep_crosswalk, by = c("departamento" = "department")) |>
    relocate(id_dep, departamento, area, nivel) |>
    select(-provincia, -municipio)
}

pop_muni           <- add_ine_code_muni(pop_muni)
pop_muni_urb_rural <- add_ine_code_muni(pop_muni_urb_rural)
pop_prov           <- add_ine_code_prov(pop_prov)
pop_prov_urb_rural <- add_ine_code_prov(pop_prov_urb_rural)
pop_dept           <- add_ine_code_dept(pop_dept)
pop_dept_urb_rural <- add_ine_code_dept(pop_dept_urb_rural)
```

### Quick checks

```{r checks-population}
cat("Country rows:           ", nrow(pop_country),       "\n")
cat("Dept totals:            ", nrow(pop_dept),           "\n")
cat("Dept urban+rural rows:  ", nrow(pop_dept_urb_rural), "\n")
cat("Province totals:        ", nrow(pop_prov),           "\n")
cat("Province urban+rural:   ", nrow(pop_prov_urb_rural), "\n")
cat("Municipality totals:    ", nrow(pop_muni),           "\n")
cat("Municipality urb+rural: ", nrow(pop_muni_urb_rural), "\n")
```

```{r preview-dept}
pop_dept |>
  select(departamento, pop_2024_total, pop_2024_hombre, pop_2024_mujer)
```

```{r preview-muni}
pop_muni |>
  select(departamento, provincia, municipio, pop_2024_total, pop_2024_hombre, pop_2024_mujer) |>
  head(12)
```

### Save population outputs

```{r prep-for-saving}
#| eval: false
#| echo: true

# These have all Spanish names except for "pop" which should be "pob"
es_var_name_fix <- function(x) {
  x |> rename_with(~ stringr::str_replace(., "^pop", "pob"), starts_with("pop"))
}

pop_raw_es <- es_var_name_fix(pop_raw)
pop_country_es <- es_var_name_fix(pop_country)
pop_dept_es <- es_var_name_fix(pop_dept)
pop_dept_urb_rural_es <- es_var_name_fix(pop_dept_urb_rural)
pop_prov_es <- es_var_name_fix(pop_prov)
pop_prov_urb_rural_es <- es_var_name_fix(pop_prov_urb_rural)
pop_muni_es <- es_var_name_fix(pop_muni)
pop_muni_urb_rural_es <- es_var_name_fix(pop_muni_urb_rural)

english_var_name_fix <- function(x) {
  x |>
    rename(
      department = any_of("departamento"),
      province = any_of("provincia"),
      municipality = any_of("municipio"),
      level = any_of("nivel")
    ) |>
    rename_with(~ stringr::str_replace(., "_mujer", "_female"), ends_with("mujer")) |>
    rename_with(~ stringr::str_replace(., "_hombre", "_male"), ends_with("hombre"))
}

pop_raw_en <- english_var_name_fix(pop_raw)
pop_country_en <- english_var_name_fix(pop_country)
pop_dept_en <- english_var_name_fix(pop_dept)
pop_dept_urb_rural_en <- english_var_name_fix(pop_dept_urb_rural)
pop_prov_en <- english_var_name_fix(pop_prov)
pop_prov_urb_rural_en <- english_var_name_fix(pop_prov_urb_rural)
pop_muni_en <- english_var_name_fix(pop_muni)
pop_muni_urb_rural_en <- english_var_name_fix(pop_muni_urb_rural)

```

```{r checks-output}
cat("English output datasets:\n")
cat("Country rows:           ", nrow(pop_country_en),        "\n")
cat("Dept totals:            ", nrow(pop_dept_en),            "\n")
cat("Dept urban+rural rows:  ", nrow(pop_dept_urb_rural_en),  "\n")
cat("Province totals:        ", nrow(pop_prov_en),            "\n")
cat("Province urban+rural:   ", nrow(pop_prov_urb_rural_en),  "\n")
cat("Municipality totals:    ", nrow(pop_muni_en),            "\n")
cat("Municipality urb+rural: ", nrow(pop_muni_urb_rural_en),  "\n")

cat("\nSpanish output datasets:\n")
cat("Country rows:           ", nrow(pop_country_es),        "\n")
cat("Dept totals:            ", nrow(pop_dept_es),            "\n")
cat("Dept urban+rural rows:  ", nrow(pop_dept_urb_rural_es),  "\n")
cat("Province totals:        ", nrow(pop_prov_es),            "\n")
cat("Province urban+rural:   ", nrow(pop_prov_urb_rural_es),  "\n")
cat("Municipality totals:    ", nrow(pop_muni_es),            "\n")
cat("Municipality urb+rural: ", nrow(pop_muni_urb_rural_es),  "\n")
```

```{r save-population data }
#| eval: false
#| echo: true

# This code will not be run, since the data is already saved.
saveRDS(pop_raw_es,           "data/cpv2024/poblacion_raw.rds")
saveRDS(pop_country_es,       "data/cpv2024/poblacion_pais.rds")
saveRDS(pop_dept_es,          "data/cpv2024/poblacion_departamento.rds")
saveRDS(pop_dept_urb_rural_es,"data/cpv2024/poblacion_departamento_urb_rural.rds")
saveRDS(pop_prov_es,          "data/cpv2024/poblacion_provincia.rds")
saveRDS(pop_prov_urb_rural_es,"data/cpv2024/poblacion_provincia_urb_rural.rds")
saveRDS(pop_muni_es,          "data/cpv2024/poblacion_municipio.rds")
saveRDS(pop_muni_urb_rural_es,"data/cpv2024/poblacion_municipio_urb_rural.rds")

# This code will not be run, since the data is already saved.
saveRDS(pop_raw_en,           "data/cpv2024/population_raw.rds")
saveRDS(pop_country_en,       "data/cpv2024/population_country.rds")
saveRDS(pop_dept_en,          "data/cpv2024/population_department.rds")
saveRDS(pop_dept_urb_rural_en,"data/cpv2024/population_department_urb_rural.rds")
saveRDS(pop_prov_en,          "data/cpv2024/population_province.rds")
saveRDS(pop_prov_urb_rural_en,"data/cpv2024/population_province_urb_rural.rds")
saveRDS(pop_muni_en,          "data/cpv2024/population_municipality.rds")
saveRDS(pop_muni_urb_rural_en,"data/cpv2024/population_municipality_urb_rural.rds")

```


---

## Part 2: Ethnic Self-Identification (`AUTOIDENTIFICACION.xlsx`, sheet 3)

Source: sheet "3" of `AUTOIDENTIFICACION.xlsx`. This table reports, for each geographic
unit and urban/rural subdivision, the number of people who declared belonging to each of
133 indigenous nations, pueblos, and ethnic categories in the 2024 census. The `total`
column is the count of all declarations (a person who identified with multiple groups is
counted once per declaration).

The header spans two rows (Excel rows 5–6):

- **Row 5** geographic keys + `TOTAL` + merged header "DECLARACIONES DE AUTOIDENTIFICACIÓN
  CENSO 2024" spanning all ethnic-group columns.
- **Row 6** individual ethnic-group names (133 names, cols 7–139).

Data begins at row 7. Footnote rows at the bottom of the sheet (with long `nivel` strings)
are removed during cleaning.

```{r load-autoidentificacion}
auto_path <- "../bolivia-data/censo_2024/AUTOIDENTIFICACION.xlsx"

# Row 6 of the sheet holds the 133 ethnic-group names; those become the trailing
# column names. Read it separately before reading the data.
ethnic_group_names <- read_excel(
  auto_path,
  sheet     = "3",
  col_names = FALSE,
  skip      = 5,   # skip rows 1–5; row 6 is the one we want
  n_max     = 1
) |>
  unlist(use.names = FALSE)

# Assemble the full column-name vector: 6 geographic keys + 133 ethnic groups.
auto_col_names <- c(
  "nivel", "departamento", "provincia", "municipio", "area", "total",
  ethnic_group_names
)

# Read data rows (starting at Excel row 7).
auto_raw <- read_excel(
  auto_path,
  sheet     = "3",
  col_names = FALSE,
  skip      = 6
)
names(auto_raw) <- auto_col_names

auto_raw <- auto_raw |>
  mutate(across(c(total, all_of(ethnic_group_names)), as.numeric)) |>
  # Drop footnote rows: they have very long nivel strings, not short level names
  filter(!is.na(nivel), nchar(nivel) <= 20)
```

### Disaggregate by geographic level

The row structure mirrors Part 1: each geographic entity appears three times (total,
Urbana, Rural). The `nivel` labels used here are identical to those in `POBLACION.xlsx`.

```{r disaggregate-autoidentificacion}
auto_country        <- auto_raw |> filter(nivel == "País",            !area %in% subdivisions)

auto_dept           <- auto_raw |>
  filter(nivel == "Departamento", !area %in% subdivisions)

auto_dept_urb_rural <- auto_raw |>
  filter(nivel == "Departamento",  area %in% subdivisions)

auto_prov           <- auto_raw |>
  filter(nivel == "Provincia",     !area %in% subdivisions)

auto_prov_urb_rural <- auto_raw |>
  filter(nivel == "Provincia",      area %in% subdivisions)

auto_muni           <- auto_raw |>
  filter(nivel == "Municipio/TIOC", !area %in% subdivisions)

auto_muni_urb_rural <- auto_raw |>
  filter(nivel == "Municipio/TIOC",  area %in% subdivisions)
```

```{r add-ine-code-autoidentificacion}
auto_muni           <- add_ine_code_muni(auto_muni)
auto_muni_urb_rural <- add_ine_code_muni(auto_muni_urb_rural)
auto_prov           <- add_ine_code_prov(auto_prov)
auto_prov_urb_rural <- add_ine_code_prov(auto_prov_urb_rural)
auto_dept           <- add_ine_code_dept(auto_dept)
auto_dept_urb_rural <- add_ine_code_dept(auto_dept_urb_rural)
```

### Quick checks

```{r checks-autoidentificacion}
cat("Country rows:           ", nrow(auto_country),       "\n")
cat("Dept totals:            ", nrow(auto_dept),           "\n")
cat("Dept urban+rural rows:  ", nrow(auto_dept_urb_rural), "\n")
cat("Province totals:        ", nrow(auto_prov),           "\n")
cat("Province urban+rural:   ", nrow(auto_prov_urb_rural), "\n")
cat("Municipality totals:    ", nrow(auto_muni),           "\n")
cat("Municipality urb+rural: ", nrow(auto_muni_urb_rural), "\n")
```

```{r preview-auto-dept}
# Total declarations by department (top 5 groups)
top5 <- auto_muni |>
  select(all_of(ethnic_group_names)) |>
  summarise(across(everything(), sum, na.rm = TRUE)) |>
  pivot_longer(everything(), names_to = "grupo", values_to = "n") |>
  slice_max(n, n = 5)

auto_dept |>
  select(departamento, total, Quechua, Aymara, Chiquitano, Guaraní, Mojeño) |>
  arrange(desc(total))
```

```{r preview-auto-muni}
auto_muni |>
  select(departamento, provincia, municipio, total, Quechua, Aymara) |>
  arrange(desc(total)) |> 
  head(12)
```

### Save ethnic identification outputs

```{r save-autoidentificacion}
#| eval: false
#| echo: true
# This code will not be run, since the data is already saved.
saveRDS(auto_raw,           "data/cpv2024/autoidentificacion_raw.rds")
saveRDS(auto_country,       "data/cpv2024/autoidentificacion_pais.rds")
saveRDS(auto_dept,          "data/cpv2024/autoidentificacion_departamento.rds")
saveRDS(auto_dept_urb_rural,"data/cpv2024/autoidentificacion_departamento_urb_rural.rds")
saveRDS(auto_prov,          "data/cpv2024/autoidentificacion_provincia.rds")
saveRDS(auto_prov_urb_rural,"data/cpv2024/autoidentificacion_provincia_urb_rural.rds")
saveRDS(auto_muni,          "data/cpv2024/autoidentificacion_municipio.rds")
saveRDS(auto_muni_urb_rural,"data/cpv2024/autoidentificacion_municipio_urb_rural.rds")

auto_raw_en <- english_var_name_fix(auto_raw)
auto_country_en <- english_var_name_fix(auto_country)
auto_dept_en <- english_var_name_fix(auto_dept)
auto_dept_urb_rural_en <- english_var_name_fix(auto_dept_urb_rural)
auto_prov_en <- english_var_name_fix(auto_prov)
auto_prov_urb_rural_en <- english_var_name_fix(auto_prov_urb_rural)
auto_muni_en <- english_var_name_fix(auto_muni)
auto_muni_urb_rural_en <- english_var_name_fix(auto_muni_urb_rural)

# This code will not be run, since the data is already saved.
saveRDS(auto_raw_en,           "data/cpv2024/autoidentificacion_raw.rds")
saveRDS(auto_country_en,       "data/cpv2024/autoidentificacion_country.rds")
saveRDS(auto_dept_en,          "data/cpv2024/autoidentificacion_department.rds")
saveRDS(auto_dept_urb_rural_en,"data/cpv2024/autoidentificacion_department_urb_rural.rds")
saveRDS(auto_prov_en,          "data/cpv2024/autoidentificacion_province.rds")
saveRDS(auto_prov_urb_rural_en,"data/cpv2024/autoidentificacion_province_urb_rural.rds")
saveRDS(auto_muni_en,          "data/cpv2024/autoidentificacion_municipality.rds")
saveRDS(auto_muni_urb_rural_en,"data/cpv2024/autoidentificacion_municipality_urb_rural.rds")
```
