---
title: "Bolivia CPV 2024: Population QuickStatements for Wikidata"
format: html
freeze: true
---

::: callout-note
## AI-generated text, click for process details

*This page was first generated using the Posit AI plug-in running Claude Sonnet 4.6, operating in my repository wiki-graph, which already contained the functions referenced here. The prompt I used was:*

For each of the following tasks, build functions and embed the process in a qmd that explains the needed steps.

Before we can deploy the quickstatements below, we need to clear out the small number of P1082 values. They only appear in these ID's: "Q628665" "Q1516972" "Q1814603" "Q1814613" "Q1814898" "Q1814929" "Q1952982"

We're going to build a new table building off of pop_muni, with qid values added, using muni_id_lookup_table to connect them through id_muni.

Each row will drive a series of columns with quickstatements produced by add_quick_statement_column, one for each of the three total population columns. These quickstatements will add a value for P1082 (population) for the corresponding year of the census. All will be sourced to Q139668018.

*The page reflects extensive manual and automated revisions to the code following a debugging process. Several sections below were added by hand.*
:::

## Overview

This document adds P1082 (population) statements to Wikidata items for Bolivia's 339 municipalities **and** for Bolivia's 9 departments and 112 provinces using census counts from three surveys: 2001, 2012, and 2024 (CPV 2024).

Each statement is sourced to Q139668018 (the Wikidata item for the CPV 2024 publication) and carries a P585 (point in time) qualifier for the relevant census year.

Before adding new statements, a small number of pre-existing P1082 claims on seven municipality items must be removed. Those items were tagged during a prior data-cleaning pass; their existing values are undated or unreferenced and would otherwise create duplicates.

## Setup

```{r setup}
library(tidyverse)
library(readr)
library(httr)
library(jsonlite)

source("create_quick_statement.R")
source("get_wikidata_instances.R")
```

```{r load-data}
muni_id_lookup_table <- readRDS("../ultimate-consequences/data/muni_id_lookup_table.rds")
pop_muni             <- readRDS("data/cpv2024/poblacion_municipio.rds")
pop_prov             <- readRDS("data/cpv2024/poblacion_provincia.rds")
pop_dept             <- readRDS("data/cpv2024/poblacion_departamento.rds")
```

```{r constants}
# CPV 2024 publication item used as "stated in" (P248)
SOURCE_QID <- "Q139668018"

p585_for_year <- function(year) sprintf("+%d-01-01T00:00:00Z/9", year)

# Wikidata instance-of classes
BOLIVIA_QID            <- "Q750"
DEPARTMENT_CLASS_QID   <- "Q250050"  # department of Bolivia
PROVINCE_CLASS_QID     <- "Q1062593" # province of Bolivia
MUNICIPALITY_CLASS_QID <- "Q1062710" # municipality of Bolivia
```

## Step 1: Fetch current Wikidata municipality data

A single `get_wikidata_instances()` call retrieves every Bolivian municipality item, collecting both the INE code (P14142) used for QID mapping and any existing P1082 (population) claims needed to generate the removal statements in the next step.

```{r fetch-wd-municipalities}
municipalities_wd <- get_wikidata_instances(
  MUNICIPALITY_CLASS_QID,
  property                    = "P14142",
  property_names              = "ine_code",
  numeric_list_properties     = "P1082",
  numeric_list_property_names = "population",
  country                     = BOLIVIA_QID
)

cat("Municipality items retrieved:", nrow(municipalities_wd), "\n")
cat("Items with INE code:         ", sum(!is.na(municipalities_wd$ine_code)), "\n")
cat("Items with P1082:            ", sum(municipalities_wd$population_n > 0, na.rm = TRUE), "\n")
```

## Step 2: Remove pre-existing P1082 claims (municipalities only)

Seven items carry one or more P1082 statements that were added before this structured import. They must be deleted first, or QuickStatements will create duplicate claims.

`make_p1082_removals()` extracts the current claim values from the already-fetched `municipalities_wd` tibble (no second API call needed) and emits one removal line per claim in QuickStatements batch syntax (`-QID | P1082 | value`).

```{r fn-make-removals}
#' Generate QuickStatements removal lines for all P1082 claims on a set of items.
#'
#' @param municipalities_wd Tibble returned by get_wikidata_instances(), which must
#'   include population_n, population_1 … population_N columns (numeric list
#'   property columns produced when numeric_list_properties = "P1082").
#' @param target_qids Character vector of QIDs whose P1082 claims should be removed.
#'
#' @return A tibble with columns `qid`, `label_es`, `claim_slot`, `value`, and
#'   `removal_statement` (the QuickStatements deletion line).
make_p1082_removals <- function(municipalities_wd, target_qids) {
  df <- municipalities_wd |>
    filter(qid %in% target_qids)

  if (nrow(df) == 0) stop("None of target_qids found in municipalities_wd.")

  # Identify the per-claim value columns: population_1, population_2, ...
  # (exclude population, population_n, population_*_year, population_*_ref)
  value_cols <- names(df) |>
    str_subset("^population_\\d+$")

  df |>
    select(qid, label_es, all_of(value_cols)) |>
    pivot_longer(
      cols      = all_of(value_cols),
      names_to  = "claim_slot",
      values_to = "value"
    ) |>
    filter(!is.na(value)) |>
    mutate(
      value_str         = as.character(as.integer(value)),
      removal_statement = str_c("-", qid, " | P1082 | ", value_str)
    ) |>
    select(qid, label_es, claim_slot, value, removal_statement)
}
```

```{r make-removals}
qids_to_clear <- c(
  "Q628665",  # Porco
  "Q1516972", # Pojo
  "Q1814603", # Loreto
  "Q1814613", # Exaltación
  "Q1814898", # San Buenaventura
  "Q1814929", # Corocoro
  "Q1952982"  # (label missing in WD)
)

removals <- make_p1082_removals(municipalities_wd, qids_to_clear)
removals
```

Copy the lines below into QuickStatements batch mode and run them **before** proceeding to Step 4.

```{r print-removals}
writeLines(removals$removal_statement)
```

## Step 3: Build municipality–QID population table

The join chain:

1.  **pop_muni → muni_id_lookup_table** on `municipio == muni_cpv2024` and `departamento == department`. The department key is required because many municipality names (e.g. "San Pedro", "Santa Rosa") are shared across departments; omitting it inflates the row count from 340 to 369.
2.  **muni_id_lookup_table → Wikidata** on `id_muni == ine_code` using the `municipalities_wd` tibble already in hand.

TIOC rows in `pop_muni` (e.g. "TIOC-Raqaypampa", "TIOC-Territorio Indígena Multiétnico") have no matching `muni_cpv2024` in the lookup table and will produce `NA` after the join; they are dropped and reported to the console.

```{r fn-build-pop-table}
#' Join pop_muni to Wikidata QIDs through muni_id_lookup_table.
#'
#' @param pop_muni    Tibble of municipality-level population rows from CPV 2024.
#' @param lookup      muni_id_lookup_table with columns id_muni, muni_cpv2024,
#'   and department.
#' @param ine_qid_map Tibble with columns id_muni (INE 6-digit code) and qid.
#'   Must have NA rows filtered out before calling (to avoid spurious TIOC matches).
#'
#' @return A tibble of matched rows, one per municipality, with all pop_muni
#'   columns plus id_muni and qid. Unmatched rows are dropped and messaged.
build_population_qs_table <- function(pop_muni, lookup, ine_qid_map) {
  ine_qid_map <- ine_qid_map |> filter(!is.na(id_muni))

  joined <- pop_muni |>
    left_join(
      lookup |> select(id_muni, muni_cpv2024, department),
      by = c("municipio" = "muni_cpv2024", "departamento" = "department")
    ) |>
    left_join(ine_qid_map, by = "id_muni")

  unmatched <- joined |> filter(is.na(qid))
  if (nrow(unmatched) > 0) {
    message(nrow(unmatched), " pop_muni row(s) had no QID match and will be dropped:")
    walk(unmatched$municipio, ~ message("  - ", .x))
  }

  joined |> filter(!is.na(qid))
}
```

```{r build-pop-table}
ine_qid_map <- municipalities_wd |>
  select(qid, id_muni = ine_code)

pop_qs <- build_population_qs_table(pop_muni, muni_id_lookup_table, ine_qid_map)

cat("Rows in pop_qs:", nrow(pop_qs), "\n")
cat("Distinct QIDs: ", n_distinct(pop_qs$qid), "\n")

pop_qs |>
  select(departamento, municipio, id_muni, qid,
         pob_2001_total, pob_2012_total, pob_2024_total) |>
  head(10)
```

## Step 4: Fetch Wikidata departments and provinces (with INE codes)

Unlike municipalities, we can map provinces and departments directly via the INE code (P14142), which is now present on Wikidata for these items.

- Departments: 2-digit INE code, matching `cod.dep` from the INE community classification file.
- Provinces: 4-digit INE code (`DEP` + `PRO`), matching `cod.prov`.

```{r fetch-wd-departments-provinces}
# departments_wd <- get_wikidata_instances(
#   DEPARTMENT_CLASS_QID,
#   property       = "P14142",
#   property_names = "ine_code",
#   country        = BOLIVIA_QID,
#   numeric_list_properties      = "P1082",
#   numeric_list_property_names  = "population"
# )

departments_wd <- readRDS("data/wikiadata-bo-departments.RDS")

department_ine_lookup <- departments_wd |> 
                           filter(!is.na(ine_code)) |> 
                           mutate(department=str_replace(label_en, " Department", ""),
                                  dep_qid = qid) |>
                           select(ine_code, department, dep_qid) |> 
                           arrange(ine_code)



# provinces_wd <- get_wikidata_instances(
#   PROVINCE_CLASS_QID,
#   property = c("P131", "P17", "P14142"),
#   property_names = c("located_in", "country", "ine_code"),
#   numeric_list_properties      = "P1082",
#   numeric_list_property_names  = "population"
# )
# 
# provinces_wd <- provinces_wd |> 
#   mutate(dep_ine = str_sub(ine_code, 1, 2)) |>
#   left_join(department_ine_lookup, by=join_by(dep_ine == ine_code))

provinces_wd <- readRDS("data/wikiadata-bo-provinces.RDS")

cat("Department items retrieved:", nrow(departments_wd), "\n")
cat("Departments with INE code: ", sum(!is.na(departments_wd$ine_code)), "\n")
cat("Province items retrieved:  ", nrow(provinces_wd), "\n")
cat("Provinces with INE code:   ", sum(!is.na(provinces_wd$ine_code)), "\n")
```

Wikidata already has some values for population. Let's examine them. All nine departments have two values for population. Most provinces have no population data.

```{r existing-population-data}
departments_wd <- departments_wd |> filter(!is.na(ine_code)) # filter out Litoral
     # One Bolivian department has not been controlled by Bolivia since the 19th century
     # It has no INE code.

departments_wd |> count(population_n) 

provinces_wd |> count(population_n)

departments_wd |> 
  select(qid, label_en, ine_code, population_1, population_1_year, population_1_ref,
         population_2, population_2_year, population_2_ref)

provinces_wd |> filter(population_n > 0) |> 
  select(qid, label_en, ine_code, population_1, population_1_year, 
                                  population_2, population_2_year)

province_pop_qids <- provinces_wd |> filter(population_n > 0) |> pull(qid)

```

Looking at the results we see our task: We will have to double-check departmental population values against our new sources and add infromation for 2001. We can clear the five undated data points for province populations.

```{r remove province-population-data}
qids_to_clear <- province_pop_qids
removals_prov <- make_p1082_removals(provinces_wd, qids_to_clear)
```

Copy the lines below into QuickStatements batch mode and run them **before** proceeding to Step 4.

```{r print-removals-2}
writeLines(removals_prov$removal_statement)
```

## Step 5: Build province/department population tables via INE codes

The CPV 2024 population extracts already include department/province names. We add INE codes as follows:

- **Departments:** take the first two digits of `row_num`-independent codes by grouping on `departamento` and joining to Wikidata via department INE code. (In practice `pop_dept` already has one row per department.)

- **Provinces:** take the first four digits of the 6-digit municipality code. Because `pop_prov` does not contain a province code, we compute it from the department/province order using the INE code on Wikidata: we join on `(departamento, provincia)` after extracting a clean province label from WD.

Note: Wikidata province labels are not perfectly standardized (e.g. multiple "Cercado" provinces). We therefore match primarily by INE code (P14142). Where that is missing, we fall back to a conservative name+department join and report unmatched rows.


```{r build-dept-pop-table}
# pop_dept has one row per department; create dept INE code = 2 digits.
# We derive it from the province/municipality-style codes when available, but
# for CPV 2024 tables we only need the mapping through Wikidata.
# Here we join by department Spanish label as a fallback if needed.

# Prefer matching by INE code if pop_dept already has it (it currently doesn't),
# otherwise match by department name against WD Spanish label without prefix.

wd_dept_name <- departments_wd |>
  transmute(
    qid,
    ine_code,
    dept_name = label_es |>
      str_remove("^Departamento de(l )?\\s*") |>
      str_replace("Potosí", "Potosi")
  )

# pop_dept_qs <- pop_dept |>
#   left_join(wd_dept_name, join_by(id_dep==ine_code)) |> 
#   relocate(id_dep, departamento, qid)

pop_dept_qs <- pop_dept |>
  left_join(departments_wd, join_by(id_dep==ine_code)) |> 
  relocate(id_dep, departamento, qid)

unmatched <- pop_dept_qs |> filter(is.na(qid))
if (nrow(unmatched) > 0) {
  message(nrow(unmatched), " department row(s) had no QID match:")
  walk(unmatched$departamento, ~ message("  - ", .x))
}

pop_dept_qs <- pop_dept_qs |> filter(!is.na(qid))
```

```{r build-province-pop-table}
# Province mapping strategy:
# 1) If Wikidata has INE code (P14142) for a province item, use it.
# 2) Compute province code by joining pop_prov to a lookup built from WD labels.

wd_prov_lookup <- provinces_wd |>
  transmute(
    qid,
    ine_code,
    department = label_es |>
      str_extract("\\(([^)]+)\\)$") |>
      str_remove_all("[()]") |>
      str_replace("Potosí", "Potosi"),
    province = label_es |>
      str_remove("\\s*\\([^)]+\\)$") |>
      str_remove("^Provincia de\\s+") |>
      str_remove("^Cercado$")
  )

# Some province items may not have "(Department)" disambiguation.
# Fall back to located-in labels when present would be better, but this QMD
# keeps it simple: we primarily rely on INE codes being present.

pop_prov_qs <- pop_prov |>
  mutate(
    department = departamento |> str_trim() |> str_replace("Potosí", "Potosi"),
    province   = provincia    |> str_trim()
  ) |>
  # join on (province, department) to recover province QID
  left_join(wd_prov_lookup |> select(qid, province, department, ine_code),
            by = c("province" = "province", "department" = "department"))

pop_prov_qs <- pop_prov |>
  left_join(provinces_wd, join_by(ine_code)) |> 
  relocate(ine_code, provincia, qid)

unmatched_prov <- pop_prov_qs |> filter(is.na(qid))
if (nrow(unmatched_prov) > 0) {
  message(nrow(unmatched_prov), " province row(s) had no QID match and will be dropped (inspect manually):")
  walk(head(unmatched_prov$provincia, 50), ~ message("  - ", .x))
}

pop_prov_qs <- pop_prov_qs |> filter(!is.na(qid))
```

## Step 6: Compare existing and new population data in departments

Looking at the existing population data on Wikidata for departments, we see the 
most recent year is 2024 (the census year) for all nine. The one older year
is 2012 for seven departments. These are the values that might match up.

```{r compare-dept-population}
pop_dept_qs |> 
  count(population_1_year)
pop_dept_qs |> 
  count(population_2_year)

pop_dept_qs |> 
  filter(population_1_year == 2024) |>
  select(1:3, pob_2024_total, population_1, population_1_year, population_1_ref) |>
  mutate(match = pob_2024_total == population_1)  |> 
  relocate(match, .after = qid)

pop_dept_qs |> 
  filter(population_2_year == 2012) |>
  select(1:3, pob_2012_total, population_2, population_2_year, population_2_ref) |> 
  mutate(match = pob_2012_total == population_2) |> 
  relocate(match, .after = qid)

pop_dept_qs |> 
  filter(population_2_year != 2012) |>
  select(1:3,  population_2, population_2_year, population_2_ref) 
```
The 2024 data are close, but unequal in all values. These values were initial releases by the Census and we will delete them to update them.

The 2012 data match in two cases, but none of theme are cited. We'll delete all of these as well and replace them all with new, cited information.

Finally, we have two data points, without source, for 2016 and 2020. Since we are adding authoritative numbers for 2012 and 2024, we will delete these as well to avoid confusion.

Copy the lines below into QuickStatements batch mode and run them **before** proceeding to Step 7.

```{r make-dept-removals}
qids_to_clear <- departments_wd$qid
removals_dep <- make_p1082_removals(departments_wd, qids_to_clear)
writeLines(removals_dep$removal_statement)
```


## Step 7: Generate P1082 QuickStatements (municipalities, provinces, departments)

One QuickStatements column per census year is produced using `add_quick_statement_column_q()`.

Each statement carries:

- a P585 (point in time) qualifier at year-level precision (`/9`);
- S248 sourcing to **Q139668018** (CPV 2024).

```{r fn-add-qs-col}
add_quick_statement_column_q <- function(dataframe, qid_col, property, value_col, ...){
  dataframe %>%
    rowwise() %>%
    mutate(quick_statement = create_quick_statement({{ qid_col }}, property, {{ value_col }}, ...)) %>%
    ungroup()
}
```

```{r generate-qs-muni}
pop_qs_with_statements_muni <- pop_qs |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2001_total,
    qualifiers    = list(P585 = p585_for_year(2001)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2001 = quick_statement) |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2012_total,
    qualifiers    = list(P585 = p585_for_year(2012)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2012 = quick_statement) |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2024_total,
    qualifiers    = list(P585 = p585_for_year(2024)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2024 = quick_statement)
```

```{r generate-qs-dept}
pop_qs_with_statements_dept <- pop_dept_qs |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2001_total,
    qualifiers    = list(P585 = p585_for_year(2001)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2001 = quick_statement) |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2012_total,
    qualifiers    = list(P585 = p585_for_year(2012)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2012 = quick_statement) |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2024_total,
    qualifiers    = list(P585 = p585_for_year(2024)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2024 = quick_statement)
```

```{r generate-qs-province}
pop_qs_with_statements_prov <- pop_prov_qs |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2001_total,
    qualifiers    = list(P585 = p585_for_year(2001)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2001 = quick_statement) |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2012_total,
    qualifiers    = list(P585 = p585_for_year(2012)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2012 = quick_statement) |>
  add_quick_statement_column_q(
    qid, "P1082", pob_2024_total,
    qualifiers    = list(P585 = p585_for_year(2024)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2024 = quick_statement)
```

```{r stack-statements}
muni_statements <- 
  pop_qs_with_statements_muni |> select(qid, qs_2001, qs_2012, qs_2024) |>
  distinct() |>
  arrange(qid) |>
  pivot_longer(cols = starts_with("qs_"), values_to = "statement") |>
  pull(statement)

other_statements <- bind_rows(
  pop_qs_with_statements_prov |> select(qid, qs_2001, qs_2012, qs_2024),
  pop_qs_with_statements_dept |> select(qid, qs_2001, qs_2012, qs_2024)
) |>
  distinct() |>
  arrange(qid) |>
  pivot_longer(cols = starts_with("qs_"), values_to = "statement") |>
  pull(statement)

cat("Total QuickStatements lines:", length(muni_statements)+length(other_statements), "\n")
```

## Step 7: Review and export

Preview the first 15 lines to sanity-check the format before upload.

```{r preview-statements}
writeLines(head(other_statements, 15))
```

Export to a plain-text file for pasting into the QuickStatements batch interface at <https://qs-dev.toolforge.org/>.

```{r export}
#| eval: false
# writeLines(all_statements, "data/cpv2024_population_quickstatements.txt")
writeLines(muni_statements, "data/cpv2024_population_quickstatements_municipalities.txt")
writeLines(other_statements, "data/cpv2024_population_quickstatements_dept_prov.txt")

# cat("Written to data/cpv2024_population_quickstatements.txt\n")
cat("Written municipality statements to data/cpv2024_population_quickstatements_municipalities.txt\n")
cat("Written department/province statements to data/cpv2024_population_quickstatements_dept_prov.txt\n")
```

The expected format of each line is:

```         
QID | P1082 | <count> | P585 | +YYYY-01-01T00:00:00Z/9 | S248 | Q139668018 | S813 | +YYYY-MM-DDT00:00:00Z/11
```

Run the removal batch (Step 2) first, then submit the export file.
