---
title: "Bolivia CPV 2024: Population QuickStatements for Wikidata"
format: html
---

## Overview

This document adds P1082 (population) statements to Wikidata items for Bolivia's 339
municipalities using census counts from three surveys: 2001, 2012, and 2024 (CPV 2024).
Each statement is sourced to Q139668018 (the Wikidata item for the CPV 2024 publication)
and carries a P585 (point in time) qualifier for the relevant census year.

Before adding new statements, a small number of pre-existing P1082 claims on seven items
must be removed. Those items were tagged during a prior data-cleaning pass; their existing
values are undated or unreferenced and would otherwise create duplicates.

## Setup

```{r setup}
library(tidyverse)
library(readr)
library(httr)
library(jsonlite)

source("create_quick_statement.R")
source("get_wikidata_instances.R")
```

```{r load-data}
muni_id_lookup_table <- readRDS("../ultimate-consequences/data/muni_id_lookup_table.rds")
pop_muni             <- readRDS("data/poblacion_municipio.rds")
```

## Step 1: Fetch current Wikidata municipality data

A single `get_wikidata_instances()` call retrieves every Bolivian municipality item,
collecting both the INE code (P14142) used for QID mapping and any existing P1082
(population) claims needed to generate the removal statements in the next step.

```{r fetch-wd-municipalities}
municipalities_wd <- get_wikidata_instances(
  "Q1062710",
  property                    = "P14142",
  property_names              = "ine_code",
  numeric_list_properties     = "P1082",
  numeric_list_property_names = "population"
)

cat("Municipality items retrieved:", nrow(municipalities_wd), "\n")
cat("Items with INE code:         ", sum(!is.na(municipalities_wd$ine_code)), "\n")
cat("Items with P1082:            ", sum(municipalities_wd$population_n > 0, na.rm = TRUE), "\n")
```

## Step 2: Remove pre-existing P1082 claims

Seven items carry one or more P1082 statements that were added before this
structured import. They must be deleted first, or QuickStatements will create
duplicate claims.

`make_p1082_removals()` extracts the current claim values from the already-fetched
`municipalities_wd` tibble (no second API call needed) and emits one removal line
per claim in QuickStatements batch syntax (`-QID | P1082 | value`).

```{r fn-make-removals}
#' Generate QuickStatements removal lines for all P1082 claims on a set of items.
#'
#' @param municipalities_wd Tibble returned by get_wikidata_instances(), which must
#'   include population_n, population_1 … population_N columns (numeric list
#'   property columns produced when numeric_list_properties = "P1082").
#' @param target_qids Character vector of QIDs whose P1082 claims should be removed.
#'
#' @return A tibble with columns `qid`, `label_es`, `claim_slot`, `value`, and
#'   `removal_statement` (the QuickStatements deletion line).
make_p1082_removals <- function(municipalities_wd, target_qids) {
  df <- municipalities_wd |>
    filter(qid %in% target_qids)

  if (nrow(df) == 0) stop("None of target_qids found in municipalities_wd.")

  # Identify the per-claim value columns: population_1, population_2, ...
  # (exclude population, population_n, population_*_year, population_*_ref)
  value_cols <- names(df) |>
    str_subset("^population_\\d+$")

  df |>
    select(qid, label_es, all_of(value_cols)) |>
    pivot_longer(
      cols      = all_of(value_cols),
      names_to  = "claim_slot",
      values_to = "value"
    ) |>
    filter(!is.na(value)) |>
    mutate(
      value_str         = as.character(as.integer(value)),
      removal_statement = str_c("-", qid, " | P1082 | ", value_str)
    ) |>
    select(qid, label_es, claim_slot, value, removal_statement)
}
```

```{r make-removals}
qids_to_clear <- c(
  "Q628665",  # Porco
  "Q1516972", # Pojo
  "Q1814603", # Loreto
  "Q1814613", # Exaltación
  "Q1814898", # San Buenaventura
  "Q1814929", # Corocoro
  "Q1952982"  # (label missing in WD)
)

removals <- make_p1082_removals(municipalities_wd, qids_to_clear)
removals
```

Copy the lines below into QuickStatements batch mode and run them **before** proceeding
to Step 4.

```{r print-removals}
writeLines(removals$removal_statement)
```

## Step 3: Build the municipality–QID population table

The join chain:

1. **pop_muni → muni_id_lookup_table** on `municipio == muni_cpv2024` and
   `departamento == department`. The department key is required because many
   municipality names (e.g. "San Pedro", "Santa Rosa") are shared across departments;
   omitting it inflates the row count from 340 to 369.
2. **muni_id_lookup_table → Wikidata** on `id_muni == ine_code` using the
   `municipalities_wd` tibble already in hand.

TIOC rows in `pop_muni` (e.g. "TIOC-Raqaypampa", "TIOC-Territorio Indígena
Multiétnico") have no matching `muni_cpv2024` in the lookup table and will produce
`NA` after the join; they are dropped and reported to the console.

```{r fn-build-pop-table}
#' Join pop_muni to Wikidata QIDs through muni_id_lookup_table.
#'
#' @param pop_muni    Tibble of municipality-level population rows from CPV 2024.
#' @param lookup      muni_id_lookup_table with columns id_muni, muni_cpv2024,
#'   and department.
#' @param ine_qid_map Tibble with columns id_muni (INE 6-digit code) and qid.
#'   Must have NA rows filtered out before calling (to avoid spurious TIOC matches).
#'
#' @return A tibble of matched rows, one per municipality, with all pop_muni
#'   columns plus id_muni and qid. Unmatched rows are dropped and messaged.
build_population_qs_table <- function(pop_muni, lookup, ine_qid_map) {
  ine_qid_map <- ine_qid_map |> filter(!is.na(id_muni))

  joined <- pop_muni |>
    left_join(
      lookup |> select(id_muni, muni_cpv2024, department),
      by = c("municipio" = "muni_cpv2024", "departamento" = "department")
    ) |>
    left_join(ine_qid_map, by = "id_muni")

  unmatched <- joined |> filter(is.na(qid))
  if (nrow(unmatched) > 0) {
    message(nrow(unmatched), " pop_muni row(s) had no QID match and will be dropped:")
    walk(unmatched$municipio, ~ message("  - ", .x))
  }

  joined |> filter(!is.na(qid))
}
```

```{r build-pop-table}
ine_qid_map <- municipalities_wd |>
  select(qid, id_muni = ine_code)

pop_qs <- build_population_qs_table(pop_muni, muni_id_lookup_table, ine_qid_map)

cat("Rows in pop_qs:", nrow(pop_qs), "\n")
cat("Distinct QIDs: ", n_distinct(pop_qs$qid), "\n")

pop_qs |>
  select(departamento, municipio, id_muni, qid,
         pop_2001_total, pop_2012_total, pop_2024_total) |>
  head(10)
```

## Step 4: Generate P1082 QuickStatements

One QuickStatements column per census year is produced using
`add_quick_statement_column_q()`. Each statement carries:

- a P585 (point in time) qualifier at year-level precision (`/9`);
- S248 sourcing to **Q139668018** (CPV 2024).

The three columns are stacked into a single output file ready for batch upload.

```{r fn-year-qualifier}
p585_for_year <- function(year) sprintf("+%d-01-01T00:00:00Z/9", year)
SOURCE_QID <- "Q139668018"
```

```{r generate-qs}
pop_qs_with_statements <- pop_qs |>

  add_quick_statement_column_q(
    qid, "P1082", pop_2001_total,
    qualifiers    = list(P585 = p585_for_year(2001)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2001 = quick_statement) |>

  add_quick_statement_column_q(
    qid, "P1082", pop_2012_total,
    qualifiers    = list(P585 = p585_for_year(2012)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2012 = quick_statement) |>

  add_quick_statement_column_q(
    qid, "P1082", pop_2024_total,
    qualifiers    = list(P585 = p585_for_year(2024)),
    type          = "quantity",
    reference_qid = SOURCE_QID
  ) |>
  rename(qs_2024 = quick_statement)

pop_qs_with_statements |>
  select(municipio, qid, qs_2001, qs_2012, qs_2024) |>
  head(5)
```

```{r stack-statements}
all_statements <- pop_qs_with_statements |>
  arrange(qid) |>
  select(qid, qs_2001, qs_2012, qs_2024) |>
  pivot_longer(cols = starts_with("qs_"), values_to = "statement") |>
  pull(statement)

cat("Total QuickStatements lines:", length(all_statements), "\n")
```

## Step 5: Review and export

Preview the first 15 lines to sanity-check the format before upload.

```{r preview-statements}
writeLines(head(all_statements, 15))
```

Export to a plain-text file for pasting into the QuickStatements batch interface at
<https://quickstatements.toolforge.org/>.

```{r export}
writeLines(all_statements, "data/cpv2024_population_quickstatements.txt")
cat("Written to data/cpv2024_population_quickstatements.txt\n")
```

The expected format of each line is:

```
QID | P1082 | <count> | qal585 | +YYYY-01-01T00:00:00Z/9 | S248 | Q139668018 | S813 | +YYYY-MM-DDT00:00:00Z/11
```

Run the removal batch (Step 2) first, then submit this file.
