---
title: "Bolivia 2025 General Election — Results by Geography"
author: "JCR Björk"
date: "2026-03-15"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 8
execute:
  warning: false
  message: false
  freeze: auto
---

```{r setup}
library(tidyverse)
library(sf)
library(patchwork)
```

## Data preparation

### Load raw election data and rename party columns

```{r load-data}
# Column rename map (OEP camelCase → snake_case)
geo_renames <- c(
  mesa_code   = "CodigoMesa",
  mesa_number = "NumeroMesa",
  registered  = "InscritosHabilitados",
  country_code = "CodigoPais",
  country_name = "NombrePais",
  dept_code   = "CodigoDepartamento",
  dept_name   = "NombreDepartamento",
  prov_code   = "CodigoProvincia",
  prov_name   = "NombreProvincia",
  mun_code    = "CodigoSeccion",
  mun_name    = "NombreMunicipio",
  localidad_code = "CodigoLocalidad",
  localidad_name = "NombreLocalidad",
  recinto_code   = "CodigoRecinto",
  recinto_name   = "NombreRecinto"
)

vote_renames <- c(
  valid_votes = "VotoValido",
  blank_votes = "VotoBlanco",
  null_votes  = "TotalVotoNulo",
  votes_cast  = "VotoEmitido"
)

# --- First round (all three races) -------------------------------------------

raw_1v <- read_csv(
  "data-raw/EG2025_20250824_235619_5976286028509320003.csv",
  show_col_types = FALSE
) |>
  filter(NombrePais == "BOLIVIA")

# Ballot slot → party mapping (confirmed against OEP separata)
party_renames_1v <- c(
  AP         = "Voto1",
  LyP_ADN    = "Voto2",
  APB_SUMATE = "Voto3",
  LIBRE      = "Voto5",
  FP         = "Voto6",
  MAS_IPSP   = "Voto7",
  UNIDAD     = "Voto9",
  PDC        = "Voto10"
)

elections_1v <- raw_1v |>
  rename(!!!party_renames_1v) |>
  rename(race = Descripcion) |>
  rename(!!!geo_renames) |>
  rename(!!!vote_renames)

parties_main <- c("AP", "LyP_ADN", "APB_SUMATE", "LIBRE", "FP", "MAS_IPSP", "UNIDAD", "PDC")
parties_especial <- c(parties_main, "Voto11", "Voto12")

# --- Second round (presidential only) ----------------------------------------

raw_2v <- read_csv(
  "data-raw/EG2025_2v_20251026_235911_6311285959951043675.csv",
  show_col_types = FALSE
) |>
  filter(NombrePais == "BOLIVIA")

elections_2v <- raw_2v |>
  rename(!!!geo_renames) |>
  rename(!!!vote_renames)

parties_2v <- c("PDC", "LIBRE")
```

### Load GADM boundary files and lookup tables

```{r load-gadm}
muni_lookup <- readRDS("data/muni_id_lookup_table.rds")
prov_lookup <- readRDS("data/prov_id_lookup_table.rds")

gadm1 <- st_read("data/gadm41_BOL_3.gpkg", layer = "ADM_ADM_1", quiet = TRUE) |>
  select(NAME_1, geom)
gadm2 <- st_read("data/gadm41_BOL_3.gpkg", layer = "ADM_ADM_2", quiet = TRUE) |>
  select(NAME_1, NAME_2, geom)
gadm3 <- st_read("data/gadm41_BOL_3.gpkg", layer = "ADM_ADM_3", quiet = TRUE) |>
  select(NAME_1, NAME_2, NAME_3, geom)
```

### Build OEP → INE → GADM crosswalk

OEP geographic codes map directly to INE codes: `id_prov` = `sprintf("%02d%02d",
dept, prov)` and `id_muni` = `sprintf("%02d%02d%02d", dept, prov, seccion)`.
We use the pre-built lookup tables (`prov_id_lookup_table`, `muni_id_lookup_table`)
to get the GADM name variants for spatial joins. TIOC/AIOC indigenous
autonomies created after the GADM vintage will not match.

```{r crosswalk}
dept_to_gadm <- tibble(
  dept_code = 1:9,
  dept_gadm = c("Chuquisaca", "La Paz", "Cochabamba", "Oruro", "Potosí",
                "Tarija", "Santa Cruz", "Beni", "Pando")
)

prov_to_gadm <- prov_lookup |>
  select(id_prov, province_gadm) |>
  mutate(dept_code = as.integer(substr(id_prov, 1, 2)),
         prov_code = as.integer(substr(id_prov, 3, 4))) |>
  left_join(dept_to_gadm, by = "dept_code")

muni_to_gadm <- muni_lookup |>
  select(id_muni, muni_gadm, department) |>
  mutate(dept_code = as.integer(substr(id_muni, 1, 2)),
         prov_code = as.integer(substr(id_muni, 3, 4)),
         mun_code  = as.integer(substr(id_muni, 5, 6))) |>
  left_join(dept_to_gadm, by = "dept_code")

cat("Province lookup:", nrow(prov_to_gadm), "rows,",
    sum(is.na(prov_to_gadm$province_gadm)), "missing GADM name\n")
cat("Municipality lookup:", nrow(muni_to_gadm), "rows,",
    sum(is.na(muni_to_gadm$muni_gadm)), "missing GADM name\n")
```

### Aggregation functions

```{r agg-functions}
aggregate_results <- function(data, ..., parties) {
  data |>
    group_by(...) |>
    summarise(
      mesas = n(),
      registered = sum(registered),
      votes_cast = sum(votes_cast),
      valid_votes = sum(valid_votes),
      null_votes = sum(null_votes),
      blank_votes = sum(blank_votes),
      across(all_of(parties), sum),
      .groups = "drop"
    ) |>
    mutate(
      turnout_pct = votes_cast / registered * 100,
      null_pct = null_votes / votes_cast * 100,
      across(all_of(parties), ~ . / valid_votes * 100, .names = "{.col}_pct")
    )
}
```

## First-round presidential results {#sec-presidente}

### Department level

```{r pres-dept}
pres_1v <- elections_1v |> filter(race == "PRESIDENTE")

pres_dept <- aggregate_results(pres_1v, dept_code, dept_name,
                               parties = parties_main)
pres_dept |>
  select(dept_name, mesas, registered, turnout_pct, null_pct,
         all_of(paste0(parties_main, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(registered))
```

### Province level

```{r pres-prov}
pres_prov <- aggregate_results(
  pres_1v,
  dept_code, dept_name,
  prov_code, prov_name,
  parties = parties_main
)

pres_prov |>
  select(dept_name, prov_name, valid_votes,
         all_of(paste0(parties_main, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(dept_name, desc(valid_votes)) |>
  head(20)
```

### Municipality level

```{r pres-mun}
pres_mun <- aggregate_results(
  pres_1v,
  dept_code, dept_name,
  prov_code, prov_name,
  mun_code, mun_name,
  parties = parties_main
)

pres_mun |>
  select(dept_name, mun_name, valid_votes,
         all_of(paste0(parties_main, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(valid_votes)) |>
  head(20)
```

### Presidential choropleths

```{r pres-choropleths}
#| fig-height: 40
#| fig-width: 10

# --- Choropleth helpers -------------------------------------------------------
party_colors <- c(
  AP = "#3B8CC4", LyP_ADN = "#1A1A1A", APB_SUMATE = "#6B2D8B",
  LIBRE = "#D42E2E", FP = "#A8DCF0", MAS_IPSP = "#1B3D8F",
  UNIDAD = "#E8A838", PDC = "#2E7D7B",
  Voto11 = "#7B5B3A", Voto12 = "#5A7B3A"
)

party_labels <- c(
  AP = "AP", LyP_ADN = "LyP ADN", APB_SUMATE = "APB Súmate",
  LIBRE = "LIBRE", FP = "FP", MAS_IPSP = "MAS-IPSP",
  UNIDAD = "UNIDAD", PDC = "PDC",
  Voto11 = "Voto 11", Voto12 = "Voto 12"
)

make_choropleth <- function(geo_sf, data, join_cols, pct_col, title,
                            fill_color = "#2E7D7B") {
  df_joined <- geo_sf |>
    left_join(data, by = join_cols)

  ggplot(df_joined) +
    geom_sf(aes(fill = .data[[pct_col]]), color = "white", linewidth = 0.15) +
    scale_fill_gradient(low = "white", high = fill_color,
                        name = "% vote share", na.value = "grey85",
                        limits = c(0, NA)) +
    labs(title = title) +
    theme_void(base_size = 9) +
    theme(legend.position = "bottom",
          legend.key.width = unit(1.2, "cm"),
          plot.title = element_text(face = "bold", size = 10))
}

choropleth_panel <- function(agg_dept, agg_prov, agg_mun,
                             parties,
                             race_label = "Presidential") {
  dept_joined <- agg_dept |>
    left_join(dept_to_gadm, by = "dept_code")

  prov_joined <- agg_prov |>
    left_join(prov_to_gadm, by = c("dept_code", "prov_code"))

  mun_joined <- agg_mun |>
    left_join(muni_to_gadm, by = c("dept_code", "prov_code", "mun_code"))

  plots <- list()

  for (party in parties) {
    pct <- paste0(party, "_pct")
    color <- party_colors[[party]] %||% "#555555"
    label <- party_labels[[party]] %||% party

    # Shared scale: 0 to max across all three geographic levels
    pct_max <- max(
      c(dept_joined[[pct]], prov_joined[[pct]], mun_joined[[pct]]),
      na.rm = TRUE
    )
    shared_limits <- c(0, pct_max)

    p_dept <- make_choropleth(
      gadm1, dept_joined,
      c("NAME_1" = "dept_gadm"),
      pct, paste0(label, " — Department"), color
    ) + scale_fill_gradient(low = "white", high = color, name = "%",
                            na.value = "grey85", limits = shared_limits)

    p_prov <- make_choropleth(
      gadm2, prov_joined,
      c("NAME_1" = "dept_gadm", "NAME_2" = "province_gadm"),
      pct, paste0(label, " — Province"), color
    ) + scale_fill_gradient(low = "white", high = color, name = "%",
                            na.value = "grey85", limits = shared_limits)

    p_mun <- make_choropleth(
      gadm3, mun_joined,
      c("NAME_1" = "dept_gadm", "NAME_3" = "muni_gadm"),
      pct, paste0(label, " — Municipality"), color
    ) + scale_fill_gradient(low = "white", high = color, name = "%",
                            na.value = "grey85", limits = shared_limits)

    plots[[party]] <- p_dept + p_prov + p_mun +
      plot_layout(ncol = 3, widths = c(1, 1, 1))
  }

  wrap_plots(plots, ncol = 1) +
    plot_annotation(title = paste(race_label, "vote share by party"),
                    theme = theme(plot.title = element_text(face = "bold", size = 14)))
}

choropleth_panel(pres_dept, pres_prov, pres_mun,
                 parties_main, "Presidential (first round)")
```

## Diputado Uninominal results {#sec-uninominal}

### Department level

```{r uni-dept}
uni <- elections_1v |> filter(race == "DIPUTADO CIRCUNSCRIPCIÓN UNINOMINAL")

uni_dept <- aggregate_results(uni, dept_code, dept_name,
                              parties = parties_main)
uni_dept |>
  select(dept_name, mesas, registered, turnout_pct, null_pct,
         all_of(paste0(parties_main, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(registered))
```

### Province level

```{r uni-prov}
uni_prov <- aggregate_results(
  uni,
  dept_code, dept_name,
  prov_code, prov_name,
  parties = parties_main
)
```

### Municipality level

```{r uni-mun}
uni_mun <- aggregate_results(
  uni,
  dept_code, dept_name,
  prov_code, prov_name,
  mun_code, mun_name,
  parties = parties_main
)

uni_mun |>
  select(dept_name, mun_name, valid_votes,
         all_of(paste0(parties_main, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(valid_votes)) |>
  head(20)
```

### Uninominal choropleths

```{r uni-choropleths}
#| fig-height: 40
#| fig-width: 10

choropleth_panel(uni_dept, uni_prov, uni_mun,
                 parties_main, "Diputado Uninominal")
```

## Diputado Circunscripción Especial results {#sec-especial}

The 7 indigenous circunscripciones especiales cover specific regions. Voto11
and Voto12 are additional parties that only contest this race.

### Department level

```{r esp-dept}
esp <- elections_1v |> filter(race == "DIPUTADO CIRCUNSCRIPCIÓN ESPECIAL")

esp_dept <- aggregate_results(esp, dept_code, dept_name,
                              parties = parties_especial)
esp_dept |>
  select(dept_name, mesas, valid_votes,
         all_of(paste0(parties_especial, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(valid_votes))
```

### Province level

```{r esp-prov}
esp_prov <- aggregate_results(
  esp,
  dept_code, dept_name,
  prov_code, prov_name,
  parties = parties_especial
)
```

### Municipality level

```{r esp-mun}
esp_mun <- aggregate_results(
  esp,
  dept_code, dept_name,
  prov_code, prov_name,
  mun_code, mun_name,
  parties = parties_especial
)
```

### Especial choropleths

The especial race only covers municipalities within the 7 indigenous
circunscripciones — most of the country will appear grey.

```{r esp-choropleths}
#| fig-height: 50
#| fig-width: 10

choropleth_panel(esp_dept, esp_prov, esp_mun,
                 parties_especial, "Diputado Circunscripción Especial")
```

## Null vote rate {#sec-nulos}

The null vote rate was remarkably high nationally (19.9%), driven by
Cochabamba (33.3%). These maps show the geographic pattern.

```{r nulo-choropleths}
#| fig-height: 6
#| fig-width: 10

nulo_dept <- pres_dept |>
  left_join(dept_to_gadm, by = "dept_code")
nulo_prov <- pres_prov |>
  left_join(prov_to_gadm, by = c("dept_code", "prov_code"))
nulo_mun <- pres_mun |>
  left_join(muni_to_gadm, by = c("dept_code", "prov_code", "mun_code"))

p_nulo_dept <- make_choropleth(
  gadm1, nulo_dept,
  c("NAME_1" = "dept_gadm"),
  "null_pct", "Null vote % — Department", "#8B0000"
) + scale_fill_gradient(low = "white", high = "#8B0000", name = "%",
                        na.value = "grey85")

p_nulo_prov <- make_choropleth(
  gadm2, nulo_prov,
  c("NAME_1" = "dept_gadm", "NAME_2" = "province_gadm"),
  "null_pct", "Null vote % — Province", "#8B0000"
) + scale_fill_gradient(low = "white", high = "#8B0000", name = "%",
                        na.value = "grey85")

p_nulo_mun <- make_choropleth(
  gadm3, nulo_mun,
  c("NAME_1" = "dept_gadm", "NAME_3" = "muni_gadm"),
  "null_pct", "Null vote % — Municipality", "#8B0000"
) + scale_fill_gradient(low = "white", high = "#8B0000", name = "%",
                        na.value = "grey85")

p_nulo_dept + p_nulo_prov + p_nulo_mun +
  plot_layout(ncol = 3) +
  plot_annotation(title = "Presidential null vote rate (first round)",
                  theme = theme(plot.title = element_text(face = "bold", size = 14)))
```

## Vote in 30 largest municipalities {#sec-top30}

```{r top30-setup}
# Name mapping: display name → OEP election data name
top30_display <- c(
  "Santa Cruz de la Sierra", "El Alto", "La Paz", "Cochabamba",
  "Oruro", "Sucre", "Tarija", "Sacaba", "Potosí", "Quillacollo",
  "Warnes", "La Guardia", "Trinidad", "Montero", "Viacha", "Riberalta",
  "Cotoca", "Yacuiba", "Villa Tunari", "San Ignacio de Velasco",
  "Colcapirhua", "Puerto Villarroel", "Tiquipaya", "Caranavi",
  "Yapacaní", "Vinto", "Sipesipe", "El Torno", "Cobija", "Pailón"
)

oep_name_map <- c(
  "Santa Cruz de la Sierra" = "Santa Cruz de La Sierra",
  "La Paz" = "Nuestra Señora de La Paz",
  "San Ignacio de Velasco" = "San Ignacio"
)

top30_oep <- ifelse(top30_display %in% names(oep_name_map),
                    oep_name_map[top30_display],
                    top30_display)

top30_dept <- pres_mun |>
  filter(mun_name %in% top30_oep) |>
  # Disambiguate San Ignacio: de Velasco is in Santa Cruz (dept 7)
  filter(!(mun_name == "San Ignacio" & dept_code == 8)) |>
  select(dept_code, dept_name, mun_name) |>
  mutate(display_name = names(oep_name_map)[match(mun_name, oep_name_map)],
         display_name = coalesce(display_name, mun_name))

top30_ordered <- top30_dept |>
  mutate(orig_pos = match(display_name, top30_display)) |>
  arrange(dept_code, orig_pos)

group1_names <- top30_ordered$mun_name[1:15]
group2_names <- top30_ordered$mun_name[16:30]
group1_display <- top30_ordered$display_name[1:15]
group2_display <- top30_ordered$display_name[16:30]
group1_depts <- top30_ordered$dept_name[1:15]
group2_depts <- top30_ordered$dept_name[16:30]

pres_top30 <- pres_mun |>
  filter(mun_name %in% top30_oep) |>
  filter(!(mun_name == "San Ignacio" & dept_code == 8))

cat_colors <- c(
  setNames(party_colors[parties_main], party_labels[parties_main]),
  "Blank" = "grey70", "Null" = "grey40"
)

make_top30_chart <- function(data, oep_names, display_names, dept_names,
                             parties, subtitle) {
  df <- data |>
    filter(mun_name %in% oep_names) |>
    pivot_longer(cols = c(all_of(parties), null_votes, blank_votes),
                 names_to = "category", values_to = "votes") |>
    mutate(pct = votes / votes_cast * 100) |>
    mutate(
      category = recode(category, !!!party_labels,
                        null_votes = "Null", blank_votes = "Blank"),
      category = factor(category,
                        levels = c(rev(party_labels[parties]), "Blank", "Null")),
      display_name = names(oep_name_map)[match(mun_name, oep_name_map)],
      display_name = coalesce(display_name, mun_name)
    )

  label_lookup <- setNames(
    paste0(display_names, " (", dept_names, ")"),
    display_names
  )

  df <- df |>
    mutate(display_label = label_lookup[display_name],
           display_label = factor(display_label, levels = rev(label_lookup)))

  ggplot(df, aes(x = pct, y = display_label, fill = category)) +
    geom_col(position = "stack", width = 0.75) +
    scale_fill_manual(values = cat_colors, name = NULL) +
    scale_x_continuous(labels = \(x) paste0(x, "%")) +
    labs(x = "% of votes cast", y = NULL, subtitle = subtitle) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "right")
}
```

```{r top30-chart1}
#| fig-height: 7
#| fig-width: 9

make_top30_chart(pres_top30, group1_names, group1_display, group1_depts,
                 parties_main, "Municipalities 1–15 (sorted by department)")
```

```{r top30-chart2}
#| fig-height: 7
#| fig-width: 9

make_top30_chart(pres_top30, group2_names, group2_display, group2_depts,
                 parties_main, "Municipalities 16–30 (sorted by department)")
```

## Vote by Albó municipality category {#sec-albo-groups}

Municipalities are classified A through I based on their CEL (Condición
Étnico-Lingüística) profile from the 2024 census, following Albó & Romero's
framework as implemented in `albo-restudy-2024.qmd`. Category A represents
the most strongly indigenous-language municipalities; category I the least.

```{r albo-vote-by-category}
#| fig-height: 6
#| fig-width: 9

cel_geo <- readRDS("../bolivia-data/Censo 2024/base_datos_csv_2024/cel_geo.rds")

albo_groups <- cel_geo |>
  filter(!is.na(municipality), is.na(urban_rural)) |>
  select(department, province, municipality, cel_2plus, cel_4plus, cel_5plus) |>
  mutate(
    albo_group = case_when(
      cel_5plus >= 90 & cel_4plus >= 90 & cel_2plus >= 90 ~ "A",
      cel_5plus >  67 & cel_4plus >  80 & cel_2plus >  90 ~ "B",
      cel_5plus >  50 & cel_4plus >  67 & cel_2plus >  67 ~ "C",
      cel_5plus >  33 & cel_4plus >  50 & cel_2plus >  67 ~ "D",
      cel_2plus >  75                                      ~ "E",
      cel_2plus >  50                                      ~ "F",
      cel_2plus >  33                                      ~ "G",
      cel_2plus >  10                                      ~ "H",
      TRUE                                                 ~ "I"
    )
  )

pres_by_albo <- elections_1v |>
  filter(race == "PRESIDENTE") |>
  left_join(albo_groups |> select(department, province, municipality, albo_group),
            by = c("dept_code" = "department",
                   "prov_code" = "province",
                   "mun_code" = "municipality")) |>
  filter(!is.na(albo_group)) |>
  group_by(albo_group) |>
  summarise(
    across(all_of(parties_main), sum),
    null_votes = sum(null_votes),
    blank_votes = sum(blank_votes),
    votes_cast = sum(votes_cast),
    .groups = "drop"
  )

pres_albo_long <- pres_by_albo |>
  pivot_longer(cols = c(all_of(parties_main), null_votes, blank_votes),
               names_to = "category", values_to = "votes") |>
  mutate(pct = votes / votes_cast * 100) |>
  mutate(
    category = recode(category, !!!party_labels,
                      null_votes = "Null", blank_votes = "Blank"),
    category = factor(category,
                      levels = c(rev(party_labels[parties_main]), "Blank", "Null"))
  )

n_munis <- albo_groups |> count(albo_group)

ggplot(pres_albo_long, aes(x = albo_group, y = pct, fill = category)) +
  geom_col(position = "stack", width = 0.75) +
  geom_text(
    data = n_munis,
    aes(x = albo_group, y = -1.5, label = paste0("n=", n), fill = NULL),
    size = 2.8, color = "grey40"
  ) +
  scale_fill_manual(values = cat_colors, name = NULL) +
  scale_y_continuous(labels = \(x) paste0(x, "%"), expand = expansion(add = c(3, 1))) +
  labs(
    x = "Albó municipality category",
    y = "% of votes cast",
    title = "Presidential vote composition by Albó CEL category (first round)",
    subtitle = "Municipalities classified A (most indigenous-language) to I (least)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

## Second-round presidential results {#sec-segunda-vuelta}

```{r pres2-dept}
pres_2v <- elections_2v

pres2_dept <- aggregate_results(pres_2v, dept_code, dept_name,
                                parties = parties_2v)
pres2_dept |>
  select(dept_name, mesas, registered, turnout_pct, null_pct,
         all_of(paste0(parties_2v, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(registered))
```

```{r pres2-prov}
pres2_prov <- aggregate_results(
  pres_2v,
  dept_code, dept_name,
  prov_code, prov_name,
  parties = parties_2v
)
```

```{r pres2-mun}
pres2_mun <- aggregate_results(
  pres_2v,
  dept_code, dept_name,
  prov_code, prov_name,
  mun_code, mun_name,
  parties = parties_2v
)

pres2_mun |>
  select(dept_name, mun_name, valid_votes,
         all_of(paste0(parties_2v, "_pct"))) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  arrange(desc(valid_votes)) |>
  head(20)
```

### Second-round choropleths

```{r pres2-choropleths}
#| fig-height: 12
#| fig-width: 10

choropleth_panel(pres2_dept, pres2_prov, pres2_mun,
                 parties_2v, "Presidential (second round)")
```

### Second-round null vote rate

```{r nulo2-choropleths}
#| fig-height: 6
#| fig-width: 10

nulo2_dept <- pres2_dept |>
  left_join(dept_to_gadm, by = "dept_code")
nulo2_prov <- pres2_prov |>
  left_join(prov_to_gadm, by = c("dept_code", "prov_code"))
nulo2_mun <- pres2_mun |>
  left_join(muni_to_gadm, by = c("dept_code", "prov_code", "mun_code"))

p_nulo2_dept <- make_choropleth(
  gadm1, nulo2_dept,
  c("NAME_1" = "dept_gadm"),
  "null_pct", "Null vote % — Department", "#8B0000"
) + scale_fill_gradient(low = "white", high = "#8B0000", name = "%",
                        na.value = "grey85")

p_nulo2_prov <- make_choropleth(
  gadm2, nulo2_prov,
  c("NAME_1" = "dept_gadm", "NAME_2" = "province_gadm"),
  "null_pct", "Null vote % — Province", "#8B0000"
) + scale_fill_gradient(low = "white", high = "#8B0000", name = "%",
                        na.value = "grey85")

p_nulo2_mun <- make_choropleth(
  gadm3, nulo2_mun,
  c("NAME_1" = "dept_gadm", "NAME_3" = "muni_gadm"),
  "null_pct", "Null vote % — Municipality", "#8B0000"
) + scale_fill_gradient(low = "white", high = "#8B0000", name = "%",
                        na.value = "grey85")

p_nulo2_dept + p_nulo2_prov + p_nulo2_mun +
  plot_layout(ncol = 3) +
  plot_annotation(title = "Presidential null vote rate (second round)",
                  theme = theme(plot.title = element_text(face = "bold", size = 14)))
```

### Second-round top 30 municipalities

```{r pres2-top30-setup}
pres2_top30 <- pres2_mun |>
  filter(mun_name %in% top30_oep) |>
  filter(!(mun_name == "San Ignacio" & dept_code == 8))
```

```{r pres2-top30-chart1}
#| fig-height: 7
#| fig-width: 9

make_top30_chart(pres2_top30, group1_names, group1_display, group1_depts,
                 parties_2v, "Second round — Municipalities 1–15 (sorted by department)")
```

```{r pres2-top30-chart2}
#| fig-height: 7
#| fig-width: 9

make_top30_chart(pres2_top30, group2_names, group2_display, group2_depts,
                 parties_2v, "Second round — Municipalities 16–30 (sorted by department)")
```

### Second-round vote by Albó municipality category

```{r albo-vote-2v}
#| fig-height: 6
#| fig-width: 9

pres2_by_albo <- elections_2v |>
  left_join(albo_groups |> select(department, province, municipality, albo_group),
            by = c("dept_code" = "department",
                   "prov_code" = "province",
                   "mun_code" = "municipality")) |>
  filter(!is.na(albo_group)) |>
  group_by(albo_group) |>
  summarise(
    across(all_of(parties_2v), sum),
    null_votes = sum(null_votes),
    blank_votes = sum(blank_votes),
    votes_cast = sum(votes_cast),
    .groups = "drop"
  )

pres2_albo_long <- pres2_by_albo |>
  pivot_longer(cols = c(all_of(parties_2v), null_votes, blank_votes),
               names_to = "category", values_to = "votes") |>
  mutate(pct = votes / votes_cast * 100) |>
  mutate(
    category = recode(category, !!!party_labels,
                      null_votes = "Null", blank_votes = "Blank"),
    category = factor(category,
                      levels = c(rev(party_labels[parties_2v]), "Blank", "Null"))
  )

ggplot(pres2_albo_long, aes(x = albo_group, y = pct, fill = category)) +
  geom_col(position = "stack", width = 0.75) +
  geom_text(
    data = n_munis,
    aes(x = albo_group, y = -1.5, label = paste0("n=", n), fill = NULL),
    size = 2.8, color = "grey40"
  ) +
  scale_fill_manual(values = cat_colors, name = NULL) +
  scale_y_continuous(labels = \(x) paste0(x, "%"), expand = expansion(add = c(3, 1))) +
  labs(
    x = "Albó municipality category",
    y = "% of votes cast",
    title = "Presidential vote composition by Albó CEL category (second round)",
    subtitle = "Municipalities classified A (most indigenous-language) to I (least)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```
