---
title: "Ethiopia: Admin Boundary–Wikidata Crosswalk"
author: "Carwil Bjork-James"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
---

## Overview

Ethiopia has three administrative levels below the national level:

| Level | Name | Count |
|-------|------|-------|
| Admin 1 | Region | 15 |
| Admin 2 | Zone | 107 |
| Admin 3 | Woreda (District) | 1,148 |

This document builds a crosswalk between the official OCHA/UN admin boundary
file (`maps/eth_admin_boundaries.xlsx`) and Wikidata items for each level.
Working top-down (Region → Zone → Woreda) lets us use parent-level QID matches
to disambiguate child items that share names across regions.

The ultimate goal is a three-level crosswalk table
`data/eth_admin_crosswalk.rds` linking each official pcode to its Wikidata QID.

---

## Setup

```{r setup}
library(tidyverse)
library(readxl)
library(httr)
library(jsonlite)

source("R/get_wikidata_instances.R")
```

### Reusable helper functions

The functions below are designed to be country-agnostic and can be reused for
other countries with an OCHA-style admin boundary Excel file.

```{r helpers}
# ---- load_admin_sheet -------------------------------------------------------
#' Load one admin sheet from an OCHA-style boundaries Excel file.
#' Saves the result as an RDS for downstream use and returns it invisibly.
#'
#' @param xlsx_path  Path to the Excel workbook.
#' @param sheet      Sheet name (e.g. "eth_admin1").
#' @param rds_path   Where to write the cached RDS.
#' @param key_cols   Character vector of columns to keep (NULL = keep all).
load_admin_sheet <- function(xlsx_path, sheet, rds_path, key_cols = NULL) {
  df <- read_excel(xlsx_path, sheet = sheet)
  if (!is.null(key_cols)) df <- select(df, any_of(key_cols))
  saveRDS(df, rds_path)
  message("Saved ", nrow(df), " rows → ", rds_path)
  invisible(df)
}

# ---- fetch_wikidata_cached --------------------------------------------------
#' Fetch Wikidata instances for an admin class, with RDS caching.
#'
#' Arguments mirror get_wikidata_instances(); the result is saved to
#' `cache_path` on first run and reloaded on subsequent runs.
fetch_wikidata_cached <- function(class_qid,
                                  cache_path,
                                  property       = NULL,
                                  property_names = NULL,
                                  languages      = c("en", "am", "om", "so", "ti", "aa"),
                                  limit          = 1000,
                                  batch_size     = 30,
                                  force          = FALSE) {
  if (!force && file.exists(cache_path)) {
    cached <- readRDS(cache_path)
    # Re-fetch automatically if this cache predates alias support
    if (any(str_starts(names(cached), "aliases_"))) {
      message("Loading cached data from ", cache_path)
      return(cached)
    }
    message("Cache at ", cache_path, " predates alias support; re-fetching...")
  }
  result <- get_wikidata_instances(
    class_qid      = class_qid,
    property       = property,
    property_names = property_names,
    languages      = languages,
    limit          = limit,
    batch_size     = batch_size
  )
  saveRDS(result, cache_path)
  message("Saved ", nrow(result), " rows → ", cache_path)
  result
}

# ---- normalize_place_name ---------------------------------------------------
#' Normalize a place-name vector for fuzzy matching.
#'
#' Strips parenthetical disambiguation, comma-suffixes (", Region"),
#' and a configurable set of common administrative-level words
#' ("woreda", "town", "special", etc.).
#'
#' @param x          Character vector of names.
#' @param strip_terms Extra words to remove (in addition to defaults).
normalize_place_name <- function(x, strip_terms = NULL) {
  default_strip <- c("woreda", "town", "special", "zone", "district",
                     "region", "city", "municipality")
  all_strip <- paste(
    c(default_strip, strip_terms),
    collapse = "|"
  )
  pattern <- paste0("\\b(", all_strip, ")\\b")
  x |>
    str_to_lower() |>
    str_remove_all("\\s*\\(.*?\\)") |>   # remove (TG), (Oromia), etc.
    str_remove_all(",.*$") |>            # remove ", Somali", ", SNNPR"
    str_remove_all(pattern) |>
    str_squish()
}

# ---- match_admin_names ------------------------------------------------------
#' Match official admin names to Wikidata labels in up to four rounds.
#'
#' Round 1: exact lower-case match on the primary label.
#' Round 2: normalized match (strip admin-level words, disambiguations).
#' Round 3: exact lower-case match against any Wikidata alias.
#' Round 4: normalized match against any Wikidata alias.
#'
#' Rounds 3–4 require `wd_df` to contain `aliases_*` list columns
#' (produced automatically when `entity_props` includes "aliases").
#'
#' Returns the official data frame with a `qid` column and a `match_round`
#' column ("exact", "normalized", "alias_exact", "alias_normalized", or NA).
#'
#' @param official_df   Data frame with official boundary records.
#' @param official_col  Name of the column containing official place names.
#' @param wd_df         Wikidata data frame with a `qid` column.
#' @param wd_label_col  Name of the label column in wd_df (default "label_en").
#' @param strip_terms   Extra words to strip in normalization rounds.
match_admin_names <- function(official_df,
                              official_col,
                              wd_df,
                              wd_label_col = "label_en",
                              strip_terms  = NULL) {

  off_lower <- str_to_lower(official_df[[official_col]])
  wd_lower  <- str_to_lower(wd_df[[wd_label_col]])

  off_norm  <- normalize_place_name(official_df[[official_col]], strip_terms)
  wd_norm   <- normalize_place_name(wd_df[[wd_label_col]],       strip_terms)

  # Round 1: exact lower-case
  r1 <- match(off_lower, wd_lower)
  matched_qid <- wd_df$qid[r1]
  match_round <- if_else(!is.na(r1), "exact", NA_character_)

  # Round 2: normalized (only for still-unmatched)
  r2 <- match(off_norm, wd_norm)
  fill_r2 <- is.na(matched_qid) & !is.na(r2)
  matched_qid[fill_r2] <- wd_df$qid[r2[fill_r2]]
  match_round[fill_r2] <- "normalized"

  # Rounds 3 & 4: match against Wikidata aliases
  alias_cols <- names(wd_df)[str_starts(names(wd_df), "aliases_")]
  if (length(alias_cols) > 0 && any(is.na(matched_qid))) {

    # Aliases are stored as ", "-collapsed strings; split them into one row each
    alias_long <- wd_df |>
      select(qid, all_of(alias_cols)) |>
      pivot_longer(-qid, names_to = "lang_col", values_to = "aliases_str") |>
      filter(!is.na(aliases_str)) |>
      mutate(alias = str_split(aliases_str, ",\\s*")) |>
      unnest(alias) |>
      filter(nzchar(alias)) |>
      mutate(
        alias_lower = str_to_lower(alias),
        alias_norm  = normalize_place_name(alias, strip_terms)
      )

    # Round 3: exact alias match
    unmatched_idx <- which(is.na(matched_qid))
    candidate3 <- tibble(idx = unmatched_idx, name_lower = off_lower[unmatched_idx]) |>
      left_join(alias_long |> distinct(alias_lower, qid),
                by = c("name_lower" = "alias_lower")) |>
      filter(!is.na(qid)) |>
      distinct(idx, .keep_all = TRUE)
    matched_qid[candidate3$idx] <- candidate3$qid
    match_round[candidate3$idx] <- "alias_exact"

    # Round 4: normalized alias match
    unmatched_idx2 <- which(is.na(matched_qid))
    candidate4 <- tibble(idx = unmatched_idx2, name_norm = off_norm[unmatched_idx2]) |>
      left_join(alias_long |> distinct(alias_norm, qid),
                by = c("name_norm" = "alias_norm")) |>
      filter(!is.na(qid)) |>
      distinct(idx, .keep_all = TRUE)
    matched_qid[candidate4$idx] <- candidate4$qid
    match_round[candidate4$idx] <- "alias_normalized"
  }

  official_df |>
    mutate(
      qid         = matched_qid,
      match_round = match_round
    )
}

# ---- report_match_status ----------------------------------------------------
#' Print a tidy match-status summary for one admin level.
report_match_status <- function(matched_df, level_name, n_official, n_wd) {
  n_exact      <- sum(matched_df$match_round == "exact",            na.rm = TRUE)
  n_norm       <- sum(matched_df$match_round == "normalized",       na.rm = TRUE)
  n_alias_ex   <- sum(matched_df$match_round == "alias_exact",      na.rm = TRUE)
  n_alias_norm <- sum(matched_df$match_round == "alias_normalized", na.rm = TRUE)
  n_man        <- sum(matched_df$match_round == "manual",           na.rm = TRUE)
  n_total      <- n_exact + n_norm + n_alias_ex + n_alias_norm + n_man
  n_unmatched  <- sum(is.na(matched_df$qid))

  tibble(
    Level           = level_name,
    `Official`      = n_official,
    `In WD`         = n_wd,
    `Exact`         = n_exact,
    `Normalized`    = n_norm,
    `Alias exact`   = n_alias_ex,
    `Alias norm`    = n_alias_norm,
    `Manual`        = n_man,
    `Total matched` = n_total,
    `Unmatched`     = n_unmatched
  )
}
```

---

## Load official admin boundaries

All three admin levels come from a single OCHA-sourced Excel workbook.
We load each sheet, keep only the essential columns, and save as RDS so
downstream scripts can work without the xlsx dependency.

```{r load-admin-boundaries}
admin1_rds <- "data/eth_admin1.rds"
admin2_rds <- "data/eth_admin2.rds"
admin3_rds <- "data/eth_admin3.rds"

xlsx_path  <- "maps/eth_admin_boundaries.xlsx"

# Columns we care about at each level (add more if needed)
admin1_cols <- c("adm1_pcode", "adm1_name", "adm0_pcode", "adm0_name")
admin2_cols <- c("adm2_pcode", "adm2_name", "adm1_pcode", "adm1_name")
admin3_cols <- c("adm3_pcode", "adm3_name", "adm3_alt_name",
                 "adm2_pcode", "adm2_name",
                 "adm1_pcode", "adm1_name",
                 "area_sqkm")

eth_admin1 <- load_admin_sheet(xlsx_path, "eth_admin1", admin1_rds, admin1_cols)
eth_admin2 <- load_admin_sheet(xlsx_path, "eth_admin2", admin2_rds, admin2_cols)
eth_admin3 <- load_admin_sheet(xlsx_path, "eth_admin3", admin3_rds, admin3_cols)
```

Quick sanity check — does the table describe a complete three-level hierarchy?

```{r admin-counts}
tribble(
  ~Level,    ~Expected, ~Found,
  "Region",  15L,       n_distinct(eth_admin3$adm1_pcode),
  "Zone",   107L,       n_distinct(eth_admin3$adm2_pcode),
  "Woreda", 1148L,      n_distinct(eth_admin3$adm3_pcode)
) |>
  mutate(OK = if_else(Expected == Found, "✓", "✗")) |>
  knitr::kable(caption = "Official admin hierarchy counts")
```

---

## Level 1 — Regions (Admin1)

**Wikidata class:** [region of Ethiopia (Q1057504)](https://www.wikidata.org/wiki/Q1057504)

With only 15 regions, this level is small enough to match almost entirely
by name. We fetch P131 (located in / country) to verify items are genuinely
Ethiopian regions, and P17 (country) as a secondary check.

```{r fetch-regions}
regions_wd <- fetch_wikidata_cached(
  class_qid      = "Q1057504",
  cache_path     = "data/et_regions_wd.rds",
  property       = c("P131", "P17"),
  property_names = c("located_in", "country"),
  languages      = c("en", "am", "om", "so", "ti", "aa"),
  limit          = 50,
  batch_size     = 30
)

cat("Wikidata returned", nrow(regions_wd), "items for Q1057504\n")
regions_wd |> select(qid, label_en, label_am, located_in, country)
```

### Name matching

```{r match-regions}
# Build a clean lookup from the official table (one row per region)
admin1_official <- eth_admin1 |>
  distinct(adm1_pcode, adm1_name)

regions_matched <- match_admin_names(
  official_df  = admin1_official,
  official_col = "adm1_name",
  wd_df        = regions_wd
)

# Review results
regions_matched |>
  select(adm1_pcode, adm1_name, qid, match_round) |>
  arrange(adm1_pcode) |>
  knitr::kable(caption = "Admin1 (Region) name-match results")
```

### Unmatched regions

```{r regions-unmatched}
unmatched_official_r1 <- regions_matched |> filter(is.na(qid))
unmatched_wd_r1 <- regions_wd |>
  filter(!qid %in% regions_matched$qid)

cat("Official regions without a WD match:\n")
print(unmatched_official_r1 |> select(adm1_pcode, adm1_name))

cat("\nWD items not matched to any official region:\n")
print(unmatched_wd_r1 |> select(qid, label_en, label_am))
```

### Manual matches

Add pairs here as you identify them from the unmatched output above.

```{r regions-manual}
# Format: tibble(adm1_pcode = ..., qid = ...)
# Example (fill in as needed):
#   "ET07" → Q... (Central Ethiopia)
region_manual_matches <- tribble(
  ~adm1_pcode, ~qid,
  # Post-2023 SNNPR successor regions: WD labels use "Regional State" suffix
  "ET07", "Q122415622",   # Central Ethiopia  → Central Ethiopia Regional State
  "ET08", "Q122148951",   # South Ethiopia    → South Ethiopia Regional State
  "ET11", "Q109788993"    # South West Ethiopia → Southwest Ethiopia Regional State
  # ET99 (Contested) has no single WD item; intentionally left unmatched
)

# Apply manual matches
if (nrow(region_manual_matches) > 0) {
  regions_matched <- rows_patch(
    regions_matched,
    region_manual_matches |> mutate(match_round = "manual"),
    by = "adm1_pcode",
    unmatched = "ignore"
  )
}
```

### Save & report

```{r regions-save}
saveRDS(regions_matched, "data/et_regions_matched.rds")

report_match_status(regions_matched, "Region (Admin1)",
                    n_official = nrow(admin1_official),
                    n_wd       = nrow(regions_wd)) |>
  knitr::kable()
```

---

## Level 2 — Zones (Admin2)

**Wikidata class:** [zone of Ethiopia (Q219875)](https://www.wikidata.org/wiki/Q219875)

With 107 zones we have more name collisions (some zone names appear in multiple
regions), so we use the region QIDs from Level 1 to disambiguate: after round-2
name matching, we verify that the matched WD item's `located_in` agrees with
the region's QID.

```{r fetch-zones}
zones_wd <- fetch_wikidata_cached(
  class_qid      = "Q219875",
  cache_path     = "data/et_zones_wd.rds",
  property       = c("P131", "P17"),
  property_names = c("located_in", "country"),
  languages      = c("en", "am", "om", "so", "ti", "aa"),
  limit          = 200,
  batch_size     = 30
)

cat("Wikidata returned", nrow(zones_wd), "items for Q219875\n")
```

### Name matching (with region disambiguation)

```{r match-zones}
# Build official zone lookup with the parent region QID attached
admin2_official <- eth_admin2 |>
  distinct(adm2_pcode, adm2_name, adm1_pcode) |>
  left_join(regions_matched |> select(adm1_pcode, region_qid = qid),
            by = "adm1_pcode")

# Round 1 + 2 name match
zones_matched <- match_admin_names(
  official_df  = admin2_official,
  official_col = "adm2_name",
  wd_df        = zones_wd
)

# Region-consistency check: flag matches where WD located_in ≠ region QID
zones_wd_located <- zones_wd |> select(qid, wd_located_in = located_in)

zones_matched <- zones_matched |>
  left_join(zones_wd_located, by = "qid") |>
  mutate(
    region_consistent = case_when(
      is.na(qid)            ~ NA,
      is.na(region_qid)     ~ NA,     # official region has no QID yet
      is.na(wd_located_in)  ~ NA,     # WD item has no P131
      wd_located_in == region_qid ~ TRUE,
      TRUE                  ~ FALSE
    )
  )

# Summary
cat("Matched:", sum(!is.na(zones_matched$qid)), "/", nrow(admin2_official), "\n")
cat("Region-consistent:", sum(zones_matched$region_consistent == TRUE, na.rm = TRUE), "\n")
cat("Region-inconsistent:", sum(zones_matched$region_consistent == FALSE, na.rm = TRUE), "\n")
```

### Review problematic zones

```{r zones-inspect}
# Unmatched official zones
zones_matched |>
  filter(is.na(qid)) |>
  select(adm2_pcode, adm2_name, adm1_pcode) |>
  arrange(adm1_pcode) |>
  knitr::kable(caption = "Official zones with no WD match")

# Inconsistent region assignments (name matched but wrong region)
zones_matched |>
  filter(region_consistent == FALSE) |>
  select(adm2_pcode, adm2_name, region_qid, qid, wd_located_in, match_round) |>
  knitr::kable(caption = "Zone name matches with wrong parent region")

# WD zones not matched to any official zone
zones_wd |>
  filter(!qid %in% zones_matched$qid) |>
  select(qid, label_en, label_am, located_in) |>
  knitr::kable(caption = "WD zone items not matched to any official zone")
```

### Manual matches

```{r zones-manual}
zone_manual_matches <- tribble(
  ~adm2_pcode, ~qid
  # Fill from inspection above
)

if (nrow(zone_manual_matches) > 0) {
  zones_matched <- rows_patch(
    zones_matched,
    zone_manual_matches |> mutate(match_round = "manual"),
    by = "adm2_pcode",
    unmatched = "ignore"
  )
}
```

### Save & report

```{r zones-save}
saveRDS(zones_matched, "data/et_zones_matched.rds")

report_match_status(zones_matched, "Zone (Admin2)",
                    n_official = nrow(admin2_official),
                    n_wd       = nrow(zones_wd)) |>
  knitr::kable()
```

---

## Level 3 — Woredas / Districts (Admin3)

**Wikidata class:** [woreda of Ethiopia (Q690840)](https://www.wikidata.org/wiki/Q690840)

With 1,148 woredas this is the hardest level: names are most variable between
sources, and many woreda names repeat across regions. We use zone QIDs from
Level 2 as the primary disambiguation key, and also retrieve the OpenStreetMap
relation ID (P402) for later spatial cross-checks.

```{r fetch-woredas}
woredas_wd <- fetch_wikidata_cached(
  class_qid      = "Q690840",
  cache_path     = "data/et_woredas_wd.rds",
  property       = c("P131", "P402"),
  property_names = c("located_in", "openstreetmap_code"),
  languages      = c("en", "am", "om", "so", "ti", "aa"),
  limit          = 1200,
  batch_size     = 30
)

cat("Wikidata returned", nrow(woredas_wd), "items for Q690840\n")
```

### Name matching (with zone disambiguation)

```{r match-woredas}
# Official woreda table with parent zone QID
admin3_official <- eth_admin3 |>
  distinct(adm3_pcode, adm3_name, adm3_alt_name, adm2_pcode, adm1_pcode) |>
  left_join(zones_matched  |> select(adm2_pcode, zone_qid   = qid), by = "adm2_pcode") |>
  left_join(regions_matched|> select(adm1_pcode, region_qid = qid), by = "adm1_pcode")

# Round 1 + 2 name match
woredas_matched <- match_admin_names(
  official_df  = admin3_official,
  official_col = "adm3_name",
  wd_df        = woredas_wd,
  # Strip Ethiopia-specific admin-level words from comparison
  strip_terms  = c("liyu", "special", "ketema", "awraja")
)

# Zone-consistency check
wd_located <- woredas_wd |> select(qid, wd_located_in = located_in)

woredas_matched <- woredas_matched |>
  left_join(wd_located, by = "qid") |>
  mutate(
    zone_consistent = case_when(
      is.na(qid)           ~ NA,
      is.na(zone_qid)      ~ NA,
      is.na(wd_located_in) ~ NA,
      wd_located_in == zone_qid ~ TRUE,
      TRUE                 ~ FALSE
    )
  )

cat("Matched:", sum(!is.na(woredas_matched$qid)), "/ 1148\n")
cat("Zone-consistent:", sum(woredas_matched$zone_consistent == TRUE,  na.rm = TRUE), "\n")
cat("Zone-inconsistent:", sum(woredas_matched$zone_consistent == FALSE, na.rm = TRUE), "\n")
```

### Match summary by region

```{r woredas-by-region}
woredas_matched |>
  group_by(adm1_pcode) |>
  summarise(
    woredas   = n(),
    matched   = sum(!is.na(qid)),
    pct       = round(100 * matched / woredas),
    zone_ok   = sum(zone_consistent == TRUE,  na.rm = TRUE),
    zone_bad  = sum(zone_consistent == FALSE, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  left_join(eth_admin1 |> distinct(adm1_pcode, adm1_name), by = "adm1_pcode") |>
  select(adm1_name, woredas, matched, pct, zone_ok, zone_bad) |>
  arrange(desc(woredas)) |>
  knitr::kable(caption = "Woreda match coverage by region")
```

### Unmatched official woredas

```{r woredas-unmatched}
woredas_matched |>
  filter(is.na(qid)) |>
  select(adm3_pcode, adm3_name, adm3_alt_name, adm2_pcode) |>
  arrange(adm3_pcode) |>
  knitr::kable(caption = "Official woredas with no WD match (for manual review)")
```

### WD items not matched to any official woreda

These may be historical woredas, recently renamed units, or items that need
label corrections on Wikidata.

```{r woredas-wd-unmatched}
woredas_wd |>
  filter(!qid %in% woredas_matched$qid) |>
  select(qid, label_en, label_am, located_in) |>
  arrange(label_en) |>
  knitr::kable(caption = "WD woreda items not matched to any official woreda")
```

### Manual matches

```{r woredas-manual}
# This table will grow as you work through the unmatched items above.
# Format: adm3_pcode (from official) ↔ qid (from Wikidata)
woreda_manual_matches <- tribble(
  ~adm3_pcode, ~qid
  # "ET010101", "Q..."
)

if (nrow(woreda_manual_matches) > 0) {
  woredas_matched <- rows_patch(
    woredas_matched,
    woreda_manual_matches |> mutate(match_round = "manual"),
    by = "adm3_pcode",
    unmatched = "ignore"
  )
}
```

### Save & report

```{r woredas-save}
saveRDS(woredas_matched, "data/et_woredas_matched.rds")

report_match_status(woredas_matched, "Woreda (Admin3)",
                    n_official = nrow(admin3_official),
                    n_wd       = nrow(woredas_wd)) |>
  knitr::kable()
```

---

## Combined crosswalk

Once matching is reasonably complete, assemble the full three-level crosswalk
and save it as a single reference table.

```{r crosswalk}
eth_admin_crosswalk <- eth_admin3 |>
  distinct(adm3_pcode, adm3_name, adm2_pcode, adm2_name, adm1_pcode, adm1_name) |>
  left_join(
    woredas_matched  |> select(adm3_pcode, woreda_qid  = qid, woreda_match = match_round),
    by = "adm3_pcode"
  ) |>
  left_join(
    zones_matched    |> select(adm2_pcode, zone_qid    = qid, zone_match   = match_round),
    by = "adm2_pcode"
  ) |>
  left_join(
    regions_matched  |> select(adm1_pcode, region_qid  = qid, region_match = match_round),
    by = "adm1_pcode"
  )

saveRDS(eth_admin_crosswalk, "data/eth_admin_crosswalk.rds")

# Summary
eth_admin_crosswalk |>
  summarise(
    total_woredas     = n(),
    woreda_qid_found  = sum(!is.na(woreda_qid)),
    zone_qid_found    = sum(!is.na(zone_qid)),
    region_qid_found  = sum(!is.na(region_qid))
  ) |>
  pivot_longer(everything()) |>
  knitr::kable(caption = "Crosswalk coverage summary")
```

---

## Next steps

- Work through the **manual match tables** at each level, using the unmatched
  output as a guide. Start with Admin1 (15 items) and Admin2 (107 items) before
  tackling the 1,148 woredas.
- For hard name mismatches, add the Wikidata `label_am` (Amharic) column to
  the comparison — many Amharic names are more consistent with the official
  coding than English transliterations.
- Once the crosswalk is sufficiently complete, prepare **QuickStatements** to
  add OCHA pcodes (P300 or a new property) to Wikidata items.
- Consider a spatial join using OSM relation IDs (P402) or boundary geometry to
  resolve remaining unmatched items.
