---
title: "Ethiopia Municipality Wikipedia Pages"
author: "Carwil Bjork-James"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
editor: visual
---

# Overview

This document scrapes English Wikipedia for all pages in the [Districts of Ethiopia](https://en.wikipedia.org/wiki/Category:Districts_of_Ethiopia) category tree, extracts structured information from each page, and compares the coverage with Wikidata's inventory of Ethiopia's districts.

```{r setup}
#| message: false
library(tidyverse)
library(httr)
library(jsonlite)
library(WikipediR)

source("src/wikipedia-tools.R")
source("R/get_wikidata_instances.R")
```

# 1. Discover district pages from Wikipedia category

We start by listing the subcategories of `Category:Districts of Ethiopia`, keeping only those whose name begins with "Districts" (i.e. excluding "Former districts of Ethiopia" and other non-district subcategories).

```{r get-subcategories}
top_cat <- "Category:Districts of Ethiopia"
subcats <- get_subcategories(top_cat)
subcats

# Keep only subcategories beginning with "Municipalities"
muni_subcats <- subcats |>
  filter(str_detect(title, "^Category:District") | str_detect(title, "^Category:Special woredas") )

muni_subcats
```

Now collect all article pages from each subcategory, plus any pages listed directly under the top-level category.

```{r get-all-pages}
# Pages directly in the top-level category
top_pages <- get_category_pages(top_cat)

# Pages in each "Municipalities of ..." subcategory
subcat_pages <- map_dfr(muni_subcats$title, function(cat) {
  Sys.sleep(0.3)
  get_category_pages(cat) |>
    mutate(source_category = cat)
})

# Combine and deduplicate
all_muni_pages <- bind_rows(
  top_pages |> mutate(source_category = top_cat),
  subcat_pages
) |>
  distinct(title, .keep_all = TRUE) |>
  # Drop any remaining category or template pages
  filter(ns == 0)

message(nrow(all_muni_pages), " unique district article pages found")
```

# 2. Fetch page metadata (page length + Wikidata QID)

```{r page-info}
page_info <- get_page_info_batch(all_muni_pages$title)

muni_df <- all_muni_pages |>
  select(-pageid) |>
  left_join(page_info, by = "title")
```

# 3. Cache and parse wikitext

We cache each page's raw wikitext to `wikipedia-pages/bolivia/` so that subsequent runs don't re-download.

```{r cache-wikitext}
cache_dir <- "wikipedia-pages/ethiopia"
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

# Fetch / load cached wikitext for every page
muni_df <- muni_df |>
  mutate(wikitext = map_chr(title, function(t) {
    # 1. Pause for 0.5 seconds before making the API call
    Sys.sleep(0.5) 
    
    # 2. Proceed with your cached/live fetch
    txt <- cache_wikitext(t, cache_dir)
    if (is.null(txt)) NA_character_ else txt
  }, .progress = "Fetching wikitext"))
```

# 4. Extract structured fields from each page

## 4a. Character count, citations, and census years

```{r extract-counts}
muni_df <- muni_df |>
  mutate(
    n_chars = nchar(wikitext),
    n_citations = map_int(wikitext, count_citations),
    n_refs = map_int(wikitext, count_refs),
    census_years = map(wikitext, extract_census_years),
    census_years_str = map_chr(census_years, ~ paste(.x, collapse = ", "))
  )
```

## 4b. Infobox fields

Parse the main infobox from each article and extract every key–value pair. We first collect all infobox data as a list column, then pivot the most common fields into their own columns.

```{r extract-infoboxes}
muni_df <- muni_df |>
  mutate(infobox_raw = map(wikitext, extract_infobox))

muni_df <- muni_df |>
  mutate(infobox_fields = map(infobox_raw, \(x) names(x)[x != ""])) |>
  mutate(n_fields = lengths(infobox_fields)) |>
  mutate(infobox_values = map(infobox_raw, \(x) unname(x[x != ""])))

# Identify which infobox keys appear most often
all_keys <- muni_df |>
  filter(!map_lgl(infobox_raw, is.null)) |>
  pull(infobox_raw) |>
  map(names) |>
  unlist() |>
  # Normalise key names to lowercase with underscores
  str_to_lower() |>
  str_replace_all("\\s+", "_")

filled_keys <- muni_df |> 
  pull(infobox_fields) |> 
  unlist() 

key_freq <- tibble(key = filled_keys) |>
  count(key, sort = TRUE)

key_freq |> print(n = 40)


```

```{r pivot-infobox}
# For each page, create a one-row tibble of cleaned infobox values
infobox_tbl <- muni_df |>
  transmute(
    title,
    infobox_row = map(infobox_raw, function(ib) {
      if (is.null(ib)) return(tibble())
      # Normalise key names
      names(ib) <- names(ib) |>
        str_to_lower() |>
        str_replace_all("\\s+", "_")
      # Clean values
      ib <- map(ib, clean_infobox_value)
      as_tibble(ib)
    })
  ) |>
  unnest(infobox_row, keep_empty = TRUE)

# Join back to main table (keeping the infobox columns separate for clarity)
muni_result <- muni_df |>
  select(-infobox_raw, -wikitext) |>
  left_join(infobox_tbl, by = "title")

muni_result$wikidata_qid |> unique() -> et_dist_wikipedia_qids

```

# 5. Compare with Wikidata

Query Wikidata for all items that are instances of district of Ethiopia then compare with the Wikipedia category listing.

```{r wikidata-query}
# Q690840 = "district of Ethiopia"
wd_munis <- get_wikidata_instances(
  "Q690840",
  property = c("P131"),
  property_names = c("located_in"),
  country = "Q115",
  languages = c("en", "es"),
  limit = 1500
)

districts_wd <- get_wikidata_instances(
  class_qid      = "Q690840",
  property       = c("P131", "P402"),
  property_names = c("located_in", "openstreetmap_code"),
  languages      = c("en", "am", "om", "so", "ti", "aa"),
  limit          = 1200,
  batch_size     = 30
)

# Extract English Wikipedia article title from the sitelinks
# wd_munis <- wd_munis |>
districts_wd <- districts_wd |>
  mutate(
    en_wiki_title = map_chr(wikipedia_articles, function(arts) {
      if (length(arts) == 0) return(NA_character_)
      en_art <- arts[str_detect(arts, "^en:")]
      if (length(en_art) == 0) return(NA_character_)
      str_replace(en_art[1], "^en:\\s*", "")
    })
  )

wd_munis <- districts_wd

```

```{r compare-lists}
# Pages found in Wikipedia category but not in Wikidata
wp_only <- muni_result |>
  filter(!wikidata_qid %in% wd_munis$qid | is.na(wikidata_qid))

# Wikidata items with no Wikipedia category page
wd_only <- wd_munis |>
  filter(!qid %in% muni_result$wikidata_qid)

wd_only_has_article <- wd_only |>
  filter(!is.na(en_wiki_title))

wd_only_no_article <- wd_only |>
  filter(is.na(en_wiki_title))

# Both
in_both <- muni_result |>
  filter(wikidata_qid %in% wd_munis$qid)

cat(glue::glue(
  "Wikipedia category pages: {nrow(muni_result)}\n",
  "Wikidata 'district of Ethiopia' items: {nrow(wd_munis)}\n",
  "In both: {nrow(in_both)}\n",
  "Wikipedia only: {nrow(wp_only)}\n",
  "Wikidata only (no category page): {nrow(wd_only)}\n",
  "Wikidata district, Wikipedia page not in category: {nrow(wd_only_has_article)}\n",
  "Wikidata district, no Wikipedia page: {nrow(wd_only_no_article)}\n"
))
```

### Pages in Wikipedia category but not Wikidata

```{r wp-only-table}
wp_only |>
  select(title, wikidata_qid, n_chars, n_citations) |>
  arrange(title)
```

### Wikidata items with no Wikipedia category page

```{r wd-only-table}
wd_only |>
  select(qid, label_en, label_am, en_wiki_title) |>
  arrange(label_en)
```

# 6. Summary table

```{r save-results}
# Save the main result
saveRDS(muni_result, "data/ethiopia_district_wikipedia.rds")
write_csv(muni_result |> select(-census_years),
          "data/ethiopia_municipality_wikipedia.csv")

# districts_wp <- read_rds("data/ethiopia_district_wikipedia.rds")

muni_result |>
  select(title, wikidata_qid, page_length, n_chars, n_citations, n_refs,
         census_years_str) |>
  arrange(title) |>
  head(20)
```
