---
title: "Wrangling ICE Detention Data"
author: "Carwil Bjork-James"
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
editor: visual
---

# Data Sources

We will work from the data spreadsheets provided online at ICE's [Detention Management](https://www.ice.gov/detain/detention-management) page, but will always maintain our own archived versions (as data files in our repository) and check that they are stored on the Internet Archive Wayback Machine for future use.

As of February 4, 2026, we can find:

-   [ Detention FY 2026 YTD, Alternatives to Detention FY 2026 YTD and Facilities FY 2026 YTD, Footnotes (221 KB)](https://www.ice.gov/doclib/detention/FY26_detentionStats02022026.xlsx) `FY26_detentionStats02022026.xlsx`

-   [ Detention FY 2025 YTD, Alternatives to Detention FY 2025 YTD and Facilities FY 2025 YTD, Footnotes (1.5 MB)](https://www.ice.gov/doclib/detention/FY25_detentionStats09242025.xlsx) `FY25_detentionStats09242025.xlsx`

-   [ FY 2024 Detention Statistics](https://www.ice.gov/doclib/detention/FY24_detentionStats.xlsx) `FY24-detentionstats.xlsx`

-   [ FY 2023 Detention Statistics](https://www.ice.gov/doclib/detention/FY23_detentionStats.xlsx) `FY23-detentionstats.xlsx`

-   [ FY 2022 Detention Statistics](https://www.ice.gov/doclib/detention/FY22-detentionStats.xlsx) `FY22-detentionStats.xlsx`

-   [ FY 2021 Detention Statistics](https://www.ice.gov/doclib/detention/FY21-detentionstats.xlsx) `FY21-detentionStats.xlsx`

-   [ FY 2020 Detention Statistics](https://www.ice.gov/doclib/detention/FY20-detentionstats.xlsx) `FY20-detentionstats.xlsx`

-   [ FY 2019 Detention Statistics](https://www.ice.gov/doclib/detention/FY19-detentionstats.xlsx) `FY19-detentionstats.xlsx`

There are subtle changes between the Facilities page on each of these, which will have to take into account as we import them. The import code I created for the first time was built around the FY2026 spreadsheet.

The clean import script for the FY 2026 data is [shared as a gist here](https://gist.github.com/CarwilB/6bf4ab7bbd05c8c0a02552402f17d4f2). Feel free to use or modify it.

## Initial Import

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r libraries}
#| echo: false
#| message: false
library(readxl)
library(libr)
library(httr)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(tibble)
library(WikipediR)
```

This code for the download only needs to executed once. The `eval: false` option disables the execution of the code chunk.

```{r download-one-spreadsheet}
#| eval: false

# Simple single-file version
data_url <- "https://www.ice.gov/doclib/detention/FY26_detentionStats02022026.xlsx"
local_file <- "data/FY26_detentionStats02022026.xlsx"
sheet_name <- "Facilities FY26"

httr::GET(data_url, write_disk(local_file, overwrite = TRUE))
```

In scaling up to the set of spreadsheets, I had to think about which specifics from them affected my code. This metadata is what I included here: the fiscal year, filenames, name of the sheet within the .xlsx document, position of the header rows, and rightmost column of the data.

This allows for some automatically calculated data for later use, such as the first data row and the number of that rightmost column.

```{r spreadsheet-metadata}
# Tibble with multiple years
data_file_info <- tibble(
  year_name = c("FY19", "FY20", "FY21", "FY22", "FY23",
    "FY24", "FY25", "FY26"),
  year = c(2019:2026),
  url = c("https://www.ice.gov/doclib/detention/FY19-detentionstats.xlsx",
    "https://www.ice.gov/doclib/detention/FY20-detentionstats.xlsx",
    "https://www.ice.gov/doclib/detention/FY21-detentionstats.xlsx",
    "https://www.ice.gov/doclib/detention/FY22-detentionStats.xlsx",
    "https://www.ice.gov/doclib/detention/FY23_detentionStats.xlsx",
    "https://www.ice.gov/doclib/detention/FY24_detentionStats.xlsx",
    "https://www.ice.gov/doclib/detention/FY25_detentionStats09242025.xlsx",
    "https://www.ice.gov/doclib/detention/FY26_detentionStats02022026.xlsx"),
  local_file = c("data/FY19-detentionstats.xlsx",
    "data/FY20-detentionstats.xlsx",
    "data/FY21-detentionstats.xlsx",
    "data/FY22-detentionStats.xlsx",
    "data/FY23_detentionStats.xlsx",
    "data/FY24_detentionStats.xlsx",
    "data/FY25_detentionStats09242025.xlsx",
    "data/FY26_detentionStats02022026.xlsx"),
  sheet_name = c("Facilities FY19", "Facilities EOYFY20 ", "Facilities FY21 YTD",
    "Facilities FY22", "Facilities EOFY23",
    "Facilities EOFY24", "Facilities FY25", "Facilities FY26"),
  header_rows = list(c(5,7), c(5,7), c(5,7), c(5,7), c(4,6),
                  c(5,7), c(5,7), c(9,10)),
  right_column= c("AE", "AE", "AE", "AD", "AG",
    "AB", "AB", "AA")
)

LETTERS_PLUS <- c(LETTERS, paste0("A", LETTERS), paste0("B", LETTERS))

data_file_info <- data_file_info %>%
  rowwise() %>%
  mutate(
    first_header_row = unlist(header_rows)[[1]],
    second_header_row = unlist(header_rows)[[2]],
    first_data_row = max(unlist(header_rows)) + 1,
    right_column_num = which(LETTERS_PLUS == right_column)) %>%
  select(-header_rows)

```

```{r download-multiple-spreadsheets}
#| eval: false

# Download all files
for (i in 1:nrow(data_file_info)) {
  httr::GET(data_file_info$url[i],
    write_disk(data_file_info$local_file[i], overwrite = FALSE))
}
```
# Importing the variables

Next we import the two-line headers and combine them into variable names. This took a fair amount of experimentation the first time.

```{r import-variable-names}
#| message: false
header_rows <- function(data_file_info_row){
  with(data_file_info_row, {
    first_row <- read_excel(
      local_file,
      sheet = sheet_name,
      range = cell_limits(
        ul = c(first_header_row, 1),
        lr = c(first_header_row, right_column_num)
      ),
      col_names = FALSE
    )
    second_row <- read_excel(
      local_file,
      sheet = sheet_name,
      range = cell_limits(
        ul = c(second_header_row, 1),
        lr = c(second_header_row, right_column_num)
      ),
      col_names = FALSE
    )
    rbind(first_row, second_row)
  })
}

# This single call generates the headers for FY26 as a test
# raw_headers <- header_rows(data_file_info %>% filter(year_name == "FY26"))

# Create a list of headers for each spreadsheet
raw_headers_list <- map(1:nrow(data_file_info), function(i) {
  header_rows(data_file_info[i, ])
})
names(raw_headers_list) <- data_file_info$year_name
```

```{r clean-variable-names}
# 2. Create Clean Variable Names
#    - Transpose headers to columns for easier filling
#    - Fill the top header row (Row 9) downward to cover merged cells
#    - Remove "FY26"
#    - Combine Top and Bottom headers
#    - Format to snake_case (lowercase with underscores)
clean_variable_names_from_header <- function(raw_headers) {
  clean_names <- raw_headers %>%
    t() %>%
    as.data.frame() %>%
    fill(V1) %>%                                     # Propagate Row 9 headers
    mutate(
      # Remove "FY26" and extra whitespace from both header parts
      V1 = str_trim(str_remove_all(V1, "FY\\d\\d")),
      V1 = str_remove_all(V1, ":"), # Remove colons if present
      V2 = str_trim(str_remove_all(V2, "FY\\d\\d")),
      V2 = str_remove_all(V2, ":"),

      V1 = case_when( # clean extra words
        str_detect(V1, "This list") ~ "Facility",
        V1 == "Facility Information" ~ "Facility",
#        str_detect(V1, "Average Length of Stay") ~ "ALOS",
        V1 == "ADP Detainee Classification Level" ~ "ADP Detainee Classification",
        V1 == "ADP Detainee Security Level" ~ "ADP Detainee Security",
        V1 == "ADP ICE Threat Level" ~ "ADP",
        V1 == "ADP Mandatory" ~ "ADP",
        V1 == "Contract Facility Inspections Information" ~ "Inspections",

        TRUE ~ V1
      ),

      # Combine Row 9 (V1) and Row 10 (V2)
      combined = paste(V1, V2),

      # Clean: Lowercase, replace special chars with _, remove duplicates
      var_name = tolower(combined),
      var_name = str_replace_all(var_name, "[^a-z0-9]+", "_"),
      var_name = str_remove_all(var_name, "^_+|_+$") # Remove leading/trailing _
    ) %>%
    pull(var_name)
  clean_names
}

clean_names_list <- map(raw_headers_list, clean_variable_names_from_header)
names(clean_names_list) <- data_file_info$year_name
# clean_names_list$FY24 # This would print out the variables from one sheet
```

Now we need to check if the variables are the same across multiple years. (Hopefully some of our simplifying helped with that.) This graphing package may seem like overkill, but it really simplified my work here.

```{r compare-variable-names}
#| warning: false
library(UpSetR)
upset_plot <- upset(fromList(clean_names_list),
                    sets = names(clean_names_list),
                    order.by = "freq")  # Visual intersection analysis
upset_plot

# Create a dataframe showing which variables are in which lists
presence_matrix <- clean_names_list %>%
  imap_dfr(~ tibble(variable = .x, list_name = .y)) %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = list_name, values_from = present, values_fill = FALSE)
```

Looking at that graph, I see that most variables are there in all the data sheets, but there are some differences. We can use the `presence_matrix` dataframe to see which variables are missing from which years.

```{r find-variable-differences}
# Find FY23-only variables
vars_in_all <- presence_matrix %>%
  filter(rowSums(select(., -variable)) == ncol(select(., -variable))) %>%
  pull(variable)

fy23_only <- presence_matrix %>%
  filter(FY23 == TRUE & rowSums(select(., -variable, -FY23)) == 0) %>%
  pull(variable)
not_in_fy23_22 <- presence_matrix %>%
  filter(FY23 == FALSE) %>%
  filter(FY22 == FALSE) %>%
  pull(variable)
not_in_fy23 <- presence_matrix %>%
  filter(FY23 == FALSE) %>%
  pull(variable)
```

- **`r length(vars_in_all)` variables in all sheets:** `r vars_in_all`
- **Variables not in all sheets:** `r setdiff(unlist(clean_names_list), vars_in_all)`
- **`r length(fy23_only)` variables only in FY23:** `r fy23_only`

Fortunately, nearly all variables are in all years. And the differences in the variables seem to all consist in how inspections are reparted. We may want to think through how to re-create `inspections_last_inspection_date` based on the other data, but I'll ignore that problem for now.

# Importing the data

```{r import-data}
clean_facility_names <- function(x) {
  # 1. Base conversion to Title Case
  x <- str_to_title(x)

  # 2. Typos and Truncations found in your specific list
  x <- str_replace_all(x, "Facili\\b", "Facility")       # Fixes #3
  x <- str_replace_all(x, "Processsing", "Processing")   # Fixes #155

  # 3. Expansion Map (Abbreviations -> Full Words)
  #    Using word boundaries (\\b) to prevent replacing inside other words.
  expansions <- c(
    "\\bDept\\.?\\b" = "Department",
    "\\bCtr\\.?\\b"  = "Center",
    "\\bCorr\\.?\\b" = "Correctional",
    "\\bInst\\.?\\b" = "Institution",
    "\\bFed\\.?\\b"  = "Federal",
    "\\bDet\\.?\\b"  = "Detention",
    "\\bFac\\.?\\b"  = "Facility",
    "\\bCo\\.?\\b"   = "County"
  )

  x <- str_replace_all(x, expansions)

  # 4. Acronym Restoration (Re-capitalizing)
  #    These were turned into "Ice", "Mdc" etc. by str_to_title
  acronyms <- c(
    "\\bOf\\b" = "of", # keep "of" lowercase
    "\\bUs\\b" = "US", # keep "US" uppercase; this would'nt work on general text
    "\\bIce\\b" = "ICE",
    "\\bEro\\b" = "ERO",
    "\\bMdc\\b" = "MDC",
    "\\bCca\\b" = "CCA",
    "\\bFci\\b" = "FCI",
    "\\bFdc\\b" = "FDC",
    "\\bJtf\\b" = "JTF",
    "\\bSsm\\b" = "SSM",
    "\\bTgk\\b" = "TGK",
    "\\bIpc\\b" = "IPC",
    "\\bSpc\\b" = "SPC",
    "\\bIah\\b" = "IAH",
    "\\bClipc\\b" = "CLIPC",
    "\\bDigsa\\b" = "DIGSA",
    "\\bIgsa\\b" = "IGSA"
  )

  x <- str_replace_all(x, acronyms)

  # 5. State Abbreviation Expansions
  #    Handles both (FL) and comma formats like ", NM"
  states <- c(
    "\\(Fl\\)" = "(Florida)",
    "\\(In\\)" = "(Indiana)",
    "\\(Mo\\)" = "(Missouri)",
    "\\(Mt\\)" = "(Montana)",
    "\\(Ne\\)" = "(Nebraska)",
    "\\(Ny\\)" = "(New York)",
    "\\(Tx\\)" = "(Texas)",
    "\\(Ut\\)" = "(Utah)",
    ", Nm\\b"  = ", New Mexico",
    ", Mt\\b"  = ", Montana"
  )

  x <- str_replace_all(x, states)
  
  # Facility Name Specific Fixes
  names <- c(
    "CCA, Florence Correctional Center" = "Central Arizona Florence Correctional Complex",
    "Eloy Federal Contract Facility" = "Eloy Detention Center",
    "T Don Hutto Detention Center" = "T. Don Hutto Detention Center",
    "Torrance/Estancia, New Mexico" = "Torrance County Detention Facility",
    "Houston Contract Detention Facility" = "Houston Processing Center",
    "Elizabeth Contract Detention Facility" = "Elizabeth Detention Center",
    "Folkston D Ray ICE Processing Center" = "Folkston ICE Processing Center",
     "Robert A Deyton Detention Facility" = "Robert A. Deyton Detention Center",
    "ERO El Paso Camp East Montana" = "Camp East Montana",
    "Port Isabel SPC" = "Port Isabel Detention Center",
    "Bluebonnet Detention Facility" = "Bluebonnet Detention Center"
  )
  
  x <- str_replace_all(x, names)

  return(x)
}


# 4. Import the "Facilities" sheet
# facilities_data <- read_xlsx(local_file, sheet = sheet_name)

read_facilities_data <- function(data_file_info_row){
  with(data_file_info_row, {
    read_xlsx(
      local_file,
      sheet = sheet_name,
      range = cell_limits(ul = c(first_data_row, 1), lr = c(NA, right_column_num)),
      col_names = clean_names_list[[year_name]] # Use precomputed clean names
                                                # A more elegant programming approach,
                                                # might incorporate them into data_file_info
    )
  })
}

fix_city_names <- function(x) {
  x <- str_replace_all(x, "Ft.lauderdale", "Fort Lauderdale")
  x <- str_replace_all(x, "Ft. Lauderdale", "Fort Lauderdale")
  x <- str_replace_all(x, "Cottonwood Fall", "Cottonwood Falls")
  x <- str_replace_all(x, "Sault Ste Marie", "Sault Ste. Marie")
  x <- str_replace_all(x, "Bunkerhill", "Bunker Hill")
  x <- str_replace_all(x, "Mcelhattan", "McElhattan") 
  x <- str_replace_all(x, "Mcfarland", "McFarland")
  x <- str_replace_all(x, "Mccook", "McCook")
  x <- str_replace_all(x, "Mccall", "McCall")
  x
}

clean_facilities_data <- function(facilities_data) {
  numerical_columns = c("facility_average_length_of_stay_alos", "adp_detainee_classification_level_a",
                        "adp_detainee_classification_level_b", "adp_detainee_classification_level_c",
                        "adp_detainee_classification_level_d", "adp_criminality_male_crim",
                        "adp_criminality_male_non_crim", "adp_criminality_female_crim",
                        "adp_criminality_female_non_crim", "adp_ice_threat_level_1",
                        "adp_ice_threat_level_2", "adp_ice_threat_level_3", "adp_no_ice_threat_level",
                        "adp_mandatory", "inspections_guaranteed_minimum")
  numerical_columns_present <- numerical_columns[numerical_columns %in% names(facilities_data)]

  facilities_data_clean <- facilities_data %>%
    mutate(
      # 1. Convert specific columns to Numeric and round to 1 decimal
      across(
        .cols = facility_average_length_of_stay_alos:adp_mandatory,
        .fns = ~ round(as.numeric(.), 1)
      ),

      # 2. Convert Guaranteed Minimum to Integer
      inspections_guaranteed_minimum = as.integer(inspections_guaranteed_minimum),

      # Standardize Last Inspection Type values
      inspections_last_inspection_type = case_when(
        inspections_last_inspection_type %in% c("PRE-OCCUPANCY", "Pre-Occupancy") ~
          "PREOCC", # Standardize Pre-Occupancy
        TRUE ~ inspections_last_inspection_type
      ),

      # Adjust capitalization
      facility_name = clean_facility_names(facility_name),
      facility_city = str_to_title(facility_city),
      facility_address = str_to_title(facility_address),
      facility_male_female = str_to_title(facility_male_female),
      
      # Some minor name clean up
      facility_city = fix_city_names(facility_city),

      # 4. Ensure all other columns are Characters (as requested)
      #    We exclude the columns we just processed to avoid overwriting them
      across(
        .cols = -numerical_columns_present,
        .fns = as.character
      )
    ) 
  guantanamo_patch <- tibble(
    facility_name = "JTF Camp Six",
    facility_city = "Guantanamo Bay",
    facility_state = "Cuba",
    facility_address = "Bldg 2144, U.S. Naval Station Guantanamo Bay",
  )
  facilities_data_clean <- facilities_data_clean %>%
    rows_update(guantanamo_patch, by = "facility_name", unmatched = "ignore")
      # ignore because the Guantanamo patch doesn't apply to earlier years
    
  facilities_data_clean
}

aggregate_facilties_data <- function(facilities_data) {
  facilities_data <- facilities_data %>%
    mutate(
      sum_classification_levels = rowSums(
        select(., starts_with("adp_detainee_classification_level_")),
        na.rm = TRUE
      ),
      sum_criminality_levels = rowSums(
        select(., starts_with("adp_criminality_")),
        na.rm = TRUE
      ),
      sum_threat_levels = rowSums(
        select(., starts_with("adp_ice_threat_level_")|(starts_with("adp_no_ice_threat_level"))),
        na.rm = TRUE
      )
    ) %>%
    mutate(
      share_non_crim = (adp_criminality_male_non_crim + adp_criminality_female_non_crim) / sum_criminality_levels,
      share_no_threat = ( adp_no_ice_threat_level) / sum_threat_levels
    ) %>%
  # Assign type
    mutate(
      facility_type_wiki = case_when(
        str_detect(facility_type_detailed, "IGSA") ~ "Jail",
        str_detect(facility_type_detailed, "USMS IGA") ~ "Jail",
        str_detect(facility_type_detailed, "DIGSA") ~ "Dedicated Migrant Detention Center",
        str_detect(facility_type_detailed, "STATE") ~ "State Migrant Detention Center",
        str_detect(facility_type_detailed, "BOP") ~ "Federal Prison",
        str_detect(facility_type_detailed, "Family") ~ "Private Family Detention Center",
        str_detect(facility_type_detailed, "CDF") ~ "Private Migrant Detention Center",
        str_detect(facility_type_detailed, "USMS CDF") ~ "Private Migrant Detention Center",
        str_detect(facility_type_detailed, "SPC") ~ "ICE Migrant Detention Center",
        str_detect(facility_type_detailed, "STAGING") ~ "ICE Short-Term Migrant Detention Center",
        str_detect(facility_type_detailed, "DOD") ~ "Military Detention Center",
        TRUE ~ "Other"
      )
    )
}

# facilities_data_list <- map(1:nrow(data_file_info), function(i)
#   read_facilities_data(data_file_info[i, ])
# )
# names(facilities_data_list) <- 

facilities_data_list <- map(1:nrow(data_file_info), function(i)
  read_facilities_data(data_file_info[i, ]) %>%
    clean_facilities_data() %>%
    aggregate_facilties_data()
)
names(facilities_data_list) <- data_file_info$year_name

```

```{r facility-name-alignment}
# Check for facility name alignment across years
facility_names_list <- map(facilities_data_list, ~ {
  if ("facility_name" %in% names(.x)) {
    unique(.x$facility_name)
  } else {
    character(0)
  }
})

names(facility_names_list) <- names(clean_names_list)

upset_plot_2 <- upset(fromList(facility_names_list),
                    sets = names(facility_names_list),
                    keep.order = T,
                    order.by = "degree")  # Visual intersection analysis
upset_plot_2

presence_matrix_facilities <- facility_names_list %>%
  imap_dfr(~ tibble(facility_name = .x, list_name = .y)) %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = list_name, values_from = present, values_fill = FALSE)

presence_matrix_facilities <- presence_matrix_facilities %>%
  mutate(num_lists_present = rowSums(select(., -facility_name)))

presence_matrix_facilities %>% count(num_lists_present)

presence_matrix_facilities %>%
  filter((num_lists_present == 1) & (FY26)) %>%
  pull(facility_name)

all_facility_names <- unlist(facility_names_list, use.names = FALSE)

current_facility_names <- unique(facility_names_list$FY26)

```

# Harmonize Names with Wikipedia Page
```{r harmonize-with-wikipedia}
library(httr)
library(rvest)

# 1. Define URL and Fetch Page
url <- "https://en.wikipedia.org/w/index.php?title=List_of_immigrant_detention_sites_in_the_United_States&oldid=1334453072"
response <- GET(url)
page <- read_html(content(response, as = "text"))

# 2. Select the target table node
table_node <- html_element(page, "table.wikitable")

# 3. Extract the standard table text
w_detention_sites_tb <- html_table(table_node)

# 4. Extract links from the first column specifically
# We select all rows, then extract the 'href' from the 'a' tag in the first 'td'
rows <- html_elements(table_node, "tr")

# Note: We use map_chr to ensure we handle rows without links (like headers) correctly
links <- map_chr(rows, function(row) {
  # Look for the first data cell, then an anchor tag inside it
  link_node <- html_element(row, "td:first-child a")
  href <- html_attr(link_node, "href")
  
  if (is.na(href)) {
    return(NA_character_)
  } else {
    # Construct full URL
    return(paste0("https://en.wikipedia.org", href))
  }
})

# 5. Clean and Merge
# Remove the first element of links if it corresponds to the header row
if (length(links) > nrow(w_detention_sites_tb)) {
  links <- links[(length(links) - nrow(w_detention_sites_tb) + 1):length(links)]
}

w_detention_sites_tb <- w_detention_sites_tb %>%
  add_column(link = links, .after = 1)%>%
  mutate(linked_article = case_when(
    str_detect(link, "/wiki/") ~ {
      # Extract everything after /wiki/
      slug <- str_extract(link, "(?<=/wiki/).*")
      str_replace_all(slug, "_", " ")
    },
    TRUE ~ NA_character_),
    #extract text befoe any parentheses or brackets for the primary name
    primary_name = str_extract(Name, "^[^\\(]+") # Extract text before any parentheses
  )

clean_variable_names <- function(df) {
  df %>%
    rename_with(~ str_replace_all(., "[^a-zA-Z0-9]+", "_")) %>%
    rename_with(~ str_remove_all(., "^_+|_+$")) %>%
    rename_with(tolower)
}

w_detention_sites_tb <- clean_variable_names(w_detention_sites_tb) %>%
  mutate(
    wiki_slug = case_when(
      str_detect(link, "/wiki/") ~ {
        # Extract everything after /wiki/
        slug <- str_extract(link, "(?<=/wiki/).*")
        str_replace_all(slug, "_", " ")
      },
      TRUE ~ ""  # Return empty string if /wiki/ is not found
    )
  ) %>%
  separate(location, into = c("city", "state"), 
             sep = ",\\s*", extra = "merge", fill = "right")

# Optional: Final cleanup of text and brackets
w_detention_sites_tb <- w_detention_sites_tb %>%
  mutate(across(where(is.character), ~ gsub("\\[\\d+\\]", "", .)))

print(w_detention_sites_tb)

```

```{r check-wikipedia-alignment}
compare_ice_wiki_duplicates <- function(ice_df, wiki_df) {
  
  # 1. Prepare ICE Data
  ice_clean <- ice_df %>%
    transmute(
      original_name = facility_name,
      clean_name  = str_to_upper(str_trim(facility_name)),
      clean_city  = str_to_upper(str_trim(facility_city)),
      # Standardizing state to full name upper case to match wiki
      clean_state = str_to_upper(state.name[match(facility_state, state.abb)]),
      source      = "ICE"
    )

  # 2. Prepare Wikipedia Data
  wiki_clean <- wiki_df %>%
#    separate(location, into = c("city", "state"), 
#             sep = ",\\s*", extra = "merge", fill = "right") %>%
    transmute(
      original_name = name, # Use lowercase 'name'
      clean_name  = str_to_upper(str_trim(name)),
      clean_city  = str_to_upper(str_trim(city)),
      clean_state = str_to_upper(str_trim(state)),
      source      = "Wiki"
    )

  # 3. Combine and find location-based variants
  comparison <- bind_rows(ice_clean, wiki_clean) %>%
    filter(!is.na(clean_city), !is.na(clean_state)) %>%
    group_by(clean_city, clean_state) %>%
    summarise(
      num_variants = n_distinct(clean_name),
      in_both = all(c("ICE", "Wiki") %in% source),
      
      # Extract names assigned to ICE source
      ice_variants = paste(unique(original_name[source == "ICE"]), collapse = " | "),
      
      # Extract names assigned to Wiki source
      wiki_variants = paste(unique(original_name[source == "Wiki"]), collapse = " | "),
      
      # Keep the combined version for reference
      all_name_variants = paste(unique(original_name), collapse = " | "),
      
      sources = paste(unique(source), collapse = " & "),
      .groups = "drop"
    ) %>%
    filter(num_variants > 1) %>%
    mutate(location = paste(clean_city, clean_state, sep = ", ")) %>%
    select(location, num_variants, in_both, ice_variants, wiki_variants, all_name_variants, sources) %>%
    arrange(desc(num_variants), desc(in_both))

  return(comparison)
}

names_with_wikilinks <- w_detention_sites_tb %>%
  filter(wiki_slug != "") %>%
  pull(name)

names_with_management_info <- w_detention_sites_tb %>%
  filter(management != "") %>%
  pull(name)

duplicate_table <- compare_ice_wiki_duplicates(facilities_data_list$FY26,
                                               w_detention_sites_tb) %>%
  filter(in_both) %>% 
  filter(num_variants >= 2) %>% 
  select(ice_variants, wiki_variants, location) %>%
  mutate(has_wiki_link = case_when(
    str_detect(wiki_variants, "\\|") ~ FALSE,  # "multiple wiki variants",
    str_trim(wiki_variants) %in% names_with_wikilinks ~ TRUE,
    TRUE ~ FALSE)) %>%
  mutate(has_management_info = case_when(
    str_detect(wiki_variants, "\\|") ~ FALSE,  # "multiple wiki
    str_trim(wiki_variants) %in% names_with_management_info ~ TRUE,
    TRUE ~ FALSE))

wiki_name_match_table <- duplicate_table %>%
  filter(!str_detect(ice_variants, "\\|")) %>%
  filter(!str_detect(wiki_variants, "\\|")) %>% 
  filter(has_wiki_link | has_management_info)

# hand verified
does_not_match <- c("Farmville Detention Center", 
                    "Dallas County Jail - Lew Sterrett Justice Center",
                    "Leavenworth US Penitentiary",
                    "IAH Secure Adult Detention Facility (Polk)",
                    "FDC Philadelphia",
                    "El Valle Detention Facility",
                    "Greene County Jail")

wiki_name_match_table <- wiki_name_match_table %>% 
  filter(!(ice_variants %in% does_not_match))

nrow(wiki_name_match_table) 

```

Current status: 
- 36 city/state matches, most of which are the same facility with different names (`duplicate_table %>% filter(in_both) %>% filter(num_variants ==2) %>% select(4, 5) %>% print(n=100)`)
- 50 with the same name in FY26 ICE data and on Wikipedia (` which(current_facility_names %in% w_detention_sites_tb$Name)`)

But 223 are listed for FY26 altogether and 395 are on the Wikipedia page. So a lot more to characterize. (Most of these might by in other places and the city/state data might be enough for that; just 17 on the Wikipedia page lack a location `w_detention_sites_tb %>% filter(Location=="")`, but 15 of these have a link.)

Next step is to check for how many of the items with Wikipedia links are in these sets, because those matches are most important.

Further step is to merge these tables together and integrate the Wikipedia links and useful Wikipedia data, which is mostly the Management column.

```{r search-wikipedia-for-matches-to-fy26}
source("src/add-wikipedia-matches.R")

# To generate this table, we're actually inserting this code
# AFTER the next chunk (add-match-table-to-fy26-ice-data), but
# it's easier to insert the results here instead of making a 
# more complex case_when statement.

# Heads up: this is a very slow function call!
# facilities_fy26_wiki_matches <- facilities_data_merged$FY26 %>%
#  add_wikipedia_matches(name_col = "facility_name")
# saveRDS(facilities_fy26_wiki_matches, here::here("data", "facilities_fy26_wiki_matches.rds"))
#
facilities_fy26_wiki_matches <- readRDS(here::here("data", "facilities_fy26_wiki_matches.rds"))

facilities_fy26_wiki_matches %>% 
  count(!is.na(wiki_match), wikipedia_match) %>%
  rename(previous_match = 1,
         match_by_wikipedia_search = 2)
  

facilities_fy26_wiki_matches %>% 
  filter(wikipedia_match) %>% 
  select(1, facility_state, 36) 

exclude_list <- c("Boone County Jail",
                  "Davis County",
                  "Douglas County",
                  "Franklin County Jail", # needs new page for new jail in C, PA
                  "Jefferson County Jail", # no page for Idaho jail
                  "Lawrence County",
                  "Pottawattamie County Jail", # references historic site; needs new page
                  "Southern Regional"
                  )

facilities_fy26_wiki_match_list <- facilities_fy26_wiki_matches %>%
  filter(wikipedia_match) %>% 
  filter(!is.na(wiki_match)) %>% #new matches only
  filter(!facility_name %in% exclude_list) %>%
  select(1, wikipedia_title) 



# Notes on better names:
# "Davis County" > "Davis County Correctional Facility" (Utah) / no wiki page
# "Southern Regional" > "Southern Regional Jail and Correctional Facility"

```

```{r add-match-table-to-fy26-ice-data}
# Generate a match table based on city/state matches and Wikipedia links


# Add a match column to the FY26 ICE data 
# indicating if there's a Wikipedia match
facilities_data_merged <- list()
facilities_data_merged$FY26 <- facilities_data_list$FY26 %>%
  mutate(
    clean_state = state.name[match(facility_state, state.abb)]
  ) %>%
  mutate(wiki_match = case_when(
    # Condition 1: Exact match in the main wiki table
    facility_name %in% w_detention_sites_tb$name ~ facility_name,
    
    # Condition 2: Match via the lookup table
    facility_name %in% wiki_name_match_table$ice_variants ~ 
      wiki_name_match_table$wiki_variants[match(facility_name, wiki_name_match_table$ice_variants)],
    
    # Condition 3: Match via the Wikipedia search function
    facility_name %in% facilities_fy26_wiki_match_list$facility_name ~
      facilities_fy26_wiki_match_list$wikipedia_title[match(facility_name, facilities_fy26_wiki_match_list$facility_name)],
    
    TRUE ~ NA_character_
  )) %>%
  left_join(
    select(w_detention_sites_tb, name, state, wiki_slug, management),
    by = c("wiki_match" = "name", "clean_state" = "state")
  ) 

facilities_data_merged$FY26 %>% 
  filter(!is.na(wiki_match)) %>%
  relocate(facility_name, wiki_match, wiki_slug, management) %>% print(n=100)
  
# Hunting for extra matches to perfect this patch
#
# ice_fy26_locations <- facilities_data_merged$FY26 %>% 
#   transmute(facility_name, location = paste0(facility_city, ", ", facility_state))
# 
# contractors_fy26_patch %>%
#   mutate(matched = facility_name %in% facilities_data_merged$FY26$facility_name) %>% 
#   filter(!matched) %>% 
#   mutate(matched_location = location %in% ice_fy26_locations$location) %>%
#   mutate(matched_ice_facility = ice_fy26_locations$facility_name[match(location, ice_fy26_locations$location)]) %>%
#   arrange(matched_location) %>%
#   filter(matched_location) %>%
#   select(1, 6, 2) -> contractors_fy26_location_matches
# 
# contractors_fy26_location_matches[c(5, 8, 16), c(1,2)] %>% dput

contractors_fy26_patch <-
  read.csv(here::here("data","contractors-patch-fy26.csv")) 
names(contractors_fy26_patch) <- c("facility_name", "location", "management")

name_fixes <- tibble(facility_name = c("Dallas Transitional Center", 
                       "West Tennessee Detention Facility"),
                     matched_ice_facility = c("Dallas County Jail - Lew Sterrett Justice Center", 
                            "Western Tennessee Detention Facility")
                     )

contractors_fy26_patch <- contractors_fy26_patch %>%
  left_join(name_fixes, by = "facility_name") %>%
  mutate(facility_name = coalesce(matched_ice_facility, facility_name)) %>%
  select(-matched_ice_facility)


facilities_data_merged$FY26 <- facilities_data_merged$FY26 %>%
  rows_update(select(contractors_fy26_patch, facility_name, management),
              by ="facility_name",
              unmatched = "ignore") 
```



```{r format-wikitable}
facilities_wikitable_from_merged <- function(facilties_data, year_name) {
  if (!"wiki_match" %in% names(facilties_data)) {
    facilities_data <- facilities_data %>%
      mutate(
        wiki_match = NA_character_
      )
  }
  
  wiki_table <- facilties_data %>%
    mutate(
      facility_name_wiki = case_when(
        facility_name == wiki_slug ~ paste0("[[", facility_name, "]]"),
        wiki_slug != "" ~ paste0("[[", wiki_slug, "|", facility_name, "]]"),
        (wiki_match != "") & (!is.na(wiki_match)) ~ paste0("[[", wiki_match, "|", facility_name, "]]"),
        TRUE ~ facility_name
      ),
      city_state = paste0(facility_city, ", ", facility_state),
      city_state_wiki = paste0("[[", city_state, "]]"),
      status = paste0("In use (", year_name, ")"),
      location = city_state_wiki,
      authority = facility_type_detailed,
      #           management = "",
      average_daily_population = round(sum_criminality_levels),
      minimum_capacity = inspections_guaranteed_minimum,
      demographics = facility_male_female) %>%
    select(
      facility_name_wiki,
      status,
      location,
      facility_type_wiki,
      authority,
      management,
      average_daily_population,
      minimum_capacity,
      demographics)
} 


  
# Function to print a data frame as a MediaWiki table

table_column_names <- c(
    "Facility Name",
    "Status (year)",
    "Location",
    "Facility Type",
    "Authority",
    "Management",
    "Average Daily Population",
    "Minimum Capacity",
    "Demographics"
  )

get_wikitable <- function(df, caption = NULL, class = "wikitable sortable",
                            column_names = NULL) {
  # Initialize an empty character vector to store lines
  out <- c()

  # 1. Start Table
  out <- c(out, paste0('{| class="', class, '"'))

  # 2. Add Caption (Optional)
  if (!is.null(caption)) {
    out <- c(out, paste0("|+ ", caption))
  }

  # 3. Header Row
  # Use provided column_names if available, otherwise use dataframe names
  if (is.null(column_names)) {
    column_names <- names(df)
  }
  
  # Use ! for headers, separating with !!
  out <- c(out, paste0("! ", paste(column_names, collapse = " !! ")))

  # 4. Data Rows
  formatted_rows <- apply(df, 1, function(row) {
    # Convert to character and escape pipes
    row_clean <- gsub("\\|", "|", as.character(row))
    # Replace NAs with empty string
    row_clean[is.na(row_clean)] <- ""

    paste0("| ", paste(row_clean, collapse = " || "))
  })

  # Add each row preceded by the row separator |-
  # Interleave "|-" with the formatted row strings
  out <- c(out, paste0("|-\n", formatted_rows))

  # 5. End Table
  out <- c(out, "|}")

  # Combine all elements into a single string separated by newlines
  return(paste(out, collapse = "\n"))
}
```

```{r enable-copy-to-wikitext, echo=FALSE}
source(here::here("src","gh-copy-assets.R"), local = knitr::knit_global())
include_table_assets() # This injects the CSS and JS once
```

```{r output-wikitable}
facilities_wikitable <- facilities_data_merged$FY26 %>% 
  facilities_wikitable_from_merged(year_name = "FY26")





year_name <- "FY26"
get_wikitable(facilities_wikitable, 
                  caption = paste0("ICE Detention Facilities in ", year_name),
                  column_names = table_column_names) %>%
  copy_to_clipboard_button_wt()
```

Update complete.

```{r treemap}
library(treemap)
library(ggplot2)

facilities_data_current <- facilities_data_merged$FY26 %>%
  mutate(
    adp = rowSums(select(., starts_with("adp_criminality_")), na.rm = TRUE),
    management = replace_na(management, "")
  )

facilities_data_current %>%
  group_by(facility_type_wiki, facility_state) %>%
  summarise(total_adp = sum(adp))

library(treemapify)
library(dplyr)

# 1. Prepare your data
# We group by both levels to get the aggregate 'value' for each state within each type
library(dplyr)
library(tidyr) # for replace_na

plot_data <- facilities_data_current %>%
  # 1. First, handle the NAs in management so the logic works
  mutate(management = replace_na(as.character(management), "")) %>%
  
  # 2. Group by ALL variables you need for the final label
  group_by(facility_type_wiki, facility_state, facility_name, management) %>%
  
  # 3. Sum the ADP
  summarise(total_adp = sum(adp, na.rm = TRUE), .groups = "drop") %>%
  
  # 4. Filter and Round
  filter(total_adp > 0) %>%
  mutate(total_adp = round(total_adp)) %>%
  
  # 5. Create the label now that management is present in the dataframe
  mutate(display_label = case_when(
    management == "" ~ paste0(facility_name, " (", facility_state, ")\n", 
                              format(total_adp, big.mark = ",")),
    TRUE             ~ paste0(facility_name, " (", management, " / ", facility_state, ")\n", 
                              format(total_adp, big.mark = ","))
  ))

# 2. Create the Plot
p <- ggplot(plot_data, aes(
    area = total_adp, 
    fill = facility_type_wiki, 
    label = display_label, 
    subgroup = facility_type_wiki
  )) +
  # Main tiles - thin white lines for the "leaf" nodes
  geom_treemap(color = "white", size = 0.5) + 
  
  # Thicker borders to define the Facility Type groups
  geom_treemap_subgroup_border(color = "white", size = 2.5) +
  
  # Child labels (State + ADP) in top-left
  geom_treemap_text(
    colour = "black", 
    place = "topleft", 
    size = 9, 
    grow = FALSE, 
    reflow = TRUE,
    padding.x = unit(1.5, "mm"),
    padding.y = unit(1.5, "mm"),
    min.size = 4 # Smallest font size to allow before hiding text
  ) +
  
  # Visual Styling
  scale_fill_brewer(palette = "Set3") + # Soft, categorical D3-like palette
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = "Average Daily Population of detainees by facility type",
    fill = "Facility Type",
    caption = "Box area is proportional to Total ADP"
  )

print(p)

```

```{r treemap-2}
# 1. Data Preparation
plot_data <- facilities_data_current %>%
  # Handle NAs immediately
  mutate(management = replace_na(as.character(management), "")) %>%
  # Group by the three nesting levels + management for the label
  group_by(facility_state, facility_type_wiki, facility_name, management) %>%
  summarise(total_adp = sum(adp, na.rm = TRUE), .groups = "drop") %>%
  filter(total_adp > 0) %>%
  mutate(
    total_adp = round(total_adp),
    # Constructing the requested label
    display_label = case_when(
      management == "" ~ paste0(facility_name, " (", facility_state, ")\n", 
                                format(total_adp, big.mark = ",")),
      TRUE             ~ paste0(facility_name, " (", management, " ", facility_state, ")\n", 
                                format(total_adp, big.mark = ","))
    )
  )

# 2. Create the Triple-Nested Treemap
p <- ggplot(plot_data, aes(
    area = total_adp, 
    fill = facility_state,          # Color by State for high-level visual grouping
    label = display_label, 
    subgroup = facility_state,      # Level 1: State
    subgroup2 = facility_type_wiki  # Level 2: Type
  )) +
  # Main tiles for individual facilities
  geom_treemap(color = "white", size = 0.5) + 
  
  # Thick borders for States
  geom_treemap_subgroup_border(color = "white", size = 4) +
  
  # Thinner borders for Facility Types within states
  geom_treemap_subgroup2_border(color = "white", size = 1.5, linetype = "dashed") +
  
  # State Labels (Upper Left of the state block)
  geom_treemap_subgroup_text(
    place = "topleft", 
    grow = FALSE, 
    alpha = 0.7, 
    colour = "white", 
    fontface = "bold",
    size = 14
  ) +
  
  # Individual Facility Labels (The detailed labels)
  geom_treemap_text(
    colour = "black", 
    place = "topleft", 
    size = 9,             # Smaller size for high density
    grow = FALSE, 
    reflow = TRUE,
    padding.x = unit(1, "mm"),
    padding.y = unit(1, "mm"),
    min.size = 3
  ) +
  
  # Styling
  scale_fill_viridis_d(option = "turbo", guide = "none") + 
  theme_minimal() +
  labs(
    title = "Facility ADP: Nested by State and Type",
    subtitle = "Solid lines: State | Dashed lines: Facility Type",
    caption = "Area represents Total ADP"
  )

print(p)
```
