---
title: "Indigenous identity in LAPOP 2023 and the Bolivia Census 2024"
format:
  html:
    toc: true
    toc-depth: 2
    df-print: kable
execute:
  warning: false
  message: false
---

## Background

The LAPOP AmericasBarometer 2023 and the Bolivia Census 2024 (CPV 2024) both
ask about indigenous identity and language, but with different instruments,
populations, and sample sizes. This document maps the questions across the two
sources and compares the resulting distributions, looking for where they agree
and where they diverge.

LAPOP interviewed 1,706 Bolivian adults (18+) face-to-face in 2023. The census
enumerated 11.4 million people in 2024, of whom 7.7 million are 18 or older.
All census comparisons below restrict to the 18+ population to match LAPOP's
sampling frame.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
library(arrow)
library(tidyverse)

DATA_DIR  <- "../bolivia-data/Censo 2024/base_datos_csv_2024"
CENSO_DIR <- "../bolivia-data/Censo 2024"

lapop <- readRDS("data/lapop_bol_2023.rds")
idioma_cats <- readRDS("data/idioma_cats.rds")
ds <- open_dataset(file.path(DATA_DIR, "Persona_CPV-2024.csv"),
                   format = "csv", delimiter = ";")
```

------------------------------------------------------------------------

## The LAPOP identity and language questions

LAPOP asks five questions that touch on indigenous identity and language. They
are asked in sequence during the interview, and two of them (`boletidnewb`,
`leng4`) are conditional on earlier answers.

### 1. Racial/ethnic self-classification (`etid`)

> *¿Usted se considera una persona blanca, mestiza, indígena u originaria,
> negra, mulata, u otra?*

This is a closed-ended question with six response options. It measures
**perceived racial/ethnic category** — not belonging to a specific indigenous
people.

```{r}
#| label: tbl-etid
#| tbl-cap: "LAPOP: Racial/ethnic self-classification"
lapop |>
  filter(!is.na(etid)) |>
  count(etid_f) |>
  mutate(pct = n / sum(n) * 100) |>
  rename(Category = etid_f, N = n, `%` = pct)
```

Two-thirds of respondents identify as *mestiza*. Only 14% choose
*indígena-originaria* — but as we will see, a much larger share says "yes" to
belonging to an indigenous people.

### 2. Indigenous belonging (`boletidnew`)

> *¿Usted se considera perteneciente a algún pueblo indígena u originario o
> afroboliviano?*

This is a yes/no question that is asked of everyone, regardless of their answer
to `etid`. It captures **identification with a named indigenous people**, which
is conceptually broader than the racial self-categorization above.

```{r}
#| label: tbl-boletidnew
#| tbl-cap: "LAPOP: Indigenous belonging"
lapop |>
  filter(!is.na(boletidnew)) |>
  count(indig_id_f) |>
  mutate(pct = n / sum(n) * 100) |>
  rename(Response = indig_id_f, N = n, `%` = pct)
```

54% say yes — nearly four times the share that chose *indígena-originaria* in
`etid`. This gap is well documented: many Bolivians who consider themselves
mestizo in racial terms simultaneously identify as belonging to a Quechua or
Aymara people. The cross-tabulation confirms this:

```{r}
#| label: tbl-etid-indig
#| tbl-cap: "LAPOP: Cross-tabulation of racial category and indigenous belonging"
lapop |>
  filter(!is.na(etid), !is.na(boletidnew)) |>
  count(etid_f, indig_id_f) |>
  group_by(etid_f) |>
  mutate(pct_row = n / sum(n) * 100) |>
  ungroup() |>
  select(etid_f, indig_id_f, n, pct_row) |>
  pivot_wider(names_from = indig_id_f, values_from = c(n, pct_row)) |>
  rename(`Race/ethnicity` = etid_f)
```

Half of all self-identified *mestizos* also say they belong to an indigenous
people. Even 41% of those who chose *blanca* do so — suggesting the two
questions tap quite different dimensions of identity.

### 3. Indigenous group (`boletidnewb`)

> *¿A cuál de los siguientes pueblos indígenas u originarios o afroboliviano
> pertenece?*

Asked only of those who answered "Sí" to `boletidnew`. Response options:
Quechua, Aymara, Guaraní, Chiquitano, Mojeño, Afroboliviano, Otro, or
"no especifica."

```{r}
#| label: tbl-boletidnewb
#| tbl-cap: "LAPOP: Indigenous group (among self-identifiers)"
lapop |>
  filter(boletidnew == 1, !is.na(boletidnewb)) |>
  count(indig_group_f) |>
  mutate(pct = n / sum(n) * 100) |>
  arrange(desc(n)) |>
  rename(Group = indig_group_f, N = n, `%` = pct)
```

### 4. Mother tongue (`leng1`)

> *¿Cuál es su lengua materna, es decir, el primer idioma que aprendió en su
> niñez?*

Six response categories: Castellano, Quechua, Aymara, Guaraní, Otro (nativo),
Otro extranjero.

```{r}
#| label: tbl-leng1
#| tbl-cap: "LAPOP: Mother tongue"
lapop |>
  filter(!is.na(leng1)) |>
  count(leng_materna_f) |>
  mutate(pct = n / sum(n) * 100) |>
  rename(`Mother tongue` = leng_materna_f, N = n, `%` = pct)
```

### 5. Parents' languages (`leng4`)

> *¿Qué idiomas hablaban normalmente tus padres en el hogar cuando eras niño/a?*

This question captures intergenerational language transmission rather than the
respondent's own language. It has five categories.

```{r}
#| label: tbl-leng4
#| tbl-cap: "LAPOP: Parents' language in childhood home"
lapop |>
  filter(!is.na(leng4)) |>
  count(leng_padres_f) |>
  mutate(pct = n / sum(n) * 100) |>
  rename(`Parents' language` = leng_padres_f, N = n, `%` = pct)
```

59% of respondents grew up in homes where at least one indigenous language was
spoken (categories 2 + 3), even though only 25% report an indigenous mother
tongue. This gap captures the common pattern of passive bilingualism — parents
spoke Quechua or Aymara, but the child acquired Castellano as their first
language.

------------------------------------------------------------------------

## Census 2024 equivalents

The census asks comparable questions but with finer granularity and a much
larger population. The main equivalences are:

| Concept | LAPOP variable | Census variable |
|---------|---------------|----------------|
| Indigenous self-identification | `boletidnew` | `p32_pueblo_per` |
| Indigenous group | `boletidnewb` (8 options) | `p32_pueblos` (58 groups) |
| Mother tongue | `leng1` (6 categories) | `idioma_mat` (83 languages) |
| Racial/ethnic category | `etid` | *(not asked)* |
| Parents' language | `leng4` | *(not asked)* |

::: {.callout-note}
The census does not ask about racial self-classification (`etid`) — this is a
LAPOP-only question. Conversely, the census asks about *current spoken
languages* (`p331`–`p333`) and *language of greatest daily use*
(`idioma_mayor_uso`), which LAPOP does not.
:::

------------------------------------------------------------------------

## Comparison: Indigenous self-identification {#sec-selfid}

```{r}
#| label: census-selfid
census_selfid <- ds |>
  filter(p26_edad >= 18, p32_pueblo_per %in% c(1L, 2L)) |>
  group_by(p32_pueblo_per) |>
  summarise(n = n()) |>
  collect() |>
  mutate(pct = n / sum(n) * 100)
```

```{r}
#| label: tbl-selfid-compare
#| tbl-cap: "Indigenous self-identification: LAPOP 2023 vs Census 2024 (18+)"
lapop_selfid <- lapop |>
  filter(boletidnew %in% c(1, 2)) |>
  count(boletidnew) |>
  mutate(pct = n / sum(n) * 100,
         response = ifelse(boletidnew == 1, "Self-identifies", "Does not"))

comparison_selfid <- tibble(
  Response = c("Self-identifies", "Does not"),
  `LAPOP %` = c(
    lapop_selfid$pct[lapop_selfid$response == "Self-identifies"],
    lapop_selfid$pct[lapop_selfid$response == "Does not"]
  ),
  `Census %` = c(
    census_selfid$pct[census_selfid$p32_pueblo_per == 1],
    census_selfid$pct[census_selfid$p32_pueblo_per == 2]
  )
)

comparison_selfid |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

LAPOP yields a substantially higher self-identification rate (54% vs 42%) than
the census. Several factors likely contribute:

1. **Question wording**: LAPOP asks about *belonging* to a pueblo
   ("¿se considera perteneciente...?"), while the census asks whether the
   respondent *self-identifies* with one ("¿se autoidentifica...?"). The
   framing of belonging may be more inclusive.
2. **Interviewer effects**: Face-to-face LAPOP interviews involve a trained
   interviewer who reads the question aloud, while the census is
   self-administered or enumerator-assisted with many other questions. Social
   desirability effects may differ.
3. **Timing**: The LAPOP fieldwork (2023) and the census (2024) are close in
   time, so temporal changes are unlikely to explain the full gap.

------------------------------------------------------------------------

## Comparison: Indigenous group

The census `p32_pueblos` variable distinguishes 58 indigenous groups (plus
generic and foreign categories). LAPOP collapses these into 8. To compare, we
map the census groups to the LAPOP scheme:

- Census Quechua → LAPOP Quechua
- Census Aymara → LAPOP Aymara
- Census Guaraní → LAPOP Guaraní
- Census Chiquitano → LAPOP Chiquitano
- Census Mojeño + Mojeño Ignaciano + Mojeño Trinitario → LAPOP Mojeño
- Census Afroboliviano → LAPOP Afroboliviano
- Census "Otras declaraciones" → LAPOP "No especifica, indígena nomás"
- All others → LAPOP "Otro"

```{r}
#| label: census-group
pueblos_cats <- tribble(
  ~code, ~label,
  1,  "Afroboliviano",       2,  "Araona",            3,  "Aymara",
  4,  "Ayoreo",              5,  "Baure",             6,  "Canichana",
  7,  "Cavineño",            8,  "Cayubaba",          9,  "Chácobo",
  10, "Charka Qhara Qhara",  11, "Chichas",          12, "Chiquitano",
  13, "Chuwi",               14, "Ese Ejja",          15, "Guaraní",
  16, "Guarasu'we",          17, "Gwarayu",           18, "Itonama",
  19, "Jach'a Carangas",     20, "Jalq'a",            21, "Joaquiniano",
  22, "Kallawaya",           23, "Killacas",           24, "Leco",
  25, "Lípez",               26, "Lupaca",             27, "Machineri",
  28, "Maropa",              29, "Mojeño",             30, "Mojeño Ignaciano",
  31, "Mojeño Trinitario",   32, "Moré",               33, "Mosetén",
  34, "Movima",              35, "Pacahuara",          36, "Pakajaqi",
  37, "Paunaca",             38, "Puquina",            39, "Qhapaq Uma Suyu",
  40, "Quechua",             41, "Qullas",             42, "Raqaypampa",
  43, "Sirionó",             44, "Sora",               45, "Tacana",
  46, "Tapiete",             47, "Toromona",           48, "Tsimane'",
  49, "Uru-Chipaya",         50, "Weenhayek",          51, "Yaminawa",
  52, "Yampara",             53, "Yuqui",              54, "Yurakaré",
  55, "Quechua - Aymara",    56, "Más de una descripción",
  57, "Otras declaraciones", 58, "Otros pueblos indígenas extranjeros",
  98, "No se autoidentifica", 99, "Sin respuesta"
)

census_group <- ds |>
  filter(p26_edad >= 18, p32_pueblo_per == 1L) |>
  group_by(p32_pueblos) |>
  summarise(n = n()) |>
  collect() |>
  left_join(pueblos_cats, by = c("p32_pueblos" = "code")) |>
  mutate(
    lapop_group = case_when(
      label == "Quechua" ~ "Quechua",
      label == "Aymara" ~ "Aymara",
      label == "Guaraní" ~ "Guaraní",
      label == "Chiquitano" ~ "Chiquitano",
      label %in% c("Mojeño", "Mojeño Ignaciano", "Mojeño Trinitario") ~ "Mojeño",
      label == "Afroboliviano" ~ "Afroboliviano",
      label == "Otras declaraciones" ~ "No especifica",
      TRUE ~ "Otro"
    )
  ) |>
  group_by(lapop_group) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(pct_census = n / sum(n) * 100)

lapop_group <- lapop |>
  filter(boletidnew == 1, !is.na(boletidnewb)) |>
  count(indig_group_f) |>
  mutate(
    pct_lapop = n / sum(n) * 100,
    lapop_group = recode(as.character(indig_group_f),
                         "No especifica, indígena nomás" = "No especifica")
  )
```

```{r}
#| label: tbl-group-compare
#| tbl-cap: "Indigenous group composition: LAPOP 2023 vs Census 2024 (18+ self-identifiers)"
comparison_group <- lapop_group |>
  select(lapop_group, pct_lapop) |>
  full_join(census_group |> select(lapop_group, pct_census),
            by = "lapop_group") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(desc(pct_census)) |>
  rename(Group = lapop_group, `LAPOP %` = pct_lapop, `Census %` = pct_census)

comparison_group
```

The group distributions are broadly similar, with Quechua and Aymara comprising
roughly three-quarters of identifiers in both sources. Two notable differences:

- **Quechua vs Aymara balance**: LAPOP shows Quechua at 47% and Aymara at 29%,
  while the census has them nearly tied (39% vs 38%). This may reflect LAPOP's
  stratified sampling, which groups departments into regional strata rather
  than sampling each department proportionally.
- **"No especifica" / generic identifiers**: Both sources show 3–12% using
  generic terms. The census figure (12%) is higher, likely because its
  open-ended response allows more vague answers that are subsequently
  reclassified by INE into the "Otras declaraciones" category.

------------------------------------------------------------------------

## Comparison: Mother tongue

LAPOP's `leng1` and the census's `idioma_mat` both ask about the first language
learned in childhood. LAPOP offers 6 categories; the census distinguishes 83
languages. We map the census to LAPOP's categories:

- Census code 6 (Castellano) → Castellano
- Census code 27 (Quechua) → Quechua
- Census code 2 (Aymara) → Aymara
- Census code 12 (Guaraní) → Guaraní
- Census codes 1–37 (other indigenous) → Otro nativo
- Census codes ≥ 40 (foreign, including unlabeled) → Otro extranjero

```{r}
#| label: census-leng
census_leng <- ds |>
  filter(p26_edad >= 18, !is.na(idioma_mat),
         !idioma_mat %in% c(998L, 994L)) |>
  group_by(idioma_mat) |>
  summarise(n = n()) |>
  collect() |>
  mutate(
    lapop_leng = case_when(
      idioma_mat == 6 ~ "Castellano",
      idioma_mat == 27 ~ "Quechua",
      idioma_mat == 2 ~ "Aymara",
      idioma_mat == 12 ~ "Guaraní",
      idioma_mat >= 40 ~ "Otro extranjero",
      TRUE ~ "Otro nativo"
    )
  ) |>
  group_by(lapop_leng) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(pct_census = n / sum(n) * 100)

lapop_leng <- lapop |>
  filter(!is.na(leng1)) |>
  mutate(
    lapop_leng = case_when(
      leng1 == 1001 ~ "Castellano",
      leng1 == 1002 ~ "Quechua",
      leng1 == 1003 ~ "Aymara",
      leng1 == 1006 ~ "Guaraní",
      leng1 == 1004 ~ "Otro nativo",
      leng1 == 1005 ~ "Otro extranjero"
    )
  ) |>
  count(lapop_leng) |>
  mutate(pct_lapop = n / sum(n) * 100)
```

```{r}
#| label: tbl-leng-compare
#| tbl-cap: "Mother tongue: LAPOP 2023 vs Census 2024 (18+)"
comparison_leng <- lapop_leng |>
  select(lapop_leng, pct_lapop) |>
  full_join(census_leng |> select(lapop_leng, pct_census),
            by = "lapop_leng") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(desc(pct_census)) |>
  rename(`Mother tongue` = lapop_leng, `LAPOP %` = pct_lapop,
         `Census %` = pct_census)

comparison_leng
```

Mother tongue distributions are remarkably consistent across the two sources,
with Castellano at 73–75%, Quechua at 16%, and Aymara at 8–10%. This
convergence is reassuring — language is a less ambiguous question than identity,
and both instruments capture it similarly.

------------------------------------------------------------------------

## Cross-tabulation: identity × mother tongue {#sec-crosstab}

The most informative comparison is the relationship between indigenous
self-identification and mother tongue. In both sources, a substantial share of
indigenous identifiers are Castellano mother-tongue speakers — but the two
sources differ in degree.

```{r}
#| label: compute-crosstabs

# LAPOP cross-tab
lapop_cross <- lapop |>
  filter(boletidnew %in% c(1, 2), !is.na(leng1)) |>
  mutate(
    identifies = ifelse(boletidnew == 1, "Self-identifies", "Does not"),
    madre = case_when(
      leng1 == 1001 ~ "Castellano",
      leng1 == 1002 ~ "Quechua",
      leng1 == 1003 ~ "Aymara",
      leng1 == 1006 ~ "Guaraní",
      TRUE ~ "Other"
    )
  ) |>
  count(identifies, madre) |>
  group_by(identifies) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup()

# Census cross-tab
census_cross <- ds |>
  filter(p26_edad >= 18, p32_pueblo_per %in% c(1L, 2L),
         !is.na(idioma_mat), !idioma_mat %in% c(998L, 994L)) |>
  group_by(p32_pueblo_per, idioma_mat) |>
  summarise(n = n(), .groups = "drop") |>
  collect() |>
  mutate(
    identifies = ifelse(p32_pueblo_per == 1, "Self-identifies", "Does not"),
    madre = case_when(
      idioma_mat == 6 ~ "Castellano",
      idioma_mat == 27 ~ "Quechua",
      idioma_mat == 2 ~ "Aymara",
      idioma_mat == 12 ~ "Guaraní",
      TRUE ~ "Other"
    )
  ) |>
  group_by(identifies, madre) |>
  summarise(n = sum(n), .groups = "drop") |>
  group_by(identifies) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup()
```

```{r}
#| label: tbl-cross-lapop
#| tbl-cap: "LAPOP 2023: Mother tongue by indigenous identification"
lapop_cross |>
  select(identifies, madre, pct) |>
  pivot_wider(names_from = madre, values_from = pct, values_fill = 0) |>
  rename(`Identification status` = identifies)
```

```{r}
#| label: tbl-cross-census
#| tbl-cap: "Census 2024 (18+): Mother tongue by indigenous identification"
census_cross |>
  select(identifies, madre, pct) |>
  pivot_wider(names_from = madre, values_from = pct, values_fill = 0) |>
  rename(`Identification status` = identifies)
```

```{r}
#| label: fig-crosstab
#| fig-cap: "Mother tongue composition among indigenous identifiers and non-identifiers"
#| fig-height: 5
#| fig-width: 8
bind_rows(
  lapop_cross |> mutate(source = "LAPOP 2023"),
  census_cross |> mutate(source = "Census 2024")
) |>
  mutate(
    madre = factor(madre, levels = c("Castellano", "Quechua", "Aymara",
                                     "Guaraní", "Other")),
    panel = paste(source, identifies, sep = "\n")
  ) |>
  ggplot(aes(x = madre, y = pct)) +
  geom_col() +
  facet_grid(identifies ~ source) +
  labs(x = NULL, y = "% within group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The cross-tabulations reveal two key patterns:

1. **Castellano dominance among identifiers**: In LAPOP, 63% of indigenous
   identifiers report Castellano as their mother tongue. In the census, the
   figure is even higher at 48% — but note that the census captures many more
   indigenous-language speakers overall (20% Aymara + 30% Quechua in the census
   vs 13% + 23% in LAPOP among identifiers).

2. **Non-identifiers with indigenous mother tongues**: In both sources, roughly
   8–10% of people who *do not* self-identify as indigenous report an
   indigenous mother tongue (mostly Quechua). In the census this amounts to
   roughly 330,000 adults — people who grew up speaking Quechua or Aymara but
   do not claim indigenous belonging.

------------------------------------------------------------------------

## Summary

| Measure | LAPOP 2023 | Census 2024 (18+) | Gap |
|---------|:----------:|:------------------:|:---:|
| Indigenous self-identification | 54.2% | 42.2% | +12 pp |
| Quechua mother tongue | 16.2% | 16.0% | ≈0 |
| Aymara mother tongue | 7.9% | 9.6% | −1.7 pp |
| Castellano mother tongue | 75.3% | 72.6% | +2.7 pp |
| Castellano among identifiers | 63.1% | 48.2% | +15 pp |

The largest discrepancy between sources is in indigenous self-identification
(+12 pp in LAPOP). Language distributions, by contrast, are strikingly
consistent. This suggests that the identity gap is driven primarily by
differences in question wording and survey context rather than by sampling
bias — if LAPOP's sample were simply over-representing indigenous populations,
we would expect to see higher indigenous-language rates as well.

The high rate of Castellano mother-tongue speakers among indigenous identifiers
in both sources (48–63%) underscores a point central to the broader census
analysis: indigenous identity in Bolivia is increasingly decoupled from
indigenous language use. The LAPOP `leng4` variable reinforces this — 59% of
respondents report that their parents spoke an indigenous language at home, far
exceeding the 25% who learned one as their own mother tongue.
```

<function_calls>
<invoke name="todoWrite">
<parameter name="todos">[{"content": "Examine census comparable variables and existing aggregates", "status": "completed"}, {"content": "Compute LAPOP rates for identity and language questions", "status": "completed"}, {"content": "Compute census 2024 comparable rates (18+ population)", "status": "completed"}, {"content": "Write the QMD document", "status": "completed"}]