---
title: "Wikidata QID ↔ OCHA Pcode Crosswalk"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  cache: true
  warning: false
  message: false
knitr:
  opts_chunk:
    cache.path: "data/cache/knitr-pcode/"
---

## Overview

This document builds and evaluates a crosswalk between Wikidata QIDs and OCHA
p-codes from the Global Administrative Boundaries GDB
(`../humdata/maps/global_admin_boundaries_matched_latest.gdb`), which contains:

- **ADM2**: 19,705 features (provinces, departments, cantons, municipalities)
- **ADM3**: 55,622 features (municipalities, districts, communes, parishes)

Six countries that appear in the Wikipedia presence analysis are **absent from
the GDB** (Argentina, Brazil, Paraguay, Suriname, Uruguay, Cuba, Puerto Rico)
and are excluded throughout. The 15 remaining countries in both datasets are
the focus here.

Two strategies are applied in priority order:

1. **Code-based** — for countries where a Wikidata external-ID property
   directly encodes the pcode suffix (e.g. P14142 for Bolivia, P7325 for
   Colombia). These are the most reliable matches and are preferred wherever
   available.
2. **Name-based** — normalized string matching between Wikidata labels and GDB
   name columns, with accent and admin-type prefix stripping. For countries
   with many duplicate sub-unit names (Venezuela, Panama), a hierarchical
   second pass uses P131 (parent admin unit) to anchor items before matching.

The pipeline is implemented in `src/pcode-crosswalk.R`. QuickStatements to
add P7590 (OCHA P-code) to matched items are generated at the end of this
document but are **not executed** — a manual quality review is expected first.

---

## Setup

```{r setup}
#| cache: false
library(sf)
library(httr)
library(jsonlite)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
library(tibble)
library(stringi)
library(ggplot2)

source("src/pcode-crosswalk.R")
source("R/create_quick_statement.R")
```

---

## Data

### GDB layers

```{r load-gdb}
GDB_PATH <- "../humdata/maps/global_admin_boundaries_matched_latest.gdb"

adm2_sf <- st_read(GDB_PATH, layer = "admin2", quiet = TRUE)
adm3_sf <- st_read(GDB_PATH, layer = "admin3", quiet = TRUE)

cat("ADM2:", nrow(adm2_sf), "features\n")
cat("ADM3:", nrow(adm3_sf), "features\n")
```

### Countries present in both datasets

The 15 countries present in both the GDB and the Wikipedia presence analysis,
with their ISO 3166-1 alpha-2 codes used to join the two sources:

```{r country-map}
# Countries in the Wikipedia presence analysis (from wikipedia-admin-presence.qmd)
# that also appear in the GDB
our_countries_gdb <- tribble(
  ~country,              ~country_qid, ~iso2,
  "Bolivia",             "Q750",       "BO",
  "Chile",               "Q298",       "CL",
  "Colombia",            "Q739",       "CO",
  "Costa Rica",          "Q800",       "CR",
  "Dominican Republic",  "Q786",       "DO",
  "Ecuador",             "Q736",       "EC",
  "El Salvador",         "Q792",       "SV",
  "Guatemala",           "Q774",       "GT",
  "Haiti",               "Q790",       "HT",
  "Honduras",            "Q783",       "HN",
  "Mexico",              "Q96",        "MX",
  "Nicaragua",           "Q811",       "NI",
  "Panama",              "Q804",       "PA",
  "Peru",                "Q419",       "PE",
  "Venezuela",           "Q717",       "VE"
)

knitr::kable(
  our_countries_gdb |> select(country, iso2),
  caption = "Countries present in both the GDB and the Wikipedia presence analysis"
)
```

### GDB working tables

Strip geometry, attach country labels, and select the columns used for
matching and name lookups.

```{r build-gdb-tables}
adm2_gdb <- adm2_sf |>
  st_drop_geometry() |>
  inner_join(our_countries_gdb |> select(country, iso2), by = "iso2") |>
  select(country, iso2, adm0_pcode, adm1_pcode, adm2_pcode,
         adm2_name, adm2_name1, adm2_name2, adm2_name3)

adm3_gdb <- adm3_sf |>
  st_drop_geometry() |>
  inner_join(our_countries_gdb |> select(country, iso2), by = "iso2") |>
  select(country, iso2, adm1_pcode, adm2_pcode, adm3_pcode,
         adm3_name, adm3_name1, adm3_name2, adm3_name3)

cat("ADM2 rows in scope:", nrow(adm2_gdb), "\n")
cat("ADM3 rows in scope:", nrow(adm3_gdb), "\n")
```

### Wikidata items

`all_data` is loaded from the cached presence-matrix pipeline in
`wikipedia-admin-presence.qmd`. It provides one row per Wikidata item with
its English and Spanish labels and admin level.

```{r load-all-data}
# Load presence data from the admin presence pipeline cache
sa_adm2_all   <- readRDS("data/cache/sa_adm2_all.rds")
sa_adm3_all   <- readRDS("data/cache/sa_adm3_all.rds")
newc_adm2_all <- readRDS("data/cache/newc_adm2_all.rds")
newc_adm3_all <- readRDS("data/cache/newc_adm3_all.rds")

brazil_cache <- "data/cache/sa_adm2_brazil_df.rds"
all_tiers <- list(sa_adm2_all, sa_adm3_all, newc_adm2_all, newc_adm3_all)
if (file.exists(brazil_cache)) all_tiers <- c(all_tiers, list(readRDS(brazil_cache)))

all_data <- bind_rows(all_tiers)

# Keep only the columns used by the crosswalk pipeline
all_data <- all_data |>
  select(qid, country, adm_level, label_en, label_es) |>
  filter(country %in% our_countries_gdb$country)

cat("Wikidata items in scope:", nrow(all_data), "\n")
cat("Countries:", n_distinct(all_data$country), "\n")
all_data |> count(adm_level)
```

---

## Code-based property reference

The following table records which Wikidata property encodes the pcode suffix
for each country. The pcode is reconstructed as `paste0(iso2, property_value)`.
Property coverage was verified by sampling items and comparing constructed
codes against the GDB pcode lists.

```{r code-properties-table}
knitr::kable(
  PCODE_CODE_PROPERTIES |>
    mutate(adm_levels = map_chr(adm_levels, paste, collapse = " + ")) |>
    select(Country = country, Property = pid, `ISO2 prefix` = iso2,
           `Admin levels` = adm_levels),
  caption = "Wikidata properties used for code-based pcode matching"
)
```

Properties investigated but not adopted:

| Country | Property | Reason not used |
|---------|----------|-----------------|
| Argentina | P4833 (INDEC) | Argentina absent from GDB |
| Colombia | P2082 | Not populated on sampled items; P7325 preferred |
| Haiti | P1370 | Very sparse — most arrondissements lack it |
| All | P7590 (OCHA P-code) | Not yet populated — this crosswalk will add it |

---

## Name-matching methodology

For countries without a reliable code property, labels are normalized and
compared to GDB name columns:

1. **Normalize**: lowercase → Latin-ASCII transliteration (strip accents) →
   collapse non-alphanumeric characters to spaces → trim.
2. **Strip admin-type prefix**: remove leading tokens such as "Cantón",
   "Provincia de", "Municipio de" etc. (both languages, both normalized and
   stripped variants are compared).
3. **Match within country**: inner join on `(country, normalized_name)`.
   Only one-to-one matches are kept (`n_gdb_hits == 1`).
4. **Hierarchical pass (Venezuela, Panama)**: for ADM3 items with ambiguous
   country-level matches, fetch the P131 parent via SPARQL, look the parent
   up in the ADM2 crosswalk, and re-match within the confirmed parent unit.

---

## Build the crosswalk

```{r build-crosswalk}
#| cache: true
#| cache-extra: !expr digest::digest(readLines("src/pcode-crosswalk.R"))
crosswalk <- build_pcode_crosswalk(
  all_data               = all_data,
  adm2_gdb               = adm2_gdb,
  adm3_gdb               = adm3_gdb,
  hierarchical_countries = c("Venezuela", "Panama"),
  sparql_batch_size      = 300,
  sparql_delay           = 0.4,
  cache_file             = "data/cache/crosswalk_qid_pcode.rds"
)

cat("Total matched items:", nrow(crosswalk), "\n")
cat("Unique QIDs:        ", n_distinct(crosswalk$qid), "\n")
```

---

## Coverage summary

```{r coverage-table}
gdb_countries <- our_countries_gdb$country

total_in_scope <- all_data |>
  count(country, adm_level, name = "n_wikidata")

coverage <- crosswalk |>
  count(country, adm_level, match_method, name = "n_matched") |>
  left_join(total_in_scope, by = c("country", "adm_level")) |>
  mutate(pct = round(100 * n_matched / n_wikidata, 1))

# Roll up to country × adm_level for a clean summary
coverage_summary <- crosswalk |>
  count(country, adm_level, name = "n_matched") |>
  left_join(total_in_scope, by = c("country", "adm_level")) |>
  mutate(
    pct            = round(100 * n_matched / n_wikidata, 1),
    primary_method = case_when(
      country == "Bolivia"    ~ "code (P14142)",
      country == "Colombia"   ~ "code (P7325)",
      country == "Mexico"     ~ "code (P3801)",
      country == "Costa Rica" & adm_level == "ADM3" ~ "code (P281)",
      country == "Chile"      & adm_level == "ADM3" ~ "code (P6929)",
      TRUE ~ "name match"
    )
  ) |>
  arrange(country, adm_level)

knitr::kable(
  coverage_summary |>
    select(Country = country, Level = adm_level,
           `WD items` = n_wikidata, Matched = n_matched,
           `%` = pct, `Primary method` = primary_method),
  caption = "Crosswalk coverage by country and administrative level"
)
```

```{r fig-coverage}
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Crosswalk coverage (%) by country and admin level, coloured by match strategy"
coverage_summary |>
  mutate(
    strategy = if_else(str_starts(primary_method, "code"), "Code-based", "Name-based"),
    country  = reorder(country, pct, FUN = mean)
  ) |>
  ggplot(aes(x = pct, y = country, fill = strategy)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0.15)) +
  geom_vline(xintercept = 100, linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~ adm_level) +
  scale_fill_manual(values = c("Code-based" = "#2166AC", "Name-based" = "#F4A582")) +
  scale_x_continuous(limits = c(0, 105), breaks = seq(0, 100, 25)) +
  labs(x = "% of Wikidata items matched", y = NULL, fill = "Strategy") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Unmatched items

```{r unmatched-summary}
unmatched <- all_data |>
  anti_join(crosswalk, by = "qid") |>
  count(country, adm_level, name = "n_unmatched") |>
  left_join(total_in_scope, by = c("country", "adm_level")) |>
  mutate(pct_unmatched = round(100 * n_unmatched / n_wikidata, 1)) |>
  filter(n_unmatched > 0) |>
  arrange(desc(n_unmatched))

knitr::kable(
  unmatched |>
    select(Country = country, Level = adm_level,
           Unmatched = n_unmatched, `Total WD` = n_wikidata,
           `% unmatched` = pct_unmatched),
  caption = "Unmatched Wikidata items by country and level"
)
```

The largest gaps are:

- **Venezuela ADM2/ADM3**: municipality names repeat heavily across states;
  a full state-level hierarchical pass via P131 would recover many of the
  remaining ~400 unmatched items.
- **Ecuador ADM3**: many parish names are non-unique across cantons and
  Wikidata coverage of parishes is incomplete.
- **Chile ADM3**: P6929 covers ~58%; the remainder have no P6929 value in
  Wikidata.
- **Panama ADM3**: corregimiento names repeat within districts.

---

## Quality check

### Match method breakdown

```{r method-breakdown}
crosswalk |>
  count(match_method, name = "n") |>
  mutate(
    pct      = round(100 * n / nrow(crosswalk), 1),
    strategy = if_else(str_starts(match_method, "code"), "Code-based", "Name-based")
  ) |>
  arrange(strategy, match_method) |>
  knitr::kable(caption = "Crosswalk entries by match method")
```

### Spot-check: name-matched items

The function below pulls a stratified sample of name-matched items and shows
the Wikidata label alongside the GDB name it matched. Scan for false positives
(same normalized string, different real-world unit).

```{r spot-check-fn}
#| cache: false
spot_check_name_matches <- function(crosswalk, all_data, adm2_gdb, adm3_gdb,
                                    country_filter = NULL, n = 20, seed = 5501) {
  set.seed(seed)

  xw <- crosswalk |>
    filter(str_starts(match_method, "name")) |>
    left_join(all_data |> select(qid, label_en, label_es), by = "qid")

  if (!is.null(country_filter)) xw <- xw |> filter(country %in% country_filter)

  adm2_names <- adm2_gdb |> distinct(adm2_pcode, gdb_name = adm2_name)
  adm3_names <- adm3_gdb |> distinct(adm3_pcode, gdb_name = adm3_name)

  xw |>
    left_join(adm2_names, by = "adm2_pcode") |>
    left_join(adm3_names, by = "adm3_pcode") |>
    mutate(gdb_name = coalesce(gdb_name.x, gdb_name.y)) |>
    select(qid, country, adm_level, match_method,
           wd_label_en = label_en, wd_label_es = label_es,
           gdb_name, pcode) |>
    slice_sample(n = min(n, nrow(xw)))
}
```

```{r spot-check-display}
spot_check_name_matches(crosswalk, all_data, adm2_gdb, adm3_gdb, n = 30) |>
  knitr::kable(caption = "Random sample of name-matched items for review")
```

### Code-matched sample

Code matches are structurally reliable (property value = numeric pcode suffix),
but it is worth confirming a few against the GDB name to catch any systematic
encoding errors.

```{r code-match-sample}
set.seed(1783)

adm2_names <- adm2_gdb |> distinct(adm2_pcode, gdb_name_adm2 = adm2_name)
adm3_names <- adm3_gdb |> distinct(adm3_pcode, gdb_name_adm3 = adm3_name)

crosswalk |>
  filter(str_starts(match_method, "code")) |>
  left_join(all_data |> select(qid, label_en), by = "qid") |>
  left_join(adm2_names, by = "adm2_pcode") |>
  left_join(adm3_names, by = "adm3_pcode") |>
  mutate(gdb_name = coalesce(gdb_name_adm2, gdb_name_adm3)) |>
  select(qid, country, adm_level, pcode, match_method, label_en, gdb_name) |>
  group_by(match_method) |>
  slice_sample(n = 5) |>
  ungroup() |>
  knitr::kable(caption = "Sample of code-matched items (5 per property)")
```

### Hierarchical ADM3 matches

These 210 items were ambiguous at the country level but resolved via P131.
A higher proportion of these merit individual review.

```{r hierarchical-sample}
set.seed(2290)

crosswalk |>
  filter(match_method == "name_adm3_hierarchical") |>
  left_join(all_data |> select(qid, label_en, label_es), by = "qid") |>
  left_join(adm3_names, by = "adm3_pcode") |>
  select(qid, country, pcode, adm2_pcode, wd_label_en = label_en,
         wd_label_es = label_es, gdb_name = gdb_name_adm3) |>
  slice_sample(n = 20) |>
  knitr::kable(caption = "Sample of hierarchically resolved ADM3 matches")
```

### Items with duplicate GDB names (potential false positives)

Name matches where the GDB contains more than one unit with the same
normalized name in the same country are by construction excluded from the
crosswalk (`n_gdb_hits == 1` filter). This section documents how many items
were dropped for this reason per country.

```{r ambiguous-counts}
#| cache: false

# Re-run the name join without the n_gdb_hits filter to count ambiguous cases
normalize_admin_name <- function(x) {
  x |> str_to_lower() |> stri_trans_general("Latin-ASCII") |>
    str_replace_all("[^a-z0-9]+", " ") |> str_squish()
}

wd_adm2_labels <- all_data |>
  filter(adm_level == "ADM2") |>
  pivot_longer(c(label_en, label_es), values_to = "raw", names_to = "lang") |>
  filter(!is.na(raw)) |>
  mutate(norm = normalize_admin_name(raw)) |>
  distinct(qid, country, norm)

gdb_adm2_norm <- adm2_gdb |>
  pivot_longer(starts_with("adm2_name"), values_to = "raw", names_to = "nv") |>
  filter(!is.na(raw), raw != "") |>
  mutate(norm = normalize_admin_name(raw)) |>
  distinct(country, adm2_pcode, norm)

wd_adm2_labels |>
  inner_join(gdb_adm2_norm, by = c("country", "norm"),
             relationship = "many-to-many") |>
  group_by(qid, country) |>
  summarise(n_gdb_hits = n_distinct(adm2_pcode), .groups = "drop") |>
  filter(n_gdb_hits > 1) |>
  count(country, name = "n_ambiguous") |>
  arrange(desc(n_ambiguous)) |>
  knitr::kable(caption = "ADM2 items dropped due to ambiguous name match (>1 GDB unit)")
```

---

## Unmatched entities — high-coverage countries {#sec-unmatched}

For country × level combinations with ≥ 85% match rate the residual gap is
small enough to inspect unit by unit. Both sides of the join are shown: Wikidata
items with no pcode, and GDB units with no QID.

```{r high-coverage-strata}
#| cache: false
high_coverage <- coverage_summary |>
  filter(pct >= 85) |>
  select(country, adm_level, n_matched, n_wikidata, pct)

knitr::kable(
  high_coverage |>
    select(Country = country, Level = adm_level,
           Matched = n_matched, `WD total` = n_wikidata, `%` = pct),
  caption = "Country × level strata with ≥ 85% match rate"
)
```

### Unmatched Wikidata items

These items exist in Wikidata but have no GDB pcode in the crosswalk. Likely
causes in high-coverage countries: the unit was created, renamed, or split
after the GDB snapshot; the Wikidata label is in an unexpected language or
spelling; or the Wikidata item covers a slightly different geographic entity
(e.g. a city item vs. a municipality item).

```{r unmatched-wd}
#| cache: false
unmatched_wd <- all_data |>
  semi_join(high_coverage, by = c("country", "adm_level")) |>
  anti_join(crosswalk,  by = "qid") |>
  mutate(
    wd_link = paste0("[", qid, "](https://www.wikidata.org/wiki/", qid, ")")
  ) |>
  arrange(country, adm_level, label_en) |>
  select(Country = country, Level = adm_level,
         QID = wd_link, `Label (en)` = label_en, `Label (es)` = label_es)

cat(nrow(unmatched_wd), "unmatched Wikidata items across",
    n_distinct(paste(unmatched_wd$Country, unmatched_wd$Level)), "strata\n")

knitr::kable(unmatched_wd,
             caption = "Wikidata items with no GDB pcode (high-coverage strata)")
```

### Unmatched GDB units

These GDB units have no matched Wikidata item. Likely causes: the unit has no
Wikidata item yet (a gap in Wikidata coverage); the GDB reflects a more recent
or more granular boundary dataset than what Wikidata tracks; or the pcode
belongs to a sub-division that Wikidata models at a different level.

```{r unmatched-gdb}
#| cache: false
# ADM2 GDB units with no crosswalk entry, for high-coverage countries
unmatched_gdb_adm2 <- adm2_gdb |>
  semi_join(high_coverage |> filter(adm_level == "ADM2"), by = "country") |>
  filter(!adm2_pcode %in% crosswalk$pcode) |>
  mutate(adm_level = "ADM2") |>
  select(Country = country, Level = adm_level, Pcode = adm2_pcode,
         `ADM1 pcode` = adm1_pcode, `GDB name` = adm2_name)

# ADM3 GDB units with no crosswalk entry, for high-coverage countries
unmatched_gdb_adm3 <- adm3_gdb |>
  semi_join(high_coverage |> filter(adm_level == "ADM3"), by = "country") |>
  filter(!adm3_pcode %in% crosswalk$pcode) |>
  mutate(adm_level = "ADM3") |>
  select(Country = country, Level = adm_level, Pcode = adm3_pcode,
         `ADM1 pcode` = adm1_pcode, `GDB name` = adm3_name)

unmatched_gdb <- bind_rows(unmatched_gdb_adm2, unmatched_gdb_adm3) |>
  arrange(Country, Level, Pcode)

cat(nrow(unmatched_gdb), "unmatched GDB units across",
    n_distinct(paste(unmatched_gdb$Country, unmatched_gdb$Level)), "strata\n")

knitr::kable(unmatched_gdb,
             caption = "GDB units with no matched Wikidata QID (high-coverage strata)")
```

### Summary: gap by country

```{r gap-summary}
#| cache: false
wd_gap <- unmatched_wd |>
  count(Country, Level, name = "unmatched_wd")

gdb_gap <- unmatched_gdb |>
  count(Country, Level, name = "unmatched_gdb")

high_coverage |>
  left_join(wd_gap,  by = c("country" = "Country", "adm_level" = "Level")) |>
  left_join(gdb_gap, by = c("country" = "Country", "adm_level" = "Level")) |>
  replace_na(list(unmatched_wd = 0L, unmatched_gdb = 0L)) |>
  mutate(
    `Unmatched WD`  = unmatched_wd,
    `Unmatched GDB` = unmatched_gdb
  ) |>
  select(Country = country, Level = adm_level, `%` = pct,
         `Unmatched WD`, `Unmatched GDB`) |>
  knitr::kable(caption = "Residual gap on each side for high-coverage strata")
```

---

## QuickStatements

The function `make_pcode_quickstatements()` generates Wikidata QuickStatements
to add **P7590** (OCHA P-code) to matched items. It takes the crosswalk, a
country name, and optional reference metadata, and returns a tibble with one
row per item.

**No statements are executed here.** Run the quality checks above first, then
call the function with appropriate `methods` and `reference_*` arguments for
each country.

### Function signature

```r
make_pcode_quickstatements(
  crosswalk,
  country,                        # e.g. "Bolivia"
  pid            = "P7590",       # OCHA P-code property
  adm_levels     = c("ADM2","ADM3"),
  methods        = NULL,          # NULL = all; or e.g. c("code_P14142")
  reference_qid  = NULL,          # Wikidata item for "stated in" (P248)
  reference_url  = NULL,          # URL for P854
  retrieved_date = Sys.Date(),
  output_file    = NULL           # write TSV if provided
)
```

### Preview: Bolivia (code-matched items only)

Bolivia is used as the preview country because all its matches are
code-based (P14142) and therefore require no additional name-match review.

```{r qs-preview}
qs_bolivia <- make_pcode_quickstatements(
  crosswalk     = crosswalk,
  country       = "Bolivia",
  methods       = c("code_P14142"),       # code-confirmed only
  reference_url = "https://data.humdata.org/dataset/global-administrative-unit-layers"
)

cat("Statements to generate:", nrow(qs_bolivia), "\n\n")
cat("Sample (first 5):\n")
cat(head(qs_bolivia$quick_statement, 5), sep = "\n")
```

```{r qs-preview-table}
qs_bolivia |>
  select(qid, adm_level, pcode, match_method, quick_statement) |>
  head(10) |>
  knitr::kable(caption = "First 10 QuickStatements for Bolivia (not yet uploaded)")
```

### Generating statements for other countries

Once the quality review is complete for a given country, write its statements
to a file and review in the QuickStatements batch tool before submitting:

```r
# Example — do not run until quality review is done
make_pcode_quickstatements(
  crosswalk     = crosswalk,
  country       = "Mexico",
  methods       = "code_P3801",
  reference_url = "https://data.humdata.org/dataset/global-administrative-unit-layers",
  output_file   = "data/qs_mexico_pcode.tsv"
)
```

| Country | Recommended methods before upload | Notes |
|---------|----------------------------------|-------|
| Bolivia | `code_P14142` | 100% coverage, no review needed |
| Colombia | `code_P7325` | 99% coverage, spot-check recommended |
| Mexico | `code_P3801` | 99% ADM2 coverage |
| Costa Rica ADM3 | `code_P281` | 99% ADM3; ADM2 by name needs review |
| Chile ADM3 | `code_P6929` | 58% coverage; name matches need review |
| All others | `name_adm2`, `name_adm3` | Review spot-check table above first |

---

## Session info

```{r session-info}
#| cache: false
sessionInfo()
```
