---
title: "Importing Wikipedia References into Zotero"
subtitle: "A walkthrough using `wiki-refs-to-zotero.R`"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
---

## Overview

`wiki-refs-to-zotero.R` extracts every citation template from a Wikipedia
article and imports the references into your Zotero library as a new
collection.  It handles the two Wikipedia citation styles — inline
`{{Cite book|...}}` templates and `{{harvc}}` book-section entries in a
"Sources" section — as well as mixed articles that use both.

The workflow has three logical phases:

1. **Parse** — fetch the wikitext and extract citation templates into a tidy
   tibble.
2. **Enrich** *(optional)* — look up DOIs and ISBNs via the c2z package to
   fill in metadata that Wikipedia citations sometimes omit (e.g. publisher,
   full date, correct author names).
3. **Import** — post the references to a new Zotero collection via the Zotero
   REST API.

A **dry run** separates phases 1–2 from phase 3.  Use it to inspect and clean
up the parsed references before committing anything to Zotero.  Because dry
runs return the fully-parsed tibble, you can complete the import later with a
single function call — no second Wikipedia fetch required.

## Prerequisites

```{r}
#| label: setup
library(WikipediR)   # MediaWiki API client
library(tidyverse)
library(stringi)
library(httr2)
library(c2z)         # Zotero API helpers (ZoteroDoi, ZoteroIsbn, ZoteroPost)

source("wiki-refs-to-zotero.R")
```

You also need a Zotero API key with **write** access.  Store it as an
environment variable (e.g. in `.Renviron`) so you never hardcode credentials:

```
ZOTERO_USER_ID = 1234567
ZOTERO_API_KEY = aBcDeFgHiJkLmNoPqRsTuVwX
```

Your user ID appears in your Zotero profile URL:
`https://www.zotero.org/`**`yourusername`**.

## Example page

Throughout this vignette we use the article
[Meiō incident](https://en.wikipedia.org/wiki/Me%C3%B6_incident)
(`"Meiō_incident"`), a well-referenced article with a mix of journal articles,
books, and book sections.  Swap it for any Wikipedia page title.

## Step 1 — Dry run

`wiki_refs_pipeline()` defaults to `dry_run = TRUE`, so the call below fetches
and parses everything without touching Zotero.

```{r}
#| label: dry-run
refs <- wiki_refs_pipeline(
  page_name     = "Meiō_incident",
  enrich        = TRUE,       # resolve DOIs and ISBNs via c2z
  zotero_import = TRUE,       # prepare for import (but don't post yet)
  user_id       = Sys.getenv("ZOTERO_USER_ID"),
  api_key       = Sys.getenv("ZOTERO_API_KEY"),
  dry_run       = TRUE        # <-- inspect only; nothing is posted
)
```

You will see progress messages like:

```
Fetching wikitext for: Meiō_incident
Extracting citations...
Found 48 citation templates (before dedup)
Deduplicating...
Unique references: 44
Enriching via c2z: 11 DOI(s), 8 ISBN(s) to look up
Enrichment: 9 via DOI, 7 via ISBN, 3 failed, 25 skipped (no identifier)
DRY RUN: Would create collection 'Meiō incident' with 44 new item(s) and 0 moved item(s).
```

## Step 2 — Inspect the results

```{r}
#| label: inspect
# Overview of reference types
refs |> count(itemType, sort = TRUE)

# Quick look at the first few rows
refs |>
  select(itemType, first_author, year, title) |>
  head(10)

# Check enrichment outcomes
refs |> count(.enrich_status)
```

Look for:

- **Missing titles** (`is.na(title)`) — these may be bare `<ref>` notes rather
  than bibliographic entries, or templates with unusual field names.
- **`itemType == "document"`** rows — the catch-all for templates the parser
  did not recognise.
- **`enrich_status == "failed"`** — the DOI or ISBN was present but the lookup
  returned nothing (often an older or non-CrossRef DOI).

```{r}
#| label: problem-rows
# Flag anything that might need manual review
refs |>
  filter(is.na(title) | itemType == "document" | .enrich_status == "failed") |>
  select(itemType, first_author, year, title, DOI, ISBN, .enrich_status)
```

::: {.callout-note}
The `refs` tibble is a plain tibble — you can edit it with standard dplyr
verbs before posting.  For example, to fix a title:

```r
refs <- refs |>
  mutate(title = if_else(
    first_author == "Smith" & year == "2003",
    "The corrected title",
    title
  ))
```
:::

## Step 3 — Check for existing Zotero items (optional)

If you have already imported some of these references into Zotero from a
different source, `check_library_for_refs()` will identify them and let you
choose to move the existing items into the new collection rather than
duplicating them.

```{r}
#| label: check-existing
refs <- check_library_for_refs(
  refs,
  user_id = Sys.getenv("ZOTERO_USER_ID"),
  api_key = Sys.getenv("ZOTERO_API_KEY")
)
```

The function queries your entire library, finds matches by DOI → ISBN →
normalised title + year, and shows an interactive side-by-side comparison for
each candidate.  You respond with:

- **y** — accept this match (move the existing item, don't create a new one)
- **n** — decline (create a new item as normal)
- **a** — accept this and all remaining matches
- **s** — stop prompting (treat all remaining candidates as new items)

After the loop you will see a summary:

```
Result: 3 existing item(s) to move, 41 new item(s) to create.
```

## Step 4 — Post to Zotero

Once you are satisfied with `refs`, call `post_refs_to_zotero()`.  This is the
only step that writes to Zotero; it uses the tibble already in memory and does
**not** re-fetch Wikipedia.

```{r}
#| label: post
result <- post_refs_to_zotero(
  refs,
  collection_name = "Meiō incident",
  user_id         = Sys.getenv("ZOTERO_USER_ID"),
  api_key         = Sys.getenv("ZOTERO_API_KEY")
)
```

Open Zotero (or refresh your web library) and the new collection should appear
with all references in it.

## Step 5 — Retrieve Zotero item keys (optional)

If you want to cite these items programmatically — for example, to build a
Pandoc bibliography or link references to other data — retrieve the Zotero
keys that were assigned on import:

```{r}
#| label: fetch-keys
refs_with_keys <- fetch_zotero_keys(
  refs,
  user_id        = Sys.getenv("ZOTERO_USER_ID"),
  api_key        = Sys.getenv("ZOTERO_API_KEY"),
  collection_key = "XXXXXXXX"   # 8-char key shown in Zotero URL
)

refs_with_keys |>
  select(first_author, year, title, zotero_key, zotero_match) |>
  head(10)
```

`zotero_match` tells you how each ref was identified: `"doi"`, `"isbn"`,
`"title_year"`, or `"unmatched"`.

You can find the collection key by looking at your library URL after clicking
the new collection, or with a quick httr2 call:

```{r}
#| label: find-collection-key
request("https://api.zotero.org") |>
  req_url_path_append("users", Sys.getenv("ZOTERO_USER_ID"), "collections") |>
  req_headers(
    `Zotero-API-Version` = "3",
    Authorization = paste("Bearer", Sys.getenv("ZOTERO_API_KEY"))
  ) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE) |>
  (\(x) tibble(key = x$key, name = x$data$name))() |>
  filter(name == "Meiō incident")
```

## One-step workflow (no dry run)

If you trust the parser and don't need to review the results first, you can
run the entire pipeline in a single call:

```{r}
#| label: one-step
refs <- wiki_refs_pipeline(
  page_name       = "Meiō_incident",
  enrich          = TRUE,
  zotero_import   = TRUE,
  check_existing  = TRUE,
  fetch_keys      = TRUE,
  user_id         = Sys.getenv("ZOTERO_USER_ID"),
  api_key         = Sys.getenv("ZOTERO_API_KEY"),
  dry_run         = FALSE   # <-- post immediately
)
```

When `fetch_keys = TRUE` and `dry_run = FALSE`, the pipeline automatically
calls `fetch_zotero_keys()` after import, so the returned tibble already
contains `zotero_key` and `zotero_match` columns.

## RIS export

At any point after parsing you can export to RIS (for import into Mendeley,
EndNote, or other reference managers):

```{r}
#| label: ris-export
export_ris(refs, "meio_incident.ris")
```

## Continuing after a dry run — summary

The key insight is that `wiki_refs_pipeline()` returns the fully-parsed (and
optionally enriched) tibble regardless of `dry_run`.  The dry-run flag only
suppresses the Zotero API write.  So the two-step pattern is:

```{r}
#| label: two-step-summary
# ── Step 1: fetch, parse, enrich ─────────────────────────────────────────────
refs <- wiki_refs_pipeline(
  "Meiō_incident",
  enrich        = TRUE,
  zotero_import = TRUE,
  user_id       = Sys.getenv("ZOTERO_USER_ID"),
  api_key       = Sys.getenv("ZOTERO_API_KEY"),
  dry_run       = TRUE    # nothing written to Zotero
)

# ... inspect, filter, or edit `refs` here ...

# ── Step 2: post (no second Wikipedia fetch) ─────────────────────────────────
post_refs_to_zotero(
  refs,
  collection_name = "Meiō incident",
  user_id         = Sys.getenv("ZOTERO_USER_ID"),
  api_key         = Sys.getenv("ZOTERO_API_KEY")
)
```

## Troubleshooting

| Symptom | Likely cause | Fix |
|---|---|---|
| `No citation templates found` warning | Page uses author-date style (`{{sfn}}`) without a "References" section containing `{{cite ...}}` templates | Check the raw wikitext; `{{sfn}}` articles embed full citations in a `{{cite ...}}` template in a "Sources" section — make sure that section exists |
| Many `itemType == "document"` rows | Unusual template name not in the recognised list | Check `.template_name` column and extend `cite_patterns` in the script if needed |
| `enrich_status == "failed"` for a DOI | DOI is not in CrossRef (e.g. non-CrossRef publisher) | No automated fix; edit the row manually |
| Zotero API 403 error | API key lacks write permission | Generate a new key at `zotero.org/settings/keys` with "Allow write access" |
| Duplicate items created | `check_existing` was `FALSE` or a match was declined | Delete duplicates in Zotero; re-run with `check_existing = TRUE` |
