---
title: "Bolivia Municipality Wikipedia & Wikidata Cleanup Tracker"
author: "Carwil Bjork-James"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
editor: visual
---

# Overview

This document tracks the state of English Wikipedia pages and Wikidata items for Bolivia's 340 municipalities. It checks:

1.  **Wikidata completeness**: Are all 339 municipalities represented as `instance of: municipality of Bolivia` (Q1062710)?
2.  **Wikipedia category coverage**: Which pages exist in the `Municipalities of Bolivia` category tree?
3.  **Expected page existence**: For each of 336 expected municipality page titles, does the page exist?
4.  **Wikidata item type**: For existing pages not matched to Wikidata municipality items, is the linked Wikidata item actually a province, locality, or something else?

Re-render this document after editing Wikipedia/Wikidata to update all tables.

```{r setup}
#| message: false
library(tidyverse)
library(httr2)
library(httr)
library(jsonlite)
library(WikipediR)
library(glue)
library(knitr)

source("src/wikipedia-tools.R")
source("get_wikidata_instances.R")
```

# 1. Expected municipality page names

The canonical list of 336 expected Wikipedia page titles was manually constructed from INE's 340-municipality list (`muni_id_lookup_table`), applying Wikipedia disambiguation conventions (e.g. "Municipality, Department" for shared names). Four capital-city municipalities use their city page name instead.

This mapping is saved in `data/page_dept_map.rds`.

```{r load-page-names}
page_dept_map <- readRDS("data/page_dept_map.rds")
muni_pages <- page_dept_map$page

message(length(muni_pages), " expected municipality page titles")
```

# 2. Query Wikidata for municipality items

Retrieve all Wikidata items that are `instance of: municipality of Bolivia` (Q1062710), with their English Wikipedia sitelinks.

```{r wikidata-query}
#| cache: true
wd_munis <- get_wikidata_instances(
  "Q1062710",
  property = c("P131"),
  property_names = c("located_in"),
  country = "Q750",
  languages = c("en", "es"),
  limit = 500
)

# Extract English Wikipedia article title from sitelinks
wd_munis <- wd_munis |>
  mutate(
    en_wiki_title = map_chr(wikipedia_articles, function(arts) {
      if (length(arts) == 0) return(NA_character_)
      en_art <- arts[str_detect(arts, "^en:")]
      if (length(en_art) == 0) return(NA_character_)
      str_replace(en_art[1], "^en:\\s*", "")
    })
  )

message(nrow(wd_munis), " Wikidata items with instance_of = municipality of Bolivia")
```

# 3. Discover Wikipedia category pages

Scrape all article pages listed in the `Category:Municipalities of Bolivia` tree (top-level + per-department subcategories).

```{r discover-category-pages}
#| cache: true
top_cat <- "Category:Municipalities_of_Bolivia"
subcats <- get_subcategories(top_cat)

muni_subcats <- subcats |>
  filter(str_detect(title, "^Category:Municipalities"))

# Pages directly in the top-level category
top_pages <- get_category_pages(top_cat)

# Pages in each department subcategory
subcat_pages <- map_dfr(muni_subcats$title, function(cat) {
  Sys.sleep(0.3)
  get_category_pages(cat) |>
    mutate(source_category = cat)
})

all_muni_pages <- bind_rows(
  top_pages |> mutate(source_category = top_cat),
  subcat_pages
) |>
  distinct(title, .keep_all = TRUE) |>
  filter(ns == 0)

message(nrow(all_muni_pages), " unique pages in Wikipedia municipality categories")
```

## Category page summary

```{r category-summary}
all_muni_pages |>
  count(source_category, name = "n_pages") |>
  arrange(source_category) |>
  kable()
```

# 4. Fetch page metadata

```{r page-metadata}
#| cache: true
page_info <- get_page_info_batch(all_muni_pages$title)

muni_df <- all_muni_pages |>
  select(-pageid) |>
  left_join(page_info, by = "title")

message(nrow(muni_df), " pages with metadata")
```

# 5. Compare Wikipedia categories with Wikidata

```{r compare-wp-wd}
wp_only <- muni_df |>
  filter(!wikidata_qid %in% wd_munis$qid | is.na(wikidata_qid))

wd_only <- wd_munis |>
  filter(!qid %in% muni_df$wikidata_qid)

in_both <- muni_df |>
  filter(wikidata_qid %in% wd_munis$qid)

comparison_summary <- tribble(
  ~Category, ~Count,
  "Wikipedia category pages", nrow(muni_df),
  "Wikidata municipality items", nrow(wd_munis),
  "In both", nrow(in_both),
  "Wikipedia category only", nrow(wp_only),
  "Wikidata only (no category page)", nrow(wd_only)
)

comparison_summary |> kable()
```

### Pages in Wikipedia category but not Wikidata

These are pages categorised as municipalities but whose Wikidata item is not typed as `municipality of Bolivia`:

```{r wp-only-table}
wp_only |>
  select(title, wikidata_qid, page_length) |>
  arrange(title) |>
  kable()
```

### Wikidata municipality items without a Wikipedia category page

```{r wd-only-table}
wd_only |>
  select(qid, label_en, label_es, en_wiki_title) |>
  arrange(label_en) |>
  kable()
```

# 6. Check expected page existence

For each of the 336 expected municipality page titles, check whether the page actually exists on English Wikipedia.

```{r check-page-existence}
#| cache: true
check_pages_exist_batch <- function(titles, lang = "en") {
  api_url <- paste0("https://", lang, ".wikipedia.org/w/api.php")
  batches <- split(titles, ceiling(seq_along(titles) / 50))
  results <- list()

  for (batch in batches) {
    Sys.sleep(0.2)
    resp <- httr::GET(api_url, query = list(
      action = "query",
      titles = paste(batch, collapse = "|"),
      format = "json",
      redirects = 1
    ))
    json <- fromJSON(content(resp, "text", encoding = "UTF-8"),
                     simplifyVector = FALSE)

    found_titles <- map_chr(json$query$pages, ~ .x$title %||% NA_character_)
    missing <- map_lgl(json$query$pages, ~ !is.null(.x$missing))

    for (i in seq_along(json$query$pages)) {
      results[[length(results) + 1]] <- tibble(
        title = found_titles[i],
        exists = !missing[i]
      )
    }
  }
  bind_rows(results)
}

# Query Wikipedia for all expected page names
existence_raw <- check_pages_exist_batch(muni_pages)

# Match back to original page names (redirects may change titles)
page_existence <- tibble(page = muni_pages) |>
  left_join(existence_raw, by = c("page" = "title")) |>
  mutate(exists = coalesce(exists, FALSE))
```

## Summary

```{r existence-summary}
existence_counts <- page_existence |>
  count(exists) |>
  mutate(status = if_else(exists, "Page exists", "No page")) |>
  select(status, n)

existence_counts |> kable()
```

### Pages that exist

```{r existing-pages-table}
page_existence |>
  filter(exists) |>
  left_join(page_dept_map, by = "page") |>
  select(page, department) |>
  arrange(department, page) |>
  kable()
```

### Pages that do NOT exist (need to be created)

```{r missing-pages-table}
page_existence |>
  filter(!exists) |>
  left_join(page_dept_map, by = "page") |>
  select(page, department) |>
  arrange(department, page) |>
  kable()
```

# 7. Classify existing pages by Wikidata item type

Some existing Wikipedia pages may be linked to a Wikidata item that represents a **province** or **locality** rather than a municipality. These need new dedicated municipality pages (or Wikidata corrections).

We check this for existing pages that were *not* matched to the Wikidata `municipality of Bolivia` query.

```{r identify-mismatched-pages}
existing_pages_df <- page_existence |> filter(exists)

# Pages already matched to a proper Wikidata municipality item
matched_to_wd <- existing_pages_df |>
  inner_join(wd_munis, by = c("page" = "en_wiki_title"))

# Pages that exist but aren't matched
existing_not_in_wd <- existing_pages_df |>
  anti_join(wd_munis, by = c("page" = "en_wiki_title"))

message(nrow(existing_not_in_wd),
        " existing pages not matched to a Wikidata municipality item")
```

```{r get-qids-for-unmatched}
#| cache: true
# Get Wikidata QIDs for unmatched pages via Wikipedia API
get_wikidata_qids <- function(titles) {
  resp <- request("https://en.wikipedia.org/w/api.php") |>
    req_url_query(
      action = "query",
      titles = paste(titles, collapse = "|"),
      prop = "pageprops",
      ppprop = "wikibase_item",
      format = "json",
      redirects = 1
    ) |>
    req_perform() |>
    resp_body_json()

  page_data <- resp$query$pages
  tibble(
    pageid = map_int(page_data, ~ as.integer(.x$pageid %||% NA)),
    title = map_chr(page_data, ~ .x$title %||% NA_character_),
    qid = map_chr(page_data, ~ .x$pageprops$wikibase_item %||% NA_character_)
  )
}

qid_results <- get_wikidata_qids(existing_not_in_wd$page)
```

```{r query-instance-of}
#| cache: true
# Query Wikidata for instance_of (P31) for each QID
get_instance_of <- function(qids) {
  qid_values <- paste(paste0("wd:", qids), collapse = " ")

  query <- glue('
    SELECT ?item ?itemLabel ?instanceOf ?instanceOfLabel WHERE {{
      VALUES ?item {{ {qid_values} }}
      ?item wdt:P31 ?instanceOf .
      SERVICE wikibase:label {{ bd:serviceParam wikibase:language "en". }}
    }}
  ')

  resp <- request("https://query.wikidata.org/sparql") |>
    req_url_query(query = query, format = "json") |>
    req_headers(`User-Agent` = "WikiGraph/1.0 (research project)") |>
    req_perform() |>
    resp_body_json()

  bindings <- resp$results$bindings
  tibble(
    qid = map_chr(bindings, ~ sub("http://www.wikidata.org/entity/", "",
                                   .x$item$value)),
    item_label = map_chr(bindings, ~ .x$itemLabel$value),
    instance_of_qid = map_chr(bindings, ~ sub("http://www.wikidata.org/entity/", "",
                                               .x$instanceOf$value)),
    instance_of_label = map_chr(bindings, ~ .x$instanceOfLabel$value)
  )
}

valid_qids <- qid_results$qid[!is.na(qid_results$qid)]
instance_data <- get_instance_of(valid_qids)
```

```{r classify-pages}
page_categories <- qid_results |>
  left_join(
    instance_data |>
      group_by(qid) |>
      summarise(
        all_types = paste(instance_of_label, collapse = "; "),
        is_municipality = any(instance_of_qid == "Q1062710"),
        is_province = any(instance_of_qid == "Q1062593"),
        is_locality = any(instance_of_qid %in% c(
          "Q515", "Q532", "Q486972", "Q3957",
          "Q1549591", "Q200250", "Q51929311", "Q5712544"
        )),
        .groups = "drop"
      ),
    by = "qid"
  ) |>
  mutate(
    category = case_when(
      is_municipality ~ "municipality (OK)",
      is_province ~ "province",
      is_locality ~ "locality",
      TRUE ~ "other/unknown"
    )
  )
```

## Summary by Wikidata item type

```{r type-summary}
page_categories |>
  count(category, name = "n_pages") |>
  arrange(desc(n_pages)) |>
  kable()
```

### Province items (page redirects to province article)

These existing page titles redirect to a Wikidata item typed as `province of Bolivia`. A new municipality page and Wikidata item are needed.

```{r province-pages}
page_categories |>
  filter(category == "province") |>
  select(
    `Expected page` = title,
    `Redirects to QID` = qid,
    `Wikidata types` = all_types
  ) |>
  kable()
```

### Locality items (page is about the city/town, not the municipality)

These existing page titles link to a Wikidata item for the settlement rather than the municipality. A new municipality page is needed.

```{r locality-pages}
page_categories |>
  filter(category == "locality") |>
  select(
    `Expected page` = title,
    `Current QID` = qid,
    `Wikidata types` = all_types
  ) |>
  kable()
```

### Title-mismatch municipalities (already correct in Wikidata)

These pages exist and are properly typed as municipalities in Wikidata but weren't matched due to title differences between the expected page name and the actual Wikidata English label.

```{r title-mismatch-pages}
page_categories |>
  filter(category == "municipality (OK)") |>
  left_join(existing_not_in_wd |> select(page), 
            by = c("title" = "page")) |>
  select(
    `Expected page name` = title,
    QID = qid
  ) |>
  kable()
```

# 8. Combined action list

## Municipalities needing new Wikipedia pages

This combines (a) expected pages that don't exist at all with (b) pages that exist but are linked to the wrong Wikidata item type.

```{r combined-action-list}
# Pages that don't exist
no_page <- page_existence |>
  filter(!exists) |>
  left_join(page_dept_map, by = "page") |>
  mutate(reason = "No page exists")

# Pages that exist but redirect to province or locality
wrong_type <- page_categories |>
  filter(category %in% c("province", "locality")) |>
  left_join(
    existing_not_in_wd |>
      left_join(page_dept_map, by = "page"),
    by = c("title" = "page")
  ) |>
  transmute(
    page = title,
    department,
    reason = glue("Existing page is {category} ({qid})")
  )

action_list <- bind_rows(no_page |> select(page, department, reason),
                         wrong_type) |>
  arrange(department, page)

message(nrow(action_list), " municipalities need new or corrected Wikipedia pages")
```

```{r action-list-table}
action_list |>
  select(Page = page, Department = department, Reason = reason) |>
  kable()
```

## Summary by department

```{r dept-summary}
dept_summary <- page_dept_map |>
  left_join(page_existence, by = "page") |>
  left_join(
    page_categories |> select(title, category),
    by = c("page" = "title")
  ) |>
  mutate(
    status = case_when(
      !exists ~ "No page",
      category == "province" ~ "Redirects to province",
      category == "locality" ~ "Links to locality",
      TRUE ~ "OK"
    )
  ) |>
  count(department, status) |>
  pivot_wider(names_from = status, values_from = n, values_fill = 0)

dept_summary |> kable()
```
