---
title: "INE–IGM Community Crosswalk: Methods and Results"
subtitle: "Bolivia Community Geography Project"
date: "2026-02-23"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    number-sections: true
execute:
  warning: false
  message: false
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

```{r}
#| label: setup
#| code-fold: false
library(tidyverse)
library(sf)
library(knitr)

crosswalk <- readRDS("../data/crosswalk_ine_igm.rds")
```

## Introduction

This document describes the construction of a crosswalk linking the Bolivian
National Statistics Institute (INE) community registry to point locations from
the Instituto Geográfico Militar (IGM). The goal is to assign each of the
19,418 communities in the INE registry a geographic identifier (`id_unico`)
from the IGM settlement point dataset, enabling downstream linkage to
coordinates and Wikidata entities.

The two datasets do not share a common key. Matching proceeds by community
name within department, with a multi-stage spatial disambiguation pipeline to
resolve cases where the same name appears in more than one location.

## Data Sources

Four datasets feed the pipeline.

**INE community registry** (`CLASIF_UB_GEOG_COMUNIDAD.xlsx`). The official
INE classification of urban geographical locations (*clasificación de
ubicaciones geográficas*), containing 19,418 communities. Each row carries an
11-digit `Codigo` that encodes department, province, municipality, canton, and
a community sequence number (see @sec-ine-codes). Administrative hierarchy
columns (`DEPARTAMENTO`, `PROVINCIA`, `MUNICIPIO`) and their numeric segment
codes (`DEP`, `PRO`, `MUN`) are also present.

**IGM settlement points** (`localizacion_poblaciones_2016.json`). A GeoJSON
point dataset of 23,891 Bolivian settlements compiled by the IGM in 2016. The
key fields used here are `id_unico` (a unique point identifier), `nombre_dep`
(department name), `nombre_c_1` (settlement name, uppercase), and `tipo_area`
(settlement type: `cp` = population centre, `cpd` = dispersed population
centre, `ci` = urban centre, `dis` = dispersed).

**GADM Bolivia level-3 boundaries** (`gadm41_BOL_3.gpkg`). Municipality
(level-3) polygon boundaries from the Global Administrative Areas database
(GADM v4.1). Used to spatially assign each IGM point to a municipality and
thereby resolve name ambiguities.

**USCA community polygons** (`etnicidad_tenencia/usca_final.shp`). Community-
level multipolygon boundaries from the Unidades Socio-Culturales Agrarias
(USCA) dataset, covering 14,426 features. Each polygon carries a 10-digit INE
code (`cod10dig`), which maps to `Codigo` after prepending a leading zero. Used
as a final spatial disambiguation layer for cases that GADM could not resolve.

## INE Code Structure {#sec-ine-codes}

Each INE community code (`Codigo`) is an 11-character string with a fixed
leading zero followed by five numeric fields:

```
0  |  DD  |  PP  |  MM  |  C  |  SSS
   dept   prov   mun   canton  seq
```

| Position | Width | Field        | Description                       |
|----------|-------|--------------|-----------------------------------|
| 1        | 1     | Leading zero | Fixed padding                     |
| 2–3      | 2     | `DD`         | Department code                   |
| 4–5      | 2     | `PP`         | Province code                     |
| 6–7      | 2     | `MM`         | Municipality code                 |
| 8        | 1     | `C`          | Canton code                       |
| 9–11     | 3     | `SSS`        | Community sequence number         |

The USCA shapefile uses a 10-digit form (the leading zero is absent). To
join USCA codes to INE `Codigo` values, prepend a zero: `paste0("0",
cod10dig)`.

## Matching Pipeline

The pipeline proceeds in eight stages, each producing a more complete and
resolved crosswalk. The central challenge is that the IGM dataset contains no
municipality field — only department and name — so names that recur within a
department cannot be disambiguated by name alone.

### Name normalization

```{r}
#| label: normalization
normalize_match_key <- function(x) {
  x |>
    str_trim() |>
    str_squish() |>
    str_to_upper() |>
    str_replace_all("[^\x20-\x7E]", "") |>
    str_replace_all("[^A-Z0-9 ]", " ") |>
    str_squish()
}
```

Both datasets are normalized before any join. The `normalize_match_key()`
function strips non-ASCII characters (encoding artifacts introduced by
accented letters in different encodings), collapses punctuation and symbols
to spaces, squishes runs of whitespace, and uppercases everything. This
handles the most common source of spurious non-matches: encoding differences
between the INE Excel file and the IGM GeoJSON.

A small number of INE names are known to differ from their IGM counterparts
for reasons beyond encoding — abbreviations, historical name changes, or
alternative place names. These are corrected with explicit overrides before
the normalized key is computed:

| INE name | IGM name | Reason |
|---|---|---|
| `OKINAWA 1` | `OKINAWA UNO` | Abbreviated ordinal |
| `TAIPIPLAYA` | `TANIPLAYA` | Alternative spelling |
| `MUYUPAMPA` | `VILLA VACA GUZMAN` | Historical name change |
| `VACAS K UCHU` | `VACAS KUCHU` | Spacing artifact |
| `VALLE DE CONCEPCION` | `CONCEPCION` | Short-form name in IGM |

### Initial join: department + name

```{r}
#| label: initial-join
ine_geog_2013 <- readxl::read_excel("data/CLASIF_UB_GEOG_COMUNIDAD.xlsx") |>
  rename(
    department  = DEPARTAMENTO,
    province    = PROVINCIA,
    municipality = MUNICIPIO
  ) |>
  mutate(
    cod.prov = paste0(DEP, PRO),
    cod.mun  = paste0(DEP, PRO, MUN),
    cod.com  = Codigo,
    cod.dep  = DEP
  )

geo <- st_read("data/localizacion_poblaciones_2016.json", quiet = TRUE)

geo_norm <- geo |>
  st_drop_geometry() |>
  mutate(
    nombre_c_1 = str_trim(nombre_c_1),
    match_key  = normalize_match_key(nombre_c_1)
  )

ine_norm <- ine_geog_2013 |>
  mutate(
    com_name  = str_trim(`CIUDAD/COMUNIDAD`),
    match_key = case_when(
      com_name == "OKINAWA 1"                      ~ normalize_match_key("OKINAWA UNO"),
      com_name == "3o GRUPO VALLE HERMOSO"          ~ normalize_match_key("3O GRUPO VALLE HERMOSO"),
      com_name == "TAIPIPLAYA"                      ~ normalize_match_key("TANIPLAYA"),
      com_name == "SINDICATO AGRARIO IMILLA IMILLA" ~ normalize_match_key("SINDICATO AGRARIO IMILLA"),
      com_name == "VALLE SACTA (Disperso)"           ~ normalize_match_key("VALLE SACTA (DISPERSO)"),
      com_name == "VALLE DE CONCEPCION"             ~ normalize_match_key("CONCEPCION"),
      com_name == "VACAS K UCHU"                    ~ normalize_match_key("VACAS KUCHU"),
      com_name == "MUYUPAMPA"                       ~ normalize_match_key("VILLA VACA GUZMAN"),
      TRUE                                          ~ normalize_match_key(com_name)
    )
  )

crosswalk_raw <- ine_norm |>
  select(Codigo, department, municipality, com_name, match_key) |>
  left_join(
    geo_norm |> select(nombre_dep, nombre_c_1, id_unico, match_key),
    by           = c("department" = "nombre_dep", "match_key"),
    relationship = "many-to-many"
  )

match_counts <- crosswalk_raw |>
  filter(!is.na(id_unico)) |>
  group_by(Codigo) |>
  summarise(n_geo = n_distinct(id_unico), .groups = "drop")
```

The initial join matches on the pair `(department, normalized_name)`. Because
the IGM dataset carries no municipality information, any community name that
appears in two or more municipalities within the same department produces
multiple candidate matches. These are flagged as *ambiguous* and carried
forward to the disambiguation stages.

After this stage, records fall into three preliminary categories:

- **Unique** — exactly one IGM point matches the name within the department.
- **Ambiguous** — two or more IGM points match.
- **Unmatched** — no IGM point with that name exists in the department.

### Stage 1: Direct unique matches

```{r}
#| label: stage1
crosswalk_geo_ine_s1 <- crosswalk_raw |>
  left_join(match_counts, by = "Codigo") |>
  mutate(
    n_geo        = coalesce(n_geo, 0L),
    match_status = case_when(
      is.na(id_unico) ~ "unmatched",
      n_geo == 1       ~ "unique",
      n_geo > 1        ~ "ambiguous"
    )
  ) |>
  select(Codigo, department, municipality, com_name, id_unico, match_status, n_geo)
```

Records with exactly one matching IGM point are immediately assigned
`match_status = "unique"`. This accounts for the large majority of the
registry.

### Stage 2: GADM spatial municipality join

```{r}
#| label: stage2
gadm_path <- "../data/gadm41_BOL_3.gpkg"
mun_boundaries <- st_read(gadm_path, layer = "ADM_ADM_3", quiet = TRUE)

gadm_lookup <- mun_boundaries |>
  st_drop_geometry() |>
  select(gadm_dep = NAME_1, gadm_prov = NAME_2, gadm_mun = NAME_3) |>
  mutate(
    dep_key = normalize_match_key(gadm_dep),
    mun_key = normalize_match_key(gadm_mun)
  )

ine_mun_lookup <- ine_geog_2013 |>
  select(department, province, municipality, cod.mun) |>
  distinct() |>
  mutate(
    dep_key = normalize_match_key(department),
    mun_key = normalize_match_key(municipality)
  )
```

To resolve ambiguous matches, each IGM point is spatially assigned to a GADM
municipality polygon using `st_within`. Points that fall on or outside polygon
boundaries (e.g. border settlements) are snapped to the nearest polygon using
`st_nearest_feature` as a fallback.

Municipality names differ substantially between the INE and GADM datasets —
different spellings, abbreviations, historical renames, and cases where GADM
uses a province name where INE uses a municipality name. A manual crosswalk
of 44 municipality name pairs resolves these discrepancies, covering all 339
INE municipalities.

Once each IGM point is tagged with a municipality code, the ambiguous
crosswalk entries are filtered to keep only the candidate whose assigned
municipality matches the INE record's municipality. If this filter reduces
the candidate set to exactly one point, the record is resolved as
`unique_via_spatial`.

### Stage 3: Name-proximity fallback

```{r}
#| label: stage3
ine_name_to_mun <- ine_norm |>
  select(department, match_key, municipality, cod.mun) |>
  distinct() |>
  group_by(department, match_key) |>
  filter(n_distinct(municipality) == 1) |>
  ungroup() |>
  select(department, match_key, municipality_ine = municipality, cod.mun_ine = cod.mun)

mun_sf <- mun_boundaries |>
  mutate(
    dep_key = normalize_match_key(NAME_1),
    mun_key = normalize_match_key(NAME_3)
  ) |>
  inner_join(
    bind_rows(
      ine_mun_lookup |>
        inner_join(gadm_lookup |> select(dep_key, mun_key, gadm_mun, gadm_prov),
                   by = c("dep_key", "mun_key")) |>
        select(cod.mun, department, municipality, gadm_mun),
      tribble(
        ~department,  ~municipality,               ~gadm_mun,
        "La Paz",     "La Paz",                    "Nuestra Señora de La Paz",
        "La Paz",     "Callapa",                   "Santiago de Callapa",
        "Beni",       "Rurrenabaque",              "Puerto Menor de Rurrenabaque",
        "Chuquisaca", "Muyupampa",                 "Villa Vaca Guzmán"
      ) |>
        left_join(ine_mun_lookup |> select(department, municipality, cod.mun),
                  by = c("department", "municipality"))
    ) |>
      mutate(
        dep_key = normalize_match_key(department),
        mun_key = normalize_match_key(gadm_mun)
      ),
    by = c("dep_key", "mun_key")
  ) |>
  select(cod.mun, department, municipality)
```

A small number of IGM points are placed outside their true GADM polygon by
the spatial join — typically because they lie very close to a border or
because the GADM boundaries do not perfectly align with the IGM point
locations. For INE records whose community name is unambiguous *within* the
department (i.e., the name maps to only one municipality in the INE list),
the correct municipality is known even without spatial containment. In these
cases, the candidate IGM point closest to that municipality's centroid is
selected, resolving the record as `unique_via_name`.

### Stage 4: USCA polygon containment

```{r}
#| label: stage4
usca_valid <- st_make_valid(
  st_read("../data/etnicidad_tenencia/usca_final.shp", quiet = TRUE)
)

igm_spatial_ine <- st_join(
  geo |> select(id_unico),
  usca_valid |> select(cod10dig),
  join = st_within
) |>
  st_drop_geometry() |>
  distinct(id_unico, .keep_all = TRUE) |>
  mutate(
    spatial_ine = if_else(!is.na(cod10dig), paste0("0", cod10dig), NA_character_)
  ) |>
  select(id_unico, spatial_ine)
```

For records still ambiguous after Stages 2 and 3, the USCA community polygon
layer provides a finer spatial filter. Each IGM point is matched to the USCA
polygon it falls within (using `st_within`), and the resulting USCA code is
compared directly to the INE `Codigo`. If exactly one of the candidate points
falls inside the USCA polygon corresponding to the INE code, that point is
assigned as `unique_via_usca`.

The USCA dataset uses 10-digit codes; a leading zero is prepended before
comparison.

### Stage 5: Canton-split detection

```{r}
#| label: stage5
id_set_sig <- crosswalk |>
  filter(match_status == "ambiguous_canton_split" | match_status == "ambiguous") |>
  filter(!is.na(id_unico)) |>
  group_by(Codigo) |>
  summarise(id_key = paste(sort(unique(id_unico)), collapse = "|"), .groups = "drop")

canton_split_map <- id_set_sig |>
  group_by(id_key) |>
  filter(n() > 1) |>
  mutate(canton_split_group = min(Codigo)) |>
  ungroup() |>
  select(Codigo, canton_split_group)
```

Bolivia's canton boundaries were reorganised after the 2001 census. As a
result, some communities appear in the INE registry under two or more
`Codigo` values that differ only in the canton digit (position 8) — with
the department, province, municipality, and community sequence number
unchanged. These sibling codes all match the same pool of IGM candidate
points.

To detect these groups, each remaining ambiguous `Codigo` is fingerprinted
by the sorted set of its candidate `id_unico` values. Any two codes that
share an identical fingerprint are placed in the same canton-split group,
identified by the lowest `Codigo` in the group (`canton_split_group`). These
records are labelled `ambiguous_canton_split`.

The many-to-one mapping (multiple INE codes → same IGM points) is **expected
and correct** for these groups. To build a deduplicated community key, use:

```r
community_key <- coalesce(canton_split_group, Codigo)
```

### Stage 6: Dispersed settlement reclassification

```{r}
#| label: stage6
pure_dis_codes <- crosswalk |>
  filter(match_status %in% c("ambiguous", "ambiguous_no_spatial")) |>
  filter(!is.na(id_unico)) |>
  left_join(
    geo |> st_drop_geometry() |> select(id_unico, tipo_area),
    by = "id_unico"
  ) |>
  group_by(Codigo) |>
  filter(all(tipo_area == "dis")) |>
  pull(Codigo) |>
  unique()
```

The IGM dataset uses `tipo_area = "dis"` to flag dispersed settlements —
rural communities spread across a wide area rather than concentrated at a
single point. For such communities, the IGM frequently records multiple
points, each representing a cluster or hamlet within the broader settlement.

When all candidate IGM points for an ambiguous INE record carry `tipo_area
== "dis"`, no single point is more authoritative than the others. These
records are reclassified as `ambiguous_dispersed`, signalling to downstream
users that the full candidate pool should be treated as a coordinate
envelope rather than forcing a single selection.

### Stage 7: Cross-municipality adjacency splitting

```{r}
#| label: stage7
cross_mun_summary <- crosswalk |>
  filter(match_status == "ambiguous_no_spatial") |>
  group_by(department, com_name) |>
  summarise(n_mun = n_distinct(municipality), .groups = "drop") |>
  count(n_mun)
```

Some records are labelled `ambiguous_no_spatial` because the same community
name appears in two municipalities within the same department and the GADM
spatial join could not assign the points unambiguously. However, if the two
municipalities are geographically non-adjacent, the candidate IGM points can
be partitioned by which municipality polygon they fall within — and that
partition resolves the ambiguity.

For each `ambiguous_no_spatial` group spanning exactly two municipalities,
the pipeline tests whether those municipalities are adjacent (`st_touches`).
For non-adjacent pairs, the candidate points are spatially split between the
two municipality polygons. Each INE record then receives only the points
assigned to its municipality:

- If the split leaves a single point: `unique_via_spatial`
- If multiple points remain (all dispersed): `ambiguous_dispersed`

### Stage 8: Mixed-type resolution

```{r}
#| label: stage8
type_resolution_note <- crosswalk |>
  filter(match_status == "unique_via_spatial") |>
  nrow()
```

For any ambiguous records still remaining after Stage 7, the `tipo_area`
field provides a final resolution heuristic. Named population centres
(`cp`, `cpd`) and urban centres (`ci`) are more specific geographic entities
than dispersed points (`dis`). If a candidate pool contains exactly one
point of the highest-priority type, that point is selected:

| Priority | `tipo_area` | Description |
|----------|-------------|-------------|
| 1 | `cp`, `cpd` | Population centre / dispersed population centre |
| 2 | `ci`        | Urban/city centre |
| 3 | `dis`       | Dispersed settlement |

A code is resolved as `unique_via_spatial` when the top-ranked type appears
exactly once. If multiple candidates share the top rank, the record remains
ambiguous.

## Results

```{r}
#| label: results-table
status_counts <- crosswalk |>
  distinct(Codigo, match_status) |>
  count(match_status, name = "n_codes") |>
  arrange(desc(n_codes))

status_counts |>
  mutate(
    pct = scales::percent(n_codes / sum(n_codes), accuracy = 0.1)
  ) |>
  rename(
    `Match status`   = match_status,
    `INE codes`      = n_codes,
    `Share`          = pct
  ) |>
  kable(align = c("l", "r", "r"))
```

```{r}
#| label: results-plot
#| fig-height: 4
status_counts |>
  mutate(
    match_status = fct_reorder(match_status, n_codes),
    resolved     = str_starts(match_status, "unique")
  ) |>
  ggplot(aes(x = n_codes, y = match_status, fill = resolved)) +
  geom_col() +
  scale_fill_manual(
    values = c("TRUE" = "#2a9d8f", "FALSE" = "#e9c46a"),
    labels = c("TRUE" = "Resolved", "FALSE" = "Ambiguous / unmatched"),
    name   = NULL
  ) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = "Number of INE codes",
    y = NULL,
    title = "INE–IGM crosswalk match status",
    subtitle = "19,418 communities across Bolivia"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
#| label: resolved-summary
n_resolved <- crosswalk |>
  distinct(Codigo, match_status) |>
  filter(str_starts(match_status, "unique")) |>
  nrow()

n_total <- n_distinct(crosswalk$Codigo)
```

Of the `r scales::comma(n_total)` INE community codes,
`r scales::comma(n_resolved)`
(`r scales::percent(n_resolved / n_total, accuracy = 0.1)`) are resolved to
a unique IGM point by at least one matching stage.

## Match Status Reference

```{r}
#| label: status-reference
status_ref <- tribble(
  ~Status,                  ~Description,
  "unique",                 "Direct 1-to-1 name match within department.",
  "unique_via_spatial",     "Ambiguous name resolved by GADM municipality boundary, cross-municipality adjacency splitting, or population-centre type priority.",
  "unique_via_name",        "Ambiguous name resolved by selecting the IGM point closest to the correct municipality centroid.",
  "unique_via_usca",        "Ambiguous name resolved by USCA community polygon containment.",
  "ambiguous_canton_split", "Same community listed under multiple canton codes due to administrative reorganisation. All sibling codes share the same IGM candidate pool; many-to-one mapping is expected.",
  "ambiguous_dispersed",    "All IGM candidates are tipo_area == 'dis' (dispersed). The IGM recorded multiple points for the same settlement. Treat all candidates as a valid coordinate pool.",
  "ambiguous",              "Genuinely repeated name within the same municipality. The correct point cannot be determined automatically.",
  "ambiguous_no_spatial",   "Name recurs across municipalities; no spatial disambiguation succeeded.",
  "unmatched",              "No IGM point with a matching name found."
)

status_ref |>
  left_join(
    crosswalk |>
      distinct(Codigo, match_status) |>
      count(match_status, name = "N"),
    by = c("Status" = "match_status")
  ) |>
  mutate(N = coalesce(N, 0L)) |>
  rename(`Match status` = Status, `N (INE codes)` = N) |>
  kable()
```

**Canton-split groups** can be identified and deduplicated as follows:

```r
# Build a deduplicated community key (one per physical place)
community_key <- coalesce(canton_split_group, Codigo)
```

## Usage

### Loading the crosswalk

```{r}
#| label: usage-load
#| eval: false
crosswalk <- readRDS("data/crosswalk_ine_igm.rds")
# or
crosswalk <- read_csv("data/crosswalk_ine_igm.csv")
```

### Joining to the INE community list

```{r}
#| label: usage-join
#| eval: false
library(tidyverse)

ine_geog_2013 |>
  left_join(
    crosswalk |> select(Codigo, id_unico, match_status, canton_split_group),
    by = "Codigo"
  )
```

### Joining to IGM coordinates

```{r}
#| label: usage-coords
#| eval: false
library(sf)

igm_sf <- st_read("data/localizacion_poblaciones_2016.json", quiet = TRUE)

# Resolved records only
crosswalk |>
  filter(str_starts(match_status, "unique")) |>
  left_join(igm_sf |> st_drop_geometry(), by = "id_unico")
```

### Looking up a single INE code

The `ine_match_status()` function (defined in `ine_community_mapping.R`)
returns a human-readable summary for any INE code, accepting both the
11-digit `Codigo` form and the 10-digit USCA `cod10dig` form:

```r
source("ine_community_mapping.R")
ine_match_status("01010101001")
```
