---
title: "INE–IGM Community Crosswalk Verification"
subtitle: "Analysis of mapping ambiguities, duplicates, and data quality issues"
author: ""
date: today
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

# Load the crosswalk
source(here::here("ine_igm_community_crosswalk.R"))
```

## Overview

The INE–IGM community crosswalk maps 11-digit INE community codes (`Codigo`) to geographic points in the IGM `localizacion_poblaciones_2016` dataset (identified by `id_unico`). This analysis examines mapping ambiguities and data quality issues revealed by the crosswalk construction process.

**Key metrics**:
- Total INE communities: {{nrow(distinct(crosswalk_geo_ine, Codigo))}}
- Unique IGM points involved in crosswalk: {{n_distinct(crosswalk_geo_ine$id_unico)}}
- Total crosswalk rows (many-to-many): {{nrow(crosswalk_geo_ine)}}
- Duplicate mappings (id_unico with n>1): {{nrow(crosswalk_geo_ine |> count(id_unico) |> filter(n > 1))}}

## Match Status Distribution

```{r}
#| label: match-status-summary

match_summary <- crosswalk_geo_ine |>
  count(match_status) |>
  arrange(desc(n)) |>
  mutate(
    pct = round(100 * n / sum(n), 1),
    match_status = fct_reorder(match_status, n)
  )

match_summary |>
  ggplot(aes(y = match_status, x = n, fill = match_status)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%d (%.1f%%)", n, pct)), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Match Status Distribution",
    x = "Number of INE Codes",
    y = NULL
  ) +
  theme_minimal()

match_summary
```

## Duplicate Mappings Analysis

When a single `id_unico` maps to multiple INE `Codigo` entries, it indicates either:
1. **Intentional** many-to-one mapping (canon splits, dispersed settlements)
2. **Unintentional** convergence through ambiguity resolution heuristics

### Duplicate Counts by Match Status

```{r}
#| label: duplicates-by-status

duplicates <- crosswalk_geo_ine |>
  count(id_unico) |>
  filter(n > 1)

dup_by_status <- crosswalk_geo_ine |>
  semi_join(duplicates, by = "id_unico") |>
  group_by(match_status) |>
  summarise(
    n_rows = n(),
    n_unique_ids = n_distinct(id_unico),
    n_unique_codigos = n_distinct(Codigo),
    avg_codigos_per_id = round(mean(n()), 2),
    .groups = "drop"
  ) |>
  arrange(desc(n_rows))

dup_by_status |>
  ggplot(aes(y = fct_reorder(match_status, n_rows), x = n_rows, fill = match_status)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_rows), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Duplicate Mappings by Resolution Method",
    subtitle = "Total rows where one id_unico maps to multiple Codigos",
    x = "Number of rows",
    y = NULL
  ) +
  theme_minimal()

dup_by_status
```

### Breakdown

**Intentional duplicates (3,426 rows / 86%)**:

- **Dispersed settlements** (2,254 rows): 939 IGM points representing the same dispersed community. The IGM dataset intentionally recorded multiple points for dispersed settlements.
- **Canton splits** (914 rows): 436 IGM points shared by 421 different INE codes because the same community was renumbered during canton boundary reorganization.
- **Name matches at all levels** (258 rows): Multiple INE codes with identical community name, municipality, province, and department. Codes were assigned as new cantons were added to a municipality; the last two digits of each Codigo (the community sequence number within the canton) are typically identical across the group. These almost certainly correspond to a single community. See Section 4.1 for examples.

**Unintended duplicates (509 rows / 13%)**:

- **Spatial resolution** (401 rows): Ambiguous names where multiple Codigos converged on the same point after GADM municipality boundary filtering.
- **USCA polygon** (108 rows): Multiple Codigos resolved via USCA community boundary spatial join.

**Edge cases (40 rows / 1%)**:

- **Ambiguous (unresolved)** (28 rows): Still-ambiguous after all spatial steps.
- **Other** (12 rows): Rare cases.

## Same Name, Same ID Convergence

How many cases have different Codigos with identical `com_name` mapping to the same `id_unico`?

```{r}
#| label: same-name-convergence

same_name_same_id <- crosswalk_geo_ine |>
  group_by(id_unico, com_name) |>
  summarise(
    n_codigos = n_distinct(Codigo),
    .groups = "drop"
  ) |>
  filter(n_codigos > 1)

convergence_by_status <- crosswalk_geo_ine |>
  group_by(id_unico, com_name, match_status) |>
  summarise(
    n_codigos = n_distinct(Codigo),
    .groups = "drop"
  ) |>
  filter(n_codigos > 1) |>
  group_by(match_status) |>
  summarise(
    n_pairs = n(),
    total_duplicate_codigos = sum(n_codigos),
    avg_codigos_per_pair = round(mean(n_codigos), 2),
    max_codigos = max(n_codigos),
    .groups = "drop"
  ) |>
  arrange(desc(n_pairs))

convergence_by_status |>
  ggplot(aes(y = fct_reorder(match_status, n_pairs), x = n_pairs, fill = match_status)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_pairs), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Convergence Pairs by Method",
    subtitle = "Cases where different Codigos with same com_name → same id_unico",
    x = "Number of (id_unico, com_name) pairs",
    y = NULL
  ) +
  theme_minimal()

convergence_by_status
```

**Total pairs with multiple Codigos**: {{nrow(same_name_same_id)}}

Of these, {{nrow(convergence_by_status)}} match_status groups account for the convergence. The largest contributors are:
- **ambiguous_dispersed**: {{convergence_by_status |> filter(match_status == "ambiguous_dispersed") |> pull(n_pairs)}} pairs — inherent to IGM's multi-point recording of dispersed settlements
- **ambiguous_canton_split**: {{convergence_by_status |> filter(match_status == "ambiguous_canton_split") |> pull(n_pairs)}} pairs — expected due to historical canton reorganization

## Cross-Municipality vs Same-Municipality Convergence

### unique_via_name Cases

**Name matches at all levels (258 rows, 127 pairs)**: These are multiple INE codes with the same names for community, municipality, province, and department. They have been recoded as new cantons are added to a municipality. Often these codes have the same two final digits. It is highly likely they correspond to a single community.

All 127 pairs are within the same municipality — none span municipal boundaries — which is consistent with canton renumbering rather than genuine geographic duplication.

**Five representative examples** (note identical last two digits within each group):

```{r}
#| label: via-name-examples

crosswalk_geo_ine |>
  filter(
    match_status == "unique_via_name",
    id_unico %in% c("2182984311-D", "2574485871-D", "3462128806-D",
                    "4175616540-D", "4543857044-D")
  ) |>
  mutate(`Last 2` = str_sub(Codigo, -2)) |>
  select(Codigo, `Last 2`, `Community` = com_name, Municipality = municipality) |>
  arrange(Community, Codigo) |>
  knitr::kable()
```

### unique_via_spatial Cases

```{r}
#| label: via-spatial-analysis

via_spatial_multi_mun <- crosswalk_geo_ine |>
  filter(match_status == "unique_via_spatial") |>
  group_by(id_unico, com_name) |>
  summarise(
    n_codigos = n_distinct(Codigo),
    n_municipalities = n_distinct(municipality),
    municipalities = paste(sort(unique(municipality)), collapse = " | "),
    .groups = "drop"
  ) |>
  filter(n_codigos > 1)

multi_mun_spatial <- via_spatial_multi_mun |>
  filter(n_municipalities > 1)

cat("unique_via_spatial (n =", nrow(via_spatial_multi_mun), "pairs with multiple Codigos)\n")
cat("  - Cross-municipality:", nrow(multi_mun_spatial), "(", 
    round(100 * nrow(multi_mun_spatial) / nrow(via_spatial_multi_mun), 1), "%)\n", sep = "")
cat("  - Same municipality:", nrow(via_spatial_multi_mun) - nrow(multi_mun_spatial), "(", 
    round(100 * (nrow(via_spatial_multi_mun) - nrow(multi_mun_spatial)) / nrow(via_spatial_multi_mun), 1), "%)\n", sep = "")
```

**79% are cross-municipality convergences**, where the same place name appears in multiple municipalities but IGM only has one geopoint. This is the most problematic pattern for downstream geocoding work.

**Top examples** by number of municipalities involved:

```{r}
#| label: via-spatial-examples

via_spatial_multi_mun |>
  arrange(desc(n_municipalities), desc(n_codigos)) |>
  slice(1:12) |>
  select(com_name, n_codigos, n_municipalities, municipalities)
```

### Risk Assessment

```{r}
#| label: risk-assessment

risk_table <- tibble(
  category = c(
    "unique_via_spatial (cross mun)",
    "unique_via_name",
    "ambiguous_dispersed",
    "ambiguous_canton_split"
  ),
  n_pairs = c(
    48,
    127,
    658,
    436
  ),
  risk_level = c(
    "High",
    "Low",
    "Low",
    "Low"
  ),
  reason = c(
    "Same name across municipalities → one IGM point; cannot disambiguate",
    "Intentional: same community recoded as new cantons added; same last two digits",
    "Intentional: IGM recorded multiple dispersed points",
    "Intentional: documented canton reorganization"
  )
)

risk_table
```

## Specific Problem Cases

### VILUYO (unique_via_spatial, 10 municipalities)

```{r}
#| label: viluyo-case

viluyo <- crosswalk_geo_ine |>
  filter(id_unico == "3548609828-D", match_status == "unique_via_spatial")

viluyo |>
  select(Codigo, municipality, com_name) |>
  arrange(municipality)
```

**Issue**: "VILUYO" appears in 10 different municipalities across Potosí department. All 10 INE codes map to a single IGM geopoint. This could indicate:
- VILUYO is a common/generic place name (like "New Village")
- IGM only documented one instance
- Data quality issue: INE duplicated a location

**Recommendation**: Manual verification needed if precise geocoding is required.

### CONDOR NAZA (unique_via_name, 4 codes in same municipality)

```{r}
#| label: condor-case

condor <- crosswalk_geo_ine |>
  filter(id_unico == "4543857044-D", match_status == "unique_via_name")

condor |>
  mutate(`Last 2` = str_sub(Codigo, -2)) |>
  select(Codigo, `Last 2`, municipality, com_name)
```

**Explanation**: Four INE codes in S.P. De Buena Vista (Potosí) all record the same community "CONDOR NAZA" with identical final digits (`28`). The codes differ only in the canton segment, reflecting successive recodings as cantons were subdivided. This is the same canton-renumbering pattern as `ambiguous_canton_split`, identified here through the name-proximity resolution path.

## Summary and Recommendations

### Data Quality Findings

1. **86% of convergences are intentional**: canton splits, dispersed settlements, and same-community recodings across new cantons (`unique_via_name`). All are well-documented by `match_status` labeling.

2. **13% involve unintended convergence** through disambiguation heuristics:
   - **48 cases** (`unique_via_spatial`, cross-municipality): Same name appears across different municipalities but IGM has only one geopoint — high risk for precise geocoding.
   - **13 cases** (`unique_via_spatial`, same-municipality): Likely further canton-renumbering cases not caught by earlier steps.

3. **Geographic generics**: Common place names (VILUYO, PAMPA GRANDE, SAN ANTONIO) appear repeatedly across municipalities. Whether these are:
   - Genuinely distinct locations poorly covered by IGM
   - The same community recorded multiple times in INE
   - Locally common names used independently in different municipalities
   
   ...cannot be determined automatically.

### For Downstream Use

- **For census/analysis work**: Use the full crosswalk; `match_status` indicates confidence level
- **For precise geocoding**: Filter to `match_status %in% c("unique", "unique_via_spatial", "unique_via_usca", "unique_via_name")` and manually verify `unique_via_spatial` cross-municipality cases (see Section 5)
- **For geographic-dispersed communities**: Treat `ambiguous_dispersed` and `ambiguous_canton_split` as coordinate pools, not single points

### Suggested Verification Steps

1. Spot-check high-priority cases (VILUYO, PAMPA GRANDE, etc.) against local administrative maps
2. Cross-reference INE canton-split documentation for expected duplications
3. For census analysis, confirm `match_status` distribution hasn't changed in latest INE release
