---
title: "INE–IGM Spatial Inspection: Multi-Code Points"
subtitle: "Manual review of IGM points claimed by multiple INE community codes via unique_via_spatial resolution"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    self-contained: true
    code-fold: true
    code-summary: "Show code"
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| cache: true

library(tidyverse)
library(sf)
library(ggplot2)

# Crosswalk (loads crosswalk_geo_ine)
source(here::here("ine_community_mapping.R"))
 
# Spatial helper functions (point_to_municipality, inspect_shared_point)
source(here::here("src", "point_municipality_lookup.R"))

# IGM geographic points
geo <- st_read(here::here("data", "geocode", "localizacion_poblaciones_2016.json"), quiet = TRUE)

# GADM municipality boundaries (with INE cod.mun, saved from ine_community_mapping.R)
mun_sf <- readRDS(here::here("data", "mun_sf.rds"))
```

```{r}
#| label: build-id-vectors

# Reproduce the same_mun_ids / cross_mun_ids vectors used in the main analysis
multi_code_id <- crosswalk_geo_ine |>
  filter(match_status == "unique_via_spatial") |>
  count(id_unico) |>
  filter(n > 1) |>
  pull(id_unico)

same_mun_ids <- crosswalk_geo_ine |>
  filter(match_status == "unique_via_spatial", id_unico %in% multi_code_id) |>
  group_by(id_unico) |>
  filter(n_distinct(municipality) == 1) |>
  pull(id_unico) |>
  unique()

cross_mun_ids <- crosswalk_geo_ine |>
  filter(match_status == "unique_via_spatial", id_unico %in% multi_code_id) |>
  group_by(id_unico) |>
  filter(n_distinct(municipality) > 1) |>
  pull(id_unico) |>
  unique()
```

## Background

During crosswalk construction, `unique_via_spatial` codes were resolved by selecting
the single IGM candidate whose GADM municipality polygon contained the INE code's
municipality. When multiple INE codes were resolved this way and all converged on the
same IGM point, that point appears here.

There are two distinct patterns:

- **Same-municipality cases** (`r length(same_mun_ids)` points): all converging codes
  are in the same INE municipality. Pattern is consistent with canton renumbering,
  analogous to `unique_via_name` duplicates.

- **Cross-municipality cases** (`r length(cross_mun_ids)` points): converging codes
  span *different* INE municipalities. A single IGM point serves as the resolved
  location for communities in multiple municipalities — whether due to a common
  place name or incomplete IGM coverage is unresolved.

**Important**: Bolivia expanded from roughly 100 to 340 municipalities between 1994
and 2010. The GADM boundaries represent the current (~340-municipality) structure;
INE codes in this crosswalk may reflect earlier, broader municipal units. No single
answer is assumed correct. Each map below shows:

- **Colored polygons**: all GADM municipality boundaries referenced by INE codes
  mapping to this point
- **Thick gray border**: the municipality currently containing the point (GADM)
- **Red dot**: the IGM point itself

## Same-Municipality Cases (`r length(same_mun_ids)` points)

All converging codes are within a single INE municipality. Likely canton
renumbering — the same community appears under multiple codes as canton
boundaries were redrawn over time.

```{r}
#| label: same-mun-maps
#| message: false
#| warning: false
#| cache: true
#| results: asis

same_mun_results <- purrr::map(same_mun_ids, inspect_shared_point)

for (result in same_mun_results) {
  cat("\n\n###", result$heading, "\n\n")
  cat("**Current GADM municipality:**", result$current_mun_label, "\n\n")
  if (length(result$unmatched_muns) > 0) {
    cat(paste0(
      "**\u26a0 Municipality names not in current GADM (historical boundaries):** ",
      paste(result$unmatched_muns, collapse = ", "), "\n\n"
    ))
  }
  cat(knitr::kable(result$table, format = "pipe") |> paste(collapse = "\n"))
  cat("\n\n")
  print(result$map)
  cat("\n\n---\n\n")
}
```

## Cross-Municipality Cases (`r length(cross_mun_ids)` points)

Converging codes span *different* INE municipalities. Each map shows how far
apart the referenced municipalities are, which helps assess whether convergence
is plausible (nearby municipalities, shifted boundaries) or suspicious (distant
municipalities, probably a common generic name with incomplete IGM coverage).

```{r}
#| label: cross-mun-maps
#| message: false
#| warning: false
#| cache: true
#| results: asis

# Sort by descending number of municipalities to show most complex cases first
cross_mun_order <- crosswalk_geo_ine |>
  filter(match_status == "unique_via_spatial", id_unico %in% cross_mun_ids) |>
  group_by(id_unico) |>
  summarise(n_mun = n_distinct(municipality), .groups = "drop") |>
  arrange(desc(n_mun)) |>
  pull(id_unico)

cross_mun_results <- purrr::map(cross_mun_order, inspect_shared_point)

for (result in cross_mun_results) {
  cat("\n\n###", result$heading, "\n\n")
  cat("**Current GADM municipality:**", result$current_mun_label, "\n\n")
  if (length(result$unmatched_muns) > 0) {
    cat(paste0(
      "**\u26a0 Municipality names not in current GADM (historical boundaries):** ",
      paste(result$unmatched_muns, collapse = ", "), "\n\n"
    ))
  }
  cat(knitr::kable(result$table, format = "pipe") |> paste(collapse = "\n"))
  cat("\n\n")
  print(result$map)
  cat("\n\n---\n\n")
}
```
